---
title: "GI Cancer: Data Cleaning"
bibliography: references.bib
---
# Load Data and Libraries
```{r}
#| label: setup
#| include: false

source("R/setup.R")

# Packages
library(readxl)     # Read xlsx files
```

```{r}
#| label: load-data

gic_raw <- read_xlsx(here("data", "IPROTECTARetrospecti_DATA_2025-08-19_2157 FINAL SHEET.xlsx"))
```

# Data Cleaning

## Check columns that had the same names in the excel file

```{r}
#| label: duplicate-column-names

# Get list of names which were identical in excel file
dupe_names <- gic_raw |> 
  select(contains("...")) |> 
  names()

# Create tibble of names to compare from `dupe_names`
check_pairs <- tibble(
  col_a = c(
    "ANC.3...108",
    "HyperK+.3...114",
    "Febrile.Neutropenia.6...189",
    "HypoNa+.6...191",
    "HypoK+.6...192",
    "HyperK+.6...193"
  ),
  col_b = c(
    "ANC.3...109",
    "HyperK+.3...115",
    "Febrile.Neutropenia.6...215",
    "HypoNa+.6...217",
    "HypoK+.6...218",
    "HyperK+.6...219"
  )
)

# Compare to see if contents are identical
pair_audit <- check_pairs |> 
  rowwise() |> 
  mutate(
    are_identical = identical(gic_raw[[col_a]], gic_raw[[col_b]]),
    status = if_else(are_identical,
                     "Identical (Redundant)",
                     "Different (Distinct)")
  ) |> 
  ungroup()

# Create table
ident_rows <- which(pair_audit$are_identical == TRUE)
diff_rows  <- which(pair_audit$are_identical == FALSE)

pair_audit |> 
  select(-are_identical) |> 
  kbl(col.names = c("Column A", "Column B", "Result"),
      booktabs = TRUE) |> 
  kable_styling(full_width = TRUE) |> 
  row_spec(ident_rows, background = cb_light[1], color = "black") |> 
  row_spec(diff_rows, background = cb_light[2], color = "black")
```

:::{.callout-note collapse="true"}
### Actions Taken: Duplicate name resolution

**Issue:** The raw Excel file contained several columns with identical or mislabeled headers.

**Resolved:** [2025-11-16] Renamed/deleted duplicates in `01_data_cleaning.qmd` (Chunk: `fix_duplicate-column-names`).

* `ANC.3...108` and `ANC.3...109` were identical. 
    * `ANC.3...109` was removed, and the remaining column was renamed to `ANC.3`.
* Two columns were named `HyperK+.3`, with no `HypoK+.3` present. 
    * The first instance was renamed to `HypoK+.3`.
* A block of variables (`Febrile.Neutropenia`, `HypoNa+`, `HypoK+`, `HyperK+`) appeared with a `.6` suffix in the Cycle 7 section. These were distinct from the actual Cycle 6 data.
    * The variables were renamed to fit in Cycle 7. (Dr. Hendricks stated that Dr. Mazhindu told her this was correct.)
:::


```{r}
#| label: fix_duplicate-column-names

gic_renamed <- gic_raw |> 
  # Fix identically names columns
  rename(
    "ANC.3" = "ANC.3...108",
    "HypoK+.3" = "HyperK+.3...114",
    "HyperK+.3" = "HyperK+.3...115",
    "Febrile.Neutropenia.6" = "Febrile.Neutropenia.6...189",
    "Febrile.Neutropenia.7" = "Febrile.Neutropenia.6...215",
    "HypoNa+.6" = "HypoNa+.6...191",
    "HypoNa+.7" = "HypoNa+.6...217",
    "HypoK+.6" = "HypoK+.6...192",
    "HypoK+.7" = "HypoK+.6...218",
    "HyperK+.6" = "HyperK+.6...193",
    "HyperK+.7" = "HyperK+.6...219"
  ) |> 
  select(
    - "ANC.3...109"
  )
```


## Check columns that have identical contents

```{r}
#| label: duplicate-column-contents

# Check for variables that have the exact same contents
dupe_cont <- names(gic_renamed)[duplicated(as.list(gic_renamed))]

duplicate_map <- map_dfr(dupe_cont, function(d_name) {
  match_idx <- detect_index(gic_renamed, ~ identical(.x, gic_renamed[[d_name]]))
  
  tibble(
    duplicate_column = d_name,
    original_source = names(gic_renamed)[match_idx]
  )
})

# List duplications
duplicate_map |> 
  kbl(col.names = c("Duplicate Column", "Original Source")) |> 
  kable_styling() |> 
  scroll_box(height = "400px")
```

:::{.callout-warning collapse="false"}
### Anomalies flagged for validation

**Issues:** The following issues were observed and require clarification or investigation in the validation phase:

* No patients have kidney disease, COPD, or epilepsy recorded. 
    * Check for accuracy
* `Albumin.bilirubin..score`, `Baseline.Creatinine.clearance`, and `Dbil.12` are completely empty.
* `HypoNa+.1` and `HypoK+.1`/`HyperK+.1` are identical to their respective `Baseline` counterparts.
* `blood.units.needed` and `Blood.transfusion` are identical for cycles 4, 7, 8, 9, and 11. 
    * Implies all transfusions were exactly 1 unit.
* `Hospitalization...10` is identical to `Was.the.cycle.delayed...10`.
    * Possibly valid, I'm not sure of the exact relationship though
:::

:::{.callout-note collapse="true"}
### Actions Taken: Remove redundant data

**Issues:** Several columns identified as duplicates or artifacts.

**Resolved:** [2025-11-16] The columns were removed in `01_data_cleaning.qmd` (Chunk: `fix_duplicate-column-contents`).

* `Febrile.Neutropenia` (duplicate of `ANC`) and `HyperK+` (duplicate of `HypoK+`) removed for all cycles.
* `Hypernatramia.Na+.3` (identical to `Hyponatramia.Na+.3`) removed.
* All `Cycle #` spacer columns removed.
:::

```{r}
#| label: fix_duplicate-column-contents

# Remove `Cycle` columns, and redundant `Febrile.Neutropenia` and `HyperK+` columns
gic_renamed <- gic_renamed |> 
  select(
    -starts_with("Febrile.Neutropenia"),
    -starts_with("HyperK+"),
    -`Hypernatramia.Na+.3`,
    -starts_with("Cycle")
  )
```

## Check consistency of cycle-specific variables
```{r}
#| label: cycle-var_consistency

# Filter for names ending in a cycle number
cycle_vars <- names(gic_renamed)[str_detect(names(gic_renamed), "\\.\\d{1,2}$")]

# Remove `Sex` variable from consideration since it was encoded similarly
cycle_vars <- cycle_vars[cycle_vars != "Sex.Male.1/Female.2"]

# Build function to audit the names
create_cycle_audit <- function(cycle_vars) {
  # Extract variable stems and cycles
  parsed_data <- tibble(raw_name = cycle_vars) |> 
    extract(raw_name,
            into = c("stem", "cycle"),
            regex = "(.*)[\\._](\\d+)$",
            remove = FALSE) |> 
    mutate(cycle = as.integer(cycle))
  
  # Report if any cycles aren't in [1, 12] or baseline/0
  out_of_bounds <- parsed_data |> 
    filter(cycle < 0 | cycle > 12 | is.na(cycle))
  
  if(nrow(out_of_bounds > 0)) {
    message("The following variables have unexpected cycle numbers:")
    print(out_of_bounds)    
  } else {
    message("All variables have valid cycle numbers.")
  }
  
  # Create the audit grid
  audit_grid <- parsed_data |> 
    filter(cycle >= 0, cycle <= 12) |> 
    select(stem, cycle) |> 
    distinct() |> 
    mutate(exists = 1) |> 
    pivot_wider(
      names_from = "cycle",
      values_from = "exists",
      values_fill = 0
    ) |> 
    arrange(stem) |> 
    rowwise() |> 
    mutate(
      all_ok = as.integer(all(c_across(any_of(as.character(1:12))) == 1))
    ) |> 
    ungroup() |> 
    select(stem,
           all_ok,
           any_of(as.character(0:12)))
  
  return(audit_grid)
}

# Function to generate table from audit output
format_audit_table <- function(grid_df) {
  
  has_baseline <- if("0" %in% names(grid_df)) grid_df$`0` == 1 else FALSE
  cycles_complete <- grid_df$all_ok == 1
  
  cycle_sum <- rowSums(grid_df[as.character(1:12)], na.rm = TRUE)
  cycles_empty <- cycle_sum == 0
  
  # Sort into cases
  ## Case A: Baseline + all cycles
  rows_complete <- which(has_baseline & cycles_complete)
  ## Case B: No baseline + all cycles
  rows_longitudinal <- which(!has_baseline & cycles_complete)
  ## Case C: Baseline, no cycles
  rows_baseline <- which(has_baseline & cycles_empty)
  ## Case D: Everything else
  rows_problem <- setdiff(1:nrow(grid_df), c(rows_complete, rows_longitudinal, rows_baseline))
  
  
  # Return the kable object
  grid_df |> 
    kbl(booktabs = TRUE) |> 
    kable_styling(full_width = FALSE) |> 
    row_spec(rows_complete,     background = cb_light[1]) |> 
    row_spec(rows_longitudinal, background = cb_light[5]) |> 
    row_spec(rows_baseline,     background = cb_light[6]) |> 
    row_spec(rows_problem,      background = cb_light[2])
}


grid_data <- create_cycle_audit(cycle_vars)
format_audit_table(grid_data)
```

:::{.callout-note collapse="true"}
### Actions Taken: Harmonize variable naming

**Issues:** Several cycle variables had minor inconsistencies, missing suffixes, or typos in their naming.

**Resolved:** [2025-11-28] Harmonized naming conventions in `01_data_cleaning.qmd` (Chunk: `fix_cycle-var-consistency`).

* `DBIL` for cycle 1, `Dbil` for the rest
    * Renamed all to `DBIL`
* `HB` for cycles 1-3, and 8, `Hb` for the rest
    * Renamed all to `HB`
* `Hospitalization.days` missing for cycle 2
    * Renamed `Hospitalization.days` to `Hospitalization.days.2`
* `Hospitalization.required.due.to.toxicity?` for cycle 1, but `Hospitalization.required.dur.to.toxicity?` for the rest
    * Renamed to fix "dur" typo where present
* `HypoNatramia.Na+` for cycles 2 and 3, but `HypoNa+` for the rest
    * Renamed all to `HypoNa+`
* `TBIL` for cycle 1, `TBil` for cycle 3, `Tbil` for the rest
    * Renamed all to `TBIL`
:::

```{r}
#| label: fix_cycle-var_consistency

gic_renamed <- gic_renamed |> 
  # Fix DBIL
  rename_with(
    .cols = matches("^dbil", ignore.case = TRUE),
    .fn = ~ str_replace(., regex("^dbil", ignore_case = TRUE), "DBIL")
  ) |> 
  
  # Fix HB
  rename_with(
    .cols = matches("^hb", ignore.case = TRUE),
    .fn = ~ str_replace(., regex("^hb", ignore_case = TRUE), "HB")
  ) |> 
  
  # Fix Hospitalization.days
  rename(`Hospitalization.days.2` = `Hospitalization.days`) |> 
  
  # Fix `Hospitalization.required.dur.to.toxicity?` typo
  rename_with(
    .cols = contains("dur.to.toxicity?"),
    .fn = ~ str_replace(., "dur\\.to", "due.to")
  ) |> 
  
  # Fix HypoNa+
  rename_with(
    .cols = matches("^(hypona|hyponatramia)", ignore.case = TRUE),
    .fn = ~ . |> 
      str_replace(regex("^hyponatramia\\.na\\+", ignore_case = TRUE), "HypoNa+") |> 
      str_replace(regex("^hypona\\+", ignore_case = TRUE), "HypoNa+")
  ) |> 
  
  # Fix TBIL
  rename_with(
    .cols = matches("^tbil", ignore.case = TRUE),
    .fn = ~ str_replace(., regex("^tbil", ignore_case = TRUE), "TBIL")
  )

# Recheck
cycle_vars <- names(gic_renamed)[str_detect(names(gic_renamed), "\\.\\d{1,2}$")]
cycle_vars <- cycle_vars[cycle_vars != "Sex.Male.1/Female.2"]
grid_data <- create_cycle_audit(cycle_vars)
format_audit_table(grid_data)
```

## Rename Variables

```{r}
#| label: rename

gic_renamed <- gic_renamed |> 
  # Name standardization
  rename(
    # Basic Variables
    hiv_status = "HIV.Status:.1,.Negative.|.2,.Positive",
    sex = "Sex.Male.1/Female.2",
    residence = "Primary.Residence.1.urban/.2.rural/3.peri-urban.",
    cancer_site = "Cancer.site:.1,.Esophagus.|.2,.Gastric.|.3,.Gastro-esophageal.junction.|.4,.Gall.bladder.,.Biliary.&.Pancreas.|.9,.Colorectal.|.11,.Anal.",
    histology = "Histology.type:.1,.adenocarcinoma/carcinoma.|.2,.Squamous.carcinoma.|.3,.Sarcoma.|.4,.Carcinoid.|.5,.Lymphoma.|.7,.Others",
    cancer_stage = "Cancer.stage:.stage.1,.I.|.2,.II.|.3,.III.|.4,.IV.|.4,.Unknown/Undocumented",
    
    # Comorbidities
    comorbid_htn = "Hypertension",
    comorbid_dm = "Diabetes.mellitus.",
    comorbid_tb = ".Tuberculosis",
    comorbid_asthma = "Asthma",
    comorbid_chd = "Cardiac.disease.",
    comorbid_copd = "COPD.",
    comorbid_ckd = "Kidney.disease.",
    comorbid_epilepsy = "Epilepsy",
    comorbid_none = "No.cormobidities",
    
    # History
    hx_etoh = "drink_alcohol_history_of",
    hx_tobacco = "tobacco_smoker_history_of",
    hx_cancer = "family_history_of_cancer",
    
    # Baseline Measurements
    weight_0 = "Weight.(kg)",
    height_0 = "Height.(cm)",
    bmi_0 = "Baseline.BMI",
    wbc_0 = "Baseline.WBC",
    anc_0 = "Baseline.ANC",
    plt_0 = "Baseline.PLT",
    cd4_count_0 = "Baseline.CD4.count",
    hb_0 = "Baseline.Hb",
    mcv_0 = "Baseline.MCV",
    mch_0 = "Baseline.MCH",
    sodium_0 = "Baseline.Na+",
    potassium_0 = "Baseline.K+",
    urea_0 = "Baseline.Urea",
    creat_0 = "Baseline.Creat",
    cr_clear_0 = "Baseline.Creatinine.clearance",
    tp_0 = "Baseline.TP",
    alb_0 = "Baseline.ALB",
    tbil_0 = "Baseline.Tbil",
    alp_0 = "Baseline.ALP",
    ast_0 = "Baseline.AST",
    alt_0 = "Baseline.ALT",
    ggt_0 = "Baseline.GGT",
    albi_score_0 = "Albumin.bilirubin..score",
    cea_0 = "Baseline.CEA",
    ca199_0 = "Baseline.CA19_9",
    
    # Treatment Info
    tx_concurrent_rtx = "chemotherapy_given_with.concurrent.radiotherapy?",
    tx_intent = "Treatment.intention-.1,.Curative.|.2,.Palliative",
    chemo_seq = "Sequence.of.chemotherapy-.1,.Neoadjuvant.|.2,.Adjuvant.|.3,.Definitive",
    chemo_type = "chemotherapy.given:.1,Platinum/Paclitaxel..|.2,.Platinum/5-FU..3,.LCV/5-FU.|.4,.CapeOX.|.5,.FOLFOX4.|.6,.Capecitabine.|.20,.Irinotecan.|.21,.Others",
    cycles_prescribed = "Total.number.chemo.cycles.prescribed",
    cycles_given = "Total.number.of.chemotherapy.administered",
    completion_rate = "Chemotherapy.completion.rates.(%)",
    death_date = "date_of_death_if_such_outc"
  ) |> 
  
  # Standardize names of cycle variables
  rename_with(
    .cols = matches("\\.\\d{1,2}$"),
    .fn = ~ .x |> 
      # Fix specific longer names
      str_replace("^Was\\.the\\.cycle\\.delayed[^0-9]*", "cycle_delayed.") |> 
      str_replace("^Hospitalization\\.required[^0-9]*", "hosp_toxicity.") |> 
      str_replace("^Hospitalization\\.days", "hosp_days") |> 
      str_replace("^Blood\\.transfusion", "transfusion_given") |> 
      str_replace("^blood\\.units\\.needed", "transfusion_units") |> 
      str_replace("^CSF-G\\.Indication", "csfg_indicated") |>
      str_replace("HypoNa\\+", "sodium") |> 
      str_replace("HypoK\\+", "potassium") |> 
      # Convert everything to snake case
      tolower() |> 
      str_replace("\\.(\\d{1,2})$", "_\\1")
  )
```

:::{.callout-note collapse="true"}
### Action Taken: Rename variables

**Issues:** Raw variable names contained factor levels and long strings.

**Resolved:** [2025-11-29] Converted all variables to standardized snake_case format in `01_data_cleaning.qmd` (Chunk: `rename`).

* Standardized naming conventions
    * Simplified names and converted all to snake case
    * Baseline measurements denoted with `_0` suffix
    * Cycle measurements denoted with `_#` suffix for the given cycle number
:::

## Renaming Audit
### Variable exists for each cycle
```{r}
#| label: check_rename-cycle-vars

# Recheck cycle vars
final_cycle_vars <- names(gic_renamed)[str_detect(names(gic_renamed), "\\_\\d{1,2}$")]
final_name_audit <- create_cycle_audit(final_cycle_vars)
format_audit_table(final_name_audit)
```

### Check for lost rows, incorrect deletions
```{r}
#| label: integrity-check_renaming_basic

# Row integrity
if(nrow(gic_raw) != nrow(gic_renamed)) {
  stop("Row count mismatch, rows were lost or added.")
} else {
  cat("Row count preserved:", nrow(gic_renamed), "patients.")
}

# Deletion verification
deleted_patterns <- c("Febrile.Neutropenia", "HyperK\\+", "Cycle \\d", "\\.\\.\\.\\d*")
found_trash <- map(deleted_patterns, ~ grep(., names(gic_renamed), value = TRUE))

if (length(unlist(found_trash)) > 0) {
  cat("\nSome columns intended for deletion are still present.")
} else {
  cat("\nAll redundant columns removed.")
}
```

### Check static variables against original names
```{r}
#| label: integrity-check_static-vars

# Static variable integrity verification
## Load Glossary
glossary_df <- read_xlsx(here("docs", "glossary.xlsx"))

## Find the start of the Longitudinal Section
longitudinal_start_index <- which(str_detect(glossary_df$Variable, "Longitudinal"))[1]

if (is.na(longitudinal_start_index)) {
  stop("Could not find 'Longitudinal' header.")
} else {
  static_vars <- glossary_df |> slice(1:(longitudinal_start_index - 1))
}

## Prepare checklist
check_list <- static_vars |> 
  filter(!str_detect(Variable, "^\\*\\*")) |> 
  select(clean_col = Variable, raw_col = `Original Name (Example)`)

## Run stationary var audit
static_audit <- check_list |> 
  rowwise() |> 
  mutate(
    # Check if clean version of var names exist
    clean_exists = clean_col %in% names(gic_renamed),
    
    # Get rid of backticks on variable names
    raw_col = str_remove_all(raw_col, "`"), 
    raw_exists = raw_col %in% names(gic_raw), 
    data_match = if (clean_exists && raw_exists) {
      identical(gic_raw[[raw_col]], gic_renamed[[clean_col]])
      } else {
        NA
        }) |> 
  ungroup() |> 
  mutate(
    Status = case_when(
      !clean_exists ~ "Clean var missing",
      !raw_exists   ~ "Raw var missing",
      data_match    ~ "Match",
      TRUE          ~ "Data mismatch"
    )
  )

## Define row indices for better visualization
static_green <- which(static_audit$Status == "Match")
static_red <- which(static_audit$Status != "Match")


## Display static audit
static_audit |> 
  arrange(Status) |> 
  select(clean_col, raw_col, Status) |> 
  kbl(booktabs = TRUE) |> 
  kable_styling(full_width = TRUE) |> 
  row_spec(static_green, background = cb_light[1]) |> 
  row_spec(static_red, background = cb_light[2]) |> 
  scroll_box(height = "400px")
```

### Check cycle variables against originals
```{r}
#| label: integrity-check_cycle-vars

# Cycle variable integrity validation
## Get all variables, both baseline and cyclic
clean_cycle_vars <- names(gic_renamed) |> str_subset("_[1-9]\\d*$")

## Create function to match against raw data/cols
find_raw_match <- function(clean_col_name) {
  clean_vec <- gic_renamed[[clean_col_name]]
  
  # Check for empty columns
  if (all(is.na(clean_vec))) {
    return(tibble(raw_match = "-", status = "All NA"))
  }
  
  # Check against every column in raw data
  matches <- names(gic_raw)[map_lgl(gic_raw, ~ identical(., clean_vec))]
  
  if (length(matches) == 0) {
    return(tibble(raw_match = NA_character_, status = "No match"))
  } else if (length(matches) == 1) {
    return(tibble(raw_match = matches, status = "Unique match"))
  } else {
    # If multiple raw columns match (like the deleted duplicates), list them
    match_list <- paste(matches, collapse = ", ")
    return(tibble(raw_match = match_list, status = "Multiple raw matches"))
  }
}

## Run cycle audit
cycle_audit <- tibble(clean_name = clean_cycle_vars) |> 
  rowwise() |> 
  mutate(result = list(find_raw_match(clean_name))) |> 
  unnest(result) |> 
  ungroup()

## Extract metadata for sorting
cycle_audit_final <- cycle_audit |> 
  extract(clean_name, into = c("stem", "cycle"), regex = "(.*)_(\\d+)$", remove = FALSE) |> 
  mutate(cycle = as.integer(cycle)) |> 
  arrange(stem, cycle)

## Define row indices for better visualization
cycle_good <- which(cycle_audit_final$status == "Unique match")
cycle_multiple <- which(cycle_audit_final$status == "Multiple raw matches")
cycle_empty <- which(cycle_audit_final$status == "All NA")
cycle_bad <- which(cycle_audit_final$status == "No match")


## Print cycle audit
cycle_audit_final |> 
  select(clean_name, raw_match, status) |> 
  kbl(booktabs = TRUE, col.names = c("Clean Name", "Matched Raw Column(s)", "Status")) |> 
  kable_styling(full_width = TRUE) |> 
  row_spec(cycle_good, background = cb_light[1]) |> 
  row_spec(cycle_multiple, background = cb_light[6]) |> 
  row_spec(cycle_empty, background = cb_light[8]) |> 
  row_spec(cycle_bad, background = cb_light[2]) |> 
  scroll_box(height = "400px")
```

# Save Cleaned Data
```{r}
#| label: save-data

saveRDS(gic_renamed, here("data", "processed", "gic_renamed.rds"))
```
