---
title: "GI Cancer: Data Cleaning"
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false

source("R/setup.R")

# Packages
library(readxl)    # Read xlsx files
```

```{r}
#| label: load-data

gic_raw <- read_xlsx(here("data", "IPROTECTARetrospecti_DATA_2025-08-19_2157 FINAL SHEET.xlsx"))
```

# Data Cleaning

## Check columns that had the same names in the excel file

```{r}
#| label: duplicate-column-names

dupe_names <- gic_raw |> 
  select(contains("...")) |> 
  names()
print(dupe_names)

# `ANC.3...108`/`ANC.3...109`
identical(gic_raw$`ANC.3...108`, gic_raw$`ANC.3...109`)
## Identical. `ANC.3...108` -> `ANC.3`, delete `ANC.3...109`


# `HyperK+.3...114`/`HyperK+.3...115`
identical(gic_raw$`HyperK+.3...114`, gic_raw$`HyperK+.3...115`)
## Identical: `HyperK+.3...114` -> `HypoK+.3`, `HyperK+.3...115` -> `HyperK+.3`


# `Febrile.Neutropenia.6...189`/`Febrile.Neutropenia.6...215`
identical(gic_raw$`Febrile.Neutropenia.6...189`, gic_raw$`Febrile.Neutropenia.6...215`)
## Not identical: `Febrile.Neutropenia.6...215` -> `Febrile.Neutropenia.7`, `Febrile.Neutropenia.6...189` -> `Febrile.Neutropenia.6`


# `HypoNa+6...191`/`HypoNa+.6...217`
identical(gic_raw$`HypoNa+6...191`, gic_raw$`HypoNa+.6...217`)
## Not identical: `HypoNa+6...191` -> `HypoNa+6`, `HypoNa+.6...217` -> `HypoNa+.7`


# `HypoK+.6...192`/`HypoK+.6...218`
identical(gic_raw$`HypoK+.6...192`, gic_raw$`HypoK+.6...218`)
## Not identical: `HypoK+.6...192` -> `HypoK+.6`, `HypoK+.6...218` -> `HypoK+.7`


# `HyperK+.6...193`/`HyperK+.6...219`
identical(gic_raw$`HyperK+.6...193`, gic_raw$`HyperK+.6...219`)
## Not identical: `HyperK+.6...193` -> `HyperK+.6`, `HyperK+.6...219` -> `HyperK+.7`
```

```{r}
#| label: duplicate-column-names-imp-changes

gic_tmp <- gic_raw |> 
  rename(
    `ANC.3` = `ANC.3...108`,
    `HypoK+.3` = `HyperK+.3...114`,
    `HyperK+.3` = `HyperK+.3...115`,
    `Febrile.Neutropenia.6` = `Febrile.Neutropenia.6...189`,
    `Febrile.Neutropenia.7` = `Febrile.Neutropenia.6...215`,
    `HypoNa+.6` = `HypoNa+.6...191`,
    `HypoNa+.7` = `HypoNa+.6...217`,
    `HypoK+.6` = `HypoK+.6...192`,
    `HypoK+.7` = `HypoK+.6...218`,
    `HyperK+.6` = `HyperK+.6...193`,
    `HyperK+.7` = `HyperK+.6...219`
  ) |> 
  select(
    - `ANC.3...109`,
  )
```


## Check columns that have identical contents

```{r}
#| label: duplicate-column-contents


dupe_cont <- names(gic_tmp)[duplicated(as.list(gic_tmp))]
duplicate_map <- map_dfr(dupe_cont, function(d_name) {
  match_idx <- detect_index(gic_tmp, ~ identical(.x, gic_tmp[[d_name]]))
  
  tibble(
    duplicate_column = d_name,
    original_source = names(gic_tmp)[match_idx]
  )
})
```

-   `r duplicate_map[[1, 1]]` and `r duplicate_map[[2, 1]]` are the same as `r duplicate_map[[1, 2]]`
    -   No patients had any of these comorbidities. Likely valid, but worth checking.
-   `r duplicate_map[[4, 1]]` is the same as `r duplicate_map[[4, 2]]`
    -   Same for `r duplicate_map[[8, 1]]`/`r duplicate_map[[8, 2]]`;       <!-- Cycle 2 -->
                 `r duplicate_map[[10, 1]]`/`r duplicate_map[[10, 2]]`;     <!-- Cycle 3 -->
                 `r duplicate_map[[14, 1]]`/`r duplicate_map[[14, 2]]`;     <!-- Cycle 4 -->
                 `r duplicate_map[[18, 1]]`/`r duplicate_map[[18, 2]]`;     <!-- Cycle 5 -->
                 `r duplicate_map[[21, 1]]`/`r duplicate_map[[21, 2]]`;     <!-- Cycle 6 -->
                 `r duplicate_map[[24, 1]]`/`r duplicate_map[[24, 2]]`;     <!-- Cycle 7 -->
                 `r duplicate_map[[28, 1]]`/`r duplicate_map[[28, 2]]`;     <!-- Cycle 8 -->
                 `r duplicate_map[[32, 1]]`/`r duplicate_map[[32, 2]]`;     <!-- Cycle 9 -->
                 `r duplicate_map[[36, 1]]`/`r duplicate_map[[36, 2]]`;     <!-- Cycle 10 -->
                 `r duplicate_map[[40, 1]]`/`r duplicate_map[[40, 2]]`;     <!-- Cycle 11 -->
                 `r duplicate_map[[44, 1]]`/`r duplicate_map[[44, 2]]`      <!-- Cycle 12 -->
-   `r duplicate_map[[5, 1]]` is the same as `r duplicate_map[[5, 2]]`
-   `r duplicate_map[[6, 1]]` and `r duplicate_map[[7, 1]]` are the same as `r duplicate_map[[6, 2]]`
-   `r duplicate_map[[6, 1]]` is the same as `r duplicate_map[[7, 1]]`
    -   Same for `r duplicate_map[[9, 1]]`/`r duplicate_map[[9, 2]]`;       <!-- Cycle 2 -->
                 `r duplicate_map[[12, 1]]`/`r duplicate_map[[12, 2]]`;     <!-- Cycle 3 -->
                 `r duplicate_map[[15, 1]]`/`r duplicate_map[[15, 2]]`;     <!-- Cycle 4 -->
                 `r duplicate_map[[19, 1]]`/`r duplicate_map[[19, 2]]`;     <!-- Cycle 5 -->
                 `r duplicate_map[[22, 1]]`/`r duplicate_map[[22, 2]]`;     <!-- Cycle 6 -->
                 `r duplicate_map[[25, 1]]`/`r duplicate_map[[25, 2]]`;     <!-- Cycle 7 -->
                 `r duplicate_map[[29, 1]]`/`r duplicate_map[[29, 2]]`;     <!-- Cycle 8 -->
                 `r duplicate_map[[33, 1]]`/`r duplicate_map[[33, 2]]`;     <!-- Cycle 9 -->
                 `r duplicate_map[[37, 1]]`/`r duplicate_map[[37, 2]]`;     <!-- Cycle 10 -->
                 `r duplicate_map[[41, 1]]`/`r duplicate_map[[41, 2]]`;     <!-- Cycle 11 -->
                 `r duplicate_map[[45, 1]]`/`r duplicate_map[[45, 2]]`      <!-- Cycle 12 -->
-   `r duplicate_map[[11, 1]]` is the same as `r duplicate_map[[11, 2]]`
-   All of the following columns are completely empty: 
                                 `r duplicate_map[[3, 1]]`,                 <!-- Almbumin.bilirubin..score -->
                                 `r duplicate_map[[3, 2]]`,                 <!-- Baseline.Creatinine.Clearance -->
                                 `r duplicate_map[[13, 1]]`,                <!-- Cycle 4 -->
                                 `r duplicate_map[[17, 1]]`,                <!-- Cycle 5 -->
                                 `r duplicate_map[[20, 1]]`,                <!-- Cycle 6 -->
                                 `r duplicate_map[[23, 1]]`,                <!-- Cycle 7 -->
                                 `r duplicate_map[[27, 1]]`,                <!-- Cycle 8 -->
                                 `r duplicate_map[[31, 1]]`,                <!-- Cycle 9 -->
                                 `r duplicate_map[[35, 1]]`,                <!-- Cycle 10 -->
                                 `r duplicate_map[[39, 1]]`,                <!-- Cycle 11 -->
                                 `r duplicate_map[[43, 1]]`,                <!-- Cycle 12 -->
                                 `r duplicate_map[[46, 1]]`                 <!-- Dbil.12 -->
    - Will get rid of the "Cycle" variables as they're just labels, the others are worth checking into.
-   `r duplicate_map[[16, 1]]` is the same as `r duplicate_map[[16, 2]]`    <!-- Cycle 4 -->
    - Same for `r duplicate_map[[26, 1]]`/`r duplicate_map[[26, 2]]`,         <!-- Cycle 7 -->
             `r duplicate_map[[30, 1]]`/`r duplicate_map[[30, 2]]`,         <!-- Cycle 8 -->
             `r duplicate_map[[34, 1]]`/`r duplicate_map[[34, 2]]`,         <!-- Cycle 9 -->
             `r duplicate_map[[42, 1]]`/`r duplicate_map[[42, 2]]`,         <!-- Cycle 11 -->
    - This is possibly valid, but worth double checking
-   `r duplicate_map[[38, 1]]` is the same as `r duplicate_map[[38, 2]]`,         <!-- Cycle 10 -->
    - Possibly valid, I'm not sure of the exact relationship though
```{r}
#| label: duplicate-column-contents-imp-changes

# Remove `Cycle` columns, and redundant `Febrile.Neutropenia` and `HyperK+` columns
gic_tmp <- gic_tmp |> 
  select(
    -starts_with("Febrile.Neutropenia"),
    -starts_with("HyperK+"),
    -`Hypernatramia.Na+.3`,
    -starts_with("Cycle")
  )
```

## Standardize naming
```{r}
#| label: name-standardization-bm-search

## Biomarkers
# Get all colum names from data
all_names <- names(gic_tmp)

# Filter for names ending in a cycle number
cycle_vars <- all_names[str_detect(all_names, "\\.\\d{1,2}$")]

# Remove Sex variable since it was encoded similarly
cycle_vars <- cycle_vars[cycle_vars != "Sex.Male.1/Female.2"]

# Build functions to audit the names and print table indicating naming consistency
create_audit_grid <- function(cycle_vars) {
  
  audit_grid <- tibble(raw_name = cycle_vars) |> 
    extract(raw_name,
            into = c("stem", "cycle"),
            regex = "(.*)\\.(\\d{1,2})$",
            remove = FALSE) |> 
    mutate(cycle = as.integer(cycle)) |> 
    filter(cycle >= 1, cycle <= 12) |> 
    select(stem, cycle) |> 
    distinct() |> 
    mutate(exists = 1) |> 
    pivot_wider(
      names_from = "cycle",
      values_from = "exists",
      values_fill = 0
    ) |> 
    arrange(stem) |> 
    rowwise() |> 
    mutate(
      all_ok = as.integer(all(c_across(any_of(as.character(1:12))) == 1))
    ) |> 
    ungroup() |> 
    select(stem,
           all_ok,
           any_of(as.character(1:12)))
  
  # Return the dataframe
  return(audit_grid)
}

format_audit_table <- function(grid_df) {
  
  green_rows <- which(grid_df$all_ok == 1)
  red_rows   <- which(grid_df$all_ok == 0)
  
  # Return the kable object
  grid_df |> 
    kbl(booktabs = TRUE) |> 
    kable_styling(full_width = FALSE) |> 
    row_spec(green_rows, background = "#D4EFDF") |> 
    row_spec(red_rows,   background = "#FADBD8")
}

grid_data <- create_audit_grid(cycle_vars)
format_audit_table(grid_data)
```

Noted issues:

- `DBIL` for cycle 1, `Dbil` for the rest
- `HB` for cycles 1-3, and 8, `Hb` for the rest
- `Hospitalization.days` missing for cycle 2
- `Hospitalization.required.due.to.toxicity?` for cycle 1, but `Hospitalization.required.dur.to.toxicity?` for the rest
- `HypoNatramia.Na+` for cycles 2 and 3, but `HypoNa+` for the rest
- `TBIL` for cycle 1, `TBil` for cycle 3, `Tbil` for the rest


```{r}
#| label: name-standardization-bm-imp-changes

gic_tmp <- gic_tmp |> 
  # Fix DBIL
  rename_with(
    .cols = matches("^dbil", ignore.case = TRUE),
    .fn = ~ str_replace(., regex("^dbil", ignore_case = TRUE), "DBIL")
  ) |> 
  
  # Fix HB
  rename_with(
    .cols = matches("^hb", ignore.case = TRUE),
    .fn = ~ str_replace(., regex("^hb", ignore_case = TRUE), "HB")
  ) |> 
  
  # Fix Hospitalization.days
  rename(`Hospitalization.days.2` = `Hospitalization.days`) |> 
  
  # Fix `Hospitalization.required.dur.to.toxicity?` typo
  rename_with(
    .cols = contains("dur.to.toxicity?"),
    .fn = ~ str_replace(., "dur\\.to", "due.to")
  ) |> 
  
  # Fix HypoNa+
  rename_with(
    .cols = matches("^(hypona|hyponatramia)", ignore.case = TRUE),
    .fn = ~ . |> 
      str_replace(regex("^hyponatramia\\.na\\+", ignore_case = TRUE), "HypoNa+") |> 
      str_replace(regex("^hypona\\+", ignore_case = TRUE), "HypoNa+")
  ) |> 
  
  # Fix TBIL
  rename_with(
    .cols = matches("^tbil", ignore.case = TRUE),
    .fn = ~ str_replace(., regex("^tbil", ignore_case = TRUE), "TBIL")
  )

# Recheck
all_names <- names(gic_tmp)
cycle_vars <- all_names[str_detect(all_names, "\\.\\d{1,2}$")]
cycle_vars <- cycle_vars[cycle_vars != "Sex.Male.1/Female.2"]
grid_data <- create_audit_grid(cycle_vars)
format_audit_table(grid_data)
```


```{r}
#| label: encode-factor-vars

gic_tmp <- gic_tmp |> 
  # Name standardization
  rename(
    # Factor variables
    hiv_status = "HIV.Status:.1,.Negative.|.2,.Positive",
    sex = "Sex.Male.1/Female.2",
    residence = "Primary.Residence.1.urban/.2.rural/3.peri-urban.",
    cancer_site = "Cancer.site:.1,.Esophagus.|.2,.Gastric.|.3,.Gastro-esophageal.junction.|.4,.Gall.bladder.,.Biliary.&.Pancreas.|.9,.Colorectal.|.11,.Anal.",
    histology = "Histology.type:.1,.adenocarcinoma/carcinoma.|.2,.Squamous.carcinoma.|.3,.Sarcoma.|.4,.Carcinoid.|.5,.Lymphoma.|.7,.Others",
    cancer_stage = "Cancer.stage:.stage.1,.I.|.2,.II.|.3,.III.|.4,.IV.|.4,.Unknown/Undocumented",
    tx_intent = "Treatment.intention-.1,.Curative.|.2,.Palliative",
    chemo_seq = "Sequence.of.chemotherapy-.1,.Neoadjuvant.|.2,.Adjuvant.|.3,.Definitive",
    chemo_type = "chemotherapy.given:.1,Platinum/Paclitaxel..|.2,.Platinum/5-FU..3,.LCV/5-FU.|.4,.CapeOX.|.5,.FOLFOX4.|.6,.Capecitabine.|.20,.Irinotecan.|.21,.Others",
    
    # Comorbidities
    comorbid_htn = "Hypertension",
    comorbid_dm = "Diabetes.mellitus.",
    comorbid_tb = ".Tuberculosis",
    comorbid_asthma = "Asthma",
    comorbid_chd = "Cardiac.disease.",
    comorbid_copd = "COPD.",
    comorbid_ckd = "Kidney.disease.",
    comorbid_epilepsy = "Epilepsy",
    comorbid_none = "No.cormobidities",
    
    # History
    hx_etoh = "drink_alcohol_history_of",
    hx_tobacco = "tobacco_smoker_history_of",
    hx_cancer = "family_history_of_cancer",
    
    # Baseline Measurements
    weight = "Weight.(kg)",
    height = "Height.(cm)",
    bmi_0 = "Baseline.BMI",
    wbc_0 = "Baseline.WBC",
    anc_0 = "Baseline.ANC",
    plt_0 = "Baseline.PLT",
    cd4_count_0 = "Baseline.CD4.count",
    hb_0 = "Baseline.Hb",
    mcv_0 = "Baseline.MCV",
    mch_0 = "Baseline.MCH",
    sodium_0 = "Baseline.Na+",
    potassium_0 = "Baseline.K+",
    urea_0 = "Baseline.Urea",
    creat_0 = "Baseline.Creat",
    cr_clear_0 = "Baseline.Creatinine.clearance",
    tp_0 = "Baseline.TP",
    alb_0 = "Baseline.ALB",
    tbil_0 = "Baseline.Tbil",
    alp_0 = "Baseline.ALP",
    ast_0 = "Baseline.AST",
    alt_0 = "Baseline.ALT",
    ggt_0 = "Baseline.GGT",
    albi_score_0 = "Albumin.bilirubin..score",
    cea_0 = "Baseline.CEA",
    ca199_0 = "Baseline.CA19_9",
    
    # Treatment info
    tx_concurent_rtx = "chemotherapy_given_with.concurrent.radiotherapy?",
    cycles_prescribed = "Total.number.chemo.cycles.prescribed",
    cycles_given = "Total.number.of.chemotherapy.administered",
    completion_rate_pct = "Chemotherapy.completion.rates.(%)"
  ) |> 
  
  # Encode factor variables correctly
  mutate(
    hiv_status = factor(hiv_status,
                        levels = c(1, 2), labels = c("Negative", "Positive")),
    sex = factor(sex,
                 levels = c(1, 2), labels = c("Male", "Female")),
    residence = factor(residence,
                       levels = c(1, 2, 3), labels = c("Urban", "Rural", "Peri-urban")),
    cancer_site = factor(cancer_site,
                         levels = c(1, 2, 3, 4, 9, 11), 
                         labels = c("Esophageal", "Gastric", "Gastro-esophageal Junction", "Gall Bladder, Biliary & Pancreas", "Colorectal", "Anal")),
    histology = factor(histology,
                       levels = c(1, 2, 3, 4, 5, 7), 
                       labels = c("Adenocarcinoma/carcinoma", "Squamous carcinoma", "Sarcoma", "Carcinoid", "Lymphoma", "Other")),
    cancer_stage = factor(cancer_stage,
                          levels = c(1, 2, 3, 4),
                          labels = c("I", "II", "III", "IV/Unknown")),
    tx_intent = factor(tx_intent,
                       levels = c(1, 2),
                       labels = c("Curative", "Palliative")),
    chemo_seq = factor(chemo_seq,
                       levels = c(1, 2, 3),
                       labels = c("Neoadjuvant", "Adjuvant", "Definitive")),
    chemo_type = factor(chemo_type,
                        levels = c(1, 2, 3, 4, 5, 6, 20, 21),
                        labels = c("Platinum/Paclitaxel", "Platinum/5-FU", "LCV/5-FU", "CapeOX", "FOLFOX4", "Capecitabine", "Irinotecan", "Other"))
  )
```





# Notes
Things I have noted and changed:

- The `ANC.3...108` and `ANC.3...109` columns were identical, I assume due to accidental duplication.
    - Removed `ANC.3...109`, renamed `ANC.3...108` to `ANC.3`
- There were two identical `HyperK+.3` columns and no `HypoK+.3` column for cycle 3
    - Renamed the first to be `HypoK+.3`
- In with all of the cycle 7 biomarker measurements, the `Febrile.Neutropenia`, `HypoNa+`, `HypoK+`, and `HyperK+` columns were labelled as if they belonged to cycle 6. They were not identical to their cycle 6 counterparts, so I assume this was a numbering mistake
    - Renamed them to reflect that they belong in cycle 7
- The `ANC` and `Febrile.Neutropenia` columns within a given cycle were all identical (e.g., `ANC.3` was identical to `Febrile.Neutropenia.3`)
    - Removed all of the `Febrile.Neutropenia` columns
- The `HypoK+` and `HyperK+` columns within a given cycle were all identical (e.g., `HypoK+.3` was identical to `HyperK+.3`)
    - Removed all of the `HyperK+` columns and renamed all of the `HypoK+` columns to `K+`
- Cycle 3 included `Hypernatramia.Na+.3` and `Hyponatramia.Na+.3` which were identical
    - Removed `Hypernatrama.Na+.3`
- There were several columns that were empty other than the cycle name (e.g., `Cycle 5`)
    - Removed all of these columns
- Standardized naming conventions
    - `DBIL`/`Dbil` -> `DBIL`
    - `HB`/`Hb` -> `HB`
    - `Hospitalization.days` -> `Hospitalization.days.1` for cycle 1
    - Fixed `Hospitalization.required.dur.to.toxicity?`typo
    - `HypoNatramia.Na+`/`HypoNa+` -> `HypoNa+`
    - `TBIL`/`TBil`/`Tbil` -> `TBIL`

Things I want to address before I change anything:

- There are no patients marked as having COPD, Kidney disease, or epilepsy
    - I assume this is correct, but want to check that there wasn't some sort of accidental duplication error
- `HypoNa+.1` is the same as `Baseline.Na+` and `HypoK+.1` is the same as `Baseline.K+`
    - Need to verify if this is correct, since none of the other baseline biomarkers are the same as their cycle 1 counterparts
- For Cancer Stage, based on the column name it looks like both stage IV cancer and unknown/undocumented stage are both denoted by 4


Things I need to dig into before I know how to handle them:

- `Dbil12` is empty. I'm pretty sure it's odd for just one of a set of biomarker measurements to be missing
- For cycles 4, 7, 8, 9, and 11 the `Blood.transfusion` and `blood.units.needed` columns are identical within a given column
    - I assume this is correct, but wanted to check because it implies that every person who received a blood transfusion in those cycles was given exactly 1 unit of blood.
    - Look at a distribution of blood units given by cycle to see how plausible this is
- `Hospitalization.requred.dur.to.toxicity?.10` is identical to `Was.the.cycle.delayed.due..to.toxicity.10`
    - Not sure if this is reasonable or not, need to look at others and compare.


# Changes To Incorporate Into Notes
Converting factor variables, renaming non-cycle variables, standardizing cycle variables
