---
title: "GI Cancer: Data Cleaning"
bibliography: references.bib
---
# Load Data and Libraries
```{r}
#| label: setup
#| include: false

source("R/setup.R")

# Packages
library(readxl)     # Read xlsx files
library(lubridate)  # Handle date data
```

```{r}
#| label: load-data

gic_raw <- read_xlsx(here("data", "IPROTECTARetrospecti_DATA_2025-08-19_2157 FINAL SHEET.xlsx"))
```

```{r}
#| label: apply-manual-patches

gic_raw <- gic_raw |> 
  mutate(
    # [DAT-017] Fix invalid sex level (3 -> 1)
    `Sex.Male.1/Female.2` = if_else(record_id == 94, 1, `Sex.Male.1/Female.2`),
    
    # [DAT-016] Fix AST1 character string (pn -> 28)
    `AST.1` = if_else(record_id == 64, "28", `AST.1`),
    
    # [DAT-020] Fix height decimal error (1498 -> 149.8)
    `Height.(cm)` = if_else(record_id == 614, 149.8, `Height.(cm)`),
    `Baseline.BMI` = if_else(record_id == 614, `Weight.(kg)` / (`Height.(cm)` / 100)^2, `Baseline.BMI`),
    
    # [DAT-008] Apply transfusion misentry fixes from table provided by Dr. Mazhindu
    `blood.units.needed.3` = if_else(record_id == 160, 1, `blood.units.needed.3`),
    `Blood.transfusion.1` = if_else(record_id == 238, 1, `Blood.transfusion.1`),
    `Blood.transfusion.10` = if_else(record_id == 437, 0, `Blood.transfusion.10`),
    `Blood.transfusion.3` = if_else(record_id == 454, 0, `Blood.transfusion.3`),
    `Blood.transfusion.2` = if_else(record_id == 505, 1, `Blood.transfusion.2`),
    `Blood.transfusion.5` = if_else(record_id == 567, 1, `Blood.transfusion.5`),
    
    # [DAT-019] Remove erroneous cycle 12 data,  match cycles given to match data presence, update completion rate
    across(ends_with(".12"), ~ if_else(record_id == 146, NA, .x)),
    `Total.number.of.chemotherapy.administered` = if_else(record_id == 27, 6, `Total.number.of.chemotherapy.administered`),
    `Chemotherapy.completion.rates.(%)` = if_else(record_id == 27, 100 * `Total.number.of.chemotherapy.administered` / `Total.number.chemo.cycles.prescribed`, `Chemotherapy.completion.rates.(%)`),
    `Total.number.of.chemotherapy.administered` = if_else(record_id == 498, 8, `Total.number.of.chemotherapy.administered`),
    `Chemotherapy.completion.rates.(%)` = if_else(record_id == 498, 100 * `Total.number.of.chemotherapy.administered` / `Total.number.chemo.cycles.prescribed`, `Chemotherapy.completion.rates.(%)`),
    `Total.number.of.chemotherapy.administered` = if_else(record_id == 766, 2, `Total.number.of.chemotherapy.administered`),
    `Chemotherapy.completion.rates.(%)` = if_else(record_id == 766, 100 * `Total.number.of.chemotherapy.administered` / `Total.number.chemo.cycles.prescribed`, `Chemotherapy.completion.rates.(%)`)
  )
```


# Structural Cleaning
## Duplicate Column Names
```{r}
#| label: assess-duplicate-names

# Get list of names which were identical in excel file
dupe_names <- gic_raw |> 
  select(contains("...")) |> 
  names()

# Create tibble of names to compare from `dupe_names`
check_pairs <- tibble(
  col_a = c(
    "ANC.3...108",
    "HyperK+.3...114",
    "Febrile.Neutropenia.6...189",
    "HypoNa+.6...191",
    "HypoK+.6...192",
    "HyperK+.6...193"
  ),
  col_b = c(
    "ANC.3...109",
    "HyperK+.3...115",
    "Febrile.Neutropenia.6...215",
    "HypoNa+.6...217",
    "HypoK+.6...218",
    "HyperK+.6...219"
  )
)

# Compare to see if contents are identical
pair_audit <- check_pairs |> 
  rowwise() |> 
  mutate(
    are_identical = identical(gic_raw[[col_a]], gic_raw[[col_b]]),
    status = if_else(are_identical,
                     "Identical (Redundant)",
                     "Different (Distinct)")
  ) |> 
  ungroup()

# Create table
ident_rows <- which(pair_audit$are_identical == TRUE)
diff_rows  <- which(pair_audit$are_identical == FALSE)

pair_audit |> 
  select(-are_identical) |> 
  kbl(col.names = c("Column A", "Column B", "Result"),
      booktabs = TRUE) |> 
  kable_styling(full_width = TRUE) |> 
  row_spec(ident_rows, background = cb_light[1], color = "black") |> 
  row_spec(diff_rows, background = cb_light[2], color = "black")
```

:::{.callout-note collapse="false"}
### Actions Taken: Duplicate name resolution (DAT-001, DAT-021, DAT-022)

**Summary:** The raw Excel file contained several columns with identical or mislabeled headers, requiring manual renaming before processing.

**Resolved:** [2025-11-16] in `01_data_cleaning.qmd` (Chunk: `fix_duplicate-column-names`).

* **ANC Duplication (DAT-001):** `ANC.3...108` and `ANC.3...109` were identical.
    * `ANC.3...109` was removed, and the remaining column was renamed to `ANC.3`.
* **Potassium Naming (DAT-021):** Two columns were named `HyperK+.3`, with no `HypoK+.3` present.
    * The first instance was renamed to `HypoK+.3`.
* **Cycle 7 Mislabeling (DAT-022):** A block of variables (`Febrile.Neutropenia`, `HypoNa+`, etc.) appeared with a `.6` suffix in the Cycle 7 section.
    * The variables were renamed to fit in Cycle 7. (Verified by Dr. Hendricks/Dr. Mazhindu).
:::


```{r}
#| label: fix-duplicate-names

gic_renamed <- gic_raw |> 
  # Fix identically named columns
  rename(
    # [DAT-001] Fix duplicate ANC.3 column, keep `ANC.3...108` version
    "ANC.3" = "ANC.3...108",
    
    # [DAT-021] Fix duplicate Potassium variable names
    "HypoK+.3" = "HyperK+.3...114",
    "HyperK+.3" = "HyperK+.3...115",
    
    # [DAT-022] Fix cycle 7 variable mislabeling
    "Febrile.Neutropenia.6" = "Febrile.Neutropenia.6...189",
    "Febrile.Neutropenia.7" = "Febrile.Neutropenia.6...215",
    "HypoNa+.6" = "HypoNa+.6...191",
    "HypoNa+.7" = "HypoNa+.6...217",
    "HypoK+.6" = "HypoK+.6...192",
    "HypoK+.7" = "HypoK+.6...218",
    "HyperK+.6" = "HyperK+.6...193",
    "HyperK+.7" = "HyperK+.6...219"
  ) |> 
  select(
    # [DAT-001] Remove redundant ANC.3 column
    - "ANC.3...109"
  )
```


## Redundant Data Content
```{r}
#| label: assess-duplicate-content

# Check for variables that have the exact same contents
dupe_cont <- names(gic_renamed)[duplicated(as.list(gic_renamed))]

duplicate_map <- map_dfr(dupe_cont, function(d_name) {
  match_idx <- detect_index(gic_renamed, ~ identical(.x, gic_renamed[[d_name]]))
  
  tibble(
    duplicate_column = d_name,
    original_source = names(gic_renamed)[match_idx]
  )
})

# List duplications
duplicate_map |> 
  kbl(col.names = c("Duplicate Column", "Original Source")) |> 
  kable_styling() |> 
  scroll_box(height = "400px")
```

:::{.callout-warning collapse="false"}
### Warning: Anomalies flagged for validation

**Issues:** The following issues were observed and require clarification or investigation in the validation phase:

* No patients have kidney disease, COPD, or epilepsy recorded. 
    * Check for accuracy
* `Albumin.bilirubin..score`, `Baseline.Creatinine.clearance`, and `Dbil.12` are completely empty.
* `HypoNa+.1` and `HypoK+.1`/`HyperK+.1` are identical to their respective `Baseline` counterparts.
* `blood.units.needed` and `Blood.transfusion` are identical for cycles 4, 7, 8, 9, and 11. 
    * Implies all transfusions were exactly 1 unit.
* `Hospitalization...10` is identical to `Was.the.cycle.delayed...10`.
    * Need to consult if this is accurate or a data entry duplication.
:::

:::{.callout-note collapse="false"}
### Actions Taken: Duplicate name resolution (DAT-001)

*Summary:* The raw Excel file contained several columns with identical or mislabeled headers.

**Resolved:** [2025-11-16] The columns were removed in `01_data_cleaning.qmd` (Chunk: `fix_duplicate-column-contents`).

* `Febrile.Neutropenia` (duplicate of `ANC`) and `HyperK+` (duplicate of `HypoK+`) removed for all cycles.
* `Hypernatramia.Na+.3` (identical to `Hyponatramia.Na+.3`) removed.
* All `Cycle #` spacer columns removed.
:::

```{r}
#| label: fix-duplicate-content

# [DAT-002] Remove unnecessary columns and cycle label columns
gic_renamed <- gic_renamed |> 
  select(
    -starts_with("Febrile.Neutropenia"),
    -starts_with("HyperK+"),
    -`Hypernatramia.Na+.3`,
    -starts_with("Cycle")
  )
```

## Longitudinal Variable Consistency
```{r}
#| label: assess-longitudinal-consistency

# Filter for names ending in a cycle number
cycle_vars <- names(gic_renamed)[str_detect(names(gic_renamed), "\\.\\d{1,2}$")]

# Remove `Sex` variable from consideration since it was encoded similarly
cycle_vars <- cycle_vars[cycle_vars != "Sex.Male.1/Female.2"]

# Build function to audit the names
create_cycle_audit <- function(cycle_vars) {
  # Extract variable stems and cycles
  parsed_data <- tibble(raw_name = cycle_vars) |> 
    extract(raw_name,
            into = c("stem", "cycle"),
            regex = "(.*)[\\._](\\d+)$",
            remove = FALSE) |> 
    mutate(cycle = as.integer(cycle))
  
  # Report if any cycles aren't in [1, 12] or baseline/0
  out_of_bounds <- parsed_data |> 
    filter(cycle < 0 | cycle > 12 | is.na(cycle))
  
  if(nrow(out_of_bounds > 0)) {
    message("The following variables have unexpected cycle numbers:")
    print(out_of_bounds)    
  } else {
    message("All variables have valid cycle numbers.")
  }
  
  # Create the audit grid
  audit_grid <- parsed_data |> 
    filter(cycle >= 0, cycle <= 12) |> 
    select(stem, cycle) |> 
    distinct() |> 
    mutate(exists = 1) |> 
    pivot_wider(
      names_from = "cycle",
      values_from = "exists",
      values_fill = 0
    ) |> 
    arrange(stem) |> 
    rowwise() |> 
    mutate(
      all_ok = as.integer(all(c_across(any_of(as.character(1:12))) == 1))
    ) |> 
    ungroup() |> 
    select(stem,
           all_ok,
           any_of(as.character(0:12)))
  
  return(audit_grid)
}

# Function to generate table from audit output
format_audit_table <- function(grid_df) {
  
  has_baseline <- if("0" %in% names(grid_df)) grid_df$`0` == 1 else FALSE
  cycles_complete <- grid_df$all_ok == 1
  
  cycle_sum <- rowSums(grid_df[as.character(1:12)], na.rm = TRUE)
  cycles_empty <- cycle_sum == 0
  
  # Sort into cases
  ## Case A: Baseline + all cycles
  rows_complete <- which(has_baseline & cycles_complete)
  ## Case B: No baseline + all cycles
  rows_longitudinal <- which(!has_baseline & cycles_complete)
  ## Case C: Baseline, no cycles
  rows_baseline <- which(has_baseline & cycles_empty)
  ## Case D: Everything else
  rows_problem <- setdiff(1:nrow(grid_df), c(rows_complete, rows_longitudinal, rows_baseline))
  
  
  # Return the kable object
  grid_df |> 
    kbl(booktabs = TRUE) |> 
    kable_styling(full_width = FALSE) |> 
    row_spec(rows_complete,     background = cb_light[1]) |> 
    row_spec(rows_longitudinal, background = cb_light[5]) |> 
    row_spec(rows_baseline,     background = cb_light[6]) |> 
    row_spec(rows_problem,      background = cb_light[2])
}


grid_data <- create_cycle_audit(cycle_vars)
format_audit_table(grid_data)
```

:::{.callout-note collapse="false"}
### Note: Harmonized variable naming

**Issues:** Several cycle variables had minor inconsistencies, missing suffixes, or typos in their naming.

**Resolved:** [2025-11-28] Harmonized naming conventions in `01_data_cleaning.qmd` (Chunk: `fix_cycle-var-consistency`).

* `DBIL` for cycle 1, `Dbil` for the rest
    * Renamed all to `DBIL`
* `HB` for cycles 1-3, and 8, `Hb` for the rest
    * Renamed all to `HB`
* `Hospitalization.days` missing for cycle 2
    * Renamed `Hospitalization.days` to `Hospitalization.days.2`
* `Hospitalization.required.due.to.toxicity?` for cycle 1, but `Hospitalization.required.dur.to.toxicity?` for the rest
    * Renamed to fix "dur" typo where present
* `HypoNatramia.Na+` for cycles 2 and 3, but `HypoNa+` for the rest
    * Renamed all to `HypoNa+`
* `TBIL` for cycle 1, `TBil` for cycle 3, `Tbil` for the rest
    * Renamed all to `TBIL`
:::

```{r}
#| label: fix-longitudinal-consistency

# [DAT-003] Harmonize longitudinal variable naming and correct typos.
gic_renamed <- gic_renamed |> 
  # Fix DBIL
  rename_with(
    .cols = matches("^dbil", ignore.case = TRUE),
    .fn = ~ str_replace(., regex("^dbil", ignore_case = TRUE), "DBIL")
  ) |> 
  
  # Fix HB
  rename_with(
    .cols = matches("^hb", ignore.case = TRUE),
    .fn = ~ str_replace(., regex("^hb", ignore_case = TRUE), "HB")
  ) |> 
  
  # Fix Hospitalization.days
  rename(`Hospitalization.days.2` = "Hospitalization.days") |> 
  
  # Fix `Hospitalization.required.dur.to.toxicity?` typo
  rename_with(
    .cols = contains("dur.to.toxicity?"),
    .fn = ~ str_replace(., "dur\\.to", "due.to")
  ) |> 
  
  # Fix HypoNa+
  rename_with(
    .cols = matches("^(hypona|hyponatramia)", ignore.case = TRUE),
    .fn = ~ . |> 
      str_replace(regex("^hyponatramia\\.na\\+", ignore_case = TRUE), "HypoNa+") |> 
      str_replace(regex("^hypona\\+", ignore_case = TRUE), "HypoNa+")
  ) |> 
  
  # Fix TBIL
  rename_with(
    .cols = matches("^tbil", ignore.case = TRUE),
    .fn = ~ str_replace(., regex("^tbil", ignore_case = TRUE), "TBIL")
  )

# Recheck
cycle_vars <- names(gic_renamed)[str_detect(names(gic_renamed), "\\.\\d{1,2}$")]
cycle_vars <- cycle_vars[cycle_vars != "Sex.Male.1/Female.2"]
grid_data <- create_cycle_audit(cycle_vars)
format_audit_table(grid_data)
```

## Standardize Variable Names

```{r}
#| label: apply-standard-names

gic_renamed <- gic_renamed |> 
  # Name standardization
  rename(
    # Basic Variables
    hiv_status = "HIV.Status:.1,.Negative.|.2,.Positive",
    sex = "Sex.Male.1/Female.2",
    residence = "Primary.Residence.1.urban/.2.rural/3.peri-urban.",
    cancer_site = "Cancer.site:.1,.Esophagus.|.2,.Gastric.|.3,.Gastro-esophageal.junction.|.4,.Gall.bladder.,.Biliary.&.Pancreas.|.9,.Colorectal.|.11,.Anal.",
    histology = "Histology.type:.1,.adenocarcinoma/carcinoma.|.2,.Squamous.carcinoma.|.3,.Sarcoma.|.4,.Carcinoid.|.5,.Lymphoma.|.7,.Others",
    cancer_stage = "Cancer.stage:.stage.1,.I.|.2,.II.|.3,.III.|.4,.IV.|.4,.Unknown/Undocumented",
    
    # Comorbidities
    comorbid_htn = "Hypertension",
    comorbid_dm = "Diabetes.mellitus.",
    comorbid_tb = ".Tuberculosis",
    comorbid_asthma = "Asthma",
    comorbid_chd = "Cardiac.disease.",
    comorbid_copd = "COPD.",
    comorbid_ckd = "Kidney.disease.",
    comorbid_epilepsy = "Epilepsy",
    comorbid_none = "No.cormobidities",
    
    # History
    hx_etoh = "drink_alcohol_history_of",
    hx_tobacco = "tobacco_smoker_history_of",
    hx_cancer = "family_history_of_cancer",
    
    # Baseline Measurements
    weight_0 = "Weight.(kg)",
    height_0 = "Height.(cm)",
    bmi_0 = "Baseline.BMI",
    wbc_0 = "Baseline.WBC",
    anc_0 = "Baseline.ANC",
    plt_0 = "Baseline.PLT",
    cd4_count_0 = "Baseline.CD4.count",
    hb_0 = "Baseline.Hb",
    mcv_0 = "Baseline.MCV",
    mch_0 = "Baseline.MCH",
    sodium_0 = "Baseline.Na+",
    potassium_0 = "Baseline.K+",
    urea_0 = "Baseline.Urea",
    creat_0 = "Baseline.Creat",
    cr_clear_0 = "Baseline.Creatinine.clearance",
    tp_0 = "Baseline.TP",
    alb_0 = "Baseline.ALB",
    tbil_0 = "Baseline.Tbil",
    alp_0 = "Baseline.ALP",
    ast_0 = "Baseline.AST",
    alt_0 = "Baseline.ALT",
    ggt_0 = "Baseline.GGT",
    albi_score_0 = "Albumin.bilirubin..score",
    cea_0 = "Baseline.CEA",
    ca199_0 = "Baseline.CA19_9",
    
    # Treatment Info
    tx_concurrent_rtx = "chemotherapy_given_with.concurrent.radiotherapy?",
    tx_intent = "Treatment.intention-.1,.Curative.|.2,.Palliative",
    chemo_seq = "Sequence.of.chemotherapy-.1,.Neoadjuvant.|.2,.Adjuvant.|.3,.Definitive",
    chemo_type = "chemotherapy.given:.1,Platinum/Paclitaxel..|.2,.Platinum/5-FU..3,.LCV/5-FU.|.4,.CapeOX.|.5,.FOLFOX4.|.6,.Capecitabine.|.20,.Irinotecan.|.21,.Others",
    cycles_prescribed = "Total.number.chemo.cycles.prescribed",
    cycles_given = "Total.number.of.chemotherapy.administered",
    completion_rate = "Chemotherapy.completion.rates.(%)",
    death_date = "date_of_death_if_such_outc"
  ) |> 
  
  # Standardize names of cycle variables
  rename_with(
    .cols = matches("\\.\\d{1,2}$"),
    .fn = ~ .x |> 
      # Fix specific longer names
      str_replace("^Was\\.the\\.cycle\\.delayed[^0-9]*", "cycle_delayed.") |> 
      str_replace("^Hospitalization\\.required[^0-9]*", "hosp_toxicity.") |> 
      str_replace("^Hospitalization\\.days", "hosp_days") |> 
      str_replace("^Blood\\.transfusion", "transfusion_given") |> 
      str_replace("^blood\\.units\\.needed", "transfusion_units") |> 
      str_replace("^CSF-G\\.Indication", "csfg_indicated") |>
      str_replace("HypoNa\\+", "sodium") |> 
      str_replace("HypoK\\+", "potassium") |> 
      # Convert everything to snake case
      tolower() |> 
      str_replace("\\.(\\d{1,2})$", "_\\1")
  )
```

:::{.callout-note collapse="false"}
### Note: Renamed variables

**Issues:** Raw variable names contained factor levels and long strings.

**Resolved:** [2025-11-29] Converted all variables to standardized snake_case format in `01_data_cleaning.qmd` (Chunk: `rename`).

* Standardized naming conventions
    * Simplified names and converted all to snake case
    * Baseline measurements denoted with `_0` suffix
    * Cycle measurements denoted with `_#` suffix for the given cycle number
:::


# Structural Integrity Audit
## Completeness Check (Grid)
```{r}
#| label: audit-structure-completeness

# Recheck cycle vars
final_cycle_vars <- names(gic_renamed)[str_detect(names(gic_renamed), "\\_\\d{1,2}$")]
final_name_audit <- create_cycle_audit(final_cycle_vars)
format_audit_table(final_name_audit)
```

## Row and Deletion Integrity
```{r}
#| label: audit-structure-integrity

# Row integrity
if(nrow(gic_raw) != nrow(gic_renamed)) {
  stop("Row count mismatch, rows were lost or added.")
} else {
  message("Row count preserved: ", nrow(gic_renamed), " patients.")
}

# Deletion verification
deleted_patterns <- c("Febrile.Neutropenia", "HyperK\\+", "Cycle \\d", "\\.\\.\\.\\d*")
found_trash <- map(deleted_patterns, ~ grep(., names(gic_renamed), value = TRUE))

if (length(unlist(found_trash)) > 0) {
  warning("Some columns intended for deletion are still present.")
} else {
  message("All redundant columns removed.")
}
```

## Traceability: Static Variables
```{r}
#| label: audit-traceability-static-vars

# Static variable integrity verification
## Load Glossary
glossary_df <- read_xlsx(here("docs", "glossary.xlsx"))

## Find the start of the Longitudinal Section
longitudinal_start_index <- which(str_detect(glossary_df$Variable, "Longitudinal"))[1]

if (is.na(longitudinal_start_index)) {
  stop("Could not find 'Longitudinal' header.")
} else {
  static_vars <- glossary_df |> slice(1:(longitudinal_start_index - 1))
}

## Prepare checklist
check_list <- static_vars |> 
  filter(!str_detect(Variable, "^\\*\\*")) |> 
  select(clean_col = Variable, raw_col = "Original Name (Example)")

## Run stationary var audit
static_audit <- check_list |> 
  rowwise() |> 
  mutate(
    # Check if clean version of var names exist
    clean_exists = clean_col %in% names(gic_renamed),
    
    # Get rid of backticks on variable names
    raw_col = str_remove_all(raw_col, "`"), 
    raw_exists = raw_col %in% names(gic_raw), 
    data_match = if (clean_exists && raw_exists) {
      identical(gic_raw[[raw_col]], gic_renamed[[clean_col]])
      } else {
        NA
        }) |> 
  ungroup() |> 
  mutate(
    Status = case_when(
      !clean_exists ~ "Clean var missing",
      !raw_exists   ~ "Raw var missing",
      data_match    ~ "Match",
      TRUE          ~ "Data mismatch"
    )
  )

## Define row indices for better visualization
static_green <- which(static_audit$Status == "Match")
static_red <- which(static_audit$Status != "Match")


## Display static audit
static_audit |> 
  arrange(Status) |> 
  select(clean_col, raw_col, Status) |> 
  kbl(booktabs = TRUE) |> 
  kable_styling(full_width = TRUE) |> 
  row_spec(static_green, background = cb_light[1]) |> 
  row_spec(static_red, background = cb_light[2]) |> 
  scroll_box(height = "400px")
```

## Traceability: Longitudinal Variables
```{r}
#| label: audit-traceability-longitudinal-vars

# Cycle variable integrity validation
## Get all variables, both baseline and cyclic
clean_cycle_vars <- names(gic_renamed) |> str_subset("_[1-9]\\d*$")

## Create function to match against raw data/cols
find_raw_match <- function(clean_col_name) {
  clean_vec <- gic_renamed[[clean_col_name]]
  
  # Check for empty columns
  if (all(is.na(clean_vec))) {
    return(tibble(raw_match = "-", status = "All NA"))
  }
  
  # Check against every column in raw data
  matches <- names(gic_raw)[map_lgl(gic_raw, ~ identical(., clean_vec))]
  
  if (length(matches) == 0) {
    return(tibble(raw_match = NA_character_, status = "No match"))
  } else if (length(matches) == 1) {
    return(tibble(raw_match = matches, status = "Unique match"))
  } else {
    # If multiple raw columns match (like the deleted duplicates), list them
    match_list <- paste(matches, collapse = ", ")
    return(tibble(raw_match = match_list, status = "Multiple raw matches"))
  }
}

## Run cycle audit
cycle_audit <- tibble(clean_name = clean_cycle_vars) |> 
  rowwise() |> 
  mutate(result = list(find_raw_match(clean_name))) |> 
  unnest(result) |> 
  ungroup()

## Extract metadata for sorting
cycle_audit_final <- cycle_audit |> 
  extract(clean_name, into = c("stem", "cycle"), regex = "(.*)_(\\d+)$", remove = FALSE) |> 
  mutate(cycle = as.integer(cycle)) |> 
  arrange(stem, cycle)

## Define row indices for better visualization
cycle_good <- which(cycle_audit_final$status == "Unique match")
cycle_multiple <- which(cycle_audit_final$status == "Multiple raw matches")
cycle_empty <- which(cycle_audit_final$status == "All NA")
cycle_bad <- which(cycle_audit_final$status == "No match")


## Print cycle audit
cycle_audit_final |> 
  select(clean_name, raw_match, status) |> 
  kbl(booktabs = TRUE, col.names = c("Clean Name", "Matched Raw Column(s)", "Status")) |> 
  kable_styling(full_width = TRUE) |> 
  row_spec(cycle_good, background = cb_light[1]) |> 
  row_spec(cycle_multiple, background = cb_light[6]) |> 
  row_spec(cycle_empty, background = cb_light[8]) |> 
  row_spec(cycle_bad, background = cb_light[2]) |> 
  scroll_box(height = "400px")
```

:::{.callout-note collapse="false"}
### Audit Findings: Structural Integrity

The renaming audit confirmed that data integrity was preserved during variable name standardization.
:::

## Checkpoint
```{r}
#| label: save-renamed-data

# Define path to file
gic_renamed_path <- here("data", "processed", "gic_renamed.rds")

# Save if file doesn't exist, or check if overwrite is intended if it does.
if (file.exists(gic_renamed_path) && interactive()) {
  # If file exists and this is being run manually in RStudio ask before overwriting
  response <- readline(prompt = paste("Warning:", basename(gic_renamed_path), "already exists. Overwrite? (y/n):"))
  
  if (tolower(substr(response, 1, 1)) == "y") {
    saveRDS(gic_renamed, gic_renamed_path)
    message("File overwritten: ", basename(gic_renamed_path))
  } else {
    message("Save operation cancelled.")
  }
} else {
  # If the file doesn't exist or document is being rendered, just save it
  saveRDS(gic_renamed, gic_renamed_path)
  message("File saved: ", basename(gic_renamed_path))
}
```

# Data Type Standardization
## Diagnostic Scan

```{r}
#| label: assess-data-types

# Helper function to get info on data types
scan_types <- function(df) {
  map_df(names(df), function(col_names) {
    vec <- df[[col_names]]
    
    # Basic stats
    curr_class <- paste(class(vec), collapse = "/")
    n_unique <- n_distinct(vec, na.rm = TRUE)
    n_miss <- sum(is.na(vec))
    
    # Content preview
    if (n_unique <= 8) {
      content <- paste(sort(unique(na.omit(vec))), collapse = ", ")
    } else if (is.numeric(vec)) {
      q <- quantile(vec, probs = c(0, 0.5, 1), na.rm = TRUE)
      content <- sprintf("Min: %s | Med: %s | Max: %s",
                         round(q[1], 2), round(q[2], 2), round(q[3], 2))
    } else {
      # Show `head()` for high cardinality strings
      content <- paste(paste(head(na.omit(vec), 4), collapse = ", "), "...")
    }
    
    tibble(
      variable = col_names,
      type = curr_class,
      unique_n = n_unique,
      missing_n = n_miss,
      sample_vals = content
    )
  })
}

# Run scan and display output
type_report <- scan_types(gic_renamed)

## Highlight anything not encoded as numeric
type_highlight <- which(type_report$type != "numeric")

type_report |> 
  kbl(booktabs = TRUE) |> 
  kable_styling(full_width = TRUE) |> 
  row_spec(type_highlight, background = cb_light[6]) |> 
  scroll_box(height = "400px")
```

:::{.callout-note collapse="false"}
### Note: Data type incosistencies resolved

**Issues:**

* **Factors Encoded as Numeric:**
	* *Issue:* Categorical variables were imported as numeric.
	* *Resolved:* [2025-11-30] Converted to factors with explicit labels based on variable names from `IPROTECTARetrospecti_DATA_2025-08-19_2157 FINAL SHEET.xlsx`.
		* *Location:* `01_data-cleaning.qmd` (Chunk: `encode-types`).
* **Empty Columns as Logical:**
	* *Issue:* `cr_clear_0`, `albi_score_0`, and `dbil_12` are stored as `logical` because they contain only `NA`.
	* *Resolved:* [2025-11-30] Coerced to `numeric`.
		* *Location:* `01_data-cleaning.qmd` (Chunk: `encode-types`).
* **Date Precision:**
	* *Issue:* `death_date` is stored as `POSIXct` (DateTime).
	* *Resolved:* [2025-11-30] Converted to `Date` class.
		* *Location:* `01_data-cleaning.qmd` (Chunk: `encode-types`).
:::

## Apply Type Coercion
```{r}
#| label: apply-type-coercion

gic_encoded <- gic_renamed |> 
  mutate(
  # Convert static factors
    sex               = factor(sex,
                         levels = c(1, 2), 
                         labels = c("Male", "Female")),
    residence         = factor(residence,
                         levels = c(0, 1, 2), 
                         labels = c("Urban", "Rural", "Peri-urban")),
    hiv_status        = factor(hiv_status,
                         levels = c(1, 2), 
                         labels = c("Negative", "Positive")),
    cancer_site       = factor(cancer_site,
                         levels = c(1, 2, 3, 4, 9, 11),
                         labels = c("Esophagus", "Gastric", 
                                    "Gastro-esophageal junction", "Gall bladder, Biliary & Pancreas", "Colorectal", "Anal")),
    histology         = factor(histology,
                          levels = c(1, 2, 3, 4, 5, 7),
                          labels = c("Adenocarcinoma/carcinoma", 
                                     "Squamous carcinoma", "Sarcoma", 
                                     "Carcinoid", "Lymphoma", "Other")),
    cancer_stage      = factor(cancer_stage,
                          levels = c(1, 2, 3, 4),
                          labels = c("Stage I", "Stage II", "Stage III", 
                                     "Stage IV")),
    comorbid_htn      = factor(comorbid_htn,
                          levels = c(0,1),
                          labels = c("No", "Yes")),
    comorbid_dm       = factor(comorbid_dm,
                          levels = c(0,1),
                          labels = c("No", "Yes")),
    comorbid_tb       = factor(comorbid_tb,
                          levels = c(0,1),
                          labels = c("No", "Yes")),
    comorbid_asthma   = factor(comorbid_asthma,
                          levels = c(0,1),
                          labels = c("No", "Yes")),
    comorbid_chd      = factor(comorbid_chd,
                          levels = c(0,1),
                          labels = c("No", "Yes")),
    comorbid_copd     = factor(comorbid_copd,
                          levels = c(0,1),
                          labels = c("No", "Yes")),
    comorbid_ckd      = factor(comorbid_ckd,
                          levels = c(0,1),
                          labels = c("No", "Yes")),
    comorbid_epilepsy = factor(comorbid_epilepsy,
                          levels = c(0,1),
                          labels = c("No", "Yes")),
    comorbid_none     = factor(comorbid_none,
                          levels = c(0,1),
                          labels = c("No", "Yes")),
    hx_etoh           = factor(hx_etoh,
                          levels = c(0,1),
                          labels = c("No", "Yes")),
    hx_tobacco        = factor(hx_tobacco,
                          levels = c(0,1),
                          labels = c("No", "Yes")),
    hx_cancer         = factor(hx_cancer,
                          levels = c(0,1),
                          labels = c("No", "Yes")),
    tx_concurrent_rtx = factor(tx_concurrent_rtx,
                          levels = c(0, 1),
                          labels = c("No", "Yes")),
    tx_intent         = factor(tx_intent,
                          levels = c(1, 2),
                          labels = c("Curative", "Palliative")),
    chemo_seq         = factor(chemo_seq,
                          levels = c(1, 2, 3),
                          labels = c("Neoadjuvant", "Adjuvant", "Definitive")),
    chemo_type        = factor(chemo_type,
                          levels = c(1, 2, 3, 4, 5, 6, 20, 21),
                          labels = c("Platinum/Paclitaxel", "Platinum/5-FU", 
                                     "LCV/5-FU", "CapeOX", "FOLFOX4", 
                                     "Capecitabine", "Irinotecan", "Other")),
    
    # Convert cycle factors
    across(
      c(starts_with("cycle_delayed_"),
        starts_with("hosp_toxicity_"),
        starts_with("transfusion_given_"),
        starts_with("csfg_indicated_")),
      ~ factor(., levels = c(0, 1), labels = c("No", "Yes"))
    ),
    
    # Convert dates
    death_date        = as_date(death_date),
    
    # Coerce logical to numeric
    cr_clear_0        = as.numeric(cr_clear_0),
    albi_score_0      = as.numeric(albi_score_0),
    dbil_12           = as.numeric(dbil_12),
    
    # Coerce `ast_1` to numeric
    ast_1             = as.numeric(ast_1)
  )
```

:::{.callout-warning collapse="false"}
### Warning: Encoding anomalies flagged

**Issues:** 
* **Numeric column as character:** `ast_1` is stored as a character vector.
	- *Investigation:* `record_id == 64` contains the value "pn" for `ast_1`.
	- *Temporary Resolution*: Coerced "pn" to `NA`.
:::

:::{.callout-note collapse="false"}
### Note: Encoded data types

**Issues:** Data type mismatches identified during scan.
**Resolved:** [2025-11-30] Corrected encoding.

* Categorical variables were imported as numeric.
  * Converted to factors with explicit labels.
* Empty numeric variables were imported as logical
  * Coerced to numeric.
* `death_date` imported as `POSIXct`.
  * Coerced to Date.
:::

# Encoding Audit
## Audit Helper Functions
```{r}
#| label: def-audit-functions

# Function to audit numeric/date variables
audit_numeric_logic <- function(df_pre, df_post, var_list) {
  map_df(var_list, function(var) {
    vec_pre <- df_pre[[var]]
    vec_post <- df_post[[var]]
    
    # Check for NA coersion
    na_diff <- sum(is.na(vec_post)) - sum(is.na(vec_pre))
    
    
    if (is.Date(vec_post)) {
      # String check for dates
      match <- all(as.character(na.omit(vec_pre)) == as.character(na.omit(vec_post)))
    } else {
      # Tolerance check for numerics
      match <- isTRUE(all.equal(vec_pre, vec_post, tolerance = 1e-10))
    }
    
    tibble(variable = var, na_diff = na_diff, match = match)
  })
}

# Function to audit factor variables
audit_factor_logic <- function(df_pre, df_post, var_list) {
  map_df(var_list, function(var) {
    raw_vec <- df_pre[[var]]
    new_vec <- df_post[[var]]
    
    tibble(
      variable = var,
      raw_value = as.character(raw_vec),
      new_label = as.character(new_vec)
    ) |> 
      count(variable, raw_value, new_label) |> 
      filter(!is.na(raw_value)) |> 
      
      mutate(is_error = is.na(new_label))
  }) |> 
    arrange(variable, as.numeric(raw_value))
}

# Function to visualize factor audit

format_factor_audit_table <- function(audit_df) {
  # Identify rows for coloring
  row_pass <- which(!audit_df$is_error)
  row_fail <- which(audit_df$is_error)
  
  audit_df |>
    arrange(variable, raw_value) |> 
    select(Variable = variable, "Raw Value" = raw_value, "Label" = new_label, 
           Count = n) |> 
    kbl(booktabs = TRUE) |> 
    kable_styling(full_width = TRUE) |> 
    
    # Apply coloring
    row_spec(row_pass, background = cb_light[1]) |> 
    row_spec(row_fail, background = cb_light[2]) |> 
    
    scroll_box(height = "400px")
}

# Function to create grid for cycle variable output
format_cycle_grid <- function(audit_df) {
  # Convert audit results (NA diff + Match) into 1/0 status
  grid_data <- audit_df |> 
    mutate(status = if_else(na_diff == 0 & match, 1, 0)) |> 
    extract(variable, into = c("stem", "cycle"), regex = "(.*)_(\\d+)$", remove = FALSE) |> 
    mutate(cycle = as.integer(cycle)) |> 
    select(stem, cycle, status) |> 
    pivot_wider(names_from = cycle, values_from = status)
  
  # Render
  tbl <- grid_data |> kbl(booktabs = T) |> kable_styling(full_width = F)
  
  for(i in 2:ncol(grid_data)) {
    col_vals <- grid_data[[i]]
    colors <- case_when(
      is.na(col_vals) ~ "white",
      col_vals == 1   ~ cb_light[1], 
      col_vals == 0   ~ cb_light[2], 
      TRUE            ~ "white"
    )
    tbl <- tbl |> column_spec(i, background = colors)
  }
  tbl |> scroll_box(height = "400px")
}

# Adapter function to convert detailed factor audit to grid status
get_cycle_factor_grid <- function(audit_df) {
  audit_df |> 
    group_by(variable) |> 
    summarize(
      # Pass (1) if ZERO data was lost
      # Fail (0) if ANY data was lost
      status = if_else(first(n_lost_total) == 0, 1, 0)
    ) |> 
    ungroup() |> 
    extract(variable, into = c("stem", "cycle"), regex = "(.*)_(\\d+)$", remove = FALSE) |> 
    mutate(cycle = as.integer(cycle)) |> 
    select(stem, cycle, status) |> 
    pivot_wider(names_from = cycle, values_from = status)
}
```

## Numeric Audits

### Audit Static Numeric, Date, and Baseline Longitudinal Variables

**Criteria:**

  - Values must match pre-cleaned data within a tolerance of $1 \times 10 ^{-10}$.
  - No new `NA`s should be introduced during type coercion.
  - String representations of dates must match.

```{r}
#| label: audit-encoding-static-numeric

sn_vars <- names(gic_encoded)[map_lgl(gic_encoded, ~ is.numeric(.) | is.Date(.))] |> 
  str_subset("_[1-9]\\d*$", negate = TRUE)

# Run Logic
results <- audit_numeric_logic(gic_renamed, gic_encoded, sn_vars)

# Formatting
row_pass <- which(results$na_diff == 0 & results$match)
row_fail <- which(results$na_diff > 0 | !results$match)

results |> 
  kbl(booktabs = TRUE, caption = "Static Numeric Integrity") |> 
  kable_styling(full_width = TRUE) |> 
  row_spec(row_pass, background = cb_light[1]) |> 
  row_spec(row_fail, background = cb_light[2]) |> 
  scroll_box(height = "400px")
```

### Audit Longitudinal Numeric Variables

**Criteria:**

  - Values must match pre-cleaned data within a tolerance of $1 \times 10 ^{-10}$.
  - No new `NA`s should be introduced during type coercion.

```{r}
#| label: audit-encode-longitudinal-numeric

cn_vars <- names(gic_encoded)[map_lgl(gic_encoded, ~ is.numeric(.) | is.Date(.))] |> 
  str_subset("_[1-9]\\d*$", negate = FALSE)

# Run Logic
results <- audit_numeric_logic(gic_renamed, gic_encoded, cn_vars)

# Print Grid
format_cycle_grid(results)
```

## Factor Audits

### Audit Static Categorical Variables

**Criteria:**

  - Each unique raw value (e.g., `1`) maps to exactly one factor label (e.g., `Male`).
  - The total count of non-missing values must remain identical before and after encoding.

```{r}
#| label: audit-encoding-static-factor

sf_vars <- names(gic_encoded)[map_lgl(gic_encoded, is.factor)] |> 
  str_subset("_\\d+$", negate = TRUE)

sf_audit <- audit_factor_logic(gic_renamed, gic_encoded, sf_vars)

if (any(sf_audit$is_error)) warning("Data loss detected in static factors.")

format_factor_audit_table(sf_audit)
```

### Audit Longitudinal Categorical Variables

**Criteria:**

  - Each unique raw value (e.g., `0`) maps to exactly one factor label (e.g., `No`).
  - The total count of non-missing values must remain identical before and after encoding.

```{r}
#| label: audit-encoding-longitudinal-factor

cf_vars <- names(gic_encoded)[map_lgl(gic_encoded, is.factor)] |> 
  str_subset("_\\d+$", negate = FALSE)

cf_audit <- audit_factor_logic(gic_renamed, gic_encoded, cf_vars)

if (any(cf_audit$is_error)) warning("Data loss detected in static factors.")

format_factor_audit_table(cf_audit)
```

:::{.callout-note collapse="false"}
### Audit Findings: Encoding

The encoding audit confirmed that data integrity was preserved during type conversion, with expected exceptions noted below:

* **Numeric Precision:** All static and longitudinal numeric variables matched pre-cleaning values within a specified tolerance to account for floating point issues.
* **Factor Consistency:** All factor variables demonstrated consistent 1:1 mapping from raw codes to labels with no data loss.
* **NA Introduction:** * **`ast_1`:** The audit flagged 1 new `NA` value.
    * *Verification:* This confirms the successful coercion of the invalid string "pn" (Record 64) to `NA`, as planned in the cleaning logic.
:::

## Export
```{r}
#| label: save-clean-data

# Define path to file
gic_encoded_path <- here("data", "processed", "gic_encoded.rds")

# Save if file doesn't exist, or check if overwrite is intended if it does.
if (file.exists(gic_encoded_path) && interactive()) {
  # If file exists and this is being run manually in RStudio ask before overwriting
  response <- readline(prompt = paste("Warning:", basename(gic_encoded_path), "already exists. Overwrite? (y/n):"))
  
  if (tolower(substr(response, 1, 1)) == "y") {
    saveRDS(gic_encoded, gic_encoded_path)
    message("File overwritten: ", basename(gic_encoded_path))
  } else {
    message("Save operation cancelled.")
  }
} else {
  # If the file doesn't exist or document is being rendered, just save it
  saveRDS(gic_encoded, gic_encoded_path)
  message("File saved: ", basename(gic_encoded_path))
}

saveRDS(gic_encoded, here("data", "processed", "gic_encoded.rds"))
```
