---
title: "GI Cancer: Data Validation"
bibliography: references.bib
---
# Load Data and Libraries
```{r}
#| label: setup
#| include: false

source("R/setup.R")


# Packages

```

```{r}
#| label: load-data

gic_encoded <- readRDS(here("data", "processed", "gic_encoded.rds"))
```

# Logic & Consistency Validation
## Cycle Consistency
Verify that `cycles_given` matches the actual presence of data in the cycle columns. Flag patients who have data present in cycles after their administered number.

```{r}
#| label: assess_extraneous-data

# Pivot longitudinal data to make finding missing data easier
real_data_map <- gic_encoded |> 
  select(record_id, pgh_number, cycles_prescribed, cycles_given, matches("_[1-9]\\d*$")) |> 
  pivot_longer(
    cols = matches("_\\d+$"),
    names_to = c("stem", "cycle"),
    names_pattern = "(.*)_(\\d+)$",
    values_to = "value",
    values_transform = list(value = as.character)
  ) |> 
  mutate(cycle = as.integer(cycle)) |> 
  left_join(var_groups, by = "stem") |> 
  mutate(type = if_else(panel == "4. Mgmt", "Mgmt", "Labs"))

# Summarize records to check if data matches cycles_given
patient_audit <- real_data_map |> 
  group_by(record_id, pgh_number, cycles_given) |> 
  summarize(
    # Check number of cycles that have labwork data
    n_lab_cycles = n_distinct(cycle[type == "Labs" & !is.na(value)]),
    # Check number of cycles that have management data
    n_mgmt_cycles = n_distinct(cycle[type == "Mgmt" & !is.na(value)]),
    # Check if number of lab cycles matches number of management cycles
    labs_mgmt_match = n_lab_cycles == n_mgmt_cycles,
    # For cycles with management data, check if all is present
    mgmt_is_complete = {
      required_vars <- expand.grid(s = mgmt_stems, c = 1:first(cycles_given)) |> 
        mutate(fullname = paste0(s, "_", c)) |> 
        pull(fullname)
      actual_vars <- paste0(stem, "_", cycle)[!is.na(value) & type == "Mgmt"]
      all(required_vars %in% actual_vars)
    },
    # Check the last cycle that had any data entered
    max_cycle_any = max(cycle[!is.na(value)], na.rm = TRUE),
    .groups = "drop"
  ) |> 
  mutate(max_cycle_any = if_else(is.infinite(max_cycle_any), 0, max_cycle_any))

# Flag pts with data that doesn't match tx hx
problem_ids <- patient_audit |> 
  filter(
    max_cycle_any > cycles_given |
    n_mgmt_cycles != cycles_given |
    n_lab_cycles != cycles_given |
    labs_mgmt_match == FALSE |
    mgmt_is_complete == FALSE
  ) |> 
  pull(record_id)

# Display audit table
if (length(problem_ids) > 0) {
  warning("Found discrepancies for ", length(problem_ids), " records.")
  
  patient_audit |> 
    filter(record_id %in% problem_ids) |> 
    kbl(booktabs = TRUE, caption = "Strict Cycle Consistency Audit") |> 
    kable_styling(full_width = FALSE)
} else {
  message("All cycle counts match data presence and management completeness.")
}
```

::: {.callout-warning collapse="false"}
### Anomaly: Extraneous Cycle Data

**Issue:** The audit identified 4 patients with data recorded for cycles greater than their `cycles_given` count.

* **Affected IDs:** 27, 146, 498, 766
* **Status:** Flagged as **DAT-019**. Investigation required to determine if the administration count is incorrect or if the data should be removed.
:::

```{r}
#| label: investigate_extraneous-data
#| results: asis

view_pt_hx <- function(pid, df, audit_data = NULL, show_empty = TRUE) {
  # Get official count and audit info for the pt
  pt_row <- df |> filter(record_id == pid)
  n_given <- pt_row |> pull(cycles_given)
  
  audit_row <- if(!is.null(audit_data)) audit_data |> filter(record_id == pid) else NULL
  
  # Extract and pivot longitudinal data
  pt_data <- pt_row |> 
    select(matches("_[1-9]\\d*$")) |> 
    pivot_longer(
      cols = everything(),
      names_to = c("stem", "cycle"),
      names_pattern = "(.*)_(\\d+)$",
      values_to = "value",
      values_transform = list(value = as.character) 
    ) |> 
    mutate(cycle = as.numeric(cycle)) |> 
    # Keep cycles based on tx hx or data presence
    filter(if(show_empty) cycle <= n_given | !is.na(value) else !is.na(value)) |> 
    left_join(var_groups, by = "stem") |>
    arrange(panel, stem)
  
  # Create wide format for kable
  pt_wide <- pt_data |> 
    select(panel, stem, cycle, value) |> 
    pivot_wider(names_from = cycle, values_from = value)
  
  cycle_cols <- as.numeric(setdiff(names(pt_wide), c("panel", "stem")))
  
  # Build table
  kb <- pt_wide |> 
    kbl(booktabs = T,
        caption = paste("Pt Data: Record ID", pid, "(Given:", n_given, ")")) |> 
    kable_styling(font_size = 11)

  # Highlight affected columns
  for(cyc in cycle_cols) {
    col_idx <- which(names(pt_wide) == as.character(cyc))
    has_any_data <- any(!is.na(pt_wide[[as.character(cyc)]]))
    
    # Check for missing management in this cycle
    mgmt_in_cycle <- pt_data |> 
      filter(cycle == cyc, panel == "4. Mgmt") |> 
      pull(value)
    is_mgmt_incomplete <- any(is.na(mgmt_in_cycle))
    
    if (cyc > n_given) {
      # Data recorded beyond the official count
      kb <- kb |> column_spec(col_idx, background = cb_light[2])
    } else if (!has_any_data) {
      # Entirely empty cycles that should have data
      kb <- kb |> column_spec(col_idx, background = cb_light[4])
    } else if (is_mgmt_incomplete) {
      # Partial management data
      kb <- kb |> column_spec(col_idx, background = cb_light[6])
    }
  }
  
  return(kb)
}

# Run audit for flagged records
walk(problem_ids, function(pid) {
  print(view_pt_hx(pid, gic_encoded, audit_data = patient_audit, show_empty = TRUE))
  cat("\n<br><br>\n") 
})
```


## Treatment Adherence Logic
Verify that the administered cycles do not exceed prescribed cycles.
```{r}
#| label: assess_cycle-counts

# Check that `cycles_given` <= `cycles_prescribed`
over_treated <- gic_encoded |> 
  filter(cycles_given > cycles_prescribed) |> 
  select(record_id, cycles_prescribed, cycles_given) |> 
  mutate(excess = cycles_given - cycles_prescribed) |> 
  arrange(desc(excess))

if (nrow(over_treated) > 0) {
  warning("Found ", nrow(over_treated), " patients with more cycles given than prescribed.")
  
  over_treated |> 
    kbl(booktabs = TRUE, caption = "Treatment Adherence Logic Check") |> 
    kable_styling(full_width = FALSE) |> 
    row_spec(1:nrow(over_treated), background = cb_light[2])
} else {
  message("Treatment adherence check passed.")
}
```

## Composite Variable Checks
Verify that pre-calculated fields match their components.
```{r}
#| label: assess_calculated-vars

# Define tolerance thresholds
bmi_tol <- 0.5
rate_tol <- 1.0

composite_audit <- gic_encoded |> 
  select(record_id, height_0, weight_0, bmi_0, 
         cycles_given, cycles_prescribed, completion_rate) |> 
  mutate(
    # 1. Calculate fresh BMI (Height is in cm, convert to m)
    bmi_calc = weight_0 / (height_0 / 100)^2,
    
    # 2. Calculate fresh Completion Rate
    rate_calc = (cycles_given / cycles_prescribed) * 100,
    
    # 3. Measure Diff
    bmi_diff = abs(bmi_0 - bmi_calc),
    rate_diff = abs(completion_rate - rate_calc)
  ) |> 
  # Filter for mismatches exceeding tolerance
  # (We use !is.na checks to ignore rows where data is missing)
  filter(
    (!is.na(bmi_diff) & bmi_diff > bmi_tol) | 
    (!is.na(rate_diff) & rate_diff > rate_tol)
  ) |> 
  arrange(desc(bmi_diff))

if (nrow(composite_audit) > 0) {
  warning("Found ", nrow(composite_audit), " records with composite variable mismatches.\n")
  
  composite_audit |> 
    select(record_id, 
           bmi_orig = bmi_0, bmi_new = bmi_calc, bmi_diff,
           rate_orig = completion_rate, rate_new = rate_calc) |> 
    kbl(caption = "Composite Variable Logic Check", digits = 1) |> 
    kable_styling() |> 
    scroll_box(height = "400px")
} else {
  message("Composite Check Passed: BMI and Completion Rates match inputs within tolerance.")
}
```
# Transfusion Logic Check
```{r}
#| label: assess_transfusions

# [DAT-008] Audit transfusion data for consistency and unit volume errors.

# Pivot to handle all cycles at once
transfusion_long <- gic_encoded |> 
  select(record_id,
         pgh_number,
         starts_with("transfusion_given_"), 
         starts_with("transfusion_units_")) |>
  pivot_longer(
    cols = -c(record_id, pgh_number),
    names_to = c(".value", "cycle"),
    names_pattern = "(.*)_(\\d+)$"
  ) |> 
  mutate(cycle = as.numeric(cycle)) |> 
  arrange(cycle)

# Summarize by Cycle
transfusion_summary <- transfusion_long |> 
  group_by(cycle) |> 
  summarize(
    n_given = sum(transfusion_given == "Yes", na.rm = TRUE),
    
    # [DAT-008] Check for missing units (Given = Yes, Units = NA/0)
    missing_units = sum(transfusion_given == "Yes" & 
                        (is.na(transfusion_units) | transfusion_units == 0), na.rm = TRUE),
    
    # [DAT-008] Check for unexpected units (Given = No, Units > 0)
    unexpected_units = sum(transfusion_units > 0 & 
                           (transfusion_given != "Yes" | is.na(transfusion_given)), na.rm = TRUE),
    
    # [DAT-008] Original redundancy check (Given == Units)
    is_identical = identical(
      as.numeric(transfusion_given == "Yes"), 
      as.numeric(transfusion_units)
    )
  )

# Print Table
error_rows <- which(transfusion_summary$missing_units > 0 | 
                    transfusion_summary$unexpected_units > 0)

transfusion_summary |> 
  kbl(booktabs = TRUE, caption = "Transfusion Logic & Consistency Check") |> 
  kable_styling(full_width = FALSE) |> 
  row_spec(error_rows, background = cb_light[2])
```
::: {.callout-warning collapse="false"}
### Anomaly: Transfusion Data Inconsistencies (DAT-008)

**Issue:** The audit revealed inconsistencies between logging if a transfusion was given and the number of units given as part of a transfusion

* **Missing Units:** 3 records are flagged as having received a transfusion ("Yes") but have zero or missing unit volume recorded. (Data incompleteness).
* **Flag Mismatch:** 3 records are flagged as NOT receiving a transfusion ("No") but contain a valid unit count (>0). (Flag error).

**Status:** Flagged for cleanup in the `fix_...` chunk. The red rows indicate cycles with **at least one** internal inconsistency.
:::

```{r}
#| label: investigate_transfusion-records

# Define the combined error condition (Given=Yes, Units=0/NA) OR (Units>0, Given=No/NA)
transfusion_error_records <- transfusion_long |> 
  filter(
    # Check A: Given but Missing/Zero Units
    (transfusion_given == "Yes" & (is.na(transfusion_units) | transfusion_units == 0)) |
    
    # Check B: Has Units but No/NA Flag
    (transfusion_units > 0 & (transfusion_given != "Yes" | is.na(transfusion_given)))
  ) |> 
  
  # Select the identifiers and context
  select(record_id, pgh_number, cycle, transfusion_given, transfusion_units) |> 
  arrange(record_id, cycle)

# Display the failed records
if (nrow(transfusion_error_records) > 0) {
  warning("Found ", nrow(transfusion_error_records), " individual cycle errors requiring review.")
  transfusion_error_records |> 
    kbl(caption = "Patient-Cycle Transfusion Inconsistencies") |> 
    kable_styling(full_width = TRUE) |> 
    scroll_box(height = "400px")
} else {
  message("All individual transfusion records are logically consistent.")
}
```

# Numeric Range Validation
## Body Measurement Outliers
Identify clinically impossible values for height and weight.

```{r}
#| label: assess_body-measurements

# Set wide limits to try and only catch errors, not outliers
limits <- list(
  ht_min = 100, ht_max = 250,
  wt_min = 30, wt_max = 200,
  bmi_min = 10, bmi_max = 50
)

# Identify outliers
body_outliers <- gic_encoded |> 
  select(record_id, height_0, weight_0, bmi_0) |> 
  filter(
    height_0 < limits$ht_min | height_0 > limits$ht_max |
    weight_0 < limits$wt_min | weight_0 > limits$wt_max |
    bmi_0 < limits$bmi_min   | bmi_0 > limits$bmi_max
  ) |> 
  arrange(height_0)

if (nrow(body_outliers) > 0) {
  warning("Found ", nrow(body_outliers), " records with impossible body measurements.")
  
  body_outliers |> 
    kbl(caption = "Body Measurement Outliers", digits = 1) |> 
    kable_styling(full_width = FALSE) |> 
    row_spec(which(body_outliers$height_0 > 250), background = cb_light[2])
} else {
  message("Body Measurement Check Passed.")
}

# Visualize results
gic_encoded |> 
  select(height_0, weight_0, bmi_0) |> 
  pivot_longer(everything(), names_to = "measure", values_to = "value") |> 
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = cb_light[3], color = "white") +
  facet_wrap(~measure, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Body Measurements", x = NULL, y = "Count")
```

::: {.callout-note}
### Action Taken [DAT-020]: Body Measurement Correction 

**Issue:** Record 614 contained an impossible height (1498 cm).
**Temporary Resolution:** Height set to 149.8 cm (suspected decimal error). BMI recalculated for this record only; all other BMI values remain as imported.
:::

## Biomarker Range Checks
Scan laboratory values for biologically impossible values (e.g., negative counts, values 10x above physiological max, etc.).
```{r}
#| label: assess_biomarker-ranges

# Generate summary stats for longitudinal variables

```

# Missingness Analysis
## Missing Data Patterns
Examine patterns of missing data.

```{r}
#| label: assess_missingness

# Visualize missingness by cycle

```

```{r}
#| label: assess_panel-missingness

# Examine missingness patterns by blood panel across cycle to identify unexpectedly missing data

```


# Feature Engineering
## Adverse Event Grading
Apply grading logic to raw laboratory values using `grading_rules.csv`.

```{r}
#| label: derive_adverse-events

# Load adverse event grading rules
rules <- read_csv(here("data", "grading_rules.csv"), show_col_types = FALSE)

# Pivot, Grade, and Pivot Back
gic_graded <- gic_encoded |> 
  # Pivot Long
  select(record_id, matches("_\\d+$")) |> 
  pivot_longer(
    cols = -record_id,
    names_to = c("stem", "cycle"),
    names_pattern = "(.*)_(\\d+)$",
    values_to = "value",
    values_transform = list(value = as.numeric) # Ensure numeric for grading math
  ) |> 
  filter(!is.na(value)) |> 
  semi_join(rules, by = "stem") |> 
  
  # Apply Logic (Min < Val <= Max)
  left_join(rules, by = "stem", relationship = "many-to-many") |> 
  filter(value > min_val & value <= max_val) |> 
  
  # Pivot Wide (Create labels)
  mutate(new_col = paste0(stem, "_grade_", cycle)) |> 
  select(record_id, new_col, label) |> 
  pivot_wider(names_from = new_col, values_from = label)

# Join back to main data
gic_validated <- gic_encoded |> 
  left_join(gic_graded, by = "record_id")

# Print a random sample of grades to see if it worked, will perform audit later
gic_validated |> 
  select(record_id, hb_1, hb_grade_1) |> 
  # Filter for patients who have data for the given cycle
  filter(!is.na(hb_1)) |> 
  head(10) |> 
  kbl(caption = "Verification: HB Cycle 1 Grading") |> 
  kable_styling(full_width = FALSE)

gic_validated |> 
  select(record_id, ast_5, ast_grade_5) |> 
  # Filter for patients who have data for the given cycle
  filter(!is.na(ast_5)) |> 
  head(10) |> 
  kbl(caption = "Verification: AST Cycle 5 Grading") |> 
  kable_styling(full_width = FALSE)

gic_validated |> 
  select(record_id, wbc_6, wbc_grade_6) |> 
  # Filter for patients who have data for the given cycle
  filter(!is.na(wbc_6)) |> 
  head(10) |> 
  kbl(caption = "Verification: WBC Cycle 6 Grading") |> 
  kable_styling(full_width = FALSE)
```

## Adverse Event Counts



# Export
```{r}
#| label: save_validated-data

# Define path to file
gic_validated_path <- here("data", "processed", "gic_validated.rds")

# Save if file doesn't exist, or check if overwrite is intended if it does.
if (file.exists(gic_validated_path) && interactive()) {
  # If file exists and this is being run manually in RStudio ask before overwriting
  response <- readline(prompt = paste("Warning:", basename(gic_validated_path), "already exists. Overwrite? (y/n):"))
  
  if (tolower(substr(response, 1, 1)) == "y") {
    saveRDS(gic_validated, gic_validated_path)
    message("File overwritten: ", basename(gic_validated_path))
  } else {
    message("Save operation cancelled.")
  }
} else {
  # If the file doesn't exist or document is being rendered, just save it
  saveRDS(gic_validated, gic_validated_path)
  message("File saved: ", basename(gic_validated_path))
}
```

