---
title: "GI Cancer: Data Validation"
bibliography: references.bib
---
# Load Data and Libraries
```{r}
#| label: setup
#| include: false

source("R/setup.R")


# Packages
```

```{r}
#| label: load-data

gic_encoded <- readRDS(here("data", "processed", "gic_encoded.rds"))
```

# Logic & Consistency Validation
## Cycle Consistency
Verify that `cycles_given` matches the actual presence of data in the cycle columns. Flag patients who have data present in cycles after their administered number.

```{r}
#| label: assess_extraneous-data

# Pivot longitudinal data to make finding missing data easier
real_data_map <- gic_encoded |> 
  select(record_id, pgh_number, cycles_prescribed, cycles_given, matches("_[1-9]\\d*$")) |> 
  pivot_longer(
    cols = matches("_\\d+$"),
    names_to = c("stem", "cycle"),
    names_pattern = "(.*)_(\\d+)$",
    values_to = "value",
    values_transform = list(value = as.character)
  ) |> 
  filter(!is.na(value)) |> 
  mutate(cycle = as.integer(cycle))

# Calculate max data cycle per patient
patient_max_cycles <- real_data_map |> 
  group_by(record_id, pgh_number, cycles_prescribed, cycles_given) |> 
  summarize(
    max_cycle_with_data = max(cycle),
    .groups = "drop"
  ) |> 
  complete(record_id = gic_encoded$record_id, fill = list(max_cycle_with_data = 0)) |> 
  left_join(select(gic_encoded, record_id, cycles_given), by = "record_id", suffix = c("", "_orig")) |> 
  mutate(cycles_given = coalesce(cycles_given, cycles_given_orig)) |> 
  select(-cycles_given_orig)

# Identify extraneous data (data beyond official count)
extraneous_data <- patient_max_cycles |> 
  filter(max_cycle_with_data > cycles_given) |> 
  mutate(excess_cycles = max_cycle_with_data - cycles_given) |> 
  arrange(desc(max_cycle_with_data))

# Display extraneous
if (nrow(extraneous_data) > 0) {
  warning("Found ", nrow(extraneous_data), " patients with data beyond their recorded cycle count.")
  extraneous_data |> 
    kbl(booktabs = TRUE, caption = "Extraneous Data Logic Check") |> 
    kable_styling()
} else {
  message("Extraneous cycle consistency check passed: No data found beyond `cycles_gi")
}

# Identify missing data
missing_cycle_data <- patient_max_cycles |> 
  filter(max_cycle_with_data < cycles_given) |> 
  mutate(missing_cycles = cycles_given - max_cycle_with_data) |> 
  arrange(desc(missing_cycles))

if (nrow(missing_cycle_data) > 0) {
  warning("Found ", nrow(missing_cycle_data), " patients missing records for completed cycles.")
  missing_cycle_data |> 
    kbl(caption = "Missing Cycle Records") |> 
    kable_styling() |> 
    scroll_box(height = "400px")
} else {
  message("Missing cycle consistency check passed: All administered cycles have data.")
}
```

::: {.callout-warning collapse="false"}
### Anomaly: Extraneous Cycle Data

**Issue:** The audit identified 4 patients with data recorded for cycles greater than their `cycles_given` count.

* **Affected IDs:** 27, 146, 498, 766
* **Status:** Flagged as **DAT-019**. Investigation required to determine if the administration count is incorrect or if the data should be removed.
:::

```{r}
#| label: investigate_extraneous-data
#| results: asis

# Define problem IDs:
problem_ids <- c(27, 146, 498, 766)

# Define panel groups for readability
var_groups <- tibble(
  stem = c(
    "hb", "mcv", "mch", "plt", "wbc", "anc", "lcc",                      # CBC
    "sodium", "potassium",                                               # Chem
    "tp", "alb", "tbil", "dbil", "alp", "alt", "ast", "ggt",             # LFT
    "cycle_delayed", "hosp_toxicity", "hosp_days", "transfusion_given",  # Mgmt
    "transfusion_units", "csfc_indicated"                                # Mgmt
  ),
  panel = c(
    rep("1. CBC", 7),
    rep("2. Chem", 2),
    rep("3. LFT", 8),
    rep("4. Mgmt", 6)
  )
)

# Function to view patient history
view_pt_hx <- function(pid, df) {
  # Get official count
  n_given <- df |> filter(record_id == pid) |> pull(cycles_given)
  
  # Extract and pivot data
  pt_data <- df |> 
    filter(record_id == pid) |> 
    select(matches("_[1-9]\\d*$")) |> 
    pivot_longer(
      cols = everything(),
      names_to = c("stem", "cycle"),
      names_pattern = "(.*)_(\\d+)$",
      values_transform = list(value = as.character) 
    ) |> 
    mutate(cycle = as.numeric(cycle)) |> 
    
    # Filter to only relevant cycles
    filter(cycle >= n_given | !is.na(value)) |> 
    
    # Join with panel info for sorting
    left_join(var_groups, by = "stem") |>
    arrange(panel, stem)
  
  # Create visuyalization
  pt_wide <- pt_data |> 
    select(panel, stem, cycle, value) |> 
    pivot_wider(names_from = cycle, values_from = value)
  cycle_cols <- setdiff(names(pt_wide), c("panel", "stem"))
  
  # 2. Filter for the bad ones
  bad_cols <- cycle_cols[map_lgl(cycle_cols, function(col) {
    # Logic: Is it a future cycle?
    is_future <- as.numeric(col) > n_given
    # Logic: Does it have data? (Not all NAs)
    has_data  <- !all(is.na(pt_wide[[col]]))
    
    return(is_future && has_data)
  })]
  
  # 3. Get indices for kable
  target_indices <- which(names(pt_wide) %in% bad_cols)
  
  # Render
  pt_wide |> 
    kbl(booktabs = T,
        caption = paste("Pt Data: Record ID", pid,
                        "(Cycles Given:", n_given, ")")) |> 
    kable_styling(font_size = 11) |> 
    column_spec(target_indices, background = cb_light[2])
}

walk(problem_ids, function(pid) {
  print(view_pt_hx(pid, gic_encoded))
  
  # Add a visual spacer between tables so they don't merge
  cat("\n<br><br>\n") 
})
```

```{r}
#| label: fix_extraneous-data

# Filter problem IDs out of working dataset until correct values determined
gic_wrk <- gic_encoded |> 
  filter(!record_id %in% problem_ids)

message("Excluded ", length(problem_ids), " records with ambiguous cycle data.")
message("Remaining number of patients: ", nrow(gic_wrk))
```

::: {.callout-note}
### Action Taken: Exclusion of Ambiguous Records

**Issue:** 4 records were identified with extraneous cycle data (data present for cycles beyond the recorded `cycles_given`).
**Investigation:** Visual inspection confirmed data inconsistencies (e.g., Record 146 has data for Cycle 1 and Cycle 12 only; Record 498 has zero-values in Cycle 8).
**Resolution:** These records have been excluded from the analysis dataset pending clarification.
:::


## Treatment Adherence Logic
Verify that the administered cycles do not exceed prescribed cycles.
```{r}
#| label: assess_cycle-counts

# Check that `cycles_given` <= `cycles_prescribed`
over_treated <- gic_wrk |> 
  filter(cycles_given > cycles_prescribed) |> 
  select(record_id, cycles_prescribed, cycles_given) |> 
  mutate(excess = cycles_given - cycles_prescribed) |> 
  arrange(desc(excess))

if (nrow(over_treated) > 0) {
  warning("Found ", nrow(over_treated), " patients with more cycles given than prescribed.")
  
  over_treated |> 
    kbl(booktabs = TRUE, caption = "Treatment Adherence Logic Check") |> 
    kable_styling(full_width = FALSE) |> 
    row_spec(1:nrow(over_treated), background = cb_light[2])
} else {
  message("Treatment adherence check passed.")
}
```

## Composite Variable Checks
Verify that pre-calculated fields match their components.
```{r}
#| label: assess_calculated-vars

# Define tolerance thresholds
bmi_tol <- 0.5
rate_tol <- 1.0

composite_audit <- gic_wrk |> 
  select(record_id, height_0, weight_0, bmi_0, 
         cycles_given, cycles_prescribed, completion_rate) |> 
  mutate(
    # 1. Calculate fresh BMI (Height is in cm, convert to m)
    bmi_calc = weight_0 / (height_0 / 100)^2,
    
    # 2. Calculate fresh Completion Rate
    rate_calc = (cycles_given / cycles_prescribed) * 100,
    
    # 3. Measure Diff
    bmi_diff = abs(bmi_0 - bmi_calc),
    rate_diff = abs(completion_rate - rate_calc)
  ) |> 
  # Filter for mismatches exceeding tolerance
  # (We use !is.na checks to ignore rows where data is missing)
  filter(
    (!is.na(bmi_diff) & bmi_diff > bmi_tol) | 
    (!is.na(rate_diff) & rate_diff > rate_tol)
  ) |> 
  arrange(desc(bmi_diff))

if (nrow(composite_audit) > 0) {
  warning("Found ", nrow(composite_audit), " records with composite variable mismatches.\n")
  
  composite_audit |> 
    select(record_id, 
           bmi_orig = bmi_0, bmi_new = bmi_calc, bmi_diff,
           rate_orig = completion_rate, rate_new = rate_calc) |> 
    kbl(caption = "Composite Variable Logic Check", digits = 1) |> 
    kable_styling() |> 
    scroll_box(height = "400px")
} else {
  message("Composite Check Passed: BMI and Completion Rates match inputs within tolerance.")
}
```

## Temporal Consistency
Verify that event dates occur within logical timeframes.
```{r}
#| label: assess_temporal-logic

# Check death_date vs study time window

```

# Numeric Range Validation
## Body Measurement Outliers
Identify clinically impossible values for height and weight.

```{r}
#| label: assess_body-measurements

# Set wide limits to try and only catch errors, not outliers
limits <- list(
  ht_min = 100, ht_max = 250,
  wt_min = 30, wt_max = 200,
  bmi_min = 10, bmi_max = 50
)

# Identify outliers
body_outliers <- gic_wrk |> 
  select(record_id, height_0, weight_0, bmi_0) |> 
  filter(
    height_0 < limits$ht_min | height_0 > limits$ht_max |
    weight_0 < limits$wt_min | weight_0 > limits$wt_max |
    bmi_0 < limits$bmi_min   | bmi_0 > limits$bmi_max
  ) |> 
  arrange(height_0)

if (nrow(body_outliers) > 0) {
  warning("Found ", nrow(body_outliers), " records with impossible body measurements.")
  
  body_outliers |> 
    kbl(caption = "Body Measurement Outliers", digits = 1) |> 
    kable_styling(full_width = FALSE) |> 
    row_spec(which(body_outliers$height_0 > 250), background = cb_light[2])
} else {
  message("Body Measurement Check Passed.")
}

# Visualize results
gic_wrk |> 
  select(height_0, weight_0, bmi_0) |> 
  pivot_longer(everything(), names_to = "measure", values_to = "value") |> 
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = cb_light[3], color = "white") +
  facet_wrap(~measure, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Body Measurements", x = NULL, y = "Count")
```

```{r}
#| label: fix_body-measurements

# Temporary resolution for DAT-020
gic_wrk <- gic_wrk |> 
  mutate(
    # Fix suspected height typo
    `height_0` = if_else(record_id == 614, 149.8, height_0),
    # Recalculate BMI
    bmi_0 = if_else(record_id == 614, weight_0 / (height_0 / 100)^2, bmi_0)
  )

```

::: {.callout-note}
### Action Taken [DAT-020]: Body Measurement Correction 

**Issue:** Record 614 contained an impossible height (1498 cm).
**Temporary Resolution:** Height set to 149.8 cm (suspected decimal error). BMI recalculated for this record only; all other BMI values remain as imported.
:::

## Biomarker Range Checks
Scan laboratory values for biologically impossible values (e.g., negative counts, values 10x above physiological max, etc.).
```{r}
#| label: assess_biomarker-ranges

# Generate summary stats for longitudinal variables

```

# Missingness Analysis
## Missing Data Patterns
Examine patterns of missing data.

```{r}
#| label: assess_missingness

# Visualize missingness by cycle

```

```{r}
#| label: assess_panel-missingness

# Examine missingness patterns by blood panel across cycle to identify unexpectedly missing data

```


# Feature Engineering
## Adverse Event Grading
Apply grading logic to raw laboratory values using `grading_rules.csv`.

```{r}
#| label: derive_adverse-events

# Load adverse event grading rules
rules <- read_csv(here("data", "grading_rules.csv"), show_col_types = FALSE)

# Pivot, Grade, and Pivot Back
gic_graded <- gic_wrk |> 
  # Pivot Long
  select(record_id, matches("_\\d+$")) |> 
  pivot_longer(
    cols = -record_id,
    names_to = c("stem", "cycle"),
    names_pattern = "(.*)_(\\d+)$",
    values_to = "value",
    values_transform = list(value = as.numeric) # Ensure numeric for grading math
  ) |> 
  filter(!is.na(value)) |> 
  semi_join(rules, by = "stem") |> 
  
  # Apply Logic (Min < Val <= Max)
  left_join(rules, by = "stem", relationship = "many-to-many") |> 
  filter(value > min_val & value <= max_val) |> 
  
  # Pivot Wide (Create labels)
  mutate(new_col = paste0(stem, "_grade_", cycle)) |> 
  select(record_id, new_col, label) |> 
  pivot_wider(names_from = new_col, values_from = label)

# Join back to main data
gic_validated <- gic_wrk |> 
  left_join(gic_graded, by = "record_id")

# Print a random sample of grades to see if it worked, will perform audit later
gic_validated |> 
  select(record_id, hb_1, hb_grade_1) |> 
  # Filter for patients who have data for the given cycle
  filter(!is.na(hb_1)) |> 
  head(10) |> 
  kbl(caption = "Verification: HB Cycle 1 Grading") |> 
  kable_styling(full_width = FALSE)

gic_validated |> 
  select(record_id, ast_5, ast_grade_5) |> 
  # Filter for patients who have data for the given cycle
  filter(!is.na(ast_5)) |> 
  head(10) |> 
  kbl(caption = "Verification: AST Cycle 5 Grading") |> 
  kable_styling(full_width = FALSE)

gic_validated |> 
  select(record_id, wbc_6, wbc_grade_6) |> 
  # Filter for patients who have data for the given cycle
  filter(!is.na(wbc_6)) |> 
  head(10) |> 
  kbl(caption = "Verification: WBC Cycle 6 Grading") |> 
  kable_styling(full_width = FALSE)
```

# Export
```{r}
#| label: save_final-data

# Define path to file
gic_validated_path <- here("data", "processed", "gic_validated.rds")

# Save if file doesn't exist, or check if overwrite is intended if it does.
if (file.exists(gic_validated_path) && interactive()) {
  # If file exists and this is being run manually in RStudio ask before overwriting
  response <- readline(prompt = paste("Warning:", basename(gic_validated_path), "already exists. Overwrite? (y/n):"))
  
  if (tolower(substr(response, 1, 1)) == "y") {
    saveRDS(gic_validated, gic_validated_path)
    message("File overwritten: ", basename(gic_validated_path))
  } else {
    message("Save operation cancelled.")
  }
} else {
  # If the file doesn't exist or document is being rendered, just save it
  saveRDS(gic_validated, gic_validated_path)
  message("File saved: ", basename(gic_validated_path))
}
```