---
title: "GI Cancer: Data Validation"
bibliography: references.bib
---
# Load Data and Libraries
```{r}
#| label: setup
#| include: false

source("R/setup.R")
```

```{r}
#| label: load-data

gic_encoded <- readRDS(here("data", "processed", "gic_encoded.rds"))
```

# Logic & Consistency Validation
## Cycle Consistency
Verify that `cycles_given` matches the actual presence of data in the cycle columns. Flag patients who have data present in cycles after their administered number.

```{r}
#| label: assess-extraneous-data

# Pivot longitudinal data to make finding missing data easier
real_data_map <- gic_encoded |> 
  select(record_id, pgh_number, cycles_prescribed, cycles_given, matches("_[1-9]\\d*$")) |> 
  pivot_longer(
    cols = matches("_\\d+$"),
    names_to = c("stem", "cycle"),
    names_pattern = "(.*)_(\\d+)$",
    values_to = "value",
    values_transform = list(value = as.character)
  ) |> 
  mutate(cycle = as.integer(cycle)) |> 
  left_join(var_groups, by = "stem") |> 
  mutate(type = if_else(panel == "4. Mgmt", "Mgmt", "Labs"))

# Summarize records to check if data matches cycles_given
patient_audit <- real_data_map |> 
  group_by(record_id, pgh_number, cycles_given) |> 
  summarize(
    # Check number of cycles that have labwork data
    n_lab_cycles = n_distinct(cycle[type == "Labs" & !is.na(value)]),
    # Check number of cycles that have management data
    n_mgmt_cycles = n_distinct(cycle[type == "Mgmt" & !is.na(value)]),
    # Check if number of lab cycles matches number of management cycles
    labs_mgmt_match = n_lab_cycles == n_mgmt_cycles,
    # For cycles with management data, check if all is present
    mgmt_is_complete = {
      required_vars <- expand.grid(s = mgmt_stems, c = 1:first(cycles_given)) |> 
        mutate(fullname = paste0(s, "_", c)) |> 
        pull(fullname)
      actual_vars <- paste0(stem, "_", cycle)[!is.na(value) & type == "Mgmt"]
      all(required_vars %in% actual_vars)
    },
    # Check the last cycle that had any data entered
    max_cycle_any = max(cycle[!is.na(value)], na.rm = TRUE),
    .groups = "drop"
  ) |> 
  mutate(max_cycle_any = if_else(is.infinite(max_cycle_any), 0, max_cycle_any))

# Flag pts with data that doesn't match tx hx
problem_ids <- patient_audit |> 
  filter(
    max_cycle_any > cycles_given |
    n_mgmt_cycles != cycles_given |
    n_lab_cycles != cycles_given |
    labs_mgmt_match == FALSE |
    mgmt_is_complete == FALSE
  ) |> 
  pull(record_id)

# Display audit table
if (length(problem_ids) > 0) {
  warning("Found discrepancies for ", length(problem_ids), " records.")
  
  patient_audit |> 
    filter(record_id %in% problem_ids) |> 
    kbl(booktabs = TRUE, caption = "Strict Cycle Consistency Audit") |> 
    kable_styling(full_width = FALSE)
} else {
  message("All cycle counts match data presence and management completeness.")
}
```


```{r}
#| label: investigate-extraneous-data
#| output: asis

view_pt_hx <- function(pid, df, audit_data = NULL, show_empty = TRUE) {
  # Get official count and audit info for the pt
  pt_row <- df |> filter(record_id == pid)
  n_given <- pt_row |> pull(cycles_given)

  audit_row <- if (!is.null(audit_data)) audit_data |> filter(record_id == pid) else NULL

  # Extract and pivot longitudinal data
  pt_data <- pt_row |>
    select(matches("_[1-9]\\d*$")) |>
    pivot_longer(
      cols = everything(),
      names_to = c("stem", "cycle"),
      names_pattern = "(.*)_(\\d+)$",
      values_to = "value",
      values_transform = list(value = as.character)
    ) |>
    mutate(cycle = as.numeric(cycle)) |>
    # Keep cycles based on tx hx or data presence
    filter(if (show_empty) cycle <= n_given | !is.na(value) else !is.na(value)) |>
    left_join(var_groups, by = "stem") |>
    arrange(panel, stem)

  # Create wide format
  pt_wide <- pt_data |>
    select(panel, stem, cycle, value) |>
    pivot_wider(names_from = cycle, values_from = value)

  cycle_cols <- as.numeric(setdiff(names(pt_wide), c("panel", "stem")))

  # Build table
  tbl <- pt_wide |>
    gt(groupname_col = "panel", rowname_col = "stem") |>
    tab_header(
      title = paste("Pt Data: Record ID", pid, "(Given:", n_given, ")")
    ) |>
    tab_options(table.font.size = px(11))

  # Highlight affected columns
  for (cyc in cycle_cols) {
    cyc_chr <- as.character(cyc)
    has_any_data <- any(!is.na(pt_wide[[cyc_chr]]))

    # Check for missing management in this cycle
    mgmt_in_cycle <- pt_data |>
      filter(cycle == cyc, panel == "4. Mgmt") |>
      pull(value)
    is_mgmt_incomplete <- any(is.na(mgmt_in_cycle))

    fill <- if (cyc > n_given) {
      cb_light[2]          # Data recorded beyond the official count
    } else if (!has_any_data) {
      cb_light[4]          # Entirely empty cycles that should have data
    } else if (is_mgmt_incomplete) {
      cb_light[6]          # Partial management data
    } else {
      NA_character_
    }

    if (!is.na(fill)) {
      tbl <- tbl |>
        tab_style(
          style = cell_fill(color = fill),
          locations = cells_body(columns = all_of(cyc_chr))
        )
    }
  }

  tbl
}

# Run audit for flagged records
cat("::: {.panel-tabset}\n\n")

for (current_pid in problem_ids) {
  cat(knitr::knit_child(
    text = c(
      "## `r current_pid`",
      "",
      "```{r}",
      "#| echo: false",
      "view_pt_hx(current_pid, gic_encoded, audit_data = patient_audit, show_empty = TRUE)",
      "```",
      ""
    ),
    envir = environment(),
    quiet = TRUE
  ))
}

cat(":::\n\n")
```


## Treatment Adherence Logic
Verify that the administered cycles do not exceed prescribed cycles.
```{r}
#| label: assess-cycle-counts

# Check that `cycles_given` <= `cycles_prescribed`
over_treated <- gic_encoded |> 
  filter(cycles_given > cycles_prescribed) |> 
  select(record_id, cycles_prescribed, cycles_given) |> 
  mutate(excess = cycles_given - cycles_prescribed) |> 
  arrange(desc(excess))

if (nrow(over_treated) > 0) {
  warning("Found ", nrow(over_treated), " patients with more cycles given than prescribed.")
  
  over_treated |> 
    kbl(booktabs = TRUE, caption = "Treatment Adherence Logic Check") |> 
    kable_styling(full_width = FALSE) |> 
    row_spec(1:nrow(over_treated), background = cb_light[2])
} else {
  message("Treatment adherence check passed.")
}
```

## Composite Variable Checks
Verify that pre-calculated fields match their components.
```{r}
#| label: assess-calculated-vars

# Define tolerance thresholds
bmi_tol <- 0.5
rate_tol <- 1.0

composite_audit <- gic_encoded |> 
  select(record_id, height_0, weight_0, bmi_0, 
         cycles_given, cycles_prescribed, completion_rate) |> 
  mutate(
    # 1. Calculate fresh BMI (Height is in cm, convert to m)
    bmi_calc = weight_0 / (height_0 / 100)^2,
    
    # 2. Calculate fresh Completion Rate
    rate_calc = (cycles_given / cycles_prescribed) * 100,
    
    # 3. Measure Diff
    bmi_diff = abs(bmi_0 - bmi_calc),
    rate_diff = abs(completion_rate - rate_calc)
  ) |> 
  # Filter for mismatches exceeding tolerance
  # (We use !is.na checks to ignore rows where data is missing)
  filter(
    (!is.na(bmi_diff) & bmi_diff > bmi_tol) | 
    (!is.na(rate_diff) & rate_diff > rate_tol)
  ) |> 
  arrange(desc(bmi_diff))

if (nrow(composite_audit) > 0) {
  warning("Found ", nrow(composite_audit), " records with composite variable mismatches.\n")
  
  composite_audit |> 
    select(record_id, 
           bmi_orig = bmi_0, bmi_new = bmi_calc, bmi_diff,
           rate_orig = completion_rate, rate_new = rate_calc) |> 
    kbl(caption = "Composite Variable Logic Check", digits = 1) |> 
    kable_styling() |> 
    scroll_box(height = "400px")
} else {
  message("Composite Check Passed: BMI and Completion Rates match inputs within tolerance.")
}
```
# Transfusion Logic Check
```{r}
#| label: assess-transfusions

# [DAT-008] Audit transfusion data for consistency and unit volume errors.

# Pivot to handle all cycles at once
transfusion_long <- gic_encoded |> 
  select(record_id,
         pgh_number,
         starts_with("transfusion_given_"), 
         starts_with("transfusion_units_")) |>
  pivot_longer(
    cols = -c(record_id, pgh_number),
    names_to = c(".value", "cycle"),
    names_pattern = "(.*)_(\\d+)$"
  ) |> 
  mutate(cycle = as.numeric(cycle)) |> 
  arrange(cycle)

# Summarize by Cycle
# transfusion_summary <- transfusion_long |> 
#   group_by(cycle) |> 
#   summarize(
#     n_given = sum(transfusion_given == "Yes", na.rm = TRUE),
#     
#     # [DAT-008] Check for missing units (Given = Yes, Units = NA/0)
#     missing_units = sum(transfusion_given == "Yes" & 
#                         (is.na(transfusion_units) | transfusion_units == 0), na.rm = TRUE),
#     
#     # [DAT-008] Check for unexpected units (Given = No, Units > 0)
#     unexpected_units = sum(transfusion_units > 0 & 
#                            (transfusion_given != "Yes" | is.na(transfusion_given)), na.rm = TRUE),
#     
#     # [DAT-008] Original redundancy check (Given == Units)
#     is_identical = identical(
#       as.numeric(transfusion_given == "Yes"), 
#       as.numeric(transfusion_units)
#     )
#   )

transfusion_summary <- transfusion_long |> 
  group_by(cycle) |> 
  summarize(
    n_rows = n(),
    n_flag_yes = sum(transfusion_given == "Yes", na.rm = TRUE),
    n_flag_no = sum(transfusion_given == "No", na.rm = TRUE),
    n_flag_na = sum(is.na(transfusion_given)),
    
    n_units_na = sum(is.na(transfusion_units)),
    n_units_present = sum(!is.na(transfusion_units)),
    n_units_pos = sum(transfusion_units > 0, na.rm = TRUE),
    n_units_zero = sum(transfusion_units == 0, na.rm = TRUE),
    
    # Check for missing units (Given == Yes && !(Units > 0))
    flag_yes_units_missing = sum(
      transfusion_given == "Yes" & is.na(transfusion_units), na.rm = TRUE
    ),
    flag_yes_units_not_pos = sum(
      transfusion_given == "Yes" & !is.na(transfusion_units) & transfusion_units <= 0, na.rm = TRUE
    ),
    
    # Check for unexpected units (Given != Yes && Units > 0)
    flag_not_yes_units_pos = sum(
      (transfusion_given != "Yes" | is.na(transfusion_given)) &
        !is.na(transfusion_units) & transfusion_units > 0, na.rm = TRUE
    ),
    
    # Check for incomplete records (one recorded but not the other)
    flag_missing_units_filled = sum(
      is.na(transfusion_given) & !is.na(transfusion_units), na.rm = TRUE
    ),
    flag_filled_units_missing = sum(
      !is.na(transfusion_given) & is.na(transfusion_units), na.rm = TRUE
    ),
    
    # Check for invalid transfusion_units values
    units_negative = sum(transfusion_units < 0, na.rm = TRUE),
    
    # Check for cases with transfusion_given and transfusion_units are the same
    suspicious_units_equal_flag = {
      idx <- !is.na(transfusion_given) & !is.na(transfusion_units)
      given_01 <- as.integer(transfusion_given[idx] == "Yes")
      units <- transfusion_units[idx]
      length(units) > 0 && all(units %in% c(0, 1)) && all(units == given_01)
    },
    
    .groups = "drop"
  )

# Print Table
error_rows <- which(
  transfusion_summary$flag_yes_units_missing > 0 |
  transfusion_summary$flag_yes_units_not_pos > 0 |
  transfusion_summary$flag_not_yes_units_pos > 0 |
  transfusion_summary$flag_missing_units_filled > 0 |
  transfusion_summary$flag_filled_units_missing > 0 |
  transfusion_summary$units_negative > 0
)

transfusion_summary |> 
  kbl(booktabs = TRUE, caption = "Transfusion Logic & Consistency Check") |> 
  kable_styling(full_width = FALSE) |> 
  row_spec(error_rows, background = cb_light[2])
```
::: {.callout-warning collapse="false"}
### Anomaly: Transfusion Data Inconsistencies (DAT-008)

**Issue:** The audit revealed inconsistencies between logging if a transfusion was given and the number of units given as part of a transfusion

* **Missing Units:** 3 records are flagged as having received a transfusion ("Yes") but have zero or missing unit volume recorded. (Data incompleteness).
* **Flag Mismatch:** 3 records are flagged as NOT receiving a transfusion ("No") but contain a valid unit count (>0). (Flag error).

**Status:** Flagged for cleanup in the `fix_...` chunk. The red rows indicate cycles with **at least one** internal inconsistency.
:::

```{r}
#| label: investigate-transfusion-records

# Identify patient-cycle inconsistencies aligned with assess-transfusions summary logic

transfusion_error_records <- transfusion_long |>
  filter(
    # A1: Given == Yes but units missing
    (transfusion_given == "Yes" & is.na(transfusion_units)) |

    # A2: Given == Yes but units not positive (0 or negative)
    (transfusion_given == "Yes" & !is.na(transfusion_units) & transfusion_units <= 0) |

    # B: Given is No/NA (i.e., not Yes) but units > 0
    ((transfusion_given != "Yes" | is.na(transfusion_given)) &
       !is.na(transfusion_units) & transfusion_units > 0) |

    # C1: Units recorded but flag missing (counts as incomplete record)
    (is.na(transfusion_given) & !is.na(transfusion_units)) |

    # C2: Flag recorded but units missing (counts as incomplete record)
    (!is.na(transfusion_given) & is.na(transfusion_units)) |

    # D: Invalid negative units
    (!is.na(transfusion_units) & transfusion_units < 0)
  ) |>
  select(record_id, pgh_number, cycle, transfusion_given, transfusion_units) |>
  arrange(record_id, cycle)

# Display the failed records
if (nrow(transfusion_error_records) > 0) {
  warning("Found ", nrow(transfusion_error_records), " individual patient-cycle transfusion inconsistencies requiring review.")
  transfusion_error_records |>
    kbl(caption = "Patient-Cycle Transfusion Inconsistencies") |>
    kable_styling(full_width = TRUE) |>
    scroll_box(height = "400px")
} else {
  message("All individual transfusion records are logically consistent under current audit rules.")
}
```

# Numeric Range Validation
## Body Measurement Outliers
Identify clinically impossible values for height and weight.

```{r}
#| label: assess-body-measurements

# Set wide limits to try and only catch errors, not outliers
limits <- list(
  ht_min = 100, ht_max = 250,
  wt_min = 30, wt_max = 200,
  bmi_min = 10, bmi_max = 50
)

# Identify outliers
body_outliers <- gic_encoded |> 
  select(record_id, height_0, weight_0, bmi_0) |> 
  filter(
    height_0 < limits$ht_min | height_0 > limits$ht_max |
    weight_0 < limits$wt_min | weight_0 > limits$wt_max |
    bmi_0 < limits$bmi_min   | bmi_0 > limits$bmi_max
  ) |> 
  arrange(height_0)

if (nrow(body_outliers) > 0) {
  warning("Found ", nrow(body_outliers), " records with impossible body measurements.")
  
  body_outliers |> 
    kbl(caption = "Body Measurement Outliers", digits = 1) |> 
    kable_styling(full_width = FALSE) |> 
    row_spec(which(body_outliers$height_0 > 250), background = cb_light[2])
} else {
  message("Body Measurement Check Passed.")
}

# Visualize results
gic_encoded |> 
  select(height_0, weight_0, bmi_0) |> 
  pivot_longer(everything(), names_to = "measure", values_to = "value") |> 
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = cb_light[3], color = "white") +
  facet_wrap(~measure, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Body Measurements", x = NULL, y = "Count")
```

::: {.callout-note}
### Action Taken [DAT-020]: Body Measurement Correction 

**Issue:** Record 614 contained an impossible height (1498 cm).
**Temporary Resolution:** Height set to 149.8 cm (suspected decimal error). BMI recalculated for this record only; all other BMI values remain as imported.
:::

## Biomarker Range Checks
Scan laboratory values for biologically impossible values (e.g., negative counts, values 10x above physiological max, etc.).
```{r}
#| label: assess-biomarker-ranges

# Generate summary stats for longitudinal variables

```

# Missingness Analysis
## Missing Data Patterns
Examine patterns of missing data.

```{r}
#| label: assess-missingness

# Visualize missingness by cycle

```

```{r}
#| label: assess-panel-missingness

# Examine missingness patterns by blood panel across cycle to identify unexpectedly missing data

```


# Feature Engineering
## Adverse Event Grading
Apply grading logic to raw laboratory values using `grading_rules.csv`.

```{r}
#| label: derive-adverse-events

# Load adverse event grading rules
rules <- read_csv(here("data", "grading_rules.csv"), show_col_types = FALSE)

# Long-format graded data
ae_graded_long <- gic_encoded |> 
  select(record_id, matches("_\\d+$")) |> 
  pivot_longer(
    cols = -record_id,
    names_to = c("stem", "cycle"),
    names_pattern = "(.*)_(\\d+)$",
    values_to = "value",
    values_transform = list(value = as.numeric)
  ) |> 
  mutate(cycle = as.integer(cycle)) |> 
  filter(!is.na(value)) |> 
  semi_join(rules, by = "stem") |> 
  left_join(rules, by = "stem", relationship = "many-to-many") |> 
  filter(value > min_val & value <= max_val) |> 
  mutate(
    ae_type = if_else(stem %in% c("sodium", "potassium"), tolower(condition), stem)
  ) |> 
  select(record_id, stem, ae_type, cycle, value, grade, condition, label)

# Wide grade columns for human-readable auditing
gic_graded_wide <- ae_graded_long |> 
  mutate(new_col = paste0(ae_type, "_grade_", cycle)) |> 
  select(record_id, new_col, label) |> 
  pivot_wider(names_from = new_col, values_from = label)


gic_validated <- gic_encoded |> 
  left_join(gic_graded_wide, by = "record_id")

# Print a random sample of grades to see if it worked, will perform audit later
gic_validated |> 
  select(record_id, hb_1, hb_grade_1) |> 
  # Filter for patients who have data for the given cycle
  filter(!is.na(hb_1)) |> 
  head(10) |> 
  kbl(caption = "Verification: HB Cycle 1 Grading") |> 
  kable_styling(full_width = FALSE)

gic_validated |> 
  select(record_id, ast_5, ast_grade_5) |> 
  # Filter for patients who have data for the given cycle
  filter(!is.na(ast_5)) |> 
  head(10) |> 
  kbl(caption = "Verification: AST Cycle 5 Grading") |> 
  kable_styling(full_width = FALSE)

gic_validated |> 
  select(record_id, wbc_6, wbc_grade_6) |> 
  # Filter for patients who have data for the given cycle
  filter(!is.na(wbc_6)) |> 
  head(10) |> 
  kbl(caption = "Verification: WBC Cycle 6 Grading") |> 
  kable_styling(full_width = FALSE)

gic_validated |> 
  select(record_id, sodium_3, hypernatremia_grade_3) |> 
  # Filter for patients who have data for the given cycle
  filter(!is.na(sodium_3)) |> 
  head(10) |> 
  kbl(caption = "Verification: Hypernatremia Cycle 3 Grading") |> 
  kable_styling(full_width = FALSE)
```

## Adverse Event Counts



# Export
```{r}
#| label: save-validated-data

save_rds_safe(gic_validated, here("data", "processed", "gic_validated.rds"))
save_rds_safe(ae_graded_long, here("data", "processed", "ae_graded_long.rds"))
```

