---
title: "GI Cancer: Exploratory Data Analysis"
bibliography: references.bib
---

# Load Data & Libraries
```{r}
#| label: setup
#| include: false

source("R/setup.R")

# Load Data
gic_validated <- readRDS(here("data", "processed", "gic_validated.rds"))

# Load Libraries
```


```{r}
#| label: engineer-ae-variables

# Identify all adverse event columns and pivot to check per-cycle status
ae_data <- gic_validated |> 
  select(record_id, matches("grade_\\d+$")) |> 
  pivot_longer(
    cols = -record_id,
    names_to = c("ae_type", "cycle"),
    names_pattern = "(.*)_grade_(\\d+)$",
    values_to = "grade"
  ) |> 
  mutate(
    cycle = as.integer(cycle),
    grade_num = as.numeric(str_extract(grade, "\\d")),
    is_ae = !is.na(grade_num) & grade_num > 0,
    is_severe = !is.na(grade_num) & grade_num >= 3
  )

# Summarize at patient level
ae_summary <- ae_data |> 
  group_by(record_id) |> 
  summarize(
    ae_cycle_count = n_distinct(cycle[is_ae == TRUE]),
    ae_early = if_else(any(is_ae == TRUE & cycle <= 4), 1, 0),
    had_severe_ae = if_else(any(is_severe == TRUE), 1, 0),
    severe_ae_early = if_else(any(is_severe == TRUE & cycle <= 4), 1, 0),
    .groups = "drop"
  )

# Join back to main dataset
gic_models <- gic_validated |> 
  left_join(ae_summary, by = "record_id") |> 
  mutate(
    had_any_ae = if_else(ae_cycle_count > 0, 1, 0)
  )
```

```{r}
#| label: tbl-ae-summary
#| tbl-cap: "Summary of Adverse Events by HIV Status"

gic_models |> 
  group_by(hiv_status) |> 
  summarize(
    total_pts = n(),
    
    any_ae_n = sum(had_any_ae),
    any_ae_pct = (any_ae_n / total_pts) * 100,
    
    severe_ae_n = sum(had_severe_ae),
    severe_ae_pct = (severe_ae_n / total_pts) * 100,
    
    early_ae_n = sum(ae_early),
    early_ae_pct = (early_ae_n / total_pts) * 100,
    
    severe_early_ae_n = sum(severe_ae_early),
    severe_early_ae_pct = (severe_early_ae_n / total_pts) * 100,
    .groups = "drop"
  ) |> 
  mutate(
    "Any AE (%)" = paste0(any_ae_n, " (", round(any_ae_pct, 2), "%)"),
    "Severe AE (%)" = paste0(severe_ae_n, " (", round(severe_ae_pct, 2), "%)"),
    "Early AE (%)" = paste0(early_ae_n, " (", round(early_ae_pct, 2), "%)"),
    "Severe Early AE (%)" = paste0(severe_early_ae_n, " (", round(severe_early_ae_pct, 2), "%)")
  ) |> 
  select(hiv_status, total_pts, "Any AE (%)", "Severe AE (%)", "Early AE (%)", "Severe Early AE (%)") |> 
  kbl(booktabs = TRUE, align = "lcccc") |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

# Model Building
## Prelim Tables for covariates
```{r}
#| label: tbl-simple

# 1. Any AE vs HIV
table(Any_AE = gic_models$had_any_ae, HIV = gic_models$hiv_status) |> 
  as.data.frame.matrix() |> 
  kbl(booktabs = TRUE, caption = "Any AE vs HIV Status") |> 
  kable_styling(full_width = FALSE, position = "left") |> 
  add_header_above(c("Any AE (0=No, 1=Yes)" = 1, "HIV Status" = 2))

# 2. Severe AE vs HIV
table(Severe_AE = gic_models$had_severe_ae, HIV = gic_models$hiv_status) |> 
  as.data.frame.matrix() |> 
  kbl(booktabs = TRUE, caption = "Severe AE (Grades 3-4) vs HIV Status") |> 
  kable_styling(full_width = FALSE, position = "left") |> 
  add_header_above(c("Severe AE (0=No, 1=Yes)" = 1, "HIV Status" = 2))

# 3. Early AE vs HIV
table(Early_AE = gic_models$ae_early, HIV = gic_models$hiv_status) |> 
  as.data.frame.matrix() |> 
  kbl(booktabs = TRUE, caption = "Early AE (Cycles 1-4) vs HIV Status") |> 
  kable_styling(full_width = FALSE, position = "left") |> 
  add_header_above(c("Early AE (0=No, 1=Yes)" = 1, "HIV Status" = 2))

# 4. Severe Early AE vs HIV
table(Severe_Early_AE = gic_models$severe_ae_early, HIV = gic_models$hiv_status) |> 
  as.data.frame.matrix() |> 
  kbl(booktabs = TRUE, caption = "Severe Early AE (Cycles 1-4) vs HIV Status") |> 
  kable_styling(full_width = FALSE, position = "left") |> 
  add_header_above(c("Severe Early AE (0=No, 1=Yes)" = 1, "HIV Status" = 2))

# table(gic_models$cancer_stage, gic_models$hiv_status)
# 
# table(gic_models$chemo_type, gic_models$hiv_status)
# 
# table(gic_models$tx_intent, gic_models$hiv_status)
```

```{r}
#| label: tbl-toxicity-raw-counts

# Any AE vs HIV
table(Any_AE = gic_models$had_any_ae, HIV = gic_models$hiv_status) |> 
  as.data.frame.matrix() |> 
  kbl(booktabs = TRUE, caption = "Any AE vs HIV Status") |> 
  kable_styling(full_width = FALSE, position = "left") |> 
  add_header_above(c("Any AE Status" = 1, "HIV Status" = 2))

# Severe AE vs HIV
table(Severe_AE = gic_models$had_severe_ae, HIV = gic_models$hiv_status) |> 
  as.data.frame.matrix() |> 
  kbl(booktabs = TRUE, caption = "Severe AE vs HIV Status") |> 
  kable_styling(full_width = FALSE, position = "left") |> 
  add_header_above(c("Severe AE Status" = 1, "HIV Status" = 2))

# Early AE vs HIV
table(Early_AE = gic_models$ae_early, HIV = gic_models$hiv_status) |> 
  as.data.frame.matrix() |> 
  kbl(booktabs = TRUE, caption = "Early AE (C1-4) vs HIV Status") |> 
  kable_styling(full_width = FALSE, position = "left") |> 
  add_header_above(c("Early AE Status" = 1, "HIV Status" = 2))
```


:::{.callout-warning collapse="false"}
### Warning: Categorical Variables with Insufficient Patients

- For cancer stage, only 3 HIV positive patients were stage II
- For chemo type, all of the types had at least 10 HIV-negative patients
  - Platinum/Paclitaxel and LCV/5-FU had 6 and 8 HIV-positive patients, respectively
  - CapeOX, FOLFOX4, and Capecitabine had 4, 2, and 2 HIV-positive patients, respectively
:::

## Simple Logistic: Full treatment period
```{r}
#| label: model-simple-full

model_any_simple <- glm(had_any_ae ~ hiv_status, 
                        data = gic_models, family = binomial)

model_severe_simple <- glm(had_severe_ae ~ hiv_status,
                           data = gic_models, family = binomial)

# summary(model_any_simple)
summary(model_severe_simple)
```

## Simple logistic: Early treatment period
```{r}
#| label: model-simple-early

model_any_early_simple <- glm(ae_early ~ hiv_status,
                              data = gic_models, family = binomial)
model_severe_early_simple <- glm(severe_ae_early ~ hiv_status,
                                 data = gic_models, family = binomial)

# summary(model_any_early_simple)
summary(model_severe_early_simple)
```

## Multivariate Logistic: Full treatment period

```{r}
#| label: model-basic-full

model_any_basic <- glm(had_any_ae ~ hiv_status + cancer_stage + chemo_type + tx_intent,
                       data = gic_models, family = binomial)

model_severe_basic <- glm(had_severe_ae ~ hiv_status + cancer_stage + chemo_type + tx_intent,
                       data = gic_models, family = binomial)

# summary(model_any_basic)
summary(model_severe_basic)
```

## Multivariate Logistic: Early treatment period

```{r}
#| label: model-basic-early

model_any_early_basic <- glm(ae_early ~ hiv_status + cancer_stage + chemo_type + tx_intent,
                       data = gic_models, family = binomial)

model_severe_early_basic <- glm(severe_ae_early ~ hiv_status + cancer_stage + chemo_type + tx_intent,
                       data = gic_models, family = binomial)

summary(model_any_early_basic)
summary(model_severe_early_basic)
```

## $\chi^2$ Tests (HIV vs Early Events)

```{r}
#| label: tests-chi-sq

# Any early AE
table_any_early <- table(gic_models$ae_early, gic_models$hiv_status)
chisq_any <- if(any(table_any_early == 0)) "Incalculable (100% rate)" else chisq.test(table_any_early)

# Severe early AE
table_severe_early <- table(gic_models$severe_ae_early, gic_models$hiv_status)
chisq_severe <- if(any(table_severe_early == 0)) "Incalculable (100% rate)" else chisq.test(table_severe_early)

print(table_any_early)
print(chisq_any)

print(table_severe_early)
print(chisq_severe)
```









