---
title: "GI Cancer: Exploratory Data Analysis"
bibliography: references.bib
---

# Load Data & Libraries
```{r}
#| label: setup
#| warning: false

source("R/setup.R")

# Load Data
gic_validated <- readRDS(here("data", "processed", "gic_validated.rds"))

# Load Libraries
library(ggcorrplot)
```

# Variable Engineering

```{r}
#| label: engineer-ae-variables

# Identify adverse event grade columns
grade_cols <- names(gic_validated)[str_detect(names(gic_validated), "_grade_([1-9]|1[0-2])$")]
### Need to determine whether baselines should be included too

# Create long table to engineer new variables
ae_long <- gic_validated |> 
  select(record_id, cycles_prescribed, cycles_given, all_of(grade_cols)) |> 
  pivot_longer(
    cols = all_of(grade_cols),
    names_to = "grade_var",
    values_to = "grade_raw"
  ) |> 
  mutate(
    cycle = as.integer(str_extract(grade_var, "\\d+$")),
    ae_type = str_remove(grade_var, "_grade_\\d+$"),
    grade = as.numeric(str_match(grade_raw, "(?i)^\\s*grade\\s*(\\d+)")[,2])
  )

# Cycle-level summaries
cycle_summary <- ae_long |>
  group_by(record_id, cycles_prescribed, cycles_given, cycle) |>
  summarize(
    ae_n_markers_measured = sum(!is.na(grade)),
    .groups = "drop"
  ) |>
  left_join(
    ae_long |>
      filter(!is.na(grade)) |>
      group_by(record_id, cycle) |>
      summarize(
        ae_n_events = sum(grade > 0),
        ae_n_severe = sum(grade >= 3),
        ae_max_grade = max(grade),
        ae_any = any(grade > 0),
        ae_severe_any = any(grade >= 3),
        .groups = "drop"
      ),
    by = c("record_id", "cycle")
  )

ae_cycle_wide <- cycle_summary |>
  pivot_wider(
    id_cols = c(record_id, cycles_prescribed, cycles_given),
    names_from = cycle,
    values_from = c(ae_n_markers_measured, ae_max_grade, ae_n_events, ae_n_severe, ae_any, ae_severe_any),
    names_glue = "{.value}_{cycle}"
  )


# Patient-level summaries
ae_summary_pt <- ae_long |> 
  group_by(record_id) |> 
  summarize(
    ae_any_early = as.integer(any(grade > 0 & cycle <= 2, na.rm = TRUE)),
    ae_severe_any_pt = as.integer(any(grade >= 3, na.rm = TRUE)),
    ae_severe_any_early = as.integer(any(grade >= 3 & cycle <= 2, na.rm = TRUE)),
    ae_any_pt = as.integer(any(grade > 0, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  left_join(
    cycle_summary |>
      filter(ae_n_markers_measured > 0) |>   # only cycles with any grading recorded
      group_by(record_id) |>
      summarize(
        ae_max_grade_pt = max(ae_max_grade, na.rm = TRUE),
        .groups = "drop"
      ),
    by = "record_id"
  )

# Merge datasets for analysis
gic_models <- gic_validated |>
  left_join(ae_summary_pt, by = "record_id") |>
  left_join(ae_cycle_wide, by = c("record_id", "cycles_prescribed", "cycles_given"))
```

## Adverse Event Variables

| Variable                  | Level   | Type         | Meaning                                                                                                          |
| ------------------------- | ------- | ------------ | ---------------------------------------------------------------------------------------------------------------- |
| `ae_any_pt`               | Patient | int (0/1)    | Flag: did the patient have any AE (`grade > 0`) at any point in treatment?                                       |
| `ae_any_early`            | Patient | int (0/1)    | Flag: did the patient have any AE (`grade > 0`) in cycles 1–2?                                                   |
| `ae_severe_any_pt`        | Patient | int (0/1)    | Flag: did the patient have any severe AE (`grade >= 3`) across all cycles?                                       |
| `ae_severe_any_early`     | Patient | int (0/1)    | Flag: did the patient have any severe AE (`grade >= 3`) in cycles 1–2?                                           |
| `ae_max_grade_pt`         | Patient | numeric / NA | Maximum adverse event grade across all treatment cycles                                                          |
| ------------------------- | ------- | ------------ | ---------------------------------------------------------------------------------------------------------------- |
| `ae_max_grade_#`          | Cycle   | numeric / NA | For cycle {#}, maximum AE grade across markers (NA if `ae_n_markers_measured_# == 0`).                           |
| `ae_n_events_#`           | Cycle   | numeric / NA | For cycle {#}, count of AEs across markers (`grade > 0`) (NA if `ae_n_markers_measured_# == 0`).                 |
| `ae_n_severe_#`           | Cycle   | numeric / NA | For cycle {#}, count of severe AEs across markers (`grade >= 3`) (NA if `ae_n_markers_measured_# == 0`).         |
| `ae_any_#`                | Cycle   | logical / NA | For cycle {#}, flag: any AE this cycle (`ae_n_events_# > 0`) (NA if `ae_n_markers_measured_# == 0`).             |
| `ae_severe_any_#`         | Cycle   | logical / NA | For cycle {#}, flag: any severe AE this cycle (`ae_n_severe_# > 0`) (NA if `ae_n_markers_measured_# == 0`).      |



## Patient-Level AE Summaries
```{r}
#| label: tbl-ae-by-pt
#| tbl-cap: "Patient-level AE summaries by HIV status."

summ_n_pct <- function(x, digits = 1) {
  n <- sum(x == 1, na.rm = TRUE)
  pct <- 100 * n / sum(!is.na(x))
  paste0(n, " (", round(pct, digits), "%)")
}

# Function to calculate mean, standard deviation, and range
summ_mean_sd_range <- function(x, digits = 1) {
  x <- x[!is.na(x)]
  m <- mean(x); s <- sd(x)
  rng <- range(x)
  paste0(
    round(m, digits), " (", round(s, digits), "); ",
    round(rng[[1]], digits), "–", round(rng[[2]], digits)
  )
}

summ_mean_sd_iqr <- function(x, digits = 1) {
  x <- x[!is.na(x)]
  m <- mean(x); s <- sd(x)
  q <- quantile(x, probs = c(0.25, 0.75), na.rm = TRUE, names = FALSE)
  paste0(
    round(m, digits), " (", round(s, digits), "); ",
    round(q[1], digits), "–", round(q[2], digits)
  )
}

summ_median_iqr <- function(x, digits = 1) {
  x <- x[!is.na(x)]
  med <- median(x)
  q <- quantile(x, probs = c(0.25, 0.75), names = FALSE)
  paste0(round(med, digits), " (", round(q[1], digits), "–", round(q[2], digits), ")")
}

pt_summ <- gic_models |>
  group_by(hiv_status) |>
  summarize(
    total_pts = as.character(n()),

    any_ae = summ_n_pct(ae_any_pt),
    any_ae_early = summ_n_pct(ae_any_early),
    any_severe_ae = summ_n_pct(ae_severe_any_pt),
    any_severe_ae_early = summ_n_pct(ae_severe_any_early),

    max_grade_pt = summ_median_iqr(ae_max_grade_pt),

    .groups = "drop"
  )

pt_wide <- pt_summ |>
  rename(
    "Total patients (N)" = total_pts,
    "Any AE (%, patients)" = any_ae,
    "Any AE in cycles 1–2 (%, patients)" = any_ae_early,
    "Any severe AE (%, patients)" = any_severe_ae,
    "Any severe AE in cycles 1–2 (%, patients)" = any_severe_ae_early,
    "Max AE grade across all cycles, median (IQR)" = max_grade_pt
  ) |>
  pivot_longer(-hiv_status, names_to = "Metric", values_to = "value") |>
  pivot_wider(names_from = hiv_status, values_from = value)

pt_wide |>
  kbl(booktabs = TRUE, align = "lcc") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "center")
```

## Cycle-Level AE Summaries
```{r}
#| label: tbl-ae-by-cycle
#| tbl-cap: "Cycle-level AE summaries by HIV status."
#| results: asis

ae_cycle_long <- gic_models |>
  select(
    record_id, hiv_status, cycles_prescribed, cycles_given,
    matches("^ae_n_markers_measured_\\d+$"),
    matches("^ae_n_events_\\d+$"),
    matches("^ae_n_severe_\\d+$"),
    matches("^ae_max_grade_\\d+$"),
    matches("^ae_any_\\d+$"),
    matches("^ae_severe_any_\\d+$")
  ) |>
  pivot_longer(
    cols = -c(record_id, hiv_status, cycles_prescribed, cycles_given),
    names_to = c("metric", "cycle"),
    names_pattern = "^(.*)_(\\d+)$",
    values_to = "value"
  ) |>
  mutate(cycle = as.integer(cycle)) |>
  pivot_wider(names_from = metric, values_from = value)

ae_cycle_summ <- ae_cycle_long |>
  group_by(hiv_status, cycle) |>
  summarize(
    measured_n = sum(ae_n_markers_measured > 0, na.rm = TRUE),
    total_pts = n(),

    `Patients (%)` = paste0(measured_n, " (", round(100 * measured_n / total_pts, 1), "%)"),

    `Any AE (%)` = {
      n <- sum(ae_any == TRUE, na.rm = TRUE)
      paste0(n, " (", round(100 * n / measured_n, 1), "%)")
    },
    `Any severe AE (%)` = {
      n <- sum(ae_severe_any == TRUE, na.rm = TRUE)
      paste0(n, " (", round(100 * n / measured_n, 1), "%)")
    },

    `# AEs, median (IQR)` = summ_median_iqr(ae_n_events),
    `# severe AEs, median (IQR)` = summ_median_iqr(ae_n_severe),
    `Max grade, median (IQR)` = summ_median_iqr(ae_max_grade, digits = 0),

    .groups = "drop"
  ) |>
  rename(Cycle = cycle) |>
  select(-measured_n, -total_pts)

# One table per HIV status
for (status in split(ae_cycle_summ, ae_cycle_summ$hiv_status)) {
  cat("\n\n### HIV status:", status$hiv_status[1], "\n\n")
  status |>
    select(-hiv_status) |>
    kbl(booktabs = TRUE, align = "lccccccc") |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) |>
    print()
}
```

```{r}
#| label: plt-ae-max-grade-pt
#| fig-cap: "Distribution of maximum AE grade by HIV status."

gic_models |>
  filter(!is.na(ae_max_grade_pt)) |>
  count(hiv_status, ae_max_grade_pt) |>
  group_by(hiv_status) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = ae_max_grade_pt, y = prop, fill = hiv_status)) +
  geom_col(position = "dodge") +
  scale_x_continuous(breaks = 0:5) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Maximum AE Grade per Patient by HIV Status",
    x = "Maximum CTCAE grade",
    y = "Proportion of patients",
    fill = "HIV status"
  ) +
  theme_minimal()
```


```{r}
#| label: prep-ae-cycle-plot-data

# Build cycle-level dataset from gic_models for plotting
ae_cycle_measured <- ae_cycle_long |> 
  filter(ae_n_markers_measured > 0) |> 
  mutate(
    pct_complete = if_else(is.na(cycles_prescribed) | cycles_prescribed <= 0,
                           NA_real_,
                           cycle / cycles_prescribed),
    pct_bin = cut(
      pct_complete,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE,
      right = TRUE
    )
  )

summarize_metric <- function(df, var) {
  df |> 
    group_by(hiv_status, cycle) |> 
    summarize(
      n = sum(!is.na(.data[[var]])),
      median = median(.data[[var]], na.rm = TRUE),
      q25 = quantile(.data[[var]], 0.25, na.rm = TRUE),
      q75 = quantile(.data[[var]], 0.75, na.rm = TRUE),
      min = min(.data[[var]], na.rm = TRUE),
      max = max(.data[[var]], na.rm = TRUE),
      .groups = "drop"
    )
}

ae_events_summ <- summarize_metric(ae_cycle_measured, "ae_n_events")
ae_severe_summ <- summarize_metric(ae_cycle_measured, "ae_n_severe")
ae_maxgrade_summ <- summarize_metric(ae_cycle_measured, "ae_max_grade")
```


```{r}
#| label: plt-ae-n-events
#| fig-cap: "Number of adverse events per treatment cycle by HIV status (median with IQR and min–max bands; measured cycles only)"

ggplot(ae_events_summ, aes(x = cycle, group = hiv_status, color = hiv_status, fill = hiv_status)) +
  geom_ribbon(aes(ymin = min, ymax = max), alpha = 0.2, color = NA) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.5, color = NA) +
  geom_line(aes(y = median), linewidth = 1) +
  geom_point(aes(y = median), size = 2) +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Adverse events per cycle",
       subtitle = "Median (line) with IQR and min–max bands among cycles with any AE grading recorded",
       x = "Treatment cycle",
       y = "Number of AEs",
       color = "HIV status", fill = "HIV status") +
  theme_minimal()
```
Initially higher trend for number of adverse events amongst HIV positive patients, but the mean drops below that of HIV negative patients after cycle 4. Need to investigate, but could have to do with high early dropout rates, or possibly related to patients with HIV generally having fewer cycles of chemotherapy prescribed.


```{r}
#| label: plt-ae-n-severe
#| fig-cap: "Number of severe adverse events (grade ≥ 3) per treatment cycle by HIV status (median with IQR band; measured cycles only"

ggplot(ae_severe_summ, aes(x = cycle, group = hiv_status, color = hiv_status, fill = hiv_status)) +
  geom_ribbon(aes(ymin = min, ymax = max), alpha = 0.2, color = NA) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.5, color = NA) +
  geom_line(aes(y = median), linewidth = 1) +
  geom_point(aes(y = median), size = 2) +
  scale_x_continuous(breaks = 1:12) +
  labs(
    title = "Severe adverse events per cycle",
    subtitle = "Grade ≥ 3; median (line) with IQR and min–max bands among measured cycles",
    x = "Treatment cycle",
    y = "Number of severe AEs",
    color = "HIV status",
    fill = "HIV status"
  ) +
  theme_minimal()
```
Not a very noticable difference based on the medians or IQRs. The HIV negative patients tend to have higher max number of severe adverse events per cycle, but as noted before, I'd guess that relates to how few cycles HIV positive patients tend to complete, and the total number of HIV negative patients compared to positive.


```{r}
#| label: plt-ae-max-grade-cycle
#| fig-cap: "Median max AE grade per cycle by HIV status among measured cycles."

ggplot(ae_maxgrade_summ, aes(x = cycle, group = hiv_status, color = hiv_status, fill = hiv_status)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.3, color = NA) +
  geom_line(aes(y = median), linewidth = 1) +
  geom_point(aes(y = median), size = 2) +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(breaks = 0:5, limits = c(0, 5)) +
  labs(
    title = "Max AE Grade per Cycle",
    subtitle = "Median (line) with IQR band among measured cycles",
    x = "Treatment cycle",
    y = "Max AE grade",
    color = "HIV status",
    fill = "HIV status"
  ) +
  theme_minimal()
```
No easily discernible visual difference between patients with respect to HIV status and maximum adverse event experienced aside from period 7 onward as previously noted.

```{r}
#| label: plt-ae-severe-prop-cycle
#| fig-cap: "Proportion of measured patients experiencing any grade 3+ AE per cycle by HIV status."

severe_prop <- ae_cycle_measured |>
  group_by(hiv_status, cycle) |>
  summarize(
    n = n(),
    prop_severe = mean(ae_severe_any, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(severe_prop, aes(x = cycle, y = prop_severe, color = hiv_status)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
  labs(
    title = "Proportion with Severe AE (Grade 3+) per Cycle",
    x = "Treatment cycle",
    y = "Proportion with any grade 3+ AE",
    color = "HIV status"
  ) +
  theme_minimal()
```
Maybe a slightly higher proportion of HIV positive patients have severe AEs than HIV negative, need to do a test to compare over cycles 1-6. As noted before, in cycles 7 and onwards, there is only a single HIV positive patient still receiving treatment.


# Descriptive Statistics
```{r}
#| label: tbl-stage-vs-hiv-status
#| tbl-cap: "Cancer stage vs HIV status"

table(Stage = gic_models$cancer_stage, HIV = gic_models$hiv_status) |>
  as.data.frame.matrix() |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left") |>
  add_header_above(c("Cancer Stage" = 1, "HIV Status" = 2))
```

## Patient & Treatment Characteristics
```{r}
#| label: tbl-chemo-type-by-hiv
#| tbl-cap: "Chemotherapy type by HIV status: patient counts, within-group proportions, and cycles prescribed"

chemo_hiv_summ <- gic_models |>
  group_by(hiv_status, chemo_type) |>
  summarize(
    n = n(),
    cycles_med_iqr = summ_median_iqr(cycles_prescribed),
    .groups = "drop"
  ) |>
  group_by(hiv_status) |>
  mutate(pct = round(100 * n / sum(n), 1),
         `n (%)` = paste0(n, " (", pct, "%)")) |>
  ungroup() |>
  select(hiv_status, chemo_type, `n (%)`, cycles_med_iqr)

neg <- chemo_hiv_summ |> filter(hiv_status == "Negative") |>
  select(chemo_type, `Neg: n (%)` = `n (%)`, `Neg: med (IQR)` = cycles_med_iqr)
pos <- chemo_hiv_summ |> filter(hiv_status == "Positive") |>
  select(chemo_type, `Pos: n (%)` = `n (%)`, `Pos: med (IQR)` = cycles_med_iqr)

chemo_levels <- levels(gic_models$chemo_type)

# Chi-squared test per chemo type for assignment proportions
chisq_pvals <- map_dbl(chemo_levels, function(ct) {
  tab <- table(
    assigned = gic_models$chemo_type == ct,
    hiv      = gic_models$hiv_status
  )
  expected <- chisq.test(tab)$expected
  if (any(expected < 5)) {
    fisher.test(tab)$p.value
  } else {
    chisq.test(tab)$p.value
  }
})

# Wilcoxon rank-sum test per chemo type for cycles prescribed
wilcox_pvals <- map_dbl(chemo_levels, function(ct) {
  dat <- gic_models |> filter(chemo_type == ct)
  if (n_distinct(dat$hiv_status) < 2) return(NA_real_)
  wilcox.test(cycles_prescribed ~ hiv_status, data = dat, exact = FALSE)$p.value
})

format_p <- function(p) {
  if (is.na(p)) return("—")
  if (p < 0.001) return(formatC(p, format = "e", digits = 3))
  as.character(round(p, 3))
}

pval_df <- tibble(
  chemo_type    = chemo_levels,
  `p (asgmt)`   = map_chr(chisq_pvals, format_p),
  `p (cycles)`  = map_chr(wilcox_pvals, format_p)
)

neg |>
  left_join(pos,     by = "chemo_type") |>
  left_join(pval_df, by = "chemo_type") |>
  rename(`Chemotherapy type` = chemo_type) |>
  select(
    `Chemotherapy type`,
    `Neg: n (%)`, `Pos: n (%)`, `p (asgmt)`,
    `Neg: med (IQR)`, `Pos: med (IQR)`, `p (cycles)`
  ) |>
  gt() |>
  tab_spanner(
    label = "Chemo assignment",
    columns = c(`Neg: n (%)`, `Pos: n (%)`, `p (asgmt)`)
  ) |>
  tab_spanner(
    label = "Cycles prescribed",
    columns = c(`Neg: med (IQR)`, `Pos: med (IQR)`, `p (cycles)`)
  ) |>
  cols_align(align = "center", columns = -`Chemotherapy type`) |>
  cols_align(align = "left",   columns = `Chemotherapy type`) |>
  tab_footnote(
    footnote = "Chi-squared test (Fisher's exact where any expected cell count < 5); Wilcoxon rank-sum test.",
    location = cells_column_labels(columns = c(`p (asgmt)`, `p (cycles)`))
  ) |>
  opt_footnote_marks(marks = "extended")
```

## Cancer Site by HIV Status

```{r}
#| label: tbl-cancer-site-hiv
#| tbl-cap: "Cancer site by HIV status"

cancer_levels <- levels(gic_models$cancer_site)
site_hiv_table <- table(gic_models$cancer_site, gic_models$hiv_status)

hiv_summ <- gic_models |>
  group_by(cancer_site, hiv_status) |>
  summarize(n = n(), .groups = "drop") |>
  group_by(cancer_site) |>
  mutate(
    pct = round(100 * n / sum(n), 1),
    `n (%)` = paste0(n, " (", pct, "%)")
  ) |>
  ungroup() |>
  select(cancer_site, hiv_status, `n (%)`) |>
  pivot_wider(names_from = hiv_status, values_from = `n (%)`,
              names_prefix = "HIV ") |>
  rename(`Cancer site` = cancer_site)

hiv_summ |>
  kbl(booktabs = TRUE, align = "lcc") |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

```{r}
#| label: tbl-cancer-site-hiv-expected
#| tbl-cap: "Expected cell counts under the null hypothesis of no association"

chisq.test(site_hiv_table)$expected |>
  round(1) |>
  as.data.frame.matrix() |>
  kbl(booktabs = TRUE, align = "lcc") |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) |>
  footnote(general = "Four cells have expected counts below 5.")
```

```{r}
#| label: tst-cancer-site-hiv-global

# Global test: is there any association between cancer site and HIV status?
fisher_global <- fisher.test(site_hiv_table)
fisher_global
```

The global Fisher's exact test indicates that HIV status is not independent of cancer site. To identify which sites are potentially causing this, each site is compared against all remaining sites using contingency tables.


```{r}
#| label: tbl-cancer-site-hiv-expected-ovr
#| tbl-cap: "Expected cell counts for one-vs-rest 2×2 tables by cancer site"

map_dfr(cancer_levels, function(cs) {
  tab <- table(
    site = gic_models$cancer_site == cs,
    hiv  = gic_models$hiv_status
  )
  expected <- suppressWarnings(chisq.test(tab)$expected)
  
  tibble(
    `Cancer site` = cs,
    `Site: HIV−` = round(expected["TRUE", "Negative"], 1),
    `Site: HIV+` = round(expected["TRUE", "Positive"], 1),
    `Rest: HIV−` = round(expected["FALSE", "Negative"], 1),
    `Rest: HIV+` = round(expected["FALSE", "Positive"], 1)
  )
}) |>
  kbl(booktabs = TRUE, align = "lcccc") |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) |>
  footnote(
    general = "Expected counts under H0 for each site vs all other sites."
  )
```


```{r}
#| label: tbl-cancer-site-hiv-posthoc
#| tbl-cap: "Post-hoc: HIV prevalence by cancer site vs all other sites"

site_posthoc <- map_dfr(cancer_levels, function(cs) {
  tab <- table(
    site = gic_models$cancer_site == cs,
    hiv  = gic_models$hiv_status
  )
  expected <- suppressWarnings(chisq.test(tab)$expected)
  use_fisher <- any(expected < 5)
  
  if (cs == "Anal") {
    p <- fisher.test(tab, alternative = "greater")$p.value
    test_used <- "Fisher's (one-sided)"
  } else if (use_fisher) {
    p <- fisher.test(tab)$p.value
    test_used <- "Fisher's Exact"
  } else {
    p <- chisq.test(tab, correct = FALSE)$p.value
    test_used <- "Chi-squared"
  }
  
  tibble(`Cancer site` = cs, p = format_p(p), Test = test_used)
})

hiv_summ |>
  left_join(site_posthoc, by = "Cancer site") |>
  kbl(booktabs = TRUE, align = "lcccc") |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) |>
  footnote(
    general = paste0(
      "Each site tested against all other sites combined. ",
      "Chi-squared used where all expected counts ≥ 5, Fisher's exact otherwise. ",
      "Anal cancer tested one-sided (greater) based on established HIV-anal cancer association."
    )
  )
```

Post-hoc tests compare HIV prevalence at each cancer site against all other cancer sites combined using contingency tables. Anal cancer is tested one-sided based on the association between anal cancer rate and HIV status.  All other comparisons are two-sided. The choice of test for each site is determined based on expected cell counts.

## Chemotherapy Assignment by Cancer Site
```{r}
#| label: tbl-cancer-site-chemo
#| tbl-cap: "Chemotherapy assignment by cancer site"

chemo_levels <- levels(gic_models$chemo_type)

gic_models |>
  group_by(cancer_site, chemo_type) |>
  summarize(n = n(), .groups = "drop") |>
  group_by(cancer_site) |>
  mutate(
    pct = round(100 * n / sum(n), 1),
    `n (%)` = paste0(n, " (", pct, "%)")
  ) |>
  ungroup() |>
  select(cancer_site, chemo_type, `n (%)`) |>
  pivot_wider(names_from = chemo_type, values_from = `n (%)`,
              values_fill = "0 (0%)") |>
  rename(`Cancer site` = cancer_site) |>
  kbl(booktabs = TRUE, align = c("l", rep("c", length(chemo_levels)))) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```


```{r}
#| label: tbl-tx-intent-vs-hiv-status
#| tbl-cap: "Treatment intent vs HIV status"

table(Intent = gic_models$tx_intent, HIV = gic_models$hiv_status) |>
  as.data.frame.matrix() |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left") |>
  add_header_above(c("Treatment Intent" = 1, "HIV Status" = 2))
```

```{r}
#| label: tbl-cycles-by-hiv
#| tbl-cap: "Chemotherapy cycles prescribed and received by HIV status"

cycles_summ <- gic_models |>
  mutate(
    completion_rate = if_else(
      is.na(cycles_prescribed) | cycles_prescribed <= 0,
      NA_real_,
      cycles_given / cycles_prescribed
    )
  ) |>
  group_by(hiv_status) |>
  summarize(
    n_patients = n(),
    cycles_prescribed = summ_mean_sd_iqr(cycles_prescribed, digits = 1),
    cycles_given = summ_mean_sd_iqr(cycles_given, digits = 1),
    completion_rate = summ_mean_sd_iqr(completion_rate, digits = 2),
    .groups = "drop"
  ) |>
  mutate(
    `HIV Status` = hiv_status,
    `Patients (N)` = n_patients,
    `Cycles prescribed, mean (SD); IQR` = cycles_prescribed,
    `Cycles given, mean (SD); IQR` = cycles_given,
    `Completion rate, mean (SD); IQR` = completion_rate
  ) |>
  select(
    `HIV Status`,
    `Patients (N)`,
    `Cycles prescribed, mean (SD); IQR`,
    `Cycles given, mean (SD); IQR`,
    `Completion rate, mean (SD); IQR`
  )

cycles_summ |>
  kbl(booktabs = TRUE, align = "lcccc") |>
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
#| label: tbl-severe-early-contingency
#| tbl-cap: "Contingency table of severe early adverse events (cycles 1-2) vs HIV status"

table_severe_early <- table(severe_early_ae = factor(gic_models$ae_severe_any_early, levels = c(0, 1), labels = c("No", "Yes")), HIV = gic_models$hiv_status)

table_severe_early |>
  as.data.frame.matrix() |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left") |>
  add_header_above(c("Severe Early AE" = 1, "HIV Status" = 2))
```

## Treatment Completion

```{r}
#| label: plt-density-cycles-rx

gic_validated |>
  count(hiv_status, cycles_prescribed) |>
  group_by(hiv_status) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = cycles_prescribed, y = prop, fill = hiv_status)) +
  geom_col(position = "dodge") +
  scale_x_continuous(name = "Cycles prescribed", breaks = 1:12) +
  ylab("Proportion") + labs(title = "Chemo Cycles Prescribed by HIV Status")
```

```{r}
#| label: plt-cycle-completion

survival_data <- gic_validated |>
  # for each patient, create a row for each cycle they were prescribed
  rowwise() |>
  mutate(cycle = list(1:cycles_prescribed)) |>
  unnest(cycle) |>
  # mark whether they actually received each cycle
  mutate(received = cycle <= cycles_given) |>
  # for each HIV status and cycle number, get the proportion who received it
  group_by(hiv_status, cycle) |>
  summarise(
    n_prescribed = n(),
    prop_received = mean(received),
    .groups = "drop"
  )

# main plot (store it)
p_main <- ggplot(survival_data, aes(x = cycle, y = prop_received, color = hiv_status)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "Chemotherapy Cycle Completion by HIV Status",
    x = NULL,
    y = "Proportion receiving cycle\n(among those prescribed at least that many)",
    color = "HIV status"
  ) +
  theme(plot.margin = margin(5, 5, 0, 5))

# table underneath
p_table <- ggplot(survival_data, aes(x = cycle, y = hiv_status, label = n_prescribed)) +
  geom_text(size = 3) +
  scale_x_continuous(breaks = 1:12, limits = range(survival_data$cycle)) +
  labs(x = NULL, y = NULL, subtitle = "Patients prescribed at least k cycles") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_text(face = "bold"),
    axis.text.x = element_blank(),
    plot.margin = margin(0, 5, 5, 5)
  )

# combine
p_main / p_table + plot_layout(heights = c(4, 1))
```



# Model Building
## Severe Adverse Events as Outcome
### Unadjusted model
```{r}
#| label: model-any-simple

model_any_simple <- glm(ae_any_pt ~ hiv_status,
                        data = gic_models, family = binomial)

# summary(model_any_simple)
```
Invalid model due to having a 100% adverse event rate in the HIV-positive group.

### Adjusted model
```{r}
#| label: model-severe-simple


model_severe_simple <- glm(ae_severe_any_pt ~ hiv_status,
                           data = gic_models, family = binomial)

summary(model_severe_simple)
```

```{r}
#| label: tbl-severe-simple-or-clean
#| tbl-cap: "Association between HIV Status and Severe Adverse Events (Grades 3-4)"

tidy(model_severe_simple, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  # Clean up the name and prepare for the reference row
  mutate(
    Term = str_replace(term, "hiv_statusPositive", "HIV Positive"),
    `OR (95% CI)` = paste0(round(estimate, 2), " (",
                           round(conf.low, 2), " - ",
                           round(conf.high, 2), ")"),
    p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))),
    std.error = round(std.error, 3)
  ) |>
  select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |>
  # Add the reference row at the top
  add_row(Term = "HIV Negative (Reference)",
          `OR (95% CI)` = "1.00",
          `Std. Error` = NA,
          `P-value` = NA,
          .before = 1) |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left")
```



### Multivariate Logistic: Full treatment period
#### Basic logistic model of any adverse events by HIV status
```{r}
#| label: model-any-basic

model_any_basic <- glm(ae_any_pt ~ hiv_status + cancer_stage + chemo_type + tx_intent,
                       data = gic_models, family = binomial)

summary(model_any_basic)
```
Invalid model due to having a 100% adverse event rate in the HIV-positive group.

#### Basic logicstic model of any severe adverse events by HIV status
```{r}
#| label: model-severe-basic

model_severe_basic <- glm(ae_severe_any_pt ~ hiv_status + cancer_stage + chemo_type + tx_intent,
                       data = gic_models, family = binomial)

summary(model_severe_basic)
```

```{r}
#| label: tbl-severe-multivariate-or
#| tbl-cap: "Adjusted Odds Ratios for Severe Adverse Events (Grades 3-4)"

model_severe_basic |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Regimen",
      tx_intent    ~ "Treatment Intent"
    ),
    pvalue_fun = label_style_pvalue(
      digits = 3
    ),
    add_estimate_to_reference_rows = TRUE
  ) |>
  bold_labels() |> 
  italicize_levels() |> 
  bold_p() |> 
  add_global_p(keep = TRUE) |> 
  add_glance_source_note(include = c(nobs, AIC))
```


### $\chi^2$ Tests (HIV vs Early Events)

```{r}
#| label: tbl-any-early-contingency
#| tbl-cap: "Contingency table of early adverse events (cycles 1-2) vs HIV status"

table_any_early <- table(early_ae = factor(gic_models$ae_any_early, levels = c(0, 1), labels = c("No", "Yes")), HIV = gic_models$hiv_status)

table_any_early |>
  as.data.frame.matrix() |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left") |>
  add_header_above(c("Any Early AE" = 1, "HIV Status" = 2))
```


<!-- ```{r} -->
<!-- #| label: tests-chi-sq -->

<!-- # Any Early AE (Cycles 1-2) -->
<!-- cat("Chi-Squared Test: Any Early AE (C1-2) vs HIV Status\n") -->
<!-- print(chisq.test(table_any_early)) -->

<!-- cat("\n\n") -->

<!-- # Severe Early AE (Cycles 1-2) -->
<!-- cat("Chi-Squared Test: Severe Early AE (C1-2) vs HIV Status\n") -->
<!-- print(chisq.test(table_severe_early)) -->
<!-- ``` -->

## HIV Status as Outcome
### Unadjusted model
```{r}
#|label: model-hiv-simple

model_hiv_simple <- glm(hiv_status ~ ae_max_grade_pt, data = gic_models, family = binomial)

summary(model_hiv_simple)

model_hiv_simple |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      ae_max_grade_pt ~ "Max CTCAE grade"
    ),
    pvalue_fun = label_style_pvalue(
      digits = 3
    ),
    add_estimate_to_reference_rows = TRUE
  ) |>
  bold_labels()
```


<!-- ```{r} -->
<!-- #| label: tbl-hiv-simple-or -->
<!-- #| tbl-cap: "Logistic regression: HIV status vs mean max AE grade (unadjusted)" -->

<!-- tidy(model_hiv_simple, conf.int = TRUE, exponentiate = TRUE) |> -->
<!--   filter(term != "(Intercept)") |> -->
<!--   mutate( -->
<!--     Term = case_when( -->
<!--       term == "ae_mean_max_grade" ~ "Mean max AE grade per measured cycle", -->
<!--       TRUE ~ term -->
<!--     ), -->
<!--     `OR (95% CI)` = paste0( -->
<!--       round(estimate, 2), " (", -->
<!--       round(conf.low, 2), " - ", -->
<!--       round(conf.high, 2), ")" -->
<!--     ), -->
<!--     p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))), -->
<!--     std.error = round(std.error, 3) -->
<!--   ) |> -->
<!--   select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") -->
<!-- ``` -->




### Adjusted model
```{r}
#| label: model-hiv-basic

model_hiv_basic <- glm(
  hiv_status ~ ae_max_grade_pt + cancer_stage + chemo_type + tx_intent,
  data = gic_models,
  family = binomial
)

summary(model_hiv_basic)
```

<!-- ```{r} -->
<!-- #| label: tbl-hiv-basic-or -->
<!-- #| tbl-cap: "Logistic regression: HIV status vs mean max AE grade (adjusted for stage, chemo type, and intent)" -->

<!-- tidy(model_hiv_basic, conf.int = TRUE, exponentiate = TRUE) |> -->
<!--   filter(term != "(Intercept)") |> -->
<!--   mutate( -->
<!--     Term = case_when( -->
<!--       term == "ae_mean_max_grade" ~ "Mean max AE grade per measured cycle", -->
<!--       TRUE ~ term -->
<!--     ), -->
<!--     `OR (95% CI)` = paste0( -->
<!--       round(estimate, 2), " (", -->
<!--       round(conf.low, 2), " - ", -->
<!--       round(conf.high, 2), ")" -->
<!--     ), -->
<!--     p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))), -->
<!--     std.error = round(std.error, 3) -->
<!--   ) |> -->
<!--   select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") -->
<!-- ``` -->

### Model comparison
```{r}
#| label: tbl-hiv-or-comparison
#| tbl-cap: "Logistic regression: HIV status — unadjusted vs adjusted odds ratios"

tbl_simple <- model_hiv_simple |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(ae_max_grade_pt ~ "Max AE grade (patient)")
  )

tbl_basic <- model_hiv_basic |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      ae_max_grade_pt ~ "Max AE grade (patient)",
      cancer_stage ~ "Cancer stage",
      chemo_type ~ "Chemo type",
      tx_intent ~ "Treatment intent"
    )
  )

tbl_merge(
  list(tbl_simple, tbl_basic),
  tab_spanner = c("Unadjusted", "Adjusted")
)
```

**Summary:** Both the unadjusted and adjusted models with HIV status as outcome indicate that each one unit increase in max AE grade across all treatment cycles is associated with between 28-34% higher odds of the patient being HIV positive. In the adjusted model, higher cancer stages were not reliably associated with being HIV positive. For chemo type there does seem to be a pretty clear difference in treatment options associated with HIV positive status. There does not appear to be any association between treatment intent and HIV status.

### Assumption checks

```{r}
#| label: chk-hiv-assumptions

# Linearity of the log-odds
gic_models |>
  filter(!is.na(ae_max_grade_pt)) |>
  group_by(ae_max_grade_pt) |>
  summarize(prop_hiv = mean(hiv_status == "Positive", na.rm = TRUE)) |>
  mutate(log_odds = log(prop_hiv / (1 - prop_hiv))) |>
  ggplot(aes(x = ae_max_grade_pt, y = log_odds)) +
  geom_point() +
  geom_line() +
  labs(x = "Max AE grade (patient)", y = "Empirical log-odds of HIV+")

print("Linearity isn't amazing, but a straight line fit would show a reasonable fit.")

# Multicollinearity
vif(model_hiv_basic)

print("No significant multicollinearity noted.")

# Influential observations
plot(model_hiv_basic, which = 4)  # Cook's distance
print("Three influential observations, need to investigate further")
```

```{r}
#| label: investigate-influential-obs

# Get row indices of most influential observations (Cook's D > threshold)
# Using 4/n as a common rule-of-thumb cutoff
n <- nrow(model_hiv_basic$model)
cooks_threshold <- 4 / n

influential_idx <- which(cooks.distance(model_hiv_basic) > cooks_threshold)
influential_ids <- gic_models$record_id[influential_idx]

cat("Influential observations (Cook's D > 4/n =", round(cooks_threshold, 4), "):\n")
cat("Row indices:", influential_idx, "\n")
cat("Record IDs:", influential_ids, "\n\n")

# Pull their covariate values to see what makes them unusual
gic_models |>
  filter(record_id %in% influential_ids) |>
  select(record_id, hiv_status, cancer_stage, chemo_type, tx_intent, ae_max_grade_pt)
```

All pts with a large Cook's D are HIV positive, but I haven't immediately spotted any other commonality amongst them to explain the group collectively being so influential, will have to do more investigation.



## Cycles Prescribed as Outcome
```{r}
#| label: model-cycles-full

# Full cycles prescribed model
model_cycles_full <- glm(
  cycles_prescribed ~ hiv_status + cancer_site + cancer_stage + chemo_type + tx_intent,
  data   = gic_models,
  family = poisson
)
```
```{r}
#| label: model-cycles-no-site

# Cycles prescribed model without cancer site
model_cycles_no_site <- glm(
  cycles_prescribed ~ hiv_status + cancer_stage + chemo_type + tx_intent,
  data   = gic_models,
  family = poisson
)
```
```{r}
#| label: tbl-cycles-models
#| tbl-cap: "Poisson regression: cycles prescribed (exponentiated coefficients = rate ratios)"

tbl_full <- model_cycles_full |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      cancer_site  ~ "Cancer Site",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Type",
      tx_intent    ~ "Treatment Intent"
    )
  ) |>
  bold_labels() |> 
  italicize_levels() |> 
  bold_p() |> 
  add_global_p(keep = TRUE) |> 
  add_vif(statistic = "aGVIF")

tbl_no_site <- model_cycles_no_site |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Type",
      tx_intent    ~ "Treatment Intent"
    )
  ) |>
  bold_labels() |> 
  italicize_levels() |> 
  bold_p() |> 
  add_global_p(keep = TRUE) |> 
  add_vif(statistic = "aGVIF")

lrt_p <- anova(model_cycles_no_site, model_cycles_full, test = "LRT")$`Pr(>Chi)`[2]


tbl_merge(
  list(tbl_no_site, tbl_full),
  tab_spanner = c("Without cancer site", "With cancer site")
) |>
  as_gt() |>
  tab_source_note(
    source_note = paste0(
      "Model without cancer site: AIC = ", round(AIC(model_cycles_no_site), 1),
      ", N = ", nobs(model_cycles_no_site),
      " | Model with cancer site: AIC = ", round(AIC(model_cycles_full), 1),
      ", N = ", nobs(model_cycles_full)
    )
  ) |>
  tab_source_note(
    source_note = paste0(
      "LRT (with vs without cancer site): p = ",
      round(anova(model_cycles_no_site, model_cycles_full, 
                  test = "LRT")$`Pr(>Chi)`[2], 3)
    )
  )
  
```


# Toxicity-Related Hospitalizations
```{r}
#| label: prep-hosp-days

hosp_cols <- names(gic_models)[str_detect(names(gic_models), "^hosp_days_([1-9]|1[0-2])$")]

hosp_long <- gic_models |> 
  select(record_id, hiv_status, cycles_given, all_of(hosp_cols)) |> 
  pivot_longer(
    cols = all_of(hosp_cols),
    names_to = "cycle_var",
    values_to = "hosp_days"
  ) |> 
  mutate(
    cycle = as.integer(str_extract(cycle_var, "\\d+"))
  )

# Patient-level summaries
hosp_pt <- hosp_long |> 
  group_by(record_id, hiv_status, cycles_given) |> 
  summarize(
    hosp_days_total = sum(hosp_days, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  mutate(
    hosp_days_per_cycle_given = if_else(is.na(cycles_given) | cycles_given <= 0,
                                        NA_real_,
                                        hosp_days_total / cycles_given)
  )

```

## Summary by HIV Status
```{r}
#| label: tbl-hosp-by-pt
#| tbl-cap: "Hospital days attributable to toxicity by HIV status"

hosp_pt |>
  group_by(hiv_status) |>
  summarize(
    n_patients = n(),

    hosp_days_total_mean_sd = summ_median_iqr(hosp_days_total, 1),
    hosp_days_per_cycle_given_mean_sd = summ_median_iqr(hosp_days_per_cycle_given, 1),
    .groups = "drop"
  ) |>
  mutate(
    `HIV Status` = hiv_status,
    `Patients (N)` = n_patients,
    `Total hosp days, mean (SD)` = hosp_days_total_mean_sd,
    `Hosp days / cycles_given, mean (SD)` = hosp_days_per_cycle_given_mean_sd
  ) |>
  select(
    `HIV Status`, `Patients (N)`,
    `Total hosp days, mean (SD)`,
    `Hosp days / cycles_given, mean (SD)`
  ) |>
  kbl(booktabs = TRUE, align = "lcccc") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
#| label: prep-hosp-cycle-data

# Identify cycle-level column groups needed for hospitalization analysis
severe_cols <- names(gic_models)[str_detect(names(gic_models), "^ae_n_severe_([1-9]|1[0-2])$")]
max_grade_cols <- names(gic_models)[str_detect(names(gic_models), "^ae_max_grade_([1-9]|1[0-2])$")]
measured_cols <- names(gic_models)[str_detect(names(gic_models), "^ae_n_markers_measured_([1-9]|1[0-2])$")]

# Pivot each set of cycle cols to long
pivot_cycle_cols <- function(df, cols, value_name) {
  df |>
    select(record_id, all_of(cols)) |>
    pivot_longer(cols = all_of(cols), names_to = "cycle_var", values_to = value_name) |>
    mutate(cycle = as.integer(str_extract(cycle_var, "\\d+$"))) |>
    select(-cycle_var)
}

severe_long <- pivot_cycle_cols(gic_models, severe_cols,    "ae_n_severe")
max_grade_long <- pivot_cycle_cols(gic_models, max_grade_cols, "ae_max_grade")
measured_long <- pivot_cycle_cols(gic_models, measured_cols,  "ae_n_markers_measured")

# Join AE variables to hospitalization data
hosp_severe_cycle <- hosp_long |>
  left_join(severe_long,   by = c("record_id", "cycle")) |>
  left_join(measured_long, by = c("record_id", "cycle")) |>
  mutate(ae_measured = ae_n_markers_measured > 0)

hosp_maxgrade_cycle <- hosp_long |>
  left_join(max_grade_long, by = c("record_id", "cycle")) |>
  left_join(measured_long,  by = c("record_id", "cycle")) |>
  mutate(ae_measured = ae_n_markers_measured > 0)
```

## Hospitalizations by AE Severity
```{r}
#| label: tbl-hosp-vs-severe

hosp_severe_cycle |> 
  filter(ae_measured) |> 
  group_by(hiv_status, ae_n_severe) |> 
  summarize(
    n_cycles = sum(!is.na(hosp_days)),
    hosp_days_mean_sd = summ_median_iqr(hosp_days),
    .groups = "drop"
  ) |> 
  mutate(
    `HIV Status` = hiv_status,
    `# severe AEs in cycle` = ae_n_severe,
    `# Cycles` = n_cycles,
    `Hosp days, mean (SD); range` = hosp_days_mean_sd
  ) |> 
  select(`HIV Status`, `# severe AEs in cycle`, `# Cycles`, `Hosp days, mean (SD); range`) |> 
  kbl(booktabs = TRUE, align = "lccc") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
#| label: plt-hosp-vs-severe
#| fig-cap: "Mean hospital days vs number of severe AEs in the cycle, by HIV status"

hosp_severe_summ <- hosp_severe_cycle |>
  filter(ae_measured) |>
  group_by(hiv_status, ae_n_severe) |>
  summarize(
    mean_hosp_days = mean(hosp_days, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(hosp_severe_summ, aes(x = ae_n_severe, y = mean_hosp_days, color = hiv_status, group = hiv_status)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    x = "Number of severe AEs in cycle",
    y = "Mean hospital days (toxicity) in cycle",
    color = "HIV status"
  ) +
  theme_minimal()
```

```{r}
#| label: tbl-hosp-vs-max-grade
#| tbl-cap: "Hospital days vs max AE grade in cycle, by HIV status (among cycles with any AE grading recorded)"

hosp_maxgrade_cycle |>
  filter(ae_measured) |>
  group_by(hiv_status, ae_max_grade) |>
  summarize(
    n_cycles = sum(!is.na(hosp_days)),
    hosp_days_median_iqr = summ_median_iqr(hosp_days, 1),
    .groups = "drop"
  ) |>
  mutate(
    `HIV Status` = hiv_status,
    `Max AE grade in cycle` = ae_max_grade,
    `# Cycles` = n_cycles,
    `Hosp days, median (IQR)` = hosp_days_median_iqr
  ) |>
  select(`HIV Status`, `Max AE grade in cycle`, `# Cycles`, `Hosp days, median (IQR)`) |>
  kbl(booktabs = TRUE, align = "lccc") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
#| label: plt-hosp-vs-max-grade
#| fig-cap: "Mean hospital days vs max AE grade in cycle, by HIV status (among measured cycles)"

hosp_maxgrade_summ <- hosp_maxgrade_cycle |>
  filter(ae_measured) |>
  group_by(hiv_status, ae_max_grade) |>
  summarize(
    mean_hosp_days = mean(hosp_days, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(hosp_maxgrade_summ, aes(x = ae_max_grade, y = mean_hosp_days, color = hiv_status, group = hiv_status)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 0:5) +
  labs(
    x = "Max AE grade in cycle",
    y = "Mean hospital days (toxicity) in cycle",
    color = "HIV status"
  ) +
  theme_minimal()
```

## Regression Models


```{r}
#| label: prep-hosp-regression


hosp_ae_cycle <- hosp_long |>
  left_join(
    ae_cycle_long |> select(record_id, cycle, ae_n_severe, ae_max_grade, ae_n_markers_measured),
    by = c("record_id", "cycle")
  ) |>
  mutate(ae_measured = ae_n_markers_measured > 0)


hosp_model_data <- hosp_ae_cycle |>
  filter(ae_measured)

hosp_model_data <- hosp_model_data |>
  mutate(any_hosp = as.integer(hosp_days > 0))
```

```{r}
#| label: model-hosp-any

# Any hospitalization this cycle (logistic)
model_any_hosp <- glmer(
  any_hosp ~ ae_max_grade + hiv_status + ae_max_grade:hiv_status + (1 | record_id),
  data = hosp_model_data,
  family = binomial
)
summary(model_any_hosp)
fixef(model_any_hosp) |> exp()
confint(model_any_hosp, method = "Wald") |> exp()
```
Each increase in max AE grade in a given cycle is associated with about a 5.5-fold increase (95% CI: 3.7-8.2) in the odds of being hospitalized. For patients with no adverse events, being HIV positive is associated with about a 3.4-fold increase in odds of being hospitalized, but the effect of being HIV positive on being hospitalized decreases by about 35% for each unit increase in max AE grade.



```{r}
# Length of stay given hospitalization
model_hosp_days <- lmer(
  hosp_days ~ ae_max_grade + hiv_status + ae_max_grade:hiv_status + (1 | record_id),
  data = hosp_model_data |> filter(hosp_days > 0)
)
summary(model_hosp_days)
```
AE severity has minimal impact on length of stay for either group once someone is already hospitalized, although HIV positive patients are expected to spend slightly longer in the hospital than HIV negative patients for each unit increase in max adverse event experienced.


# Early Treatment Dropout

```{r}
#| label: prep-early-dropout

early_dropout_cycles <- 4  # patients who stopped at or before this cycle count as early dropouts

# Add early dropout flag to gic_models
gic_models <- gic_models |>
  mutate(
    early_dropout = as.integer(
      cycles_given <= early_dropout_cycles & cycles_prescribed > early_dropout_cycles
    )
  )
```

## Descriptive Overview

This section looks at whether the types of adverse events patients experience in their first few cycles are associated with stopping treatment early, and whether that relationship looks different for HIV-positive vs HIV-negative patients.

```{r}
#| label: tbl-early-dropout-by-hiv
#| tbl-cap: "Early treatment discontinuation by HIV status at two cycle cutoffs"

map_df(c(2, 4), function(cutoff) {
  gic_models |>
    mutate(dropout = as.integer(cycles_given <= cutoff & cycles_prescribed > cutoff)) |>
    group_by(hiv_status) |>
    summarize(
      n_total     = n(),
      n_dropout   = sum(dropout == 1, na.rm = TRUE),
      pct_dropout = round(100 * n_dropout / n_total, 1),
      .groups     = "drop"
    ) |>
    mutate(
      cutoff            = cutoff,
      `Early dropout, n (%)` = paste0(n_dropout, " (", pct_dropout, "%)")
    )
}) |>
  select(hiv_status, cutoff, `Total patients` = n_total, `Early dropout, n (%)`) |>
  pivot_wider(
    names_from  = cutoff,
    values_from = `Early dropout, n (%)`,
    names_glue  = "Cutoff ≤{cutoff} cycles"
  ) |>
  rename(`HIV Status` = hiv_status, `Total patients` = `Total patients`) |>
  gt() |>
  tab_spanner(
    label   = "Early dropout, n (%)",
    columns = starts_with("Cutoff")
  ) |>
  tab_footnote(
    footnote = "Early dropout defined as cycles given ≤ cutoff among patients prescribed more than the cutoff number of cycles.",
    location = cells_column_spanners()
  ) |>
  opt_footnote_marks(marks = "extended") |>
  cols_align(align = "center", columns = -`HIV Status`) |>
  cols_align(align = "left",   columns = `HIV Status`)
```

For each AE type, the predictor used is the max grade recorded across cycles 1 to 4. Where no grade was recorded for a given AE type in that window, it's been set to 0, on the assumption that if nothing was flagged there was probably nothing to flag. That's a more reasonable assumption for the liver panel markers, which Dr. Mazhindu said are usually only ordered when there's a clinical reason to, than it is for something like hemoglobin or platelets where a missing value might just mean the data wasn't entered. It was done this way mainly because doing a complete case analysis instead would have left only 97 patients and 26 early dropout events, which isn't really enough to work with. This is worth coming back to before the analysis is finalized.

Sodium and potassium grades also don't distinguish between high and low abnormalities, so hypokalemia and hyperkalemia end up collapsed into the same grade variable. This can be revisited later, but I'm leaving it as is for the moment because the models below already have more regressors than ideal. See MET-005 and DAT-024.

## Global test for early dropout by HIV status

```{r}
#| label: tst-early-dropout-hiv
#| tbl-cap: "Early treatment discontinuation by HIV status"

# 2x2 table: early dropout x HIV status
dropout_hiv_table <- table(
  dropout = factor(gic_models$early_dropout, levels = c(0, 1), labels = c("Completed", "Early dropout")),
  HIV = gic_models$hiv_status
)

dropout_hiv_table |>
  as.data.frame.matrix() |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE) |>
  add_header_above(c(" " = 1, "HIV Status" = 2))

# Check expected counts to decide test
dropout_expected <- suppressWarnings(chisq.test(dropout_hiv_table)$expected)

if (any(dropout_expected < 5)) {
  dropout_test <- fisher.test(dropout_hiv_table)
  cat("Fisher's exact test (expected count < 5 in at least one cell)\n")
} else {
  dropout_test <- chisq.test(dropout_hiv_table, correct = FALSE)
  cat("Chi-squared test (all expected counts >= 5)\n")
}

dropout_test
```

```{r}
#| label: prep-ae-early-dropout

# Max grade per AE type across cycles 1-early_dropout_cycles
# Patients with no recorded grade for a given AE type in this window
# are excluded from that AE type's predictor (not treated as grade 0)
ae_early <- ae_long |>
  filter(cycle <= early_dropout_cycles, !is.na(grade)) |>
  group_by(record_id, ae_type) |>
  summarize(max_grade_early = max(grade, na.rm = TRUE), .groups = "drop") |>
  left_join(
    gic_models |> select(record_id, hiv_status, early_dropout),
    by = "record_id"
  ) |>
  filter(!is.na(early_dropout))

# Wide format for modeling — one row per patient, one col per AE type
ae_types <- unique(ae_early$ae_type)

ae_early_wide <- ae_early |>
  pivot_wider(
    names_from  = ae_type,
    values_from = max_grade_early
  ) |>
  mutate(
    # All missing AE grades imputed to 0 — absence of a recorded grade treated
    # as no adverse event. Less defensible for some markers (e.g. plt, hb) than
    # for liver panels, but necessary to retain full sample for LASSO.
    # Revisit imputation strategy when time permits.
    across(all_of(ae_types), ~ replace_na(.x, 0))
  ) |>
  left_join(
    gic_models |> select(record_id, hiv_status, early_dropout),
    by = c("record_id", "hiv_status", "early_dropout")
  )
```

```{r}
#| label: tbl-ae-early-dropout
#| tbl-cap: "Adverse event grades in cycles 1–4 by early dropout status and HIV status"

ae_early |>
  mutate(ae_label = ae_labels[ae_type]) |>
  group_by(hiv_status, early_dropout, ae_label) |>
  summarize(
    n_any    = sum(max_grade_early > 0, na.rm = TRUE),
    med      = median(max_grade_early, na.rm = TRUE),
    n_severe = sum(max_grade_early >= 3, na.rm = TRUE),
    .groups  = "drop"
  ) |>
  mutate(
    dropout_label = if_else(early_dropout == 1, "Early dropout", "Completed >4 cycles"),
    cell          = paste0(n_any, " / ", round(med, 1), " / ", n_severe)
  ) |>
  select(hiv_status, ae_label, dropout_label, cell) |>
  pivot_wider(
    names_from  = c(hiv_status, dropout_label),
    values_from = cell,
    names_glue  = "{hiv_status}: {dropout_label}"
  ) |>
  select(
    ae_label,
    `Negative: Completed >4 cycles`,
    `Negative: Early dropout`,
    `Positive: Completed >4 cycles`,
    `Positive: Early dropout`
  ) |>
  rename(`Adverse event` = ae_label) |>
  gt() |>
  tab_spanner(
    label   = "HIV Negative",
    columns = starts_with("Negative")
  ) |>
  tab_spanner(
    label   = "HIV Positive",
    columns = starts_with("Positive")
  ) |>
  cols_label(
    `Negative: Completed >4 cycles` = "Completed >4",
    `Negative: Early dropout`        = "Early dropout",
    `Positive: Completed >4 cycles`  = "Completed >4",
    `Positive: Early dropout`         = "Early dropout"
  ) |>
  cols_align(align = "center", columns = -`Adverse event`) |>
  cols_align(align = "left",   columns = `Adverse event`) |>
  tab_footnote(
    footnote = "Cells show: n with any AE / median max grade / n with severe AE (grade ≥ 3)",
    location = cells_column_spanners()
  ) |>
  opt_footnote_marks(marks = "extended")
```


```{r}
#| label: tst-dropout-by-ae-type
#| tbl-cap: "Association between experiencing each AE type (any grade) in cycles 1–4 and early dropout"

dropout_ae_tests <- map_dfr(ae_types, function(at) {
  dat <- ae_early_wide |>
    filter(!is.na(early_dropout)) |>
    mutate(has_ae = as.integer(.data[[at]] > 0))
  
  tab <- table(dropout = dat$early_dropout, has_ae = dat$has_ae)
  
  if (nrow(tab) < 2 | ncol(tab) < 2) {
    return(tibble(ae_type = at, p = NA_real_, test = "—",
                  n_ae = NA, n_ae_dropout = NA, n_total = NA))
  }
  
  expected <- suppressWarnings(chisq.test(tab)$expected)
  if (any(expected < 5)) {
    result <- fisher.test(tab)
    test_used <- "Fisher's"
  } else {
    result <- chisq.test(tab, correct = FALSE)
    test_used <- "Chi-squared"
  }
  
  tibble(
    ae_type = at,
    n_total = nrow(dat),
    n_ae = sum(dat$has_ae == 1),
    n_ae_dropout = sum(dat$has_ae == 1 & dat$early_dropout == 1),
    p = result$p.value,
    test = test_used
  )
}) |>
  mutate(
    ae_label = ae_labels[ae_type],
    p_adj = p.adjust(p, method = "BH"),
    `p (raw)` = map_chr(p, format_p),
    `p (adj)` = map_chr(p_adj, format_p),
    `With AE, n (dropout)` = paste0(n_ae, " (", n_ae_dropout, ")")
  )

dropout_ae_tests |>
  select(`Adverse event` = ae_label, `With AE, n (dropout)`, Test = test, `p (raw)`, `p (adj)`) |>
  gt() |>
  cols_align(align = "left", columns = `Adverse event`) |>
  cols_align(align = "center", columns = -`Adverse event`) |>
  tab_footnote(
    footnote = paste0(
      "Tests whether having any occurrence of each AE type (grade > 0) in cycles 1–",
      early_dropout_cycles,
      " is associated with early dropout, pooling HIV groups. ",
      "Chi-squared or Fisher's exact as appropriate. p-values BH-adjusted across AE types."
    ),
    locations = cells_column_spanners()
  ) |>
  opt_footnote_marks(marks = "extended")
```

Tests whether, among all patients in the early dropout analysis, is having any occurrence of a specific AE type (grade > 0) in cycles 1-4 associated with whether they droppoed out? Multiple comparrisons correction applied because testing the same population multiple times across adverse event types.


The heatmap below gives a quick visual overview of where the grade differences between early dropout and completed patients are largest, and whether those patterns look similar across HIV groups.

```{r}
#| label: plt-ae-early-dropout-heatmap
#| fig-cap: "Median max AE grade in cycles 1–4 by AE type, early dropout status, and HIV status"

ae_early |>
  group_by(hiv_status, early_dropout, ae_type) |>
  summarize(
    median_grade = median(max_grade_early, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    dropout_label = if_else(early_dropout == 1, "Early dropout", "Completed ≥5 cycles"),
    facet_label   = paste0("HIV ", hiv_status)
  ) |>
  ggplot(aes(x = dropout_label, y = ae_type, fill = median_grade)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(median_grade, 1)), size = 3) +
  scale_fill_viridis_c(option = "magma", direction = -1, 
                       limits = c(0, 4), name = "Median\nmax grade") +
  facet_wrap(~ facet_label) +
  labs(
    x = NULL,
    y = "Adverse event type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1),
    strip.text  = element_text(face = "bold")
  )
```

Before moving to the models, it's worth checking how many patients in each HIV group actually experienced each type of adverse event. AE types with very few patients in either group are excluded from both the LASSO and the univariate models since there isn't enough data to estimate anything reliable for them.

```{r}
#| label: tbl-ae-coverage
#| tbl-cap: "AE grade coverage in cycles 1–4: patients with any recorded adverse event by HIV status"

ae_early_wide |>
  select(hiv_status, early_dropout, all_of(ae_types)) |>
  pivot_longer(
    cols      = all_of(ae_types),
    names_to  = "ae_type",
    values_to = "grade"
  ) |>
  group_by(ae_type, hiv_status) |>
  summarize(
    n_total    = n(),
    n_any_ae   = sum(grade > 0, na.rm = TRUE),
    pct_any_ae = round(100 * n_any_ae / n_total, 1),
    .groups    = "drop"
  ) |>
  mutate(
    ae_label = ae_labels[ae_type],
    `n (%)` = paste0(n_any_ae, " (", pct_any_ae, "%)")
  ) |>
  select(ae_label, hiv_status, `n (%)`) |>
  pivot_wider(
    names_from  = hiv_status,
    values_from = `n (%)`,
    names_prefix = "HIV "
  ) |>
  rename(`Adverse event` = ae_label) |>
  gt() |>
  tab_spanner(
    label   = "Patients with any AE (grade > 0)",
    columns = starts_with("HIV")
  ) |>
  cols_align(align = "left",   columns = `Adverse event`) |>
  cols_align(align = "center", columns = -`Adverse event`) |>
  tab_footnote(
    footnote = paste0(
      "Among patients with cycles given ≤ ", early_dropout_cycles,
      " or prescribed > ", early_dropout_cycles, " cycles. ",
      "AE types with 0 patients in either group cannot be evaluated in LASSO or univariate models."
    ),
    location = cells_column_spanners()
  ) |>
  opt_footnote_marks(marks = "extended")
```

```{r}
#| label: prep-sparse-ae

# AE types with fewer than 5 patients with any non-zero grade in either
# HIV group — too sparse for reliable estimation
ae_cols <- intersect(ae_types, names(ae_early_wide))

sparse_ae <- ae_early_wide |>
  select(hiv_status, all_of(ae_cols)) |>
  pivot_longer(cols = all_of(ae_cols), names_to = "ae_type", values_to = "grade") |>
  group_by(ae_type, hiv_status) |>
  summarize(n_any = sum(grade > 0, na.rm = TRUE), .groups = "drop") |>
  group_by(ae_type) |>
  filter(any(n_any < 5)) |>
  pull(ae_type) |>
  unique()
```


## Univariate Logistic Models

```{r}
#| label: fit-univariate-ae-models

# One logistic model per AE type — early dropout ~ max grade in cycles 1-4 + HIV status
# Patients with no recorded grade for a given AE type are excluded from that model
univariate_results <- map_df(ae_types, function(at) {
  dat <- ae_early_wide |>
    filter(!is.na(.data[[at]]), !is.na(early_dropout), !is.na(hiv_status))

  if (sum(dat$early_dropout) < 10) {
    return(tibble(ae_type = at, note = "Insufficient early dropout cases", .rows = 1))
  }

  if (at %in% sparse_ae) {
    return(tibble(ae_type = at, note = "Sparse — excluded per coverage check", .rows = 1))
  }

  mod <- glm(
    as.formula(paste0("early_dropout ~ `", at, "` * hiv_status")),
    data   = dat,
    family = binomial
  )

  tidy(mod, conf.int = TRUE, exponentiate = TRUE) |>
    filter(str_detect(term, at)) |>
    mutate(
      ae_type    = at,
      term_label = if_else(
        str_detect(term, "hiv_status"),
        "AE grade × HIV Positive",
        "AE grade (main effect)"
      )
    )
}) |>
  filter(!is.na(estimate)) |>
  mutate(p_adj = p.adjust(p.value, method = "BH"))
```

```{r}
#| label: tbl-univariate-ae-results
#| tbl-cap: "Univariate logistic regression: association between AE type max grade and early dropout (adjusted for HIV status)"

# Sample sizes per AE type for context
ae_sample_sizes <- map_df(ae_types, function(at) {
  dat <- ae_early_wide |>
    filter(!is.na(early_dropout), !is.na(hiv_status))
  
  dat |>
    group_by(hiv_status) |>
    summarize(
      n_nonzero        = sum(.data[[at]] > 0),
      n_events_nonzero = sum(.data[[at]] > 0 & early_dropout == 1),
      .groups          = "drop"
    ) |>
    mutate(
      ae_type  = at,
      cell     = paste0(n_nonzero, " (", n_events_nonzero, ")")
    ) |>
    select(ae_type, hiv_status, cell) |>
    pivot_wider(names_from = hiv_status, values_from = cell,
                names_glue = "n_{hiv_status}")
})

# Sort labels
ae_order <- sort(ae_labels)

# Any AE types not in boot_results (dropped as sparse/zero-variance) go at the end
ae_order <- c(ae_order, setdiff(ae_labels, ae_order))

univariate_results |>
  left_join(ae_sample_sizes, by = "ae_type") |>
  mutate(
    ae_label      = ae_labels[ae_type],
    ae_label      = factor(ae_label, levels = sort(ae_labels)),
    `OR (95% CI)` = paste0(round(estimate, 2), " (", round(conf.low, 2), "–", round(conf.high, 2), ")"),
    `p (adj)`     = map_chr(p_adj, format_p),
    `p (raw)`     = map_chr(p.value, format_p)
  ) |>
  arrange(ae_label, term_label) |>
  select(
    `Adverse event`  = ae_label,
    `Term`           = term_label,
    `HIV-: n (drop)` = n_Negative,
    `HIV+: n (drop)` = n_Positive,
    `OR (95% CI)`,
    `p (raw)`,
    `p (adj)`
  ) |>
  gt() |>
  tab_row_group(
    label = "AE grade × HIV interaction",
    rows  = Term == "AE grade × HIV Positive"
  ) |>
  tab_row_group(
    label = "AE grade main effect (HIV adjusted)",
    rows  = Term == "AE grade (main effect)"
  ) |>
  cols_hide(Term) |>
  cols_align(align = "left",   columns = `Adverse event`) |>
  cols_align(align = "center", columns = -`Adverse event`) |>
  tab_footnote(
    footnote = "Main effect OR: change in odds of early dropout per one-unit increase in max AE grade, holding HIV status constant. Interaction OR: difference in that association for HIV-positive vs HIV-negative patients. N (events) = patients with any grade > 0 for this AE type / early dropouts among those patients. p-values BH-adjusted across all tests.",
    location = cells_column_labels(columns = `p (adj)`)
  ) |>
  opt_footnote_marks(marks = "extended")
```

To make the HIV comparison more concrete, the table below shows estimated ORs separately for HIV-negative and HIV-positive patients. The HIV-positive OR is derived by multiplying the main effect OR by the interaction OR, since those add on the log-odds scale. There's no direct standard error for the HIV-positive OR from the model, so treat those as directional estimates rather than something to draw firm conclusions from.

```{r}
#| label: tbl-univariate-hiv-pos-or
#| tbl-cap: "Estimated odds ratios for early dropout per unit increase in max AE grade among HIV-positive patients"

univariate_results |>
  select(ae_type, term_label, estimate, p_adj) |>
  pivot_wider(names_from = term_label, 
              values_from = c(estimate, p_adj)) |>
  mutate(
    ae_label   = factor(ae_labels[ae_type], levels = ae_order),
    or_hiv_pos = `estimate_AE grade (main effect)` * `estimate_AE grade × HIV Positive`
  ) |>
  arrange(ae_label) |>
  mutate(
    `OR (HIV-)` = round(`estimate_AE grade (main effect)`, 3),
    `p (HIV-)`  = map_chr(`p_adj_AE grade (main effect)`, format_p),
    `OR (HIV+)` = round(or_hiv_pos, 3),
    `p (interaction)` = map_chr(`p_adj_AE grade × HIV Positive`, format_p)
  ) |>
  select(
    `Adverse event`   = ae_label,
    `OR (HIV-)`,
    `p (HIV-)`,
    `OR (HIV+)`,
    `p (interaction)`
  ) |>
  gt() |>
  tab_spanner(
    label   = "HIV Negative",
    columns = c(`OR (HIV-)`, `p (HIV-)`)
  ) |>
  tab_spanner(
    label   = "HIV Positive",
    columns = c(`OR (HIV+)`, `p (interaction)`)
  ) |>
  cols_align(align = "left",   columns = `Adverse event`) |>
  cols_align(align = "center", columns = -`Adverse event`) |>
  tab_footnote(
    footnote = "ORs reflect change in odds of early dropout per unit increase in max AE grade. OR (HIV+) = main effect OR × interaction OR. p-values are BH-adjusted; p (interaction) tests whether the association differs between HIV groups.",
    location = cells_column_spanners(spanners = "HIV Positive")
  ) |>
  opt_footnote_marks(marks = "extended")
```

