---
title: "GI Cancer: Exploratory Data Analysis"
bibliography: references.bib
---

# Load Data & Libraries
```{r}
#| label: setup
#| include: false

source("R/setup.R")

# Load Data
gic_validated <- readRDS(here("data", "processed", "gic_validated.rds"))

# Load Libraries
```

# Variable Engineering

```{r}
#| label: engineer-ae-variables

# Identify adverse event grade columns
grade_cols <- names(gic_validated)[str_detect(names(gic_validated), "_grade_([1-9]|1[0-2])$")]
### Should I include baselines as well?

# Create long table to easily engineer new variables
ae_long <- gic_validated |> 
  select(record_id, cycles_prescribed, cycles_given, all_of(grade_cols)) |> 
  pivot_longer(
    cols = all_of(grade_cols),
    names_to = "grade_var",
    values_to = "grade_raw"
  ) |> 
  mutate(
    cycle = as.integer(str_extract(grade_var, "\\d+$")),
    ae_type = str_remove(grade_var, "_grade_\\d+$"),
    grade = as.numeric(str_match(grade_raw, "(?i)^\\s*grade\\s*(\\d+)")[,2])
  )

# Cycle-level summaries
cycle_summary <- ae_long |> 
  group_by(record_id, cycles_prescribed, cycles_given, cycle) |> 
  summarize(
    ae_n_markers_measured = sum(!is.na(grade)),
    
    ae_sum_grade_raw = sum(grade, na.rm = TRUE),
    ae_n_events_raw = sum(grade > 0, na.rm = TRUE),
    ae_n_severe_raw = sum(grade >= 3, na.rm = TRUE),
    
    ae_max_grade_raw = suppressWarnings(max(grade, na.rm = TRUE)),
    
    .groups = "drop"
  ) |> 
  mutate(
    ae_sum_grade = if_else(ae_n_markers_measured == 0, NA_real_, ae_sum_grade_raw),
    ae_n_events = if_else(ae_n_markers_measured == 0, NA_real_, ae_n_events_raw),
    ae_n_severe = if_else(ae_n_markers_measured == 0, NA_real_, ae_n_severe_raw),
    
    ae_max_grade = if_else(ae_n_markers_measured == 0, NA_real_, ae_max_grade_raw),
    
    ae_any = if_else(is.na(ae_n_events), NA, ae_n_events > 0),
    ae_severe_any = if_else(is.na(ae_n_severe), NA, ae_n_severe > 0)
  ) |> 
  select(-ends_with("raw"))

ae_cycle_wide <- cycle_summary |>
  pivot_wider(
    id_cols = c(record_id, cycles_prescribed, cycles_given),
    names_from = cycle,
    values_from = c(ae_n_markers_measured, ae_sum_grade, ae_max_grade, ae_n_events, ae_n_severe, ae_any, ae_severe_any),
    names_glue = "{.value}_{cycle}"
  )

# Patient-level summaries
ae_summary_pt <- ae_long |> 
  group_by(record_id) |> 
  summarize(
    ae_n_cycles_any = n_distinct(cycle[grade > 0], na.rm = TRUE),
    ae_any_early = as.integer(any(grade > 0 & cycle <= 2, na.rm = TRUE)),
    ae_severe_any_pt = as.integer(any(grade >= 3, na.rm = TRUE)),
    ae_severe_any_early = as.integer(any(grade >= 3 & cycle <= 2, na.rm = TRUE)),
    ae_any_pt = as.integer(any(grade > 0, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  left_join(
    cycle_summary |>
      filter(ae_n_markers_measured > 0) |>   # only cycles with any grading recorded
      group_by(record_id) |>
      summarize(
        ae_mean_events = mean(ae_n_events, na.rm = TRUE),
        ae_mean_max_grade = mean(ae_max_grade, na.rm = TRUE),
        ae_mean_sum_grade = mean(ae_sum_grade, na.rm = TRUE),
        .groups = "drop"
      ),
    by = "record_id"
  )

# Merge datasets for analysis
gic_models <- gic_validated |>
  left_join(ae_summary_pt, by = "record_id") |>
  left_join(ae_cycle_wide, by = c("record_id", "cycles_prescribed", "cycles_given"))
```

## Summary of new adverse event (AE) variables

| Variable                  | Level   | Type         | Meaning                                                                                                          |
| ------------------------- | ------- | ------------ | ---------------------------------------------------------------------------------------------------------------- |
| `ae_n_cycles_any`         | Patient | int (0–12)   | Number of distinct cycles where the patient had any AE (at least one `grade > 0`).                               |
| `ae_any_pt`               | Patient | int (0/1)    | Flag: did the patient have any AE (`grade > 0`) at any point in treatment?                                       |
| `ae_any_early`            | Patient | int (0/1)    | Flag: did the patient have any AE (`grade > 0`) in cycles 1–2?                                                   |
| `ae_severe_any_pt`        | Patient | int (0/1)    | Flag: did the patient have any severe AE (`grade >= 3`) across all cycles?                                       |
| `ae_severe_any_early`     | Patient | int (0/1)    | Flag: did the patient have any severe AE (`grade >= 3`) in cycles 1–2?                                           |
| `ae_mean_events`          | Patient | numeric / NA | Mean number of AEs per cycle across cycles with any AE grading recorded (`ae_n_markers_measured > 0`).           |
| `ae_mean_max_grade`       | Patient | numeric / NA | Mean of the per-cycle maximum AE grade across cycles with any AE grading recorded (`ae_n_markers_measured > 0`). |
| `ae_mean_sum_grade`       | Patient | numeric / NA | Mean of the per-cycle sum of AE grades across cycles with any AE grading recorded (`ae_n_markers_measured > 0`). |
| `ae_n_markers_measured_#` | Cycle   | int (0–12)   | For cycle #, number of AE markers with a non-missing numeric grade.                                              |
| `ae_sum_grade_#`          | Cycle   | numeric / NA | For cycle #, sum of numeric AE grades across markers (NA if `ae_n_markers_measured_# == 0`).                     |
| `ae_max_grade_#`          | Cycle   | numeric / NA | For cycle #, maximum numeric AE grade across markers (NA if `ae_n_markers_measured_# == 0`).                     |
| `ae_n_events_#`           | Cycle   | numeric / NA | For cycle #, count of AEs across markers (`grade > 0`) (NA if `ae_n_markers_measured_# == 0`).                   |
| `ae_n_severe_#`           | Cycle   | numeric / NA | For cycle #, count of severe AEs across markers (`grade >= 3`) (NA if `ae_n_markers_measured_# == 0`).           |
| `ae_any_#`                | Cycle   | logical / NA | For cycle #, flag: any AE this cycle (`ae_n_events_# > 0`) (NA if `ae_n_markers_measured_# == 0`).               |
| `ae_severe_any_#`         | Cycle   | logical / NA | For cycle #, flag: any severe AE this cycle (`ae_n_severe_# > 0`) (NA if `ae_n_markers_measured_# == 0`).        |



## Patient-Level Adverse Event Variables
```{r}
#| label: tbl-ae-by-pt
#| tbl-cap: "Patient-level AE summaries by HIV status. Means are computed across cycles with any AE grading recorded."

fmt_n_pct <- function(x, digits = 1) {
  n <- sum(x == 1, na.rm = TRUE)
  d <- sum(!is.na(x))
  pct <- if (d == 0) NA_real_ else 100 * n / d
  paste0(n, " (", round(pct, digits), "%)")
}

fmt_mean_sd <- function(x, digits = 2) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_character_)
  paste0(round(mean(x), digits), " (", round(stats::sd(x), digits), ")")
}

pt_summ <- gic_models |>
  group_by(hiv_status) |>
  summarize(
    total_pts = as.character(n()),

    any_ae = fmt_n_pct(ae_any_pt, digits = 1),
    any_ae_early = fmt_n_pct(ae_any_early, digits = 1),
    any_severe_ae = fmt_n_pct(ae_severe_any_pt, digits = 1),
    any_severe_ae_early = fmt_n_pct(ae_severe_any_early, digits = 1),

    mean_events = fmt_mean_sd(ae_mean_events),
    mean_max_grade = fmt_mean_sd(ae_mean_max_grade),
    mean_sum_grade = fmt_mean_sd(ae_mean_sum_grade),

    .groups = "drop"
  )

metric_labels <- c(
  total_pts = "Total patients (N)",
  any_ae = "Any AE (%, patients)",
  any_ae_early = "Any AE in cycles 1–2 (%, patients)",
  any_severe_ae = "Any severe AE (%, patients)",
  any_severe_ae_early = "Any severe AE in cycles 1–2 (%, patients)",
  mean_events = "Mean # AEs per measured cycle (SD)",
  mean_max_grade = "Mean max grade per measured cycle (SD)",
  mean_sum_grade = "Mean sum of grades per measured cycle (SD)"
)

pt_wide <- pt_summ |>
  pivot_longer(
    cols = -hiv_status,
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(Metric = unname(metric_labels[metric])) |>
  select(Metric, hiv_status, value) |>
  pivot_wider(names_from = hiv_status, values_from = value)

pt_wide |>
  kbl(booktabs = TRUE, align = "lcc") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## Cycle-Level Adverse Event Variables
```{r}
#| label: tbl-ae-by-cycle
#| tbl-cap: "Cycle-level AE summaries by HIV status (among patients with any AE grading recorded in the cycle)"
#| results: asis

# Function to calculate mean, standard deviation, and range
fmt_mean_sd_range <- function(x, digits = 1) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_character_)
  m <- mean(x); s <- stats::sd(x)
  rng <- range(x)
  paste0(
    round(m, digits), " (", round(s, digits), "); ",
    round(rng[[1]], digits), "–", round(rng[[2]], digits)
  )
}

# Pivot cycle-level vars back to long form for summarizing
ae_cycle_long <- gic_models |>
  select(
    record_id, hiv_status, cycles_prescribed, cycles_given,
    matches("^ae_n_markers_measured_\\d+$"),
    matches("^ae_n_events_\\d+$"),
    matches("^ae_n_severe_\\d+$"),
    matches("^ae_sum_grade_\\d+$"),
    matches("^ae_max_grade_\\d+$"),
    matches("^ae_any_\\d+$"),
    matches("^ae_severe_any_\\d+$")
  ) |>
  pivot_longer(
    cols = -c(record_id, hiv_status, cycles_prescribed, cycles_given),
    names_to = c("metric", "cycle"),
    names_pattern = "^(.*)_(\\d+)$",
    values_to = "value"
  ) |>
  mutate(cycle = as.integer(cycle)) |>
  pivot_wider(names_from = metric, values_from = value)

# Summarize by HIV status & cycle
ae_cycle_summ <- ae_cycle_long |>
  group_by(hiv_status, cycle) |>
  summarize(
    total_pts = n(),

    measured_n = sum(ae_n_markers_measured > 0, na.rm = TRUE),
    measured_pct = 100 * measured_n / total_pts,

    any_ae_n = sum(ae_any == TRUE, na.rm = TRUE),
    any_ae_pct_measured = if_else(measured_n == 0, NA_real_, 100 * any_ae_n / measured_n),

    severe_ae_n = sum(ae_severe_any == TRUE, na.rm = TRUE),
    severe_ae_pct_measured = if_else(measured_n == 0, NA_real_, 100 * severe_ae_n / measured_n),

    n_events_mean_sd = fmt_mean_sd_range(ae_n_events),
    n_severe_mean_sd = fmt_mean_sd_range(ae_n_severe),
    sum_grade_mean_sd = fmt_mean_sd_range(ae_sum_grade),
    max_grade_mean_sd = fmt_mean_sd_range(ae_max_grade, digits = 0),

    .groups = "drop"
  ) |>
  mutate(
    `Cycle` = cycle,
    `Patients w/ any AE grading recorded` = paste0(measured_n, " (", round(measured_pct, 1), "%)"),
    `Any AE in cycle (%, among measured)` = paste0(any_ae_n, " (", round(any_ae_pct_measured, 1), "%)"),
    `Any severe AE in cycle (%, among measured)` = paste0(severe_ae_n, " (", round(severe_ae_pct_measured, 1), "%)"),
    `# AEs in cycle, mean (SD); range` = n_events_mean_sd,
    `# severe AEs in cycle, mean (SD); range` = n_severe_mean_sd,
    `Sum of AE grades in cycle, mean (SD); range` = sum_grade_mean_sd,
    `Max grade in cycle, mean (SD); range` = max_grade_mean_sd
  ) |>
  select(
    hiv_status, `Cycle`,
    `Patients w/ any AE grading recorded`,
    `Any AE in cycle (%, among measured)`,
    `Any severe AE in cycle (%, among measured)`,
    `# AEs in cycle, mean (SD); range`,
    `# severe AEs in cycle, mean (SD); range`,
    `Sum of AE grades in cycle, mean (SD); range`,
    `Max grade in cycle, mean (SD); range`
  )

# Print one table per HIV status, too busy having it all in one table
ae_cycle_summ |>
  group_by(hiv_status) |>
  group_walk(\(df, key) {
    cat("\n\n### HIV status:", as.character(key$hiv_status), "\n\n")
    df |>
      select(-hiv_status) |>
      kbl(booktabs = TRUE, align = "lccccccc") |>
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) |>
      print()
  }, .keep = TRUE)
```

```{r}
#| label: prep-ae-cycle-plot-data

# Build cycle-level dataset from gic_models for plotting
ae_cycle_measured <- ae_cycle_long |> 
  filter(ae_n_markers_measured > 0) |> 
  mutate(
    pct_complete = if_else(is.na(cycles_prescribed) | cycles_prescribed <= 0,
                           NA_real_,
                           cycle / cycles_prescribed),
    pct_bin = cut(
      pct_complete,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE,
      right = TRUE
    )
  )

summarize_metric <- function(df, var) {
  df |> 
    group_by(hiv_status, cycle) |> 
    summarize(
      n = sum(!is.na(.data[[var]])),
      mean = mean(.data[[var]], na.rm = TRUE),
      q25 = quantile(.data[[var]], 0.25, na.rm = TRUE),
      q75 = quantile(.data[[var]], 0.75, na.rm = TRUE),
      min = min(.data[[var]], na.rm = TRUE),
      max = max(.data[[var]], na.rm = TRUE),
      .groups = "drop"
    )
}

ae_events_summ <- summarize_metric(ae_cycle_measured, "ae_n_events")
ae_severe_summ <- summarize_metric(ae_cycle_measured, "ae_n_severe")
ae_sumgrade_summ <- summarize_metric(ae_cycle_measured, "ae_sum_grade")
```
## Plots of Cycle-Level Adverse Event Variables
```{r}
#| label: plt-ae-n-events
#| fig-cap: "Initially higher trend for number of adverse events amongst HIV positive patients, but the mean drops below that of HIV negative patients after cycle 4. Need to investigate, but could have to do with high early dropout rates, or possibly related to patients with HIV generally having fewer cycles of chemotherapy prescribed."

ggplot(ae_events_summ, aes(x = cycle, group = hiv_status, color = hiv_status, fill = hiv_status)) +
  geom_ribbon(aes(ymin = min, ymax = max), alpha = 0.2, color = NA) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.5, color = NA) +
  geom_line(aes(y = mean), linewidth = 1) +
  geom_point(aes(y = mean), size = 2) +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Adverse events per cycle",
       subtitle = "Mean (line) with IQR and min–max bands among cycles with any AE grading recorded", 
       x = "Treatment cycle", 
       y = "Number of AEs", 
       color = "HIV status", fill = "HIV status") +
  theme_minimal()
```

```{r}
#| label: plt-ae-n-severe
#| fig-cap: "Not a very noticable difference based on the means or IQRs. The HIV negative patients tend to have higher max number of severe adverse events per cycle, but as noted before, I'd guess that relates to how few cycles HIV positive patients tend to complete, and the total number of HIV negative patients compared to positive."

ggplot(ae_severe_summ, aes(x = cycle, group = hiv_status, color = hiv_status, fill = hiv_status)) +
  geom_ribbon(aes(ymin = min, ymax = max), alpha = 0.2, color = NA) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.5, color = NA) +
  geom_line(aes(y = mean), linewidth = 1) +
  geom_point(aes(y = mean), size = 2) +
  scale_x_continuous(breaks = 1:12) +
  labs(
    title = "Severe adverse events per cycle",
    subtitle = "Grade ≥ 3; mean (line) with IQR and min–max bands among measured cycles",
    x = "Treatment cycle",
    y = "Number of severe AEs",
    color = "HIV status",
    fill = "HIV status"
  ) +
  theme_minimal()
```

```{r}
#| label: plt-ae-sum-grade
#| fig-cap: "No significant visual difference by HIV status early on other than the IQR/Max decreasing faster than for HIV negative patients, but this likely is related to the number of patients still getting chemotherapy."

ggplot(ae_sumgrade_summ, aes(x = cycle, group = hiv_status, color = hiv_status, fill = hiv_status)) +
  geom_ribbon(aes(ymin = min, ymax = max), alpha = 0.2, color = NA) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.5, color = NA) +
  geom_line(aes(y = mean), linewidth = 1) +
  geom_point(aes(y = mean), size = 2) +
  scale_x_continuous(breaks = 1:12) +
  labs(
    title = "Cumulative AE grade burden per cycle",
    subtitle = "Sum of numeric AE grades across markers; mean (line) with IQR and min–max bands",
    x = "Treatment cycle",
    y = "Sum of AE grades",
    color = "HIV status",
    fill = "HIV status"
  ) +
  theme_minimal()
```




# Model Building
## Simple Logistic:
```{r}
#|label: model-hiv-simple

model_hiv_simple <- glm(hiv_status ~ ae_mean_max_grade, data = gic_models, family = binomial)

summary(model_hiv_simple)
```
```{r}
#| label: tbl-hiv-simple-or
#| tbl-cap: "Logistic regression: HIV status vs mean max AE grade (unadjusted)"

tidy(model_hiv_simple, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    Term = case_when(
      term == "ae_mean_max_grade" ~ "Mean max AE grade per measured cycle",
      TRUE ~ term
    ),
    `OR (95% CI)` = paste0(
      round(estimate, 2), " (",
      round(conf.low, 2), " - ",
      round(conf.high, 2), ")"
    ),
    p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))),
    std.error = round(std.error, 3)
  ) |>
  select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left")
```

**Summary:** Simple logistic regression model with mean max adverse event grade as the sole regressor and HIV negative as the reference value of the outcome variable. Each one-unit increase in mean max adverse event grade (per measured cycle) is associated with a 97% increase in the odds of the patient being HIV positive.



## Basic Logistic
```{r}
#| label: model-hiv-basic

model_hiv_basic <- glm(
  hiv_status ~ ae_mean_max_grade + cancer_stage + chemo_type + tx_intent,
  data = gic_models,
  family = binomial
)

summary(model_hiv_basic)
```

```{r}
#| label: tbl-hiv-basic-or
#| tbl-cap: "Logistic regression: HIV status vs mean max AE grade (adjusted for stage, chemo type, and intent)"

tidy(model_hiv_basic, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    Term = case_when(
      term == "ae_mean_max_grade" ~ "Mean max AE grade per measured cycle",
      TRUE ~ term
    ),
    `OR (95% CI)` = paste0(
      round(estimate, 2), " (",
      round(conf.low, 2), " - ",
      round(conf.high, 2), ")"
    ),
    p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))),
    std.error = round(std.error, 3)
  ) |>
  select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left")

```



<!-- model_any_basic <- glm(ae_any_pt ~ hiv_status + cancer_stage + chemo_type + tx_intent, -->
<!--                        data = gic_models, family = binomial) -->




<!-- ## Simple Logistic: Full treatment period -->
<!-- ### Simple logistic model of any adverse events by HIV status -->
<!-- ```{r} -->
<!-- #| label: model-any_simple -->

<!-- model_any_simple <- glm(ae_any_pt ~ hiv_status, -->
<!--                         data = gic_models, family = binomial) -->

<!-- # summary(model_any_simple) -->
<!-- ``` -->
<!-- Invalid model due to having a 100% adverse event rate in the HIV-positive group. -->

<!-- ### Simple logistic model of any severe adverse events by HIV status -->
<!-- ```{r} -->
<!-- #| label: model-severe-simple -->


<!-- model_severe_simple <- glm(ae_severe_any_pt ~ hiv_status, -->
<!--                            data = gic_models, family = binomial) -->

<!-- summary(model_severe_simple) -->
<!-- ``` -->
<!-- ```{r} -->
<!-- #| label: tbl-severe-simple-or-clean -->
<!-- #| tbl-cap: "Association between HIV Status and Severe Adverse Events (Grades 3-4)" -->

<!-- tidy(model_severe_simple, conf.int = TRUE, exponentiate = TRUE) |> -->
<!--   filter(term != "(Intercept)") |> -->
<!--   # Clean up the name and prepare for the reference row -->
<!--   mutate( -->
<!--     Term = str_replace(term, "hiv_statusPositive", "HIV Positive"), -->
<!--     `OR (95% CI)` = paste0(round(estimate, 2), " (", -->
<!--                            round(conf.low, 2), " - ", -->
<!--                            round(conf.high, 2), ")"), -->
<!--     p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))), -->
<!--     std.error = round(std.error, 3) -->
<!--   ) |> -->
<!--   select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |> -->
<!--   # Add the reference row at the top -->
<!--   add_row(Term = "HIV Negative (Reference)", -->
<!--           `OR (95% CI)` = "1.00", -->
<!--           `Std. Error` = NA, -->
<!--           `P-value` = NA, -->
<!--           .before = 1) |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") -->
<!-- ``` -->



<!-- ## Multivariate Logistic: Full treatment period -->

<!-- ```{r} -->
<!-- #| label: tbl-stage-vs-hiv-status -->
<!-- #| tbl-cap: "Cancer stage vs HIV status" -->

<!-- table(Stage = gic_models$cancer_stage, HIV = gic_models$hiv_status) |> -->
<!--   as.data.frame.matrix() |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") |> -->
<!--   add_header_above(c("Cancer Stage" = 1, "HIV Status" = 2)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- #| label: tbl-chemo-type-vs-hiv-status -->
<!-- #| tbl-cap: "Chemotherapy type vs HIV status" -->

<!-- table(Stage = gic_models$chemo_type, HIV = gic_models$hiv_status) |> -->
<!--   as.data.frame.matrix() |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") |> -->
<!--   add_header_above(c("Chemo Type" = 1, "HIV Status" = 2)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- #| label: tbl-tx-intent-vs-hiv-status -->
<!-- #| tbl-cap: "Treatment intent vs HIV status" -->

<!-- table(Stage = gic_models$tx_intent, HIV = gic_models$hiv_status) |> -->
<!--   as.data.frame.matrix() |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") |> -->
<!--   add_header_above(c("Treatment Intent" = 1, "HIV Status" = 2)) -->
<!-- ``` -->

<!-- ### Basic logistic model of any adverse events by HIV status -->
<!-- ```{r} -->
<!-- #| label: model-any-basic -->

<!-- model_any_basic <- glm(ae_any_pt ~ hiv_status + cancer_stage + chemo_type + tx_intent, -->
<!--                        data = gic_models, family = binomial) -->

<!-- summary(model_any_basic) -->
<!-- ``` -->
<!-- Invalid model due to having a 100% adverse event rate in the HIV-positive group. -->

<!-- ### Basic logicstic model of any severe adverse events by HIV status -->
<!-- ```{r} -->
<!-- #| label: model-severe-basic -->

<!-- model_severe_basic <- glm(ae_severe_any_pt ~ hiv_status + cancer_stage + chemo_type + tx_intent, -->
<!--                        data = gic_models, family = binomial) -->

<!-- summary(model_severe_basic) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- #| label: tbl-severe-multivariate-or -->
<!-- #| tbl-cap: "Adjusted Odds Ratios for Severe Adverse Events (Grades 3-4)" -->

<!-- tidy(model_severe_basic, conf.int = TRUE, exponentiate = TRUE) |> -->
<!--   filter(term != "(Intercept)") |> -->
<!--   mutate( -->
<!--     # Clean up term names -->
<!--     Term = case_when( -->
<!--       term == "hiv_statusPositive" ~ "HIV Positive", -->
<!--       term == "cancer_stageStage III" ~ "Stage III", -->
<!--       term == "cancer_stageStage IV" ~ "Stage IV", -->
<!--       term == "chemo_typePlatinum/5-FU" ~ "Platinum/5-FU", -->
<!--       term == "chemo_typeLCV/5-FU" ~ "LCV/5-FU", -->
<!--       term == "chemo_typeCapeOX" ~ "CapeOX", -->
<!--       term == "chemo_typeFOLFOX4" ~ "FOLFOX4", -->
<!--       term == "chemo_typeCapecitabine" ~ "Capecitabine", -->
<!--       term == "tx_intentPalliative" ~ "Palliative Intent", -->
<!--       TRUE ~ term -->
<!--     ), -->
<!--     `OR (95% CI)` = paste0(round(estimate, 2), " (",  -->
<!--                            round(conf.low, 2), " - ",  -->
<!--                            round(conf.high, 2), ")"), -->
<!--     p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))), -->
<!--     std.error = round(std.error, 3) -->
<!--   ) |> -->
<!--   select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |> -->
<!--   # Add reference rows for clarity -->
<!--   add_row(Term = "HIV Negative (Ref)", `OR (95% CI)` = "1.00", .before = 1) |> -->
<!--   add_row(Term = "Stage II (Ref)", `OR (95% CI)` = "1.00", .before = 3) |> -->
<!--   add_row(Term = "Platinum/Paclitaxel (Ref)", `OR (95% CI)` = "1.00", .before = 6) |> -->
<!--   add_row(Term = "Curative Intent (Ref)", `OR (95% CI)` = "1.00", .before = 12) |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") |> -->
<!--   pack_rows("HIV Status", 1, 2) |> -->
<!--   pack_rows("Cancer Stage", 3, 5) |> -->
<!--   pack_rows("Chemotherapy Regimen", 6, 11) |> -->
<!--   pack_rows("Treatment Intent", 12, 13) -->
<!-- ``` -->



<!-- ## $\chi^2$ Tests (HIV vs Early Events) -->

<!-- ```{r} -->
<!-- #| label: tbl-any-early-contingency -->
<!-- #| tbl-cap: "Contingency table of early adverse events (cycles 1-4) vs HIV status" -->

<!-- table_any_early <- table(early_ae = factor(gic_models$ae_any_early, levels = c(0, 1), labels = c("No", "Yes")), HIV = gic_models$hiv_status) -->

<!-- table_any_early |>  -->
<!--   as.data.frame.matrix() |>  -->
<!--   kbl(booktabs = TRUE) |>  -->
<!--   kable_styling(full_width = FALSE, position = "left") |>  -->
<!--   add_header_above(c("Any Early AE" = 1, "HIV Status" = 2)) -->
<!-- ``` -->
<!-- ```{r} -->
<!-- #| label: tbl-severe-early-contingency -->
<!-- #| tbl-cap: "Contingency table of severe early adverse events (cycles 1-4) vs HIV status" -->

<!-- table_severe_early <- table(severe_early_ae = factor(gic_models$ae_severe_any_early, levels = c(0, 1), labels = c("No", "Yes")), HIV = gic_models$hiv_status) -->

<!-- table_severe_early |>  -->
<!--   as.data.frame.matrix() |>  -->
<!--   kbl(booktabs = TRUE) |>  -->
<!--   kable_styling(full_width = FALSE, position = "left") |>  -->
<!--   add_header_above(c("Severe Early AE" = 1, "HIV Status" = 2)) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- #| label: tests-chi-sq -->

<!-- # Any Early AE (Cycles 1-4) -->
<!-- cat("Chi-Squared Test: Any Early AE (C1-4) vs HIV Status\n") -->
<!-- print(chisq.test(table_any_early)) -->

<!-- cat("\n\n") -->

<!-- # Severe Early AE (Cycles 1-4) -->
<!-- cat("Chi-Squared Test: Severe Early AE (C1-4) vs HIV Status\n") -->
<!-- print(chisq.test(table_severe_early)) -->
<!-- ``` -->





