---
title: "GI Cancer: Exploratory Data Analysis"
bibliography: references.bib
---

# Load Data & Libraries
```{r}
#| label: setup
#| include: false

source("R/setup.R")

# Load Data
gic_validated <- readRDS(here("data", "processed", "gic_validated.rds"))

# Load Libraries
```

# Variable Engineering

```{r}
#| label: engineer-ae-variables

# Identify adverse event grade columns
grade_cols <- names(gic_validated)[str_detect(names(gic_validated), "_grade_([1-9]|1[0-2])$")]
### Should I include baselines as well?

# Create long table to easily engineer new variables
ae_long <- gic_validated |> 
  select(record_id, cycles_prescribed, cycles_given, all_of(grade_cols)) |> 
  pivot_longer(
    cols = all_of(grade_cols),
    names_to = "grade_var",
    values_to = "grade_raw"
  ) |> 
  mutate(
    cycle = as.integer(str_extract(grade_var, "\\d+$")),
    ae_type = str_remove(grade_var, "_grade_\\d+$"),
    grade = as.numeric(str_match(grade_raw, "(?i)^\\s*grade\\s*(\\d+)")[,2])
  )

# Patient-level summaries
ae_summary_pt <- ae_long |> 
  group_by(record_id) |> 
  summarize(
    # Distinct cycles with any adverse event grade > 0
    ae_n_cycles_any = n_distinct(cycle[grade > 0], na.rm = TRUE),
    
    # Early window: cycles 1-2
    ae_any_early = as.integer(any(grade > 0 & cycle <= 2, na.rm = TRUE)),
    ae_severe_any_pt = as.integer(any(grade >= 3, na.rm = TRUE)),
    ae_severe_any_early = as.integer(any(grade >= 3 & cycle <= 2, na.rm = TRUE)),
    ae_any_pt = as.integer(any(grade > 0, na.rm = TRUE)),
    .groups = "drop"
  )

# Cycle-level summaries
cycle_summary <- ae_long |> 
  group_by(record_id, cycles_prescribed, cycles_given, cycle) |> 
  summarize(
    ae_n_markers_measured = sum(!is.na(grade)),
    
    ae_sum_grade_raw = sum(grade, na.rm = TRUE),
    ae_n_events_raw = sum(grade > 0, na.rm = TRUE),
    ae_n_severe_raw = sum(grade >= 3, na.rm = TRUE),
    
    ae_max_grade_raw = suppressWarnings(max(grade, na.rm = TRUE)),
    
    .groups = "drop"
  ) |> 
  mutate(
    ae_sum_grade = if_else(ae_n_markers_measured == 0, NA_real_, ae_sum_grade_raw),
    ae_n_events = if_else(ae_n_markers_measured == 0, NA_real_, ae_n_events_raw),
    ae_n_severe = if_else(ae_n_markers_measured == 0, NA_real_, ae_n_severe_raw),
    
    ae_max_grade = if_else(ae_n_markers_measured == 0, NA_real_, ae_max_grade_raw),
    
    ae_any = if_else(is.na(ae_n_events), NA, ae_n_events > 0),
    ae_severe_any = if_else(is.na(ae_n_severe), NA, ae_n_severe > 0)
  ) |> 
  select(-ends_with("raw"))

ae_cycle_wide <- cycle_summary |>
  pivot_wider(
    id_cols = c(record_id, cycles_prescribed, cycles_given),
    names_from = cycle,
    values_from = c(ae_n_markers_measured, ae_sum_grade, ae_max_grade, ae_n_events, ae_n_severe, ae_any, ae_severe_any),
    names_glue = "{.value}_{cycle}"
  )

# Merge datasets for analysis
gic_models <- gic_validated |>
  left_join(ae_summary_pt, by = "record_id") |>
  left_join(ae_cycle_wide, by = c("record_id", "cycles_prescribed", "cycles_given"))
```

## Summary of new adverse event (AE) variables

| Variable                  | Level   | Type          | Meaning                                                                                                   |
| ------------------------- | ------- | ------------- | --------------------------------------------------------------------------------------------------------- |
| `ae_n_cycles_any`         | Patient | int (0-12)    | Number of distinct cycles where the patient had any AE (at least one `grade > 0`).                        |
| `ae_any_pt`               | Patient | int (0/1)     | Flag: did the patient have any AE (`grade > 0`) at any point in treatment?                                |
| `ae_any_early`            | Patient | int (0/1)     | Flag: did the patient have any AE (`grade > 0`) in cycles 1–2?                                            |
| `ae_severe_any_pt`        | Patient | int (0/1)     | Flag: did the patient have any severe AE (`grade >= 3`) across all cycles?                                |
| `ae_severe_any_early`     | Patient | int (0/1)     | Flag: did the patient have any severe AE (`grade >= 3`) in cycles 1–2?                                    |
| `ae_n_markers_measured_#` | Cycle   | int (0–12)    | For cycle #, number of AE markers with a non-missing numeric grade.                                       |
| `ae_sum_grade_#`          | Cycle   | numeric / NA  | For cycle #, sum of numeric AE grades across markers (NA if `ae_n_markers_measured_# == 0`).              |
| `ae_max_grade_#`          | Cycle   | numeric / NA  | For cycle #, maximum numeric AE grade across markers (NA if `ae_n_markers_measured_# == 0`).              |
| `ae_n_events_#`           | Cycle   | numeric / NA  | For cycle #, count of AEs across markers (`grade > 0`) (NA if `ae_n_markers_measured_# == 0`).            |
| `ae_n_severe_#`           | Cycle   | numeric / NA  | For cycle #, count of severe AEs across markers (`grade >= 3`) (NA if `ae_n_markers_measured_# == 0`).    |
| `ae_any_#`                | Cycle   | logical / NA  | For cycle #, flag: any AE this cycle (`ae_n_events_# > 0`) (NA if `ae_n_markers_measured_# == 0`).        |
| `ae_severe_any_#`         | Cycle   | logical / NA  | For cycle #, flag: any severe AE this cycle (`ae_n_severe_# > 0`) (NA if `ae_n_markers_measured_# == 0`). |



```{r}
#| label: tbl-ae-by-pt
#| tbl-cap: "Patient-level summary of adverse events (AEs) by HIV status. 'Early' refers to the first 2 treatment cycles."

gic_models |> 
  group_by(hiv_status) |> 
  summarize(
    total_pts = n(),
    
    any_ae_n = sum(ae_any_pt),
    any_ae_pct = (any_ae_n / total_pts) * 100,
    
    severe_ae_n = sum(ae_severe_any_pt),
    severe_ae_pct = (severe_ae_n / total_pts) * 100,
    
    early_ae_n = sum(ae_any_early),
    early_ae_pct = (early_ae_n / total_pts) * 100,
    
    severe_early_ae_n = sum(ae_severe_any_early),
    severe_early_ae_pct = (severe_early_ae_n / total_pts) * 100,
    .groups = "drop"
  ) |> 
  mutate(
    "HIV Status" = hiv_status,
    "Total Patients" = total_pts,
    "Any AE (%)" = paste0(any_ae_n, " (", round(any_ae_pct, 2), "%)"),
    "Severe AE (%)" = paste0(severe_ae_n, " (", round(severe_ae_pct, 2), "%)"),
    "Early AE (%)" = paste0(early_ae_n, " (", round(early_ae_pct, 2), "%)"),
    "Severe Early AE (%)" = paste0(severe_early_ae_n, " (", round(severe_early_ae_pct, 2), "%)")
  ) |> 
  select("HIV Status", "Total Patients", "Any AE (%)", "Severe AE (%)", "Early AE (%)", "Severe Early AE (%)") |> 
  kbl(booktabs = TRUE, align = "lcccc") |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
#| label: tbl-ae-by-cycle
#| tbl-cap: "Cycle-level AE summaries by HIV status (among patients with any AE grading recorded in the cycle)"
#| results: asis

# Function to calculate mean, standard deviation, and range
fmt_mean_sd_range <- function(x, digits = 1) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_character_)
  m <- mean(x); s <- stats::sd(x)
  rng <- range(x)
  paste0(
    round(m, digits), " (", round(s, digits), "); ",
    round(rng[[1]], digits), "–", round(rng[[2]], digits)
  )
}

# Pivot cycle-level vars back to long form for summarizing
ae_cycle_long <- gic_models |>
  select(
    record_id, hiv_status,
    matches("^ae_n_markers_measured_\\d+$"),
    matches("^ae_n_events_\\d+$"),
    matches("^ae_n_severe_\\d+$"),
    matches("^ae_sum_grade_\\d+$"),
    matches("^ae_max_grade_\\d+$"),
    matches("^ae_any_\\d+$"),
    matches("^ae_severe_any_\\d+$")
  ) |>
  pivot_longer(
    cols = -c(record_id, hiv_status),
    names_to = c("metric", "cycle"),
    names_pattern = "^(.*)_(\\d+)$",
    values_to = "value"
  ) |>
  mutate(cycle = as.integer(cycle)) |>
  pivot_wider(names_from = metric, values_from = value)

# Summarize by HIV status & cycle
ae_cycle_summ <- ae_cycle_long |>
  group_by(hiv_status, cycle) |>
  summarize(
    total_pts = n(),

    measured_n = sum(ae_n_markers_measured > 0, na.rm = TRUE),
    measured_pct = 100 * measured_n / total_pts,

    any_ae_n = sum(ae_any == TRUE, na.rm = TRUE),
    any_ae_pct_measured = if_else(measured_n == 0, NA_real_, 100 * any_ae_n / measured_n),

    severe_ae_n = sum(ae_severe_any == TRUE, na.rm = TRUE),
    severe_ae_pct_measured = if_else(measured_n == 0, NA_real_, 100 * severe_ae_n / measured_n),

    n_events_mean_sd = fmt_mean_sd_range(ae_n_events),
    n_severe_mean_sd = fmt_mean_sd_range(ae_n_severe),
    sum_grade_mean_sd = fmt_mean_sd_range(ae_sum_grade),
    max_grade_mean_sd = fmt_mean_sd_range(ae_max_grade, digits = 0),

    .groups = "drop"
  ) |>
  mutate(
    `Cycle` = cycle,
    `Patients w/ any AE grading recorded` = paste0(measured_n, " (", round(measured_pct, 1), "%)"),
    `Any AE in cycle (%, among measured)` = paste0(any_ae_n, " (", round(any_ae_pct_measured, 1), "%)"),
    `Any severe AE in cycle (%, among measured)` = paste0(severe_ae_n, " (", round(severe_ae_pct_measured, 1), "%)"),
    `# AEs in cycle, mean (SD); range` = n_events_mean_sd,
    `# severe AEs in cycle, mean (SD); range` = n_severe_mean_sd,
    `Sum of AE grades in cycle, mean (SD); range` = sum_grade_mean_sd,
    `Max grade in cycle, mean (SD); range` = max_grade_mean_sd
  ) |>
  select(
    hiv_status, `Cycle`,
    `Patients w/ any AE grading recorded`,
    `Any AE in cycle (%, among measured)`,
    `Any severe AE in cycle (%, among measured)`,
    `# AEs in cycle, mean (SD); range`,
    `# severe AEs in cycle, mean (SD); range`,
    `Sum of AE grades in cycle, mean (SD); range`,
    `Max grade in cycle, mean (SD); range`
  )

# Print one table per HIV status, too busy having it all in one table
ae_cycle_summ |>
  group_by(hiv_status) |>
  group_walk(\(df, key) {
    cat("\n\n### HIV status:", as.character(key$hiv_status), "\n\n")
    df |>
      select(-hiv_status) |>
      kbl(booktabs = TRUE, align = "lccccccc") |>
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) |>
      print()
  }, .keep = TRUE)
```


<!-- # Model Building -->

<!-- ## Simple Logistic: Full treatment period -->
<!-- ### Simple logistic model of any adverse events by HIV status -->
<!-- ```{r} -->
<!-- #| label: model-any_simple -->

<!-- model_any_simple <- glm(ae_any_pt ~ hiv_status,  -->
<!--                         data = gic_models, family = binomial) -->

<!-- # summary(model_any_simple) -->
<!-- ``` -->
<!-- Invalid model due to having a 100% adverse event rate in the HIV-positive group. -->

<!-- ### Simple logistic model of any severe adverse events by HIV status -->
<!-- ```{r} -->
<!-- #| label: model-severe-simple -->


<!-- model_severe_simple <- glm(ae_severe_any_pt ~ hiv_status, -->
<!--                            data = gic_models, family = binomial) -->

<!-- summary(model_severe_simple) -->
<!-- ``` -->
<!-- ```{r} -->
<!-- #| label: tbl-severe-simple-or-clean -->
<!-- #| tbl-cap: "Association between HIV Status and Severe Adverse Events (Grades 3-4)" -->

<!-- tidy(model_severe_simple, conf.int = TRUE, exponentiate = TRUE) |> -->
<!--   filter(term != "(Intercept)") |> -->
<!--   # Clean up the name and prepare for the reference row -->
<!--   mutate( -->
<!--     Term = str_replace(term, "hiv_statusPositive", "HIV Positive"), -->
<!--     `OR (95% CI)` = paste0(round(estimate, 2), " (",  -->
<!--                            round(conf.low, 2), " - ",  -->
<!--                            round(conf.high, 2), ")"), -->
<!--     p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))), -->
<!--     std.error = round(std.error, 3) -->
<!--   ) |> -->
<!--   select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |> -->
<!--   # Add the reference row at the top -->
<!--   add_row(Term = "HIV Negative (Reference)",  -->
<!--           `OR (95% CI)` = "1.00",  -->
<!--           `Std. Error` = NA,  -->
<!--           `P-value` = NA,  -->
<!--           .before = 1) |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") -->
<!-- ``` -->



<!-- ## Multivariate Logistic: Full treatment period -->

<!-- ```{r} -->
<!-- #| label: tbl-stage-vs-hiv-status -->
<!-- #| tbl-cap: "Cancer stage vs HIV status" -->

<!-- table(Stage = gic_models$cancer_stage, HIV = gic_models$hiv_status) |>  -->
<!--   as.data.frame.matrix() |>  -->
<!--   kbl(booktabs = TRUE) |>  -->
<!--   kable_styling(full_width = FALSE, position = "left") |>  -->
<!--   add_header_above(c("Cancer Stage" = 1, "HIV Status" = 2)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- #| label: tbl-chemo-type-vs-hiv-status -->
<!-- #| tbl-cap: "Chemotherapy type vs HIV status" -->

<!-- table(Stage = gic_models$chemo_type, HIV = gic_models$hiv_status) |>  -->
<!--   as.data.frame.matrix() |>  -->
<!--   kbl(booktabs = TRUE) |>  -->
<!--   kable_styling(full_width = FALSE, position = "left") |>  -->
<!--   add_header_above(c("Chemo Type" = 1, "HIV Status" = 2)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- #| label: tbl-tx-intent-vs-hiv-status -->
<!-- #| tbl-cap: "Treatment intent vs HIV status" -->

<!-- table(Stage = gic_models$tx_intent, HIV = gic_models$hiv_status) |>  -->
<!--   as.data.frame.matrix() |>  -->
<!--   kbl(booktabs = TRUE) |>  -->
<!--   kable_styling(full_width = FALSE, position = "left") |>  -->
<!--   add_header_above(c("Treatment Intent" = 1, "HIV Status" = 2)) -->
<!-- ``` -->

<!-- ### Basic logistic model of any adverse events by HIV status -->
<!-- ```{r} -->
<!-- #| label: model-any-basic -->

<!-- model_any_basic <- glm(ae_any_pt ~ hiv_status + cancer_stage + chemo_type + tx_intent, -->
<!--                        data = gic_models, family = binomial) -->

<!-- summary(model_any_basic) -->
<!-- ``` -->
<!-- Invalid model due to having a 100% adverse event rate in the HIV-positive group. -->

<!-- ### Basic logicstic model of any severe adverse events by HIV status -->
<!-- ```{r} -->
<!-- #| label: model-severe-basic -->

<!-- model_severe_basic <- glm(ae_severe_any_pt ~ hiv_status + cancer_stage + chemo_type + tx_intent, -->
<!--                        data = gic_models, family = binomial) -->

<!-- summary(model_severe_basic) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- #| label: tbl-severe-multivariate-or -->
<!-- #| tbl-cap: "Adjusted Odds Ratios for Severe Adverse Events (Grades 3-4)" -->

<!-- tidy(model_severe_basic, conf.int = TRUE, exponentiate = TRUE) |> -->
<!--   filter(term != "(Intercept)") |> -->
<!--   mutate( -->
<!--     # Clean up term names -->
<!--     Term = case_when( -->
<!--       term == "hiv_statusPositive" ~ "HIV Positive", -->
<!--       term == "cancer_stageStage III" ~ "Stage III", -->
<!--       term == "cancer_stageStage IV" ~ "Stage IV", -->
<!--       term == "chemo_typePlatinum/5-FU" ~ "Platinum/5-FU", -->
<!--       term == "chemo_typeLCV/5-FU" ~ "LCV/5-FU", -->
<!--       term == "chemo_typeCapeOX" ~ "CapeOX", -->
<!--       term == "chemo_typeFOLFOX4" ~ "FOLFOX4", -->
<!--       term == "chemo_typeCapecitabine" ~ "Capecitabine", -->
<!--       term == "tx_intentPalliative" ~ "Palliative Intent", -->
<!--       TRUE ~ term -->
<!--     ), -->
<!--     `OR (95% CI)` = paste0(round(estimate, 2), " (",  -->
<!--                            round(conf.low, 2), " - ",  -->
<!--                            round(conf.high, 2), ")"), -->
<!--     p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))), -->
<!--     std.error = round(std.error, 3) -->
<!--   ) |> -->
<!--   select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |> -->
<!--   # Add reference rows for clarity -->
<!--   add_row(Term = "HIV Negative (Ref)", `OR (95% CI)` = "1.00", .before = 1) |> -->
<!--   add_row(Term = "Stage II (Ref)", `OR (95% CI)` = "1.00", .before = 3) |> -->
<!--   add_row(Term = "Platinum/Paclitaxel (Ref)", `OR (95% CI)` = "1.00", .before = 6) |> -->
<!--   add_row(Term = "Curative Intent (Ref)", `OR (95% CI)` = "1.00", .before = 12) |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") |> -->
<!--   pack_rows("HIV Status", 1, 2) |> -->
<!--   pack_rows("Cancer Stage", 3, 5) |> -->
<!--   pack_rows("Chemotherapy Regimen", 6, 11) |> -->
<!--   pack_rows("Treatment Intent", 12, 13) -->
<!-- ``` -->



<!-- ## $\chi^2$ Tests (HIV vs Early Events) -->

<!-- ```{r} -->
<!-- #| label: tbl-any-early-contingency -->
<!-- #| tbl-cap: "Contingency table of early adverse events (cycles 1-4) vs HIV status" -->

<!-- table_any_early <- table(early_ae = factor(gic_models$ae_any_early, levels = c(0, 1), labels = c("No", "Yes")), HIV = gic_models$hiv_status) -->

<!-- table_any_early |>  -->
<!--   as.data.frame.matrix() |>  -->
<!--   kbl(booktabs = TRUE) |>  -->
<!--   kable_styling(full_width = FALSE, position = "left") |>  -->
<!--   add_header_above(c("Any Early AE" = 1, "HIV Status" = 2)) -->
<!-- ``` -->
<!-- ```{r} -->
<!-- #| label: tbl-severe-early-contingency -->
<!-- #| tbl-cap: "Contingency table of severe early adverse events (cycles 1-4) vs HIV status" -->

<!-- table_severe_early <- table(severe_early_ae = factor(gic_models$ae_severe_any_early, levels = c(0, 1), labels = c("No", "Yes")), HIV = gic_models$hiv_status) -->

<!-- table_severe_early |>  -->
<!--   as.data.frame.matrix() |>  -->
<!--   kbl(booktabs = TRUE) |>  -->
<!--   kable_styling(full_width = FALSE, position = "left") |>  -->
<!--   add_header_above(c("Severe Early AE" = 1, "HIV Status" = 2)) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- #| label: tests-chi-sq -->

<!-- # Any Early AE (Cycles 1-4) -->
<!-- cat("Chi-Squared Test: Any Early AE (C1-4) vs HIV Status\n") -->
<!-- print(chisq.test(table_any_early)) -->

<!-- cat("\n\n") -->

<!-- # Severe Early AE (Cycles 1-4) -->
<!-- cat("Chi-Squared Test: Severe Early AE (C1-4) vs HIV Status\n") -->
<!-- print(chisq.test(table_severe_early)) -->
<!-- ``` -->





