---
title: "GI Cancer: Exploratory Data Analysis"
bibliography: references.bib
---

# Load Data & Libraries
```{r}
#| label: setup
#| warning: false

source("R/setup.R")

# Load Data
gic_validated <- readRDS(here("data", "processed", "gic_validated.rds"))

# Load Libraries
```

# Variable Engineering

```{r}
#| label: engineer-ae-variables

# Identify adverse event grade columns
grade_cols <- names(gic_validated)[str_detect(names(gic_validated), "_grade_([1-9]|1[0-2])$")]
### Need to determine whether baselines should be included too

# Create long table to engineer new variables
ae_long <- gic_validated |> 
  select(record_id, cycles_prescribed, cycles_given, all_of(grade_cols)) |> 
  pivot_longer(
    cols = all_of(grade_cols),
    names_to = "grade_var",
    values_to = "grade_raw"
  ) |> 
  mutate(
    cycle = as.integer(str_extract(grade_var, "\\d+$")),
    ae_type = str_remove(grade_var, "_grade_\\d+$"),
    grade = as.numeric(str_match(grade_raw, "(?i)^\\s*grade\\s*(\\d+)")[,2])
  )

# Cycle-level summaries
cycle_summary <- ae_long |>
  group_by(record_id, cycles_prescribed, cycles_given, cycle) |>
  summarize(
    ae_n_markers_measured = sum(!is.na(grade)),
    .groups = "drop"
  ) |>
  left_join(
    ae_long |>
      filter(!is.na(grade)) |>
      group_by(record_id, cycle) |>
      summarize(
        ae_n_events = sum(grade > 0),
        ae_n_severe = sum(grade >= 3),
        ae_max_grade = max(grade),
        ae_any = any(grade > 0),
        ae_severe_any = any(grade >= 3),
        .groups = "drop"
      ),
    by = c("record_id", "cycle")
  )

ae_cycle_wide <- cycle_summary |>
  pivot_wider(
    id_cols = c(record_id, cycles_prescribed, cycles_given),
    names_from = cycle,
    values_from = c(ae_n_markers_measured, ae_max_grade, ae_n_events, ae_n_severe, ae_any, ae_severe_any),
    names_glue = "{.value}_{cycle}"
  )


# Patient-level summaries
ae_summary_pt <- ae_long |> 
  group_by(record_id) |> 
  summarize(
    ae_any_early = as.integer(any(grade > 0 & cycle <= 2, na.rm = TRUE)),
    ae_severe_any_pt = as.integer(any(grade >= 3, na.rm = TRUE)),
    ae_severe_any_early = as.integer(any(grade >= 3 & cycle <= 2, na.rm = TRUE)),
    ae_any_pt = as.integer(any(grade > 0, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  left_join(
    cycle_summary |>
      filter(ae_n_markers_measured > 0) |>   # only cycles with any grading recorded
      group_by(record_id) |>
      summarize(
        ae_max_grade_pt = max(ae_max_grade, na.rm = TRUE),
        .groups = "drop"
      ),
    by = "record_id"
  )

# Merge datasets for analysis
gic_models <- gic_validated |>
  left_join(ae_summary_pt, by = "record_id") |>
  left_join(ae_cycle_wide, by = c("record_id", "cycles_prescribed", "cycles_given"))
```

## Summary of new adverse event (AE) variables

| Variable                  | Level   | Type         | Meaning                                                                                                          |
| ------------------------- | ------- | ------------ | ---------------------------------------------------------------------------------------------------------------- |
| `ae_any_pt`               | Patient | int (0/1)    | Flag: did the patient have any AE (`grade > 0`) at any point in treatment?                                       |
| `ae_any_early`            | Patient | int (0/1)    | Flag: did the patient have any AE (`grade > 0`) in cycles 1–2?                                                   |
| `ae_severe_any_pt`        | Patient | int (0/1)    | Flag: did the patient have any severe AE (`grade >= 3`) across all cycles?                                       |
| `ae_severe_any_early`     | Patient | int (0/1)    | Flag: did the patient have any severe AE (`grade >= 3`) in cycles 1–2?                                           |
| `ae_max_grade_pt`         | Patient | numeric / NA | Maximum adverse event grade across all treatment cycles                                                          |
| ------------------------- | ------- | ------------ | ---------------------------------------------------------------------------------------------------------------- |
| `ae_max_grade_#`          | Cycle   | numeric / NA | For cycle {#}, maximum AE grade across markers (NA if `ae_n_markers_measured_# == 0`).                           |
| `ae_n_events_#`           | Cycle   | numeric / NA | For cycle {#}, count of AEs across markers (`grade > 0`) (NA if `ae_n_markers_measured_# == 0`).                 |
| `ae_n_severe_#`           | Cycle   | numeric / NA | For cycle {#}, count of severe AEs across markers (`grade >= 3`) (NA if `ae_n_markers_measured_# == 0`).         |
| `ae_any_#`                | Cycle   | logical / NA | For cycle {#}, flag: any AE this cycle (`ae_n_events_# > 0`) (NA if `ae_n_markers_measured_# == 0`).             |
| `ae_severe_any_#`         | Cycle   | logical / NA | For cycle {#}, flag: any severe AE this cycle (`ae_n_severe_# > 0`) (NA if `ae_n_markers_measured_# == 0`).      |



## Patient-Level Adverse Event Variables
```{r}
#| label: tbl-ae-by-pt
#| tbl-cap: "Patient-level AE summaries by HIV status."

summ_n_pct <- function(x, digits = 1) {
  n <- sum(x == 1, na.rm = TRUE)
  pct <- 100 * n / sum(!is.na(x))
  paste0(n, " (", round(pct, digits), "%)")
}

# Function to calculate mean, standard deviation, and range
summ_mean_sd_range <- function(x, digits = 1) {
  x <- x[!is.na(x)]
  m <- mean(x); s <- sd(x)
  rng <- range(x)
  paste0(
    round(m, digits), " (", round(s, digits), "); ",
    round(rng[[1]], digits), "–", round(rng[[2]], digits)
  )
}

summ_mean_sd_iqr <- function(x, digits = 1) {
  x <- x[!is.na(x)]
  m <- mean(x); s <- sd(x)
  q <- quantile(x, probs = c(0.25, 0.75), na.rm = TRUE, names = FALSE)
  paste0(
    round(m, digits), " (", round(s, digits), "); ",
    round(q[1], digits), "–", round(q[2], digits)
  )
}

summ_median_iqr <- function(x, digits = 1) {
  x <- x[!is.na(x)]
  med <- median(x)
  q <- quantile(x, probs = c(0.25, 0.75), names = FALSE)
  paste0(round(med, digits), " (", round(q[1], digits), "–", round(q[2], digits), ")")
}

pt_summ <- gic_models |>
  group_by(hiv_status) |>
  summarize(
    total_pts = as.character(n()),

    any_ae = summ_n_pct(ae_any_pt),
    any_ae_early = summ_n_pct(ae_any_early),
    any_severe_ae = summ_n_pct(ae_severe_any_pt),
    any_severe_ae_early = summ_n_pct(ae_severe_any_early),

    max_grade_pt = summ_median_iqr(ae_max_grade_pt),

    .groups = "drop"
  )

pt_wide <- pt_summ |>
  rename(
    "Total patients (N)" = total_pts,
    "Any AE (%, patients)" = any_ae,
    "Any AE in cycles 1–2 (%, patients)" = any_ae_early,
    "Any severe AE (%, patients)" = any_severe_ae,
    "Any severe AE in cycles 1–2 (%, patients)" = any_severe_ae_early,
    "Max AE grade across all cycles, median (IQR)" = max_grade_pt
  ) |>
  pivot_longer(-hiv_status, names_to = "Metric", values_to = "value") |>
  pivot_wider(names_from = hiv_status, values_from = value)

pt_wide |>
  kbl(booktabs = TRUE, align = "lcc") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## Cycle-Level Adverse Event 

```{r}
#| label: tbl-ae-by-cycle
#| tbl-cap: "Cycle-level AE summaries by HIV status."
#| results: asis

ae_cycle_long <- gic_models |>
  select(
    record_id, hiv_status, cycles_prescribed, cycles_given,
    matches("^ae_n_markers_measured_\\d+$"),
    matches("^ae_n_events_\\d+$"),
    matches("^ae_n_severe_\\d+$"),
    matches("^ae_max_grade_\\d+$"),
    matches("^ae_any_\\d+$"),
    matches("^ae_severe_any_\\d+$")
  ) |>
  pivot_longer(
    cols = -c(record_id, hiv_status, cycles_prescribed, cycles_given),
    names_to = c("metric", "cycle"),
    names_pattern = "^(.*)_(\\d+)$",
    values_to = "value"
  ) |>
  mutate(cycle = as.integer(cycle)) |>
  pivot_wider(names_from = metric, values_from = value)

ae_cycle_summ <- ae_cycle_long |>
  group_by(hiv_status, cycle) |>
  summarize(
    measured_n = sum(ae_n_markers_measured > 0, na.rm = TRUE),
    total_pts = n(),

    `Patients (%)` = paste0(measured_n, " (", round(100 * measured_n / total_pts, 1), "%)"),

    `Any AE (%)` = {
      n <- sum(ae_any == TRUE, na.rm = TRUE)
      paste0(n, " (", round(100 * n / measured_n, 1), "%)")
    },
    `Any severe AE (%)` = {
      n <- sum(ae_severe_any == TRUE, na.rm = TRUE)
      paste0(n, " (", round(100 * n / measured_n, 1), "%)")
    },

    `# AEs, median (IQR)` = summ_median_iqr(ae_n_events),
    `# severe AEs, median (IQR)` = summ_median_iqr(ae_n_severe),
    `Max grade, median (IQR)` = summ_median_iqr(ae_max_grade, digits = 0),

    .groups = "drop"
  ) |>
  rename(Cycle = cycle) |>
  select(-measured_n, -total_pts)

# One table per HIV status
for (status in split(ae_cycle_summ, ae_cycle_summ$hiv_status)) {
  cat("\n\n### HIV status:", status$hiv_status[1], "\n\n")
  status |>
    select(-hiv_status) |>
    kbl(booktabs = TRUE, align = "lccccccc") |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) |>
    print()
}
```




# Plots for Adverse Event Metrics
## Plots of Patient-Level Adverse Event Metrics

```{r}
#| label: plt-ae-max-grade-pt
#| fig-cap: "Distribution of maximum AE grade by HIV status."

gic_models |>
  filter(!is.na(ae_max_grade_pt)) |>
  count(hiv_status, ae_max_grade_pt) |>
  group_by(hiv_status) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = ae_max_grade_pt, y = prop, fill = hiv_status)) +
  geom_col(position = "dodge") +
  scale_x_continuous(breaks = 0:5) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Maximum AE Grade per Patient by HIV Status",
    x = "Maximum CTCAE grade",
    y = "Proportion of patients",
    fill = "HIV status"
  ) +
  theme_minimal()
```


## Plots of Cycle-Level Adverse Event Metrics

```{r}
#| label: prep-ae-cycle-plot-data

# Build cycle-level dataset from gic_models for plotting
ae_cycle_measured <- ae_cycle_long |> 
  filter(ae_n_markers_measured > 0) |> 
  mutate(
    pct_complete = if_else(is.na(cycles_prescribed) | cycles_prescribed <= 0,
                           NA_real_,
                           cycle / cycles_prescribed),
    pct_bin = cut(
      pct_complete,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE,
      right = TRUE
    )
  )

summarize_metric <- function(df, var) {
  df |> 
    group_by(hiv_status, cycle) |> 
    summarize(
      n = sum(!is.na(.data[[var]])),
      median = median(.data[[var]], na.rm = TRUE),
      q25 = quantile(.data[[var]], 0.25, na.rm = TRUE),
      q75 = quantile(.data[[var]], 0.75, na.rm = TRUE),
      min = min(.data[[var]], na.rm = TRUE),
      max = max(.data[[var]], na.rm = TRUE),
      .groups = "drop"
    )
}

ae_events_summ <- summarize_metric(ae_cycle_measured, "ae_n_events")
ae_severe_summ <- summarize_metric(ae_cycle_measured, "ae_n_severe")
ae_maxgrade_summ <- summarize_metric(ae_cycle_measured, "ae_max_grade")
```


```{r}
#| label: plt-ae-n-events
#| fig-cap: "Number of adverse events per treatment cycle by HIV status (median with IQR and min–max bands; measured cycles only)"

ggplot(ae_events_summ, aes(x = cycle, group = hiv_status, color = hiv_status, fill = hiv_status)) +
  geom_ribbon(aes(ymin = min, ymax = max), alpha = 0.2, color = NA) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.5, color = NA) +
  geom_line(aes(y = median), linewidth = 1) +
  geom_point(aes(y = median), size = 2) +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Adverse events per cycle",
       subtitle = "Median (line) with IQR and min–max bands among cycles with any AE grading recorded",
       x = "Treatment cycle",
       y = "Number of AEs",
       color = "HIV status", fill = "HIV status") +
  theme_minimal()
```
Initially higher trend for number of adverse events amongst HIV positive patients, but the mean drops below that of HIV negative patients after cycle 4. Need to investigate, but could have to do with high early dropout rates, or possibly related to patients with HIV generally having fewer cycles of chemotherapy prescribed.


```{r}
#| label: plt-ae-n-severe
#| fig-cap: "Number of severe adverse events (grade ≥ 3) per treatment cycle by HIV status (median with IQR band; measured cycles only"

ggplot(ae_severe_summ, aes(x = cycle, group = hiv_status, color = hiv_status, fill = hiv_status)) +
  geom_ribbon(aes(ymin = min, ymax = max), alpha = 0.2, color = NA) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.5, color = NA) +
  geom_line(aes(y = median), linewidth = 1) +
  geom_point(aes(y = median), size = 2) +
  scale_x_continuous(breaks = 1:12) +
  labs(
    title = "Severe adverse events per cycle",
    subtitle = "Grade ≥ 3; median (line) with IQR and min–max bands among measured cycles",
    x = "Treatment cycle",
    y = "Number of severe AEs",
    color = "HIV status",
    fill = "HIV status"
  ) +
  theme_minimal()
```
Not a very noticable difference based on the medians or IQRs. The HIV negative patients tend to have higher max number of severe adverse events per cycle, but as noted before, I'd guess that relates to how few cycles HIV positive patients tend to complete, and the total number of HIV negative patients compared to positive.


```{r}
#| label: plt-ae-max-grade-cycle
#| fig-cap: "Median max AE grade per cycle by HIV status among measured cycles."

ggplot(ae_maxgrade_summ, aes(x = cycle, group = hiv_status, color = hiv_status, fill = hiv_status)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.3, color = NA) +
  geom_line(aes(y = median), linewidth = 1) +
  geom_point(aes(y = median), size = 2) +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(breaks = 0:5, limits = c(0, 5)) +
  labs(
    title = "Max AE Grade per Cycle",
    subtitle = "Median (line) with IQR band among measured cycles",
    x = "Treatment cycle",
    y = "Max AE grade",
    color = "HIV status",
    fill = "HIV status"
  ) +
  theme_minimal()
```
No easily discernible visual difference between patients with respect to HIV status and maximum adverse event experienced aside from period 7 onward as previously noted.

```{r}
#| label: plt-ae-severe-prop-cycle
#| fig-cap: "Proportion of measured patients experiencing any grade 3+ AE per cycle by HIV status."

severe_prop <- ae_cycle_measured |>
  group_by(hiv_status, cycle) |>
  summarize(
    n = n(),
    prop_severe = mean(ae_severe_any, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(severe_prop, aes(x = cycle, y = prop_severe, color = hiv_status)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
  labs(
    title = "Proportion with Severe AE (Grade 3+) per Cycle",
    x = "Treatment cycle",
    y = "Proportion with any grade 3+ AE",
    color = "HIV status"
  ) +
  theme_minimal()
```
Maybe a slightly higher proportion of HIV positive patients have severe AEs than HIV negative, need to do a test to compare over cycles 1-6. As noted before, in cycles 7 and onwards, there is only a single HIV positive patient still receiving treatment.


# Descriptive Statistics
```{r}
#| label: tbl-stage-vs-hiv-status
#| tbl-cap: "Cancer stage vs HIV status"

table(Stage = gic_models$cancer_stage, HIV = gic_models$hiv_status) |>
  as.data.frame.matrix() |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left") |>
  add_header_above(c("Cancer Stage" = 1, "HIV Status" = 2))
```

```{r}
#| label: tbl-chemo-type-by-hiv
#| tbl-cap: "Chemotherapy type by HIV status: patient counts, within-group proportions, and cycles prescribed"

chemo_hiv_summ <- gic_models |>
  group_by(hiv_status, chemo_type) |>
  summarize(
    n = n(),
    cycles_med_iqr = summ_median_iqr(cycles_prescribed),
    .groups = "drop"
  ) |>
  group_by(hiv_status) |>
  mutate(pct = round(100 * n / sum(n), 1),
         `n (%)` = paste0(n, " (", pct, "%)")) |>
  ungroup() |>
  select(hiv_status, chemo_type, `n (%)`, cycles_med_iqr)

neg <- chemo_hiv_summ |> filter(hiv_status == "Negative") |>
  select(chemo_type, `Neg: n (%)` = `n (%)`, `Neg: med (IQR)` = cycles_med_iqr)

pos <- chemo_hiv_summ |> filter(hiv_status == "Positive") |>
  select(chemo_type, `Pos: n (%)` = `n (%)`, `Pos: med (IQR)` = cycles_med_iqr)

# Fisher's exact test per chemo type for assignment proportions
chemo_levels <- levels(gic_models$chemo_type)

fisher_pvals <- map_dbl(chemo_levels, function(ct) {
  tab <- table(
    assigned = gic_models$chemo_type == ct,
    hiv      = gic_models$hiv_status
  )
  fisher.test(tab)$p.value
})

# Wilcoxon rank-sum test per chemo type for cycles prescribed
wilcox_pvals <- map_dbl(chemo_levels, function(ct) {
  dat <- gic_models |> filter(chemo_type == ct)
  if (n_distinct(dat$hiv_status) < 2) return(NA_real_)
  wilcox.test(cycles_prescribed ~ hiv_status, data = dat, exact = FALSE)$p.value
})

# Apply BH correction
fisher_padj <- p.adjust(fisher_pvals, method = "BH")
wilcox_padj <- p.adjust(wilcox_pvals, method = "BH")

format_p <- function(p) {
  if (is.na(p)) return("—")
  if (p < 0.001) return("<0.001")
  as.character(round(p, 3))
}

pval_df <- tibble(
  chemo_type         = chemo_levels,
  `p (asgmt)`   = map_chr(fisher_padj, format_p),
  `p (cycles)`       = map_chr(wilcox_padj, format_p)
)


neg |>
  left_join(pos,     by = "chemo_type") |>
  left_join(pval_df, by = "chemo_type") |>
  rename(`Chemotherapy type` = chemo_type) |>
  select(
    `Chemotherapy type`,
    `Neg: n (%)`, `Pos: n (%)`, `p (asgmt)`,
    `Neg: med (IQR)`, `Pos: med (IQR)`, `p (cycles)`
  ) |>
  gt() |>
  tab_spanner(
    label = "Chemo assignment",
    columns = c(`Neg: n (%)`, `Pos: n (%)`, `p (asgmt)`)
  ) |>
  tab_spanner(
    label = "Cycles prescribed",
    columns = c(`Neg: med (IQR)`, `Pos: med (IQR)`, `p (cycles)`)
  ) |>
  cols_align(align = "center", columns = -`Chemotherapy type`) |>
  cols_align(align = "left",   columns = `Chemotherapy type`) |> 
  tab_footnote(
    footnote = "Fisher's exact tests; Wilcoxon rank-sum tests",
    location = cells_column_labels(columns = c(`p (asgmt)`, `p (cycles)`))
    ) |> 
  opt_footnote_marks(marks = "extended")
```

```{r}
#| label: tbl-cancer-site-summary
#| tbl-cap: "Cancer site by HIV status and chemotherapy assignment"

cancer_levels <- levels(gic_models$cancer_site)
chemo_levels  <- levels(gic_models$chemo_type)

# HIV status distribution by cancer site
hiv_summ <- gic_models |>
  group_by(cancer_site, hiv_status) |>
  summarize(n = n(), .groups = "drop") |>
  group_by(cancer_site) |>
  mutate(pct = round(100 * n / sum(n), 1),
         `n (%)` = paste0(n, " (", pct, "%)")) |>
  ungroup() |>
  select(cancer_site, hiv_status, `n (%)`) |>
  pivot_wider(names_from = hiv_status, values_from = `n (%)`,
              names_prefix = "HIV ") |>
  rename(`Cancer site` = cancer_site)

# Fisher's exact test per cancer site for HIV status distribution
fisher_site_pvals <- map_dbl(cancer_levels, function(cs) {
  tab <- table(
    hiv  = gic_models$hiv_status,
    site = gic_models$cancer_site == cs
  )
  if (any(tab < 5)) return(NA_real_)
  fisher.test(tab)$p.value
})
fisher_site_padj <- p.adjust(fisher_site_pvals, method = "BH")

hiv_summ <- hiv_summ |>
  mutate(`p (HIV)` = map_chr(fisher_site_padj, format_p))

# Chemo assignment by cancer site
chemo_summ <- gic_models |>
  group_by(cancer_site, chemo_type) |>
  summarize(n = n(), .groups = "drop") |>
  group_by(cancer_site) |>
  mutate(pct = round(100 * n / sum(n), 1),
         `n (%)` = paste0(n, " (", pct, "%)")) |>
  ungroup() |>
  select(cancer_site, chemo_type, `n (%)`) |>
  pivot_wider(names_from = chemo_type, values_from = `n (%)`,
              values_fill = "0 (0%)") |>
  rename(`Cancer site` = cancer_site)

# Join
hiv_summ |>
  left_join(chemo_summ, by = "Cancer site") |>
  gt() |>
  tab_spanner(
    label = "HIV status",
    columns = c(`HIV Negative`, `HIV Positive`, `p (HIV)`)
  ) |>
  tab_spanner(
    label = "Chemotherapy assignment",
    columns = all_of(chemo_levels)
  ) |>
  cols_align(align = "center", columns = -`Cancer site`) |>
  cols_align(align = "left",   columns = `Cancer site`) |>
  tab_footnote(
  footnote = "BH adjusted Fisher's exact test, only applied where all cells n ≥ 5.",
  location = cells_column_labels(columns = `p (HIV)`)
) |>
  opt_footnote_marks(marks = "extended")
```




```{r}
#| label: tbl-tx-intent-vs-hiv-status
#| tbl-cap: "Treatment intent vs HIV status"

table(Intent = gic_models$tx_intent, HIV = gic_models$hiv_status) |>
  as.data.frame.matrix() |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left") |>
  add_header_above(c("Treatment Intent" = 1, "HIV Status" = 2))
```

```{r}
#| label: tbl-cycles-by-hiv
#| tbl-cap: "Chemotherapy cycles prescribed and received by HIV status"

cycles_summ <- gic_models |>
  mutate(
    completion_rate = if_else(
      is.na(cycles_prescribed) | cycles_prescribed <= 0,
      NA_real_,
      cycles_given / cycles_prescribed
    )
  ) |>
  group_by(hiv_status) |>
  summarize(
    n_patients = n(),
    cycles_prescribed = summ_mean_sd_iqr(cycles_prescribed, digits = 1),
    cycles_given = summ_mean_sd_iqr(cycles_given, digits = 1),
    completion_rate = summ_mean_sd_iqr(completion_rate, digits = 2),
    .groups = "drop"
  ) |>
  mutate(
    `HIV Status` = hiv_status,
    `Patients (N)` = n_patients,
    `Cycles prescribed, mean (SD); IQR` = cycles_prescribed,
    `Cycles given, mean (SD); IQR` = cycles_given,
    `Completion rate, mean (SD); IQR` = completion_rate
  ) |>
  select(
    `HIV Status`,
    `Patients (N)`,
    `Cycles prescribed, mean (SD); IQR`,
    `Cycles given, mean (SD); IQR`,
    `Completion rate, mean (SD); IQR`
  )

cycles_summ |>
  kbl(booktabs = TRUE, align = "lcccc") |>
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
#| label: tbl-severe-early-contingency
#| tbl-cap: "Contingency table of severe early adverse events (cycles 1-2) vs HIV status"

table_severe_early <- table(severe_early_ae = factor(gic_models$ae_severe_any_early, levels = c(0, 1), labels = c("No", "Yes")), HIV = gic_models$hiv_status)

table_severe_early |>
  as.data.frame.matrix() |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left") |>
  add_header_above(c("Severe Early AE" = 1, "HIV Status" = 2))
```


```{r}
#| label: plt-density-cycles-rx

gic_validated |>
  count(hiv_status, cycles_prescribed) |>
  group_by(hiv_status) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = cycles_prescribed, y = prop, fill = hiv_status)) +
  geom_col(position = "dodge") +
  scale_x_continuous(name = "Cycles prescribed", breaks = 1:12) +
  ylab("Proportion") + labs(title = "Chemo Cycles Prescribed by HIV Status")
```

```{r}
#| label: plt-cycle-completion

survival_data <- gic_validated |>
  # for each patient, create a row for each cycle they were prescribed
  rowwise() |>
  mutate(cycle = list(1:cycles_prescribed)) |>
  unnest(cycle) |>
  # mark whether they actually received each cycle
  mutate(received = cycle <= cycles_given) |>
  # for each HIV status and cycle number, get the proportion who received it
  group_by(hiv_status, cycle) |>
  summarise(
    n_prescribed = n(),
    prop_received = mean(received),
    .groups = "drop"
  )

# main plot (store it)
p_main <- ggplot(survival_data, aes(x = cycle, y = prop_received, color = hiv_status)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "Chemotherapy Cycle Completion by HIV Status",
    x = NULL,
    y = "Proportion receiving cycle\n(among those prescribed at least that many)",
    color = "HIV status"
  ) +
  theme(plot.margin = margin(5, 5, 0, 5))

# table underneath
p_table <- ggplot(survival_data, aes(x = cycle, y = hiv_status, label = n_prescribed)) +
  geom_text(size = 3) +
  scale_x_continuous(breaks = 1:12, limits = range(survival_data$cycle)) +
  labs(x = NULL, y = NULL, subtitle = "Patients prescribed at least k cycles") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_text(face = "bold"),
    axis.text.x = element_blank(),
    plot.margin = margin(0, 5, 5, 5)
  )

# combine
p_main / p_table + plot_layout(heights = c(4, 1))
```



# Model Building
## Any (Severe) Adverse Events as Outcomes
### Simple Logistic: Full treatment period
#### Simple logistic model of any adverse events by HIV status
```{r}
#| label: model-any-simple

model_any_simple <- glm(ae_any_pt ~ hiv_status,
                        data = gic_models, family = binomial)

# summary(model_any_simple)
```
Invalid model due to having a 100% adverse event rate in the HIV-positive group.

#### Simple logistic model of any severe adverse events by HIV status
```{r}
#| label: model-severe-simple


model_severe_simple <- glm(ae_severe_any_pt ~ hiv_status,
                           data = gic_models, family = binomial)

summary(model_severe_simple)
```

```{r}
#| label: tbl-severe-simple-or-clean
#| tbl-cap: "Association between HIV Status and Severe Adverse Events (Grades 3-4)"

tidy(model_severe_simple, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  # Clean up the name and prepare for the reference row
  mutate(
    Term = str_replace(term, "hiv_statusPositive", "HIV Positive"),
    `OR (95% CI)` = paste0(round(estimate, 2), " (",
                           round(conf.low, 2), " - ",
                           round(conf.high, 2), ")"),
    p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))),
    std.error = round(std.error, 3)
  ) |>
  select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |>
  # Add the reference row at the top
  add_row(Term = "HIV Negative (Reference)",
          `OR (95% CI)` = "1.00",
          `Std. Error` = NA,
          `P-value` = NA,
          .before = 1) |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left")
```



### Multivariate Logistic: Full treatment period
#### Basic logistic model of any adverse events by HIV status
```{r}
#| label: model-any-basic

model_any_basic <- glm(ae_any_pt ~ hiv_status + cancer_stage + chemo_type + tx_intent,
                       data = gic_models, family = binomial)

summary(model_any_basic)
```
Invalid model due to having a 100% adverse event rate in the HIV-positive group.

#### Basic logicstic model of any severe adverse events by HIV status
```{r}
#| label: model-severe-basic

model_severe_basic <- glm(ae_severe_any_pt ~ hiv_status + cancer_stage + chemo_type + tx_intent,
                       data = gic_models, family = binomial)

summary(model_severe_basic)
```

```{r}
#| label: tbl-severe-multivariate-or
#| tbl-cap: "Adjusted Odds Ratios for Severe Adverse Events (Grades 3-4)"

model_severe_basic |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Regimen",
      tx_intent    ~ "Treatment Intent"
    ),
    pvalue_fun = label_style_pvalue(
      digits = 3
    ),
    add_estimate_to_reference_rows = TRUE
  ) |>
  bold_labels() |> 
  italicize_levels() |> 
  bold_p() |> 
  add_global_p(keep = TRUE) |> 
  add_glance_source_note(include = c(nobs, AIC))
```


### $\chi^2$ Tests (HIV vs Early Events)

```{r}
#| label: tbl-any-early-contingency
#| tbl-cap: "Contingency table of early adverse events (cycles 1-2) vs HIV status"

table_any_early <- table(early_ae = factor(gic_models$ae_any_early, levels = c(0, 1), labels = c("No", "Yes")), HIV = gic_models$hiv_status)

table_any_early |>
  as.data.frame.matrix() |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left") |>
  add_header_above(c("Any Early AE" = 1, "HIV Status" = 2))
```


<!-- ```{r} -->
<!-- #| label: tests-chi-sq -->

<!-- # Any Early AE (Cycles 1-2) -->
<!-- cat("Chi-Squared Test: Any Early AE (C1-2) vs HIV Status\n") -->
<!-- print(chisq.test(table_any_early)) -->

<!-- cat("\n\n") -->

<!-- # Severe Early AE (Cycles 1-2) -->
<!-- cat("Chi-Squared Test: Severe Early AE (C1-2) vs HIV Status\n") -->
<!-- print(chisq.test(table_severe_early)) -->
<!-- ``` -->

## HIV as Outcome
### Simple Logistic:
```{r}
#|label: model-hiv-simple

model_hiv_simple <- glm(hiv_status ~ ae_max_grade_pt, data = gic_models, family = binomial)

summary(model_hiv_simple)

model_hiv_simple |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      ae_max_grade_pt ~ "Max CTCAE grade"
    ),
    pvalue_fun = label_style_pvalue(
      digits = 3
    ),
    add_estimate_to_reference_rows = TRUE
  ) |>
  bold_labels()
```


<!-- ```{r} -->
<!-- #| label: tbl-hiv-simple-or -->
<!-- #| tbl-cap: "Logistic regression: HIV status vs mean max AE grade (unadjusted)" -->

<!-- tidy(model_hiv_simple, conf.int = TRUE, exponentiate = TRUE) |> -->
<!--   filter(term != "(Intercept)") |> -->
<!--   mutate( -->
<!--     Term = case_when( -->
<!--       term == "ae_mean_max_grade" ~ "Mean max AE grade per measured cycle", -->
<!--       TRUE ~ term -->
<!--     ), -->
<!--     `OR (95% CI)` = paste0( -->
<!--       round(estimate, 2), " (", -->
<!--       round(conf.low, 2), " - ", -->
<!--       round(conf.high, 2), ")" -->
<!--     ), -->
<!--     p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))), -->
<!--     std.error = round(std.error, 3) -->
<!--   ) |> -->
<!--   select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") -->
<!-- ``` -->




### Basic Logistic
```{r}
#| label: model-hiv-basic

model_hiv_basic <- glm(
  hiv_status ~ ae_max_grade_pt + cancer_stage + chemo_type + tx_intent,
  data = gic_models,
  family = binomial
)

summary(model_hiv_basic)
```

<!-- ```{r} -->
<!-- #| label: tbl-hiv-basic-or -->
<!-- #| tbl-cap: "Logistic regression: HIV status vs mean max AE grade (adjusted for stage, chemo type, and intent)" -->

<!-- tidy(model_hiv_basic, conf.int = TRUE, exponentiate = TRUE) |> -->
<!--   filter(term != "(Intercept)") |> -->
<!--   mutate( -->
<!--     Term = case_when( -->
<!--       term == "ae_mean_max_grade" ~ "Mean max AE grade per measured cycle", -->
<!--       TRUE ~ term -->
<!--     ), -->
<!--     `OR (95% CI)` = paste0( -->
<!--       round(estimate, 2), " (", -->
<!--       round(conf.low, 2), " - ", -->
<!--       round(conf.high, 2), ")" -->
<!--     ), -->
<!--     p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))), -->
<!--     std.error = round(std.error, 3) -->
<!--   ) |> -->
<!--   select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") -->
<!-- ``` -->
```{r}
#| label: tbl-hiv-or-comparison
#| tbl-cap: "Logistic regression: HIV status — unadjusted vs adjusted odds ratios"

tbl_simple <- model_hiv_simple |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(ae_max_grade_pt ~ "Max AE grade (patient)")
  )

tbl_basic <- model_hiv_basic |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      ae_max_grade_pt ~ "Max AE grade (patient)",
      cancer_stage ~ "Cancer stage",
      chemo_type ~ "Chemo type",
      tx_intent ~ "Treatment intent"
    )
  )

tbl_merge(
  list(tbl_simple, tbl_basic),
  tab_spanner = c("Unadjusted", "Adjusted")
)
```

**Summary:** Both the unadjusted and adjusted models with HIV status as outcome indicate that each one unit increase in max AE grade across all treatment cycles is associated with between 28-34% higher odds of the patient being HIV positive. In the adjusted model, higher cancer stages were not reliably associated with being HIV positive. For chemo type there does seem to be a pretty clear difference in treatment options associated with HIV positive status. There does not appear to be any association between treatment intent and HIV status.

### Assumption checks

```{r}
#| label: chk-hiv-assumptions

# Linearity of the log-odds
gic_models |>
  filter(!is.na(ae_max_grade_pt)) |>
  group_by(ae_max_grade_pt) |>
  summarize(prop_hiv = mean(hiv_status == "Positive", na.rm = TRUE)) |>
  mutate(log_odds = log(prop_hiv / (1 - prop_hiv))) |>
  ggplot(aes(x = ae_max_grade_pt, y = log_odds)) +
  geom_point() +
  geom_line() +
  labs(x = "Max AE grade (patient)", y = "Empirical log-odds of HIV+")

print("Linearity isn't amazing, but a straight line fit would show a reasonable fit.")

# Multicollinearity
vif(model_hiv_basic)

print("No significant multicollinearity noted.")

# Influential observations
plot(model_hiv_basic, which = 4)  # Cook's distance
print("Three influential observations, need to investigate further")
```

```{r}
#| label: investigate-influential-obs

# Get row indices of most influential observations (Cook's D > threshold)
# Using 4/n as a common rule-of-thumb cutoff
n <- nrow(model_hiv_basic$model)
cooks_threshold <- 4 / n

influential_idx <- which(cooks.distance(model_hiv_basic) > cooks_threshold)
influential_ids <- gic_models$record_id[influential_idx]

cat("Influential observations (Cook's D > 4/n =", round(cooks_threshold, 4), "):\n")
cat("Row indices:", influential_idx, "\n")
cat("Record IDs:", influential_ids, "\n\n")

# Pull their covariate values to see what makes them unusual
gic_models |>
  filter(record_id %in% influential_ids) |>
  select(record_id, hiv_status, cancer_stage, chemo_type, tx_intent, ae_max_grade_pt)
```

All pts with a large Cook's D are HIV positive, but I haven't immediately spotted any other commonality amongst them to explain the group collectively being so influential, will have to do more investigation.



## Cycles Prescribed as Outcome
```{r}
#| label: model-cycles-full

# Full cycles prescribed model
model_cycles_full <- glm(
  cycles_prescribed ~ hiv_status + cancer_site + cancer_stage + chemo_type + tx_intent,
  data   = gic_models,
  family = poisson
)
```
```{r}
#| label: model-cycles-no-site

# Cycles prescribed model without cancer site
model_cycles_no_site <- glm(
  cycles_prescribed ~ hiv_status + cancer_stage + chemo_type + tx_intent,
  data   = gic_models,
  family = poisson
)
```
```{r}
#| label: tbl-cycles-models
#| tbl-cap: "Poisson regression: cycles prescribed (exponentiated coefficients = rate ratios)"

tbl_full <- model_cycles_full |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      cancer_site  ~ "Cancer Site",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Type",
      tx_intent    ~ "Treatment Intent"
    )
  ) |>
  bold_labels() |> 
  italicize_levels() |> 
  bold_p() |> 
  add_global_p(keep = TRUE) |> 
  add_vif(statistic = "aGVIF")

tbl_no_site <- model_cycles_no_site |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Type",
      tx_intent    ~ "Treatment Intent"
    )
  ) |>
  bold_labels() |> 
  italicize_levels() |> 
  bold_p() |> 
  add_global_p(keep = TRUE) |> 
  add_vif(statistic = "aGVIF")

lrt_p <- anova(model_cycles_no_site, model_cycles_full, test = "LRT")$`Pr(>Chi)`[2]


tbl_merge(
  list(tbl_no_site, tbl_full),
  tab_spanner = c("Without cancer site", "With cancer site")
) |>
  as_gt() |>
  tab_source_note(
    source_note = paste0(
      "Model without cancer site: AIC = ", round(AIC(model_cycles_no_site), 1),
      ", N = ", nobs(model_cycles_no_site),
      " | Model with cancer site: AIC = ", round(AIC(model_cycles_full), 1),
      ", N = ", nobs(model_cycles_full)
    )
  ) |>
  tab_source_note(
    source_note = paste0(
      "LRT (with vs without cancer site): p = ",
      round(anova(model_cycles_no_site, model_cycles_full, 
                  test = "LRT")$`Pr(>Chi)`[2], 3)
    )
  )
  
```


# Exploring Hospital Stays Due to Toxicity by HIV Status
```{r}
#| label: prep-hosp-days

hosp_cols <- names(gic_models)[str_detect(names(gic_models), "^hosp_days_([1-9]|1[0-2])$")]

hosp_long <- gic_models |> 
  select(record_id, hiv_status, cycles_given, all_of(hosp_cols)) |> 
  pivot_longer(
    cols = all_of(hosp_cols),
    names_to = "cycle_var",
    values_to = "hosp_days"
  ) |> 
  mutate(
    cycle = as.integer(str_extract(cycle_var, "\\d+"))
  )

# Patient-level summaries
hosp_pt <- hosp_long |> 
  group_by(record_id, hiv_status, cycles_given) |> 
  summarize(
    hosp_days_total = sum(hosp_days, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  mutate(
    hosp_days_per_cycle_given = if_else(is.na(cycles_given) | cycles_given <= 0,
                                        NA_real_,
                                        hosp_days_total / cycles_given)
  )

```

## Table of Toxicity-Related Hospitalization by HIV status
```{r}
#| label: tbl-hosp-by-pt
#| tbl-cap: "Hospital days attributable to toxicity by HIV status"

hosp_pt |>
  group_by(hiv_status) |>
  summarize(
    n_patients = n(),

    hosp_days_total_mean_sd = summ_median_iqr(hosp_days_total, 1),
    hosp_days_per_cycle_given_mean_sd = summ_median_iqr(hosp_days_per_cycle_given, 1),
    .groups = "drop"
  ) |>
  mutate(
    `HIV Status` = hiv_status,
    `Patients (N)` = n_patients,
    `Total hosp days, mean (SD)` = hosp_days_total_mean_sd,
    `Hosp days / cycles_given, mean (SD)` = hosp_days_per_cycle_given_mean_sd
  ) |>
  select(
    `HIV Status`, `Patients (N)`,
    `Total hosp days, mean (SD)`,
    `Hosp days / cycles_given, mean (SD)`
  ) |>
  kbl(booktabs = TRUE, align = "lcccc") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
#| label: prep-hosp-cycle-data

# Identify cycle-level column groups needed for hospitalization analysis
severe_cols <- names(gic_models)[str_detect(names(gic_models), "^ae_n_severe_([1-9]|1[0-2])$")]
max_grade_cols <- names(gic_models)[str_detect(names(gic_models), "^ae_max_grade_([1-9]|1[0-2])$")]
measured_cols <- names(gic_models)[str_detect(names(gic_models), "^ae_n_markers_measured_([1-9]|1[0-2])$")]

# Pivot each set of cycle cols to long
pivot_cycle_cols <- function(df, cols, value_name) {
  df |>
    select(record_id, all_of(cols)) |>
    pivot_longer(cols = all_of(cols), names_to = "cycle_var", values_to = value_name) |>
    mutate(cycle = as.integer(str_extract(cycle_var, "\\d+$"))) |>
    select(-cycle_var)
}

severe_long <- pivot_cycle_cols(gic_models, severe_cols,    "ae_n_severe")
max_grade_long <- pivot_cycle_cols(gic_models, max_grade_cols, "ae_max_grade")
measured_long <- pivot_cycle_cols(gic_models, measured_cols,  "ae_n_markers_measured")

# Join AE variables to hospitalization data
hosp_severe_cycle <- hosp_long |>
  left_join(severe_long,   by = c("record_id", "cycle")) |>
  left_join(measured_long, by = c("record_id", "cycle")) |>
  mutate(ae_measured = ae_n_markers_measured > 0)

hosp_maxgrade_cycle <- hosp_long |>
  left_join(max_grade_long, by = c("record_id", "cycle")) |>
  left_join(measured_long,  by = c("record_id", "cycle")) |>
  mutate(ae_measured = ae_n_markers_measured > 0)
```

## Toxicity-Related Hospitalization Based on Number of Severe Adverse Events
```{r}
#| label: tbl-hosp-vs-severe

hosp_severe_cycle |> 
  filter(ae_measured) |> 
  group_by(hiv_status, ae_n_severe) |> 
  summarize(
    n_cycles = sum(!is.na(hosp_days)),
    hosp_days_mean_sd = summ_median_iqr(hosp_days),
    .groups = "drop"
  ) |> 
  mutate(
    `HIV Status` = hiv_status,
    `# severe AEs in cycle` = ae_n_severe,
    `# Cycles` = n_cycles,
    `Hosp days, mean (SD); range` = hosp_days_mean_sd
  ) |> 
  select(`HIV Status`, `# severe AEs in cycle`, `# Cycles`, `Hosp days, mean (SD); range`) |> 
  kbl(booktabs = TRUE, align = "lccc") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
#| label: plt-hosp-vs-severe
#| fig-cap: "Mean hospital days vs number of severe AEs in the cycle, by HIV status"

hosp_severe_summ <- hosp_severe_cycle |>
  filter(ae_measured) |>
  group_by(hiv_status, ae_n_severe) |>
  summarize(
    mean_hosp_days = mean(hosp_days, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(hosp_severe_summ, aes(x = ae_n_severe, y = mean_hosp_days, color = hiv_status, group = hiv_status)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    x = "Number of severe AEs in cycle",
    y = "Mean hospital days (toxicity) in cycle",
    color = "HIV status"
  ) +
  theme_minimal()
```

## Toxicity-Related Hospitalization Based on Max Adverse Event Grade
```{r}
#| label: tbl-hosp-vs-max-grade
#| tbl-cap: "Hospital days vs max AE grade in cycle, by HIV status (among cycles with any AE grading recorded)"

hosp_maxgrade_cycle |>
  filter(ae_measured) |>
  group_by(hiv_status, ae_max_grade) |>
  summarize(
    n_cycles = sum(!is.na(hosp_days)),
    hosp_days_median_iqr = summ_median_iqr(hosp_days, 1),
    .groups = "drop"
  ) |>
  mutate(
    `HIV Status` = hiv_status,
    `Max AE grade in cycle` = ae_max_grade,
    `# Cycles` = n_cycles,
    `Hosp days, median (IQR)` = hosp_days_median_iqr
  ) |>
  select(`HIV Status`, `Max AE grade in cycle`, `# Cycles`, `Hosp days, median (IQR)`) |>
  kbl(booktabs = TRUE, align = "lccc") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```




```{r}
#| label: plt-hosp-vs-max-grade
#| fig-cap: "Mean hospital days vs max AE grade in cycle, by HIV status (among measured cycles)"

hosp_maxgrade_summ <- hosp_maxgrade_cycle |>
  filter(ae_measured) |>
  group_by(hiv_status, ae_max_grade) |>
  summarize(
    mean_hosp_days = mean(hosp_days, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(hosp_maxgrade_summ, aes(x = ae_max_grade, y = mean_hosp_days, color = hiv_status, group = hiv_status)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 0:5) +
  labs(
    x = "Max AE grade in cycle",
    y = "Mean hospital days (toxicity) in cycle",
    color = "HIV status"
  ) +
  theme_minimal()
```

## Hospitalization regression


```{r}
#| label: prep-hosp-regression


hosp_ae_cycle <- hosp_long |>
  left_join(
    ae_cycle_long |> select(record_id, cycle, ae_n_severe, ae_max_grade, ae_n_markers_measured),
    by = c("record_id", "cycle")
  ) |>
  mutate(ae_measured = ae_n_markers_measured > 0)


hosp_model_data <- hosp_ae_cycle |>
  filter(ae_measured)

hosp_model_data <- hosp_model_data |>
  mutate(any_hosp = as.integer(hosp_days > 0))
```

```{r}
#| label: model-hosp-any

# Any hospitalization this cycle (logistic)
model_any_hosp <- glmer(
  any_hosp ~ ae_max_grade + hiv_status + ae_max_grade:hiv_status + (1 | record_id),
  data = hosp_model_data,
  family = binomial
)
summary(model_any_hosp)
fixef(model_any_hosp) |> exp()
confint(model_any_hosp, method = "Wald") |> exp()
```
Each increase in max AE grade in a given cycle is associated with about a 5.5-fold increase (95% CI: 3.7-8.2) in the odds of being hospitalized. For patients with no adverse events, being HIV positive is associated with about a 3.4-fold increase in odds of being hospitalized, but the effect of being HIV positive on being hospitalized decreases by about 35% for each unit increase in max AE grade.



```{r}
# Length of stay given hospitalization
model_hosp_days <- lmer(
  hosp_days ~ ae_max_grade + hiv_status + ae_max_grade:hiv_status + (1 | record_id),
  data = hosp_model_data |> filter(hosp_days > 0)
)
summary(model_hosp_days)
```
AE severity has minimal impact on length of stay for either group once someone is already hospitalized, although HIV positive patients are expected to spend slightly longer in the hospital than HIV negative patients for each unit increase in max adverse event experienced.





