---
title: "GI Cancer: Exploratory Data Analysis"
bibliography: references.bib
---

# Load Data & Libraries
```{r}
#| label: setup
#| include: false

source("R/setup.R")

# Load Data
gic_validated <- readRDS(here("data", "processed", "gic_validated.rds"))

# Load Libraries
```


```{r}
#| label: engineer-ae-variables

# Identify all adverse event columns and pivot to check per-cycle status
ae_data <- gic_validated |> 
  select(record_id, matches("grade_\\d+$")) |> 
  pivot_longer(
    cols = -record_id,
    names_to = c("ae_type", "cycle"),
    names_pattern = "(.*)_grade_(\\d+)$",
    values_to = "grade"
  ) |> 
  mutate(
    cycle = as.integer(cycle),
    grade_num = as.numeric(str_extract(grade, "\\d")),
    is_ae = !is.na(grade_num) & grade_num > 0,
    is_severe = !is.na(grade_num) & grade_num >= 3
  )

# Summarize at patient level
ae_summary <- ae_data |> 
  group_by(record_id) |> 
  summarize(
    ae_cycle_count = n_distinct(cycle[is_ae == TRUE]),
    ae_early = if_else(any(is_ae == TRUE & cycle <= 4), 1, 0),
    had_severe_ae = if_else(any(is_severe == TRUE), 1, 0),
    severe_ae_early = if_else(any(is_severe == TRUE & cycle <= 4), 1, 0),
    .groups = "drop"
  )

# Join back to main dataset
gic_models <- gic_validated |> 
  left_join(ae_summary, by = "record_id") |> 
  mutate(
    had_any_ae = if_else(ae_cycle_count > 0, 1, 0)
  )
```

```{r}
#| label: tbl-ae-summary
#| tbl-cap: "Summary of adverse events by HIV status"

gic_models |> 
  group_by(hiv_status) |> 
  summarize(
    total_pts = n(),
    
    any_ae_n = sum(had_any_ae),
    any_ae_pct = (any_ae_n / total_pts) * 100,
    
    severe_ae_n = sum(had_severe_ae),
    severe_ae_pct = (severe_ae_n / total_pts) * 100,
    
    early_ae_n = sum(ae_early),
    early_ae_pct = (early_ae_n / total_pts) * 100,
    
    severe_early_ae_n = sum(severe_ae_early),
    severe_early_ae_pct = (severe_early_ae_n / total_pts) * 100,
    .groups = "drop"
  ) |> 
  mutate(
    "HIV Status" = hiv_status,
    "Total Patients" = total_pts,
    "Any AE (%)" = paste0(any_ae_n, " (", round(any_ae_pct, 2), "%)"),
    "Severe AE (%)" = paste0(severe_ae_n, " (", round(severe_ae_pct, 2), "%)"),
    "Early AE (%)" = paste0(early_ae_n, " (", round(early_ae_pct, 2), "%)"),
    "Severe Early AE (%)" = paste0(severe_early_ae_n, " (", round(severe_early_ae_pct, 2), "%)")
  ) |> 
  select("HIV Status", "Total Patients", "Any AE (%)", "Severe AE (%)", "Early AE (%)", "Severe Early AE (%)") |> 
  kbl(booktabs = TRUE, align = "lcccc") |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

# Model Building

## Simple Logistic: Full treatment period
### Simple logistic model of any adverse events by HIV status
```{r}
#| label: model-any_simple

model_any_simple <- glm(had_any_ae ~ hiv_status, 
                        data = gic_models, family = binomial)

# summary(model_any_simple)
```
Invalid model due to having a 100% adverse event rate in the HIV-positive group.

### Simple logistic model of any severe adverse events by HIV status
```{r}
#| label: model-severe-simple


model_severe_simple <- glm(had_severe_ae ~ hiv_status,
                           data = gic_models, family = binomial)

summary(model_severe_simple)
```
```{r}
#| label: tbl-severe-simple-or-clean
#| tbl-cap: "Association between HIV Status and Severe Adverse Events (Grades 3-4)"

tidy(model_severe_simple, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  # Clean up the name and prepare for the reference row
  mutate(
    Term = str_replace(term, "hiv_statusPositive", "HIV Positive"),
    `OR (95% CI)` = paste0(round(estimate, 2), " (", 
                           round(conf.low, 2), " - ", 
                           round(conf.high, 2), ")"),
    p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))),
    std.error = round(std.error, 3)
  ) |>
  select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |>
  # Add the reference row at the top
  add_row(Term = "HIV Negative (Reference)", 
          `OR (95% CI)` = "1.00", 
          `Std. Error` = NA, 
          `P-value` = NA, 
          .before = 1) |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left")
```



## Multivariate Logistic: Full treatment period

```{r}
#| label: tbl-stage-vs-hiv-status
#| tbl-cap: "Cancer stage vs HIV status"

table(Stage = gic_models$cancer_stage, HIV = gic_models$hiv_status) |> 
  as.data.frame.matrix() |> 
  kbl(booktabs = TRUE) |> 
  kable_styling(full_width = FALSE, position = "left") |> 
  add_header_above(c("Cancer Stage" = 1, "HIV Status" = 2))
```

```{r}
#| label: tbl-chemo-type-vs-hiv-status
#| tbl-cap: "Chemotherapy type vs HIV status"

table(Stage = gic_models$chemo_type, HIV = gic_models$hiv_status) |> 
  as.data.frame.matrix() |> 
  kbl(booktabs = TRUE) |> 
  kable_styling(full_width = FALSE, position = "left") |> 
  add_header_above(c("Chemo Type" = 1, "HIV Status" = 2))
```

```{r}
#| label: tbl-tx-intent-vs-hiv-status
#| tbl-cap: "Treatment intent vs HIV status"

table(Stage = gic_models$tx_intent, HIV = gic_models$hiv_status) |> 
  as.data.frame.matrix() |> 
  kbl(booktabs = TRUE) |> 
  kable_styling(full_width = FALSE, position = "left") |> 
  add_header_above(c("Treatment Intent" = 1, "HIV Status" = 2))
```

### Basic logistic model of any adverse events by HIV status
```{r}
#| label: model-any-basic

model_any_basic <- glm(had_any_ae ~ hiv_status + cancer_stage + chemo_type + tx_intent,
                       data = gic_models, family = binomial)

# summary(model_any_basic)
```
Invalid model due to having a 100% adverse event rate in the HIV-positive group.

### Basic logicstic model of any severe adverse events by HIV status
```{r}
#| label: model-severe-basic

model_severe_basic <- glm(had_severe_ae ~ hiv_status + cancer_stage + chemo_type + tx_intent,
                       data = gic_models, family = binomial)

summary(model_severe_basic)
```

```{r}
#| label: tbl-severe-multivariate-or
#| tbl-cap: "Adjusted Odds Ratios for Severe Adverse Events (Grades 3-4)"

tidy(model_severe_basic, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    # Clean up term names
    Term = case_when(
      term == "hiv_statusPositive" ~ "HIV Positive",
      term == "cancer_stageStage III" ~ "Stage III",
      term == "cancer_stageStage IV" ~ "Stage IV",
      term == "chemo_typePlatinum/5-FU" ~ "Platinum/5-FU",
      term == "chemo_typeLCV/5-FU" ~ "LCV/5-FU",
      term == "chemo_typeCapeOX" ~ "CapeOX",
      term == "chemo_typeFOLFOX4" ~ "FOLFOX4",
      term == "chemo_typeCapecitabine" ~ "Capecitabine",
      term == "tx_intentPalliative" ~ "Palliative Intent",
      TRUE ~ term
    ),
    `OR (95% CI)` = paste0(round(estimate, 2), " (", 
                           round(conf.low, 2), " - ", 
                           round(conf.high, 2), ")"),
    p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))),
    std.error = round(std.error, 3)
  ) |>
  select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |>
  # Add reference rows for clarity
  add_row(Term = "HIV Negative (Ref)", `OR (95% CI)` = "1.00", .before = 1) |>
  add_row(Term = "Stage II (Ref)", `OR (95% CI)` = "1.00", .before = 3) |>
  add_row(Term = "Platinum/Paclitaxel (Ref)", `OR (95% CI)` = "1.00", .before = 6) |>
  add_row(Term = "Curative Intent (Ref)", `OR (95% CI)` = "1.00", .before = 12) |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left") |>
  pack_rows("HIV Status", 1, 2) |>
  pack_rows("Cancer Stage", 3, 5) |>
  pack_rows("Chemotherapy Regimen", 6, 11) |>
  pack_rows("Treatment Intent", 12, 13)
```



## $\chi^2$ Tests (HIV vs Early Events)

```{r}
#| label: tbl-any-early-contingency
#| tbl-cap: "Contingency table of early adverse events (cycles 1-4) vs HIV status"

table_any_early <- table(early_ae = factor(gic_models$ae_early, levels = c(0, 1), labels = c("No", "Yes")), HIV = gic_models$hiv_status)

table_any_early |> 
  as.data.frame.matrix() |> 
  kbl(booktabs = TRUE) |> 
  kable_styling(full_width = FALSE, position = "left") |> 
  add_header_above(c("Any Early AE" = 1, "HIV Status" = 2))
```
```{r}
#| label: tbl-severe-early-contingency
#| tbl-cap: "Contingency table of severe early adverse events (cycles 1-4) vs HIV status"

table_severe_early <- table(severe_early_ae = factor(gic_models$severe_ae_early, levels = c(0, 1), labels = c("No", "Yes")), HIV = gic_models$hiv_status)

table_severe_early |> 
  as.data.frame.matrix() |> 
  kbl(booktabs = TRUE) |> 
  kable_styling(full_width = FALSE, position = "left") |> 
  add_header_above(c("Severe Early AE" = 1, "HIV Status" = 2))
```


```{r}
#| label: tests-chi-sq

# Any Early AE (Cycles 1-4)
cat("Chi-Squared Test: Any Early AE (C1-4) vs HIV Status\n")
# Note: This may trigger a warning if any cell counts are 0
print(chisq.test(table_any_early))

cat("\n\n")

# Severe Early AE (Cycles 1-4)
cat("Chi-Squared Test: Severe Early AE (C1-4) vs HIV Status\n")
print(chisq.test(table_severe_early))
```

<!-- ## Early Period Models -->
<!-- ```{r} -->
<!-- #| label: model-simple-early -->

<!-- # Simple logistic for early period -->
<!-- model_any_early_simple <- glm(ae_early ~ hiv_status,  -->
<!--                               data = gic_models, family = binomial) -->

<!-- model_severe_early_simple <- glm(severe_ae_early ~ hiv_status,  -->
<!--                                  data = gic_models, family = binomial) -->

<!-- summary(model_any_early_simple) -->
<!-- summary(model_severe_early_simple) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- #| label: model-basic-early -->

<!-- # Multivariate logistic for early period -->
<!-- model_any_early_basic <- glm(ae_early ~ hiv_status + cancer_stage + chemo_type + tx_intent,  -->
<!--                              data = gic_models, family = binomial) -->

<!-- model_severe_early_basic <- glm(severe_ae_early ~ hiv_status + cancer_stage + chemo_type + tx_intent,  -->
<!--                                 data = gic_models, family = binomial) -->

<!-- summary(model_any_early_basic) -->
<!-- summary(model_severe_early_basic) -->
<!-- ``` -->









