---
title: "GI Cancer: Analysis"
bibliography: references.bib
---
```{r}
#| label: setup
#| warning: false

source("R/setup.R")

# Load Data
gic_descriptive <- readRDS(here("data", "processed", "gic_descriptive.rds"))
gic_models <- gic_descriptive$gic_models
ae_long <- gic_descriptive$ae_long
ae_long_grouped <- gic_descriptive$ae_long_grouped
ae_cycle_long <- gic_descriptive$ae_cycle_long
ae_cycle_measured <- gic_descriptive$ae_cycle_measured
cycle_summary <- gic_descriptive$cycle_summary
cycle_summary_grouped <- gic_descriptive$cycle_summary_grouped
grade_cols <- gic_descriptive$grade_cols
rm(gic_descriptive)
```

# Model Building
## Severe Adverse Events as Outcome
### Unadjusted model
```{r}
#| label: model-any-simple

model_any_simple <- glm(ae_any_pt ~ hiv_status,
                        data = gic_models, family = binomial)

# summary(model_any_simple)
```
Invalid model due to having a 100% adverse event rate in the HIV-positive group.

### Adjusted model
```{r}
#| label: model-severe-simple


model_severe_simple <- glm(ae_severe_any_pt ~ hiv_status,
                           data = gic_models, family = binomial)

summary(model_severe_simple)
```

```{r}
#| label: tbl-severe-simple-or-clean
#| tbl-cap: "Association between HIV Status and Severe Adverse Events (Grades 3-4)"

tidy(model_severe_simple, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  # Clean up the name and prepare for the reference row
  mutate(
    Term = str_replace(term, "hiv_statusPositive", "HIV Positive"),
    `OR (95% CI)` = paste0(round(estimate, 2), " (",
                           round(conf.low, 2), " - ",
                           round(conf.high, 2), ")"),
    p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))),
    std.error = round(std.error, 3)
  ) |>
  select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |>
  # Add the reference row at the top
  add_row(Term = "HIV Negative (Reference)",
          `OR (95% CI)` = "1.00",
          `Std. Error` = NA,
          `P-value` = NA,
          .before = 1) |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left")
```



### Multivariate Logistic: Full treatment period
#### Basic logistic model of any adverse events by HIV status
```{r}
#| label: model-any-basic

model_any_basic <- glm(ae_any_pt ~ hiv_status + cancer_stage + chemo_type + tx_intent,
                       data = gic_models, family = binomial)

summary(model_any_basic)
```
Invalid model due to having a 100% adverse event rate in the HIV-positive group.

#### Basic logicstic model of any severe adverse events by HIV status

{Deprecated}
```{r}
#| label: model-severe-basic-deprecated

model_severe_basic_deprecated <- glm(ae_severe_any_pt ~ hiv_status + cancer_stage + chemo_type + tx_intent,
                       data = gic_models, family = binomial)

summary(model_severe_basic_deprecated)
```
{Current}
```{r}
#| label: model-severe-basic

model_severe_basic <- glm(ae_severe_any_pt ~ hiv_status + cancer_stage + chemo_type + tx_intent + age + sex,
                       data = gic_models, family = binomial)

summary(model_severe_basic)
```

```{r}
#| label: tbl-severe-multivariate-or
#| tbl-cap: "Adjusted Odds Ratios for Severe Adverse Events (Grades 3-4)"

model_severe_basic |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      age ~ "Age (years)",
      sex ~ "Sex",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Regimen",
      tx_intent    ~ "Treatment Intent"
    ),
    pvalue_fun = label_style_pvalue(
      digits = 3
    ),
    add_estimate_to_reference_rows = TRUE
  ) |>
  bold_labels() |> 
  italicize_levels() |> 
  bold_p() |> 
  add_global_p(keep = TRUE) |> 
  add_glance_source_note(include = c(nobs, AIC))
```


### $\chi^2$ Tests (HIV vs Early Events)

```{r}
#| label: tbl-any-early-contingency
#| tbl-cap: "Contingency table of early adverse events (cycles 1-2) vs HIV status"

table_any_early <- table(early_ae = factor(gic_models$ae_any_early, levels = c(0, 1), labels = c("No", "Yes")), HIV = gic_models$hiv_status)

table_any_early |>
  as.data.frame.matrix() |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left") |>
  add_header_above(c("Any Early AE" = 1, "HIV Status" = 2))
```


<!-- ```{r} -->
<!-- #| label: tests-chi-sq -->

<!-- # Any Early AE (Cycles 1-2) -->
<!-- cat("Chi-Squared Test: Any Early AE (C1-2) vs HIV Status\n") -->
<!-- print(chisq.test(table_any_early)) -->

<!-- cat("\n\n") -->

<!-- # Severe Early AE (Cycles 1-2) -->
<!-- cat("Chi-Squared Test: Severe Early AE (C1-2) vs HIV Status\n") -->
<!-- print(chisq.test(table_severe_early)) -->
<!-- ``` -->

## HIV Status as Outcome
### Unadjusted model
```{r}
#|label: model-hiv-simple

model_hiv_simple <- glm(hiv_status ~ ae_max_grade_pt, data = gic_models, family = binomial)

summary(model_hiv_simple)

model_hiv_simple |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      ae_max_grade_pt ~ "Max CTCAE grade"
    ),
    pvalue_fun = label_style_pvalue(
      digits = 3
    ),
    add_estimate_to_reference_rows = TRUE
  ) |>
  bold_labels()
```


<!-- ```{r} -->
<!-- #| label: tbl-hiv-simple-or -->
<!-- #| tbl-cap: "Logistic regression: HIV status vs mean max AE grade (unadjusted)" -->

<!-- tidy(model_hiv_simple, conf.int = TRUE, exponentiate = TRUE) |> -->
<!--   filter(term != "(Intercept)") |> -->
<!--   mutate( -->
<!--     Term = case_when( -->
<!--       term == "ae_mean_max_grade" ~ "Mean max AE grade per measured cycle", -->
<!--       TRUE ~ term -->
<!--     ), -->
<!--     `OR (95% CI)` = paste0( -->
<!--       round(estimate, 2), " (", -->
<!--       round(conf.low, 2), " - ", -->
<!--       round(conf.high, 2), ")" -->
<!--     ), -->
<!--     p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))), -->
<!--     std.error = round(std.error, 3) -->
<!--   ) |> -->
<!--   select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") -->
<!-- ``` -->




### Adjusted model
{Deprecated}
```{r}
#| label: model-hiv-basic-deprecated

model_hiv_basic_deprecated <- glm(
  hiv_status ~ ae_max_grade_pt + cancer_stage + chemo_type + tx_intent,
  data = gic_models,
  family = binomial
)

summary(model_hiv_basic_deprecated)
```
{Current}
```{r}
#| label: model-hiv-basic

model_hiv_basic <- glm(
  hiv_status ~ ae_max_grade_pt + cancer_stage + chemo_type + tx_intent + sex + age,
  data = gic_models,
  family = binomial
)

summary(model_hiv_basic)
```

<!-- ```{r} -->
<!-- #| label: tbl-hiv-basic-or -->
<!-- #| tbl-cap: "Logistic regression: HIV status vs mean max AE grade (adjusted for stage, chemo type, and intent)" -->

<!-- tidy(model_hiv_basic, conf.int = TRUE, exponentiate = TRUE) |> -->
<!--   filter(term != "(Intercept)") |> -->
<!--   mutate( -->
<!--     Term = case_when( -->
<!--       term == "ae_mean_max_grade" ~ "Mean max AE grade per measured cycle", -->
<!--       TRUE ~ term -->
<!--     ), -->
<!--     `OR (95% CI)` = paste0( -->
<!--       round(estimate, 2), " (", -->
<!--       round(conf.low, 2), " - ", -->
<!--       round(conf.high, 2), ")" -->
<!--     ), -->
<!--     p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))), -->
<!--     std.error = round(std.error, 3) -->
<!--   ) |> -->
<!--   select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") -->
<!-- ``` -->

### Model comparison
```{r}
#| label: tbl-hiv-or-comparison
#| tbl-cap: "Logistic regression: HIV status — unadjusted vs adjusted odds ratios"

tbl_simple <- model_hiv_simple |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(ae_max_grade_pt ~ "Max AE grade (patient)")
  )

tbl_basic <- model_hiv_basic |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      ae_max_grade_pt ~ "Max AE grade (patient)",
      cancer_stage ~ "Cancer stage",
      chemo_type ~ "Chemo type",
      tx_intent ~ "Treatment intent",
      age ~ "Age",
      sex ~ "Sex"
    )
  )

tbl_merge(
  list(tbl_simple, tbl_basic),
  tab_spanner = c("Unadjusted", "Adjusted")
)
```

**Summary:** Both the unadjusted and adjusted models with HIV status as outcome indicate that each one unit increase in max AE grade across all treatment cycles is associated with between 28-34% higher odds of the patient being HIV positive. In the adjusted model, higher cancer stages were not reliably associated with being HIV positive. For chemo type there does seem to be a pretty clear difference in treatment options associated with HIV positive status. There does not appear to be any association between treatment intent and HIV status.

### Assumption checks

```{r}
#| label: chk-hiv-assumptions

# Linearity of the log-odds
gic_models |>
  filter(!is.na(ae_max_grade_pt)) |>
  group_by(ae_max_grade_pt) |>
  summarize(prop_hiv = mean(hiv_status == "Positive", na.rm = TRUE)) |>
  mutate(log_odds = log(prop_hiv / (1 - prop_hiv))) |>
  ggplot(aes(x = ae_max_grade_pt, y = log_odds)) +
  geom_point() +
  geom_line() +
  labs(x = "Max AE grade (patient)", y = "Empirical log-odds of HIV+")

print("Linearity isn't amazing, but a straight line fit would show a reasonable fit.")

# Multicollinearity
vif(model_hiv_basic)

print("No significant multicollinearity noted.")

# Influential observations
plot(model_hiv_basic, which = 4)  # Cook's distance
print("Three influential observations, need to investigate further")
```

```{r}
#| label: investigate-influential-obs

# Get row indices of most influential observations (Cook's D > threshold)
# Using 4/n as a common rule-of-thumb cutoff
n <- nrow(model_hiv_basic$model)
cooks_threshold <- 4 / n

influential_idx <- which(cooks.distance(model_hiv_basic) > cooks_threshold)
influential_ids <- gic_models$record_id[influential_idx]

cat("Influential observations (Cook's D > 4/n =", round(cooks_threshold, 4), "):\n")
cat("Row indices:", influential_idx, "\n")
cat("Record IDs:", influential_ids, "\n\n")

# Pull their covariate values to see what makes them unusual
gic_models |>
  filter(record_id %in% influential_ids) |>
  select(record_id, hiv_status, cancer_stage, chemo_type, tx_intent, ae_max_grade_pt)
```

All pts with a large Cook's D are HIV positive, but I haven't immediately spotted any other commonality amongst them to explain the group collectively being so influential, will have to do more investigation.



## Cycles Prescribed as Outcome
{Deprecated}
```{r}
#| label: model-cycles-full-deprecated

# Full cycles prescribed model
model_cycles_full_deprecated <- glm(
  cycles_prescribed ~ hiv_status + cancer_site + cancer_stage + chemo_type + tx_intent,
  data   = gic_models,
  family = poisson
)
```


```{r}
#| label: model-cycles-no-site-deprecated

# Cycles prescribed model without cancer site
model_cycles_no_site_deprecated <- glm(
  cycles_prescribed ~ hiv_status + cancer_stage + chemo_type + tx_intent,
  data   = gic_models,
  family = poisson
)
```

{Current}
```{r}
#| label: model-cycles-full

# Full cycles prescribed model
model_cycles_full <- glm(
  cycles_prescribed ~ hiv_status + cancer_site + cancer_stage + chemo_type + tx_intent + age,
  data   = gic_models,
  family = poisson
)
```

```{r}
#| label: model-cycles-no-site

# Cycles prescribed model without cancer site
model_cycles_no_site <- glm(
  cycles_prescribed ~ hiv_status + cancer_stage + chemo_type + tx_intent + age,
  data   = gic_models,
  family = poisson
)
```

{Deprecated}
```{r}
#| label: tbl-cycles-models-deprecated
#| tbl-cap: "Poisson regression: cycles prescribed (exponentiated coefficients = rate ratios)"

tbl_full <- model_cycles_full_deprecated |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      cancer_site  ~ "Cancer Site",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Type",
      tx_intent    ~ "Treatment Intent"
    )
  ) |>
  bold_labels() |> 
  italicize_levels() |> 
  bold_p() |> 
  add_global_p(keep = TRUE) |> 
  add_vif(statistic = "aGVIF")

tbl_no_site <- model_cycles_no_site_deprecated |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Type",
      tx_intent    ~ "Treatment Intent"
    )
  ) |>
  bold_labels() |> 
  italicize_levels() |> 
  bold_p() |> 
  add_global_p(keep = TRUE) |> 
  add_vif(statistic = "aGVIF")

lrt_p <- anova(model_cycles_no_site_deprecated, model_cycles_full_deprecated, test = "LRT")$`Pr(>Chi)`[2]


tbl_merge(
  list(tbl_no_site, tbl_full),
  tab_spanner = c("Without cancer site", "With cancer site")
) |>
  as_gt() |>
  tab_source_note(
    source_note = paste0(
      "Model without cancer site: AIC = ", round(AIC(model_cycles_no_site_deprecated), 1),
      ", N = ", nobs(model_cycles_no_site_deprecated),
      " | Model with cancer site: AIC = ", round(AIC(model_cycles_full_deprecated), 1),
      ", N = ", nobs(model_cycles_full_deprecated)
    )
  ) |>
  tab_source_note(
    source_note = paste0(
      "LRT (with vs without cancer site): p = ",
      round(anova(model_cycles_no_site_deprecated, model_cycles_full_deprecated, 
                  test = "LRT")$`Pr(>Chi)`[2], 3)
    )
  )
  
```

{Current}
```{r}
#| label: tbl-cycles-models
#| tbl-cap: "Poisson regression: cycles prescribed (exponentiated coefficients = rate ratios)"

tbl_full <- model_cycles_full |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      age ~ "Age (years)",
      cancer_site  ~ "Cancer Site",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Type",
      tx_intent    ~ "Treatment Intent"
    )
  ) |>
  bold_labels() |> 
  italicize_levels() |> 
  bold_p() |> 
  add_global_p(keep = TRUE) |> 
  add_vif(statistic = "aGVIF")

tbl_no_site <- model_cycles_no_site |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      age ~ "Age (years)",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Type",
      tx_intent    ~ "Treatment Intent"
    )
  ) |>
  bold_labels() |> 
  italicize_levels() |> 
  bold_p() |> 
  add_global_p(keep = TRUE) |> 
  add_vif(statistic = "aGVIF")

lrt_p <- anova(model_cycles_no_site, model_cycles_full, test = "LRT")$`Pr(>Chi)`[2]


tbl_merge(
  list(tbl_no_site, tbl_full),
  tab_spanner = c("Without cancer site", "With cancer site")
) |>
  as_gt() |>
  tab_source_note(
    source_note = paste0(
      "Model without cancer site: AIC = ", round(AIC(model_cycles_no_site), 1),
      ", N = ", nobs(model_cycles_no_site),
      " | Model with cancer site: AIC = ", round(AIC(model_cycles_full), 1),
      ", N = ", nobs(model_cycles_full)
    )
  ) |>
  tab_source_note(
    source_note = paste0(
      "LRT (with vs without cancer site): p = ",
      round(anova(model_cycles_no_site, model_cycles_full, 
                  test = "LRT")$`Pr(>Chi)`[2], 3)
    )
  )
  
```


# Toxicity-Related Hospitalizations
```{r}
#| label: prep-hosp-days

hosp_cols <- names(gic_models)[str_detect(names(gic_models), "^hosp_days_([1-9]|1[0-2])$")]

hosp_long <- gic_models |> 
  select(record_id, hiv_status, cycles_given, age, all_of(hosp_cols)) |> 
  pivot_longer(
    cols = all_of(hosp_cols),
    names_to = "cycle_var",
    values_to = "hosp_days"
  ) |> 
  mutate(
    cycle = as.integer(str_extract(cycle_var, "\\d+"))
  )

# Patient-level summaries
hosp_pt <- hosp_long |> 
  group_by(record_id, hiv_status, cycles_given) |> 
  summarize(
    hosp_days_total = sum(hosp_days, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  mutate(
    hosp_days_per_cycle_given = if_else(is.na(cycles_given) | cycles_given <= 0,
                                        NA_real_,
                                        hosp_days_total / cycles_given)
  )

```

## Summary by HIV Status
```{r}
#| label: tbl-hosp-by-pt
#| tbl-cap: "Hospital days attributable to toxicity by HIV status"

hosp_pt |>
  group_by(hiv_status) |>
  summarize(
    n_patients = n(),

    hosp_days_total_mean_sd = summ_median_iqr(hosp_days_total, 1),
    hosp_days_per_cycle_given_mean_sd = summ_median_iqr(hosp_days_per_cycle_given, 1),
    .groups = "drop"
  ) |>
  mutate(
    `HIV Status` = hiv_status,
    `Patients (N)` = n_patients,
    `Total hosp days, mean (SD)` = hosp_days_total_mean_sd,
    `Hosp days / cycles_given, mean (SD)` = hosp_days_per_cycle_given_mean_sd
  ) |>
  select(
    `HIV Status`, `Patients (N)`,
    `Total hosp days, mean (SD)`,
    `Hosp days / cycles_given, mean (SD)`
  ) |>
  kbl(booktabs = TRUE, align = "lcccc") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
#| label: prep-hosp-cycle-data

# Identify cycle-level column groups needed for hospitalization analysis
severe_cols <- names(gic_models)[str_detect(names(gic_models), "^ae_n_severe_([1-9]|1[0-2])$")]
max_grade_cols <- names(gic_models)[str_detect(names(gic_models), "^ae_max_grade_([1-9]|1[0-2])$")]
measured_cols <- names(gic_models)[str_detect(names(gic_models), "^ae_n_markers_measured_([1-9]|1[0-2])$")]

# Pivot each set of cycle cols to long
pivot_cycle_cols <- function(df, cols, value_name) {
  df |>
    select(record_id, all_of(cols)) |>
    pivot_longer(cols = all_of(cols), names_to = "cycle_var", values_to = value_name) |>
    mutate(cycle = as.integer(str_extract(cycle_var, "\\d+$"))) |>
    select(-cycle_var)
}

severe_long <- pivot_cycle_cols(gic_models, severe_cols,    "ae_n_severe")
max_grade_long <- pivot_cycle_cols(gic_models, max_grade_cols, "ae_max_grade")
measured_long <- pivot_cycle_cols(gic_models, measured_cols,  "ae_n_markers_measured")

# Join AE variables to hospitalization data
hosp_severe_cycle <- hosp_long |>
  left_join(severe_long,   by = c("record_id", "cycle")) |>
  left_join(measured_long, by = c("record_id", "cycle")) |>
  mutate(ae_measured = ae_n_markers_measured > 0)

hosp_maxgrade_cycle <- hosp_long |>
  left_join(max_grade_long, by = c("record_id", "cycle")) |>
  left_join(measured_long,  by = c("record_id", "cycle")) |>
  mutate(ae_measured = ae_n_markers_measured > 0)
```

## Hospitalizations by AE Severity
```{r}
#| label: tbl-hosp-vs-severe

hosp_severe_cycle |> 
  filter(ae_measured) |> 
  group_by(hiv_status, ae_n_severe) |> 
  summarize(
    n_cycles = sum(!is.na(hosp_days)),
    hosp_days_mean_sd = summ_median_iqr(hosp_days),
    .groups = "drop"
  ) |> 
  mutate(
    `HIV Status` = hiv_status,
    `# severe AEs in cycle` = ae_n_severe,
    `# Cycles` = n_cycles,
    `Hosp days, mean (SD); range` = hosp_days_mean_sd
  ) |> 
  select(`HIV Status`, `# severe AEs in cycle`, `# Cycles`, `Hosp days, mean (SD); range`) |> 
  kbl(booktabs = TRUE, align = "lccc") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
#| label: plt-hosp-vs-severe
#| fig-cap: "Mean hospital days vs number of severe AEs in the cycle, by HIV status"

hosp_severe_summ <- hosp_severe_cycle |>
  filter(ae_measured) |>
  group_by(hiv_status, ae_n_severe) |>
  summarize(
    mean_hosp_days = mean(hosp_days, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(hosp_severe_summ, aes(x = ae_n_severe, y = mean_hosp_days, color = hiv_status, group = hiv_status)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    x = "Number of severe AEs in cycle",
    y = "Mean hospital days (toxicity) in cycle",
    color = "HIV status"
  ) +
  theme_minimal()
```

```{r}
#| label: tbl-hosp-vs-max-grade
#| tbl-cap: "Hospital days vs max AE grade in cycle, by HIV status (among cycles with any AE grading recorded)"

hosp_maxgrade_cycle |>
  filter(ae_measured) |>
  group_by(hiv_status, ae_max_grade) |>
  summarize(
    n_cycles = sum(!is.na(hosp_days)),
    hosp_days_median_iqr = summ_median_iqr(hosp_days, 1),
    .groups = "drop"
  ) |>
  mutate(
    `HIV Status` = hiv_status,
    `Max AE grade in cycle` = ae_max_grade,
    `# Cycles` = n_cycles,
    `Hosp days, median (IQR)` = hosp_days_median_iqr
  ) |>
  select(`HIV Status`, `Max AE grade in cycle`, `# Cycles`, `Hosp days, median (IQR)`) |>
  kbl(booktabs = TRUE, align = "lccc") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
#| label: plt-hosp-vs-max-grade
#| fig-cap: "Mean hospital days vs max AE grade in cycle, by HIV status (among measured cycles)"

hosp_maxgrade_summ <- hosp_maxgrade_cycle |>
  filter(ae_measured) |>
  group_by(hiv_status, ae_max_grade) |>
  summarize(
    mean_hosp_days = mean(hosp_days, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(hosp_maxgrade_summ, aes(x = ae_max_grade, y = mean_hosp_days, color = hiv_status, group = hiv_status)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 0:5) +
  labs(
    x = "Max AE grade in cycle",
    y = "Mean hospital days (toxicity) in cycle",
    color = "HIV status"
  ) +
  theme_minimal()
```

## Regression Models


```{r}
#| label: prep-hosp-regression


hosp_ae_cycle <- hosp_long |>
  left_join(
    ae_cycle_long |> select(record_id, cycle, ae_n_severe, ae_max_grade, ae_n_markers_measured),
    by = c("record_id", "cycle")
  ) |>
  mutate(ae_measured = ae_n_markers_measured > 0)


hosp_model_data <- hosp_ae_cycle |>
  filter(ae_measured)

hosp_model_data <- hosp_model_data |>
  mutate(any_hosp = as.integer(hosp_days > 0))
```

{Deprecated}
```{r}
#| label: model-hosp-any-deprecated

# Any hospitalization this cycle (logistic)
model_any_hosp_deprecated <- glmer(
  any_hosp ~ ae_max_grade + hiv_status + (1 | record_id),
  data = hosp_model_data,
  family = binomial
)
summary(model_any_hosp_deprecated)
fixef(model_any_hosp_deprecated) |> exp()
confint(model_any_hosp_deprecated, method = "Wald") |> exp()
```
Each increase in max AE grade in a given cycle is associated with about a 5.5-fold increase (95% CI: 3.7-8.2) in the odds of being hospitalized. For patients with no adverse events, being HIV positive is associated with about a 3.4-fold increase in odds of being hospitalized, but the effect of being HIV positive on being hospitalized decreases by about 35% for each unit increase in max AE grade.

{Current}
```{r}
#| label: model-hosp-any

# Any hospitalization this cycle (logistic)
model_any_hosp <- glmer(
  any_hosp ~ ae_max_grade + hiv_status + age +(1 | record_id),
  data = hosp_model_data,
  family = binomial
)
summary(model_any_hosp)
fixef(model_any_hosp) |> exp()
confint(model_any_hosp, method = "Wald") |> exp()
```

```{r}
# Length of stay given hospitalization
model_hosp_days_deprecated <- lmer(
  hosp_days ~ ae_max_grade + hiv_status + (1 | record_id),
  data = hosp_model_data |> filter(hosp_days > 0)
)
summary(model_hosp_days_deprecated)
```


```{r}
# Length of stay given hospitalization
model_hosp_days <- lmer(
  hosp_days ~ ae_max_grade + hiv_status + age + (1 | record_id),
  data = hosp_model_data |> filter(hosp_days > 0)
)
summary(model_hosp_days)
```
AE severity has minimal impact on length of stay for either group once someone is already hospitalized, although HIV positive patients are expected to spend slightly longer in the hospital than HIV negative patients for each unit increase in max adverse event experienced.


# Early Treatment Dropout

```{r}
#| label: prep-early-dropout

# Add early dropout flag to gic_models
gic_models <- gic_models |>
  mutate(
    early_dropout = as.integer(
      cycles_given <= early_dropout_cycles & cycles_prescribed > early_dropout_cycles
    )
  )
```

## Descriptive Overview

This section looks at whether the types of adverse events patients experience in their first few cycles are associated with stopping treatment early, and whether that relationship looks different for HIV-positive vs HIV-negative patients.

```{r}
#| label: tbl-early-dropout-by-hiv
#| tbl-cap: "Early treatment discontinuation by HIV status at two cycle cutoffs"

map_df(c(2, 4), function(cutoff) {
  gic_models |>
    mutate(dropout = as.integer(cycles_given <= cutoff & cycles_prescribed > cutoff)) |>
    group_by(hiv_status) |>
    summarize(
      n_total     = n(),
      n_dropout   = sum(dropout == 1, na.rm = TRUE),
      pct_dropout = round(100 * n_dropout / n_total, 1),
      .groups     = "drop"
    ) |>
    mutate(
      cutoff            = cutoff,
      `Early dropout, n (%)` = paste0(n_dropout, " (", pct_dropout, "%)")
    )
}) |>
  select(hiv_status, cutoff, `Total patients` = n_total, `Early dropout, n (%)`) |>
  pivot_wider(
    names_from  = cutoff,
    values_from = `Early dropout, n (%)`,
    names_glue  = "Cutoff ≤{cutoff} cycles"
  ) |>
  rename(`HIV Status` = hiv_status, `Total patients` = `Total patients`) |>
  gt() |>
  tab_spanner(
    label   = "Early dropout, n (%)",
    columns = starts_with("Cutoff")
  ) |>
  tab_footnote(
    footnote = "Early dropout defined as cycles given ≤ cutoff among patients prescribed more than the cutoff number of cycles.",
    location = cells_column_spanners()
  ) |>
  opt_footnote_marks(marks = "extended") |>
  cols_align(align = "center", columns = -`HIV Status`) |>
  cols_align(align = "left",   columns = `HIV Status`)
```

For each AE type, the predictor used is the max grade recorded across cycles 1 to 4. Where no grade was recorded for a given AE type in that window, it's been set to 0, on the assumption that if nothing was flagged there was probably nothing to flag. That's a more reasonable assumption for the liver panel markers, which Dr. Mazhindu said are usually only ordered when there's a clinical reason to, than it is for something like hemoglobin or platelets where a missing value might just mean the data wasn't entered. It was done this way mainly because doing a complete case analysis instead would have left only 97 patients and 26 early dropout events, which isn't really enough to work with. This is worth coming back to before the analysis is finalized.

Sodium and potassium grades also don't distinguish between high and low abnormalities, so hypokalemia and hyperkalemia end up collapsed into the same grade variable. This can be revisited later, but I'm leaving it as is for the moment because the models below already have more regressors than ideal. See MET-005 and DAT-024.

## Global test for early dropout by HIV status

```{r}
#| label: tst-early-dropout-hiv
#| tbl-cap: "Early treatment discontinuation by HIV status"

# 2x2 table: early dropout x HIV status
dropout_hiv_table <- table(
  dropout = factor(gic_models$early_dropout, levels = c(0, 1), labels = c("Completed", "Early dropout")),
  HIV = gic_models$hiv_status
)

dropout_hiv_table |>
  as.data.frame.matrix() |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE) |>
  add_header_above(c(" " = 1, "HIV Status" = 2))

# Check expected counts to decide test
dropout_expected <- suppressWarnings(chisq.test(dropout_hiv_table)$expected)

if (any(dropout_expected < 5)) {
  dropout_test <- fisher.test(dropout_hiv_table)
  cat("Fisher's exact test (expected count < 5 in at least one cell)\n")
} else {
  dropout_test <- chisq.test(dropout_hiv_table, correct = FALSE)
  cat("Chi-squared test (all expected counts >= 5)\n")
}

dropout_test
```

```{r}
#| label: prep-ae-early-dropout

# Max grade per AE type across cycles 1-early_dropout_cycles
# Patients with no recorded grade for a given AE type in this window
# are excluded from that AE type's predictor (not treated as grade 0)
ae_early <- ae_long |>
  filter(cycle <= early_dropout_cycles, !is.na(grade)) |>
  group_by(record_id, ae_type) |>
  summarize(max_grade_early = max(grade, na.rm = TRUE), .groups = "drop") |>
  left_join(
    gic_models |> select(record_id, hiv_status, early_dropout),
    by = "record_id"
  ) |>
  filter(!is.na(early_dropout))

# Wide format for modeling — one row per patient, one col per AE type
ae_types <- unique(ae_early$ae_type)

ae_early_wide <- ae_early |>
  pivot_wider(
    names_from  = ae_type,
    values_from = max_grade_early
  ) |>
  mutate(
    # All missing AE grades imputed to 0 — absence of a recorded grade treated
    # as no adverse event. Less defensible for some markers (e.g. plt, hb) than
    # for liver panels, but necessary to retain full sample for LASSO.
    # Revisit imputation strategy when time permits.
    across(all_of(ae_types), ~ replace_na(.x, 0))
  ) |>
  left_join(
    gic_models |> select(record_id, hiv_status, early_dropout),
    by = c("record_id", "hiv_status", "early_dropout")
  )
```

```{r}
#| label: tbl-ae-early-dropout
#| tbl-cap: "Adverse event grades in cycles 1–4 by early dropout status and HIV status"

ae_early |>
  mutate(ae_label = ae_labels[ae_type]) |>
  group_by(hiv_status, early_dropout, ae_label) |>
  summarize(
    n_any    = sum(max_grade_early > 0, na.rm = TRUE),
    med      = median(max_grade_early, na.rm = TRUE),
    n_severe = sum(max_grade_early >= 3, na.rm = TRUE),
    .groups  = "drop"
  ) |>
  mutate(
    dropout_label = if_else(early_dropout == 1, "Early dropout", "Completed >4 cycles"),
    cell          = paste0(n_any, " / ", round(med, 1), " / ", n_severe)
  ) |>
  select(hiv_status, ae_label, dropout_label, cell) |>
  pivot_wider(
    names_from  = c(hiv_status, dropout_label),
    values_from = cell,
    names_glue  = "{hiv_status}: {dropout_label}"
  ) |>
  select(
    ae_label,
    `Negative: Completed >4 cycles`,
    `Negative: Early dropout`,
    `Positive: Completed >4 cycles`,
    `Positive: Early dropout`
  ) |>
  rename(`Adverse event` = ae_label) |>
  gt() |>
  tab_spanner(
    label   = "HIV Negative",
    columns = starts_with("Negative")
  ) |>
  tab_spanner(
    label   = "HIV Positive",
    columns = starts_with("Positive")
  ) |>
  cols_label(
    `Negative: Completed >4 cycles` = "Completed >4",
    `Negative: Early dropout`        = "Early dropout",
    `Positive: Completed >4 cycles`  = "Completed >4",
    `Positive: Early dropout`         = "Early dropout"
  ) |>
  cols_align(align = "center", columns = -`Adverse event`) |>
  cols_align(align = "left",   columns = `Adverse event`) |>
  tab_footnote(
    footnote = "Cells show: n with any AE / median max grade / n with severe AE (grade ≥ 3)",
    location = cells_column_spanners()
  ) |>
  opt_footnote_marks(marks = "extended")
```


```{r}
#| label: tst-dropout-by-ae-type
#| tbl-cap: "Association between experiencing each AE type (any grade) in cycles 1–4 and early dropout"

dropout_ae_tests <- map_dfr(ae_types, function(at) {
  dat <- ae_early_wide |>
    filter(!is.na(early_dropout)) |>
    mutate(has_ae = as.integer(.data[[at]] > 0))
  
  tab <- table(dropout = dat$early_dropout, has_ae = dat$has_ae)
  
  if (nrow(tab) < 2 | ncol(tab) < 2) {
    return(tibble(ae_type = at, p = NA_real_, test = "—",
                  n_ae = NA, n_ae_dropout = NA, n_total = NA))
  }
  
  expected <- suppressWarnings(chisq.test(tab)$expected)
  if (any(expected < 5)) {
    result <- fisher.test(tab)
    test_used <- "Fisher's"
  } else {
    result <- chisq.test(tab, correct = FALSE)
    test_used <- "Chi-squared"
  }
  
  tibble(
    ae_type = at,
    n_total = nrow(dat),
    n_ae = sum(dat$has_ae == 1),
    n_ae_dropout = sum(dat$has_ae == 1 & dat$early_dropout == 1),
    p = result$p.value,
    test = test_used
  )
}) |>
  mutate(
    ae_label = ae_labels[ae_type],
    p_adj = p.adjust(p, method = "BH"),
    `p (raw)` = map_chr(p, format_p),
    `p (adj)` = map_chr(p_adj, format_p),
    `With AE, n (dropout)` = paste0(n_ae, " (", n_ae_dropout, ")")
  )

dropout_ae_tests |>
  select(`Adverse event` = ae_label, `With AE, n (dropout)`, Test = test, `p (raw)`, `p (adj)`) |>
  gt() |>
  cols_align(align = "left", columns = `Adverse event`) |>
  cols_align(align = "center", columns = -`Adverse event`) |>
  tab_footnote(
    footnote = paste0(
      "Tests whether having any occurrence of each AE type (grade > 0) in cycles 1–",
      early_dropout_cycles,
      " is associated with early dropout, pooling HIV groups. ",
      "Chi-squared or Fisher's exact as appropriate. p-values BH-adjusted across AE types."
    ),
    locations = cells_column_spanners()
  ) |>
  opt_footnote_marks(marks = "extended")
```

Tests whether, among all patients in the early dropout analysis, is having any occurrence of a specific AE type (grade > 0) in cycles 1-4 associated with whether they dropped out? Multiple comparisons correction applied because testing the same population multiple times across adverse event types.


The heatmap below gives a quick visual overview of where the grade differences between early dropout and completed patients are largest, and whether those patterns look similar across HIV groups.

```{r}
#| label: plt-ae-early-dropout-heatmap
#| fig-cap: "Median max AE grade in cycles 1–4 by AE type, early dropout status, and HIV status"

ae_early |>
  group_by(hiv_status, early_dropout, ae_type) |>
  summarize(
    median_grade = median(max_grade_early, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    dropout_label = if_else(early_dropout == 1, "Early dropout", "Completed ≥5 cycles"),
    facet_label   = paste0("HIV ", hiv_status)
  ) |>
  ggplot(aes(x = dropout_label, y = ae_type, fill = median_grade)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(median_grade, 1)), size = 3) +
  scale_fill_viridis_c(option = "magma", direction = -1, 
                       limits = c(0, 4), name = "Median\nmax grade") +
  facet_wrap(~ facet_label) +
  labs(
    x = NULL,
    y = "Adverse event type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1),
    strip.text  = element_text(face = "bold")
  )
```

Before moving to the models, it's worth checking how many patients in each HIV group actually experienced each type of adverse event. AE types with very few patients in either group are excluded from the univariate models since there isn't enough data to estimate anything reliable for them.

```{r}
#| label: tbl-ae-coverage
#| tbl-cap: "AE grade coverage in cycles 1–4: patients with any recorded adverse event by HIV status"

ae_early_wide |>
  select(hiv_status, early_dropout, all_of(ae_types)) |>
  pivot_longer(
    cols      = all_of(ae_types),
    names_to  = "ae_type",
    values_to = "grade"
  ) |>
  group_by(ae_type, hiv_status) |>
  summarize(
    n_total    = n(),
    n_any_ae   = sum(grade > 0, na.rm = TRUE),
    pct_any_ae = round(100 * n_any_ae / n_total, 1),
    .groups    = "drop"
  ) |>
  mutate(
    ae_label = ae_labels[ae_type],
    `n (%)` = paste0(n_any_ae, " (", pct_any_ae, "%)")
  ) |>
  select(ae_label, hiv_status, `n (%)`) |>
  pivot_wider(
    names_from  = hiv_status,
    values_from = `n (%)`,
    names_prefix = "HIV "
  ) |>
  rename(`Adverse event` = ae_label) |>
  gt() |>
  tab_spanner(
    label   = "Patients with any AE (grade > 0)",
    columns = starts_with("HIV")
  ) |>
  cols_align(align = "left",   columns = `Adverse event`) |>
  cols_align(align = "center", columns = -`Adverse event`) |>
  tab_footnote(
    footnote = paste0(
      "Among patients with cycles given ≤ ", early_dropout_cycles,
      " or prescribed > ", early_dropout_cycles, " cycles. ",
      "AE types with 0 patients in either group cannot be evaluated in LASSO or univariate models."
    ),
    location = cells_column_spanners()
  ) |>
  opt_footnote_marks(marks = "extended")
```

```{r}
#| label: prep-sparse-ae

# AE types with fewer than 5 patients with any non-zero grade in either
# HIV group — too sparse for reliable estimation
ae_cols <- intersect(ae_types, names(ae_early_wide))

sparse_ae <- ae_early_wide |>
  select(hiv_status, all_of(ae_cols)) |>
  pivot_longer(cols = all_of(ae_cols), names_to = "ae_type", values_to = "grade") |>
  group_by(ae_type, hiv_status) |>
  summarize(n_any = sum(grade > 0, na.rm = TRUE), .groups = "drop") |>
  group_by(ae_type) |>
  filter(any(n_any < 5)) |>
  pull(ae_type) |>
  unique()
```


## Univariate Logistic Models

```{r}
#| label: fit-univariate-ae-models

# One logistic model per AE type — early dropout ~ max grade in cycles 1-4 + HIV status
# Patients with no recorded grade for a given AE type are excluded from that model
univariate_results <- map_df(ae_types, function(at) {
  dat <- ae_early_wide |>
    filter(!is.na(.data[[at]]), !is.na(early_dropout), !is.na(hiv_status))

  if (sum(dat$early_dropout) < 10) {
    return(tibble(ae_type = at, note = "Insufficient early dropout cases", .rows = 1))
  }

  if (at %in% sparse_ae) {
    return(tibble(ae_type = at, note = "Sparse — excluded per coverage check", .rows = 1))
  }

  mod <- glm(
    as.formula(paste0("early_dropout ~ `", at, "` * hiv_status")),
    data   = dat,
    family = binomial
  )

  tidy(mod, conf.int = TRUE, exponentiate = TRUE) |>
    filter(str_detect(term, at)) |>
    mutate(
      ae_type    = at,
      term_label = if_else(
        str_detect(term, "hiv_status"),
        "AE grade × HIV Positive",
        "AE grade (main effect)"
      )
    )
}) |>
  filter(!is.na(estimate)) |>
  mutate(p_adj = p.adjust(p.value, method = "BH"))
```

```{r}
#| label: tbl-univariate-ae-results
#| tbl-cap: "Univariate logistic regression: association between AE type max grade and early dropout (adjusted for HIV status)"

# Sample sizes per AE type for context
ae_sample_sizes <- map_df(ae_types, function(at) {
  dat <- ae_early_wide |>
    filter(!is.na(early_dropout), !is.na(hiv_status))
  
  dat |>
    group_by(hiv_status) |>
    summarize(
      n_nonzero        = sum(.data[[at]] > 0),
      n_events_nonzero = sum(.data[[at]] > 0 & early_dropout == 1),
      .groups          = "drop"
    ) |>
    mutate(
      ae_type  = at,
      cell     = paste0(n_nonzero, " (", n_events_nonzero, ")")
    ) |>
    select(ae_type, hiv_status, cell) |>
    pivot_wider(names_from = hiv_status, values_from = cell,
                names_glue = "n_{hiv_status}")
})

# Sort labels
ae_order <- sort(ae_labels)

univariate_results |>
  left_join(ae_sample_sizes, by = "ae_type") |>
  mutate(
    ae_label      = ae_labels[ae_type],
    ae_label      = factor(ae_label, levels = sort(ae_labels)),
    `OR (95% CI)` = paste0(round(estimate, 2), " (", round(conf.low, 2), "–", round(conf.high, 2), ")"),
    `p (adj)`     = map_chr(p_adj, format_p),
    `p (raw)`     = map_chr(p.value, format_p)
  ) |>
  arrange(ae_label, term_label) |>
  select(
    `Adverse event`  = ae_label,
    `Term`           = term_label,
    `HIV-: n (drop)` = n_Negative,
    `HIV+: n (drop)` = n_Positive,
    `OR (95% CI)`,
    `p (raw)`,
    `p (adj)`
  ) |>
  gt() |>
  tab_row_group(
    label = "AE grade × HIV interaction",
    rows  = Term == "AE grade × HIV Positive"
  ) |>
  tab_row_group(
    label = "AE grade main effect (HIV adjusted)",
    rows  = Term == "AE grade (main effect)"
  ) |>
  cols_hide(Term) |>
  cols_align(align = "left",   columns = `Adverse event`) |>
  cols_align(align = "center", columns = -`Adverse event`) |>
  tab_footnote(
    footnote = "Main effect OR: change in odds of early dropout per one-unit increase in max AE grade, holding HIV status constant. Interaction OR: difference in that association for HIV-positive vs HIV-negative patients. N (events) = patients with any grade > 0 for this AE type / early dropouts among those patients. p-values BH-adjusted across all tests.",
    location = cells_column_labels(columns = `p (adj)`)
  ) |>
  opt_footnote_marks(marks = "extended")
```

To make the HIV comparison more concrete, the table below shows estimated ORs separately for HIV-negative and HIV-positive patients. The HIV-positive OR is derived by multiplying the main effect OR by the interaction OR, since those add on the log-odds scale. There's no direct standard error for the HIV-positive OR from the model, so treat those as directional estimates rather than something to draw firm conclusions from.

```{r}
#| label: tbl-univariate-hiv-pos-or
#| tbl-cap: "Estimated odds ratios for early dropout per unit increase in max AE grade among HIV-positive patients"

univariate_results |>
  select(ae_type, term_label, estimate, p_adj) |>
  pivot_wider(names_from = term_label, 
              values_from = c(estimate, p_adj)) |>
  mutate(
    ae_label   = factor(ae_labels[ae_type], levels = ae_order),
    or_hiv_pos = `estimate_AE grade (main effect)` * `estimate_AE grade × HIV Positive`
  ) |>
  arrange(ae_label) |>
  mutate(
    `OR (HIV-)` = round(`estimate_AE grade (main effect)`, 3),
    `p (HIV-)`  = map_chr(`p_adj_AE grade (main effect)`, format_p),
    `OR (HIV+)` = round(or_hiv_pos, 3),
    `p (interaction)` = map_chr(`p_adj_AE grade × HIV Positive`, format_p)
  ) |>
  select(
    `Adverse event`   = ae_label,
    `OR (HIV-)`,
    `p (HIV-)`,
    `OR (HIV+)`,
    `p (interaction)`
  ) |>
  gt() |>
  tab_spanner(
    label   = "HIV Negative",
    columns = c(`OR (HIV-)`, `p (HIV-)`)
  ) |>
  tab_spanner(
    label   = "HIV Positive",
    columns = c(`OR (HIV+)`, `p (interaction)`)
  ) |>
  cols_align(align = "left",   columns = `Adverse event`) |>
  cols_align(align = "center", columns = -`Adverse event`) |>
  tab_footnote(
    footnote = "ORs reflect change in odds of early dropout per unit increase in max AE grade. OR (HIV+) = main effect OR × interaction OR. p-values are BH-adjusted; p (interaction) tests whether the association differs between HIV groups.",
    location = cells_column_spanners(spanners = "HIV Positive")
  ) |>
  opt_footnote_marks(marks = "extended")
```

