---
title: "GI Cancer: Analysis"
bibliography: references.bib
execute:
  echo: false
format:
  html:
    toc-depth: 4
    code-links:
      - text: GitHub Repo
        icon: github
        href: "https://github.com/EDukeChase/gi-cancer_project/blob/main/4%20-%20Statistical%20Analysis.qmd"
---
```{r}
#| label: setup
#| warning: false

source("R/setup.R")

# Load Data
gic_descriptive <- readRDS(onedrive_data_path("processed", "gic_descriptive.rds"))
gic_models <- gic_descriptive$gic_models
ae_long <- gic_descriptive$ae_long
ae_long_grouped <- gic_descriptive$ae_long_grouped
ae_cycle_long <- gic_descriptive$ae_cycle_long
ae_cycle_measured <- gic_descriptive$ae_cycle_measured
cycle_summary <- gic_descriptive$cycle_summary
cycle_summary_grouped <- gic_descriptive$cycle_summary_grouped
grade_cols <- gic_descriptive$grade_cols
mgmt_long <- gic_descriptive$mgmt_long
mgmt_cycle <- gic_descriptive$mgmt_cycle
rm(gic_descriptive)

# Finalized tables/figures destined for the publication doc (05), keyed by slug.
# Each save_table()/save_figure() call below also appends its result here.
pub_objects_analysis <- list()
```

# Model Building
## Severe Adverse Events as Outcome
### Unadjusted model
```{r}
#| label: model-any-simple

model_any_simple <- glm(ae_any_pt ~ hiv_status,
                        data = gic_models, family = binomial)

# summary(model_any_simple)
```
Invalid model due to having a 100% adverse event rate in the HIV-positive group.

### Adjusted model
```{r}
#| label: model-severe-simple


model_severe_simple <- glm(ae_severe_any_pt ~ hiv_status,
                           data = gic_models, family = binomial)
```

```{r}
#| label: tbl-severe-simple-or-clean
#| tbl-cap: "Association between HIV Status and Severe Adverse Events (Grades 3-4)"

tidy(model_severe_simple, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  # Clean up the name and prepare for the reference row
  mutate(
    Term = str_replace(term, "hiv_statusPositive", "HIV Positive"),
    `OR (95% CI)` = paste0(round(estimate, 2), " (",
                           round(conf.low, 2), " - ",
                           round(conf.high, 2), ")"),
    p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))),
    std.error = round(std.error, 3)
  ) |>
  select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |>
  # Add the reference row at the top
  add_row(Term = "HIV Negative (Reference)",
          `OR (95% CI)` = "1.00",
          `Std. Error` = NA,
          `P-value` = NA,
          .before = 1) |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left")
```



### Multivariate Logistic: Full treatment period
#### Basic logistic model of any adverse events by HIV status
::: {.callout-important title="Deprecated model — not a result"}
`model_any_basic` **is not a valid model and must not be read as a finding.**
Every patient in the HIV-positive group experienced at least one adverse event,
so the outcome is completely separated on HIV status and the coefficient is
uninterpretable.

It is retained only because the 100% event rate is itself worth stating in the
paper. Note that this is why age was *not* added to it under the 15-May-2026
"add age to remaining models" item: the problem is separation, not a missing
covariate, and adding covariates to a separated model makes the fit worse.

**Open decision for the team:** delete this chunk outright, or keep it as the
source of the "100% of HIV-positive patients experienced an adverse event"
sentence. If kept, it should not appear in the publication figures document.
:::

```{r}
#| label: model-any-basic

model_any_basic <- glm(ae_any_pt ~ hiv_status + cancer_stage + chemo_type + tx_intent,
                       data = gic_models, family = binomial)

summary(model_any_basic)
```

#### Basic logicstic model of any severe adverse events by HIV status


```{r}
#| label: model-severe-basic

model_severe_basic <- glm(ae_severe_any_pt ~ hiv_status + cancer_stage + chemo_type + tx_intent + age + sex,
                       data = gic_models, family = binomial)
```

```{r}
#| label: tbl-severe-multivariate-or
#| tbl-cap: "Adjusted Odds Ratios for Severe Adverse Events (Grades 3-4)"

model_severe_basic |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      age ~ "Age (years)",
      sex ~ "Sex",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Regimen",
      tx_intent    ~ "Treatment Intent"
    ),
    pvalue_fun = label_style_pvalue(
      digits = 3
    )
  ) |>
  tbl_drop_ref_rows() |>
  bold_labels() |>
  italicize_levels() |>
  bold_p() |>
  add_global_p(keep = TRUE) |>
  add_glance_source_note(include = c(nobs, AIC))
```

#### Validation: model_severe_basic
::: {.callout-note collapse="true" title="Validation: model_severe_basic"}
```{r}
#| label: chk-severe-basic-vif

check_vif(model_severe_basic)
```
:::

Publication version: adjusted odds ratios for severe AE.
```{r}
#| label: tbl-severe-or-comparison
#| tbl-cap: "Adjusted association between HIV status and severe adverse events. Reference levels are named in each variable label rather than shown as separate rows. Unadjusted estimates were closely consistent with these and are not tabulated."

tbl_severe_simple <- model_severe_simple |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(hiv_status ~ "HIV Status"),
    pvalue_fun = label_style_pvalue(digits = 3)
  ) |>
  tbl_drop_ref_rows() |>
  bold_p()

tbl_severe_adjusted <- model_severe_basic |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Regimen",
      tx_intent    ~ "Treatment Intent",
      age          ~ "Age (years)",
      sex          ~ "Sex"
    ),
    pvalue_fun = label_style_pvalue(digits = 3)
  ) |>
  tbl_drop_ref_rows() |>
  bold_p()

# 5-Jun-2026: the unadjusted estimate is very consistent with the adjusted one,
# so it is not shown. The unadjusted model is still *fit* above, because the
# manuscript states that the unadjusted results were similar and that claim
# needs to be backed by an actual fit. tbl_severe_simple is retained in the
# environment for that purpose and for the sensitivity appendix.
tbl_severe_or_comparison <- tbl_severe_adjusted |>
  bold_labels() |>
  italicize_levels()

tbl_severe_or_comparison

obj <- save_table(slug = "tbl_severe_or_comparison", gtsum = tbl_severe_or_comparison)
pub_objects_analysis$tbl_severe_or_comparison <- obj
```


### $\chi^2$ Tests (HIV vs Early Events)

```{r}
#| label: tbl-any-early-contingency
#| tbl-cap: "Contingency table of early adverse events (cycles 1-2) vs HIV status"

table_any_early <- table(early_ae = factor(gic_models$ae_any_early, levels = c(0, 1), labels = c("No", "Yes")), HIV = gic_models$hiv_status)

table_any_early |>
  as.data.frame.matrix() |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left") |>
  add_header_above(c("Any Early AE" = 1, "HIV Status" = 2))
```


<!-- ```{r} -->
<!-- #| label: tests-chi-sq -->

<!-- # Any Early AE (Cycles 1-2) -->
<!-- cat("Chi-Squared Test: Any Early AE (C1-2) vs HIV Status\n") -->
<!-- print(chisq.test(table_any_early)) -->

<!-- cat("\n\n") -->

<!-- # Severe Early AE (Cycles 1-2) -->
<!-- cat("Chi-Squared Test: Severe Early AE (C1-2) vs HIV Status\n") -->
<!-- print(chisq.test(table_severe_early)) -->
<!-- ``` -->

## HIV Status as Outcome
### Unadjusted model
```{r}
#|label: model-hiv-simple

model_hiv_simple <- glm(hiv_status ~ ae_max_grade_pt, data = gic_models, family = binomial)

model_hiv_simple |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      ae_max_grade_pt ~ "Max CTCAE grade"
    ),
    pvalue_fun = label_style_pvalue(
      digits = 3
    )
  ) |>
  tbl_drop_ref_rows() |>
  bold_labels()
```


<!-- ```{r} -->
<!-- #| label: tbl-hiv-simple-or -->
<!-- #| tbl-cap: "Logistic regression: HIV status vs mean max AE grade (unadjusted)" -->

<!-- tidy(model_hiv_simple, conf.int = TRUE, exponentiate = TRUE) |> -->
<!--   filter(term != "(Intercept)") |> -->
<!--   mutate( -->
<!--     Term = case_when( -->
<!--       term == "ae_mean_max_grade" ~ "Mean max AE grade per measured cycle", -->
<!--       TRUE ~ term -->
<!--     ), -->
<!--     `OR (95% CI)` = paste0( -->
<!--       round(estimate, 2), " (", -->
<!--       round(conf.low, 2), " - ", -->
<!--       round(conf.high, 2), ")" -->
<!--     ), -->
<!--     p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))), -->
<!--     std.error = round(std.error, 3) -->
<!--   ) |> -->
<!--   select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") -->
<!-- ``` -->




### Adjusted model

```{r}
#| label: model-hiv-basic

model_hiv_basic <- glm(
  hiv_status ~ ae_max_grade_pt + cancer_stage + chemo_type + tx_intent + sex + age,
  data = gic_models,
  family = binomial
)
```

<!-- ```{r} -->
<!-- #| label: tbl-hiv-basic-or -->
<!-- #| tbl-cap: "Logistic regression: HIV status vs mean max AE grade (adjusted for stage, chemo type, and intent)" -->

<!-- tidy(model_hiv_basic, conf.int = TRUE, exponentiate = TRUE) |> -->
<!--   filter(term != "(Intercept)") |> -->
<!--   mutate( -->
<!--     Term = case_when( -->
<!--       term == "ae_mean_max_grade" ~ "Mean max AE grade per measured cycle", -->
<!--       TRUE ~ term -->
<!--     ), -->
<!--     `OR (95% CI)` = paste0( -->
<!--       round(estimate, 2), " (", -->
<!--       round(conf.low, 2), " - ", -->
<!--       round(conf.high, 2), ")" -->
<!--     ), -->
<!--     p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))), -->
<!--     std.error = round(std.error, 3) -->
<!--   ) |> -->
<!--   select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") -->
<!-- ``` -->

### Model comparison
```{r}
#| label: tbl-hiv-or-comparison
#| tbl-cap: "Logistic regression: HIV status — unadjusted vs adjusted odds ratios"

tbl_simple <- model_hiv_simple |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(ae_max_grade_pt ~ "Max AE grade (patient)")
  )

tbl_basic <- model_hiv_basic |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      ae_max_grade_pt ~ "Max AE grade (patient)",
      cancer_stage ~ "Cancer stage",
      chemo_type ~ "Chemo type",
      tx_intent ~ "Treatment intent",
      age ~ "Age",
      sex ~ "Sex"
    )
  )

tbl_merge(
  list(tbl_simple, tbl_basic),
  tab_spanner = c("Unadjusted", "Adjusted")
)
```

**Summary:** Both the unadjusted and adjusted models with HIV status as outcome indicate that each one unit increase in max AE grade across all treatment cycles is associated with between 28-34% higher odds of the patient being HIV positive. In the adjusted model, higher cancer stages were not reliably associated with being HIV positive. For chemo type there does seem to be a pretty clear difference in treatment options associated with HIV positive status. There does not appear to be any association between treatment intent and HIV status.

### Validation: model_hiv_basic

::: {.callout-note collapse="true" title="Validation: model_hiv_basic"}
```{r}
#| label: chk-hiv-assumptions

# Linearity of the log-odds
gic_models |>
  filter(!is.na(ae_max_grade_pt)) |>
  group_by(ae_max_grade_pt) |>
  summarize(prop_hiv = mean(hiv_status == "Positive", na.rm = TRUE)) |>
  mutate(log_odds = log(prop_hiv / (1 - prop_hiv))) |>
  ggplot(aes(x = ae_max_grade_pt, y = log_odds)) +
  geom_point() +
  geom_line() +
  labs(x = "Max AE grade (patient)", y = "Empirical log-odds of HIV+")

print("Linearity isn't amazing, but a straight line fit would show a reasonable fit.")

# Multicollinearity
check_vif(model_hiv_basic)

# Influential observations
plot(model_hiv_basic, which = 4)  # Cook's distance
```

```{r}
#| label: investigate-influential-obs

# Get row indices of most influential observations (Cook's D > threshold)
# Using 4/n as a common rule-of-thumb cutoff
n <- nrow(model_hiv_basic$model)
cooks_threshold <- 4 / n

influential_idx <- which(cooks.distance(model_hiv_basic) > cooks_threshold)
influential_ids <- gic_models$record_id[influential_idx]

cat("Influential observations (Cook's D > 4/n =", round(cooks_threshold, 4), "):\n")
cat("Row indices:", influential_idx, "\n")
cat("Record IDs:", influential_ids, "\n\n")

# Pull their covariate values to see what makes them unusual
gic_models |>
  filter(record_id %in% influential_ids) |>
  select(record_id, hiv_status, cancer_stage, chemo_type, tx_intent, ae_max_grade_pt)
```

All pts with a large Cook's D are HIV positive, but I haven't immediately spotted any other commonality amongst them to explain the group collectively being so influential, will have to do more investigation.
:::



## Cycles Prescribed as Outcome
```{r}
#| label: model-cycles-full

# Full cycles prescribed model
model_cycles_full <- glm(
  cycles_prescribed ~ hiv_status + cancer_site + cancer_stage + chemo_type + tx_intent + age,
  data   = gic_models,
  family = poisson
)
```

```{r}
#| label: model-cycles-no-site

# Cycles prescribed model without cancer site
model_cycles_no_site <- glm(
  cycles_prescribed ~ hiv_status + cancer_stage + chemo_type + tx_intent + age,
  data   = gic_models,
  family = poisson
)
```

```{r}
#| label: tbl-cycles-models
#| tbl-cap: "Poisson regression: cycles prescribed (exponentiated coefficients = rate ratios)"

tbl_full <- model_cycles_full |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      age ~ "Age (years)",
      cancer_site  ~ "Cancer Site",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Type",
      tx_intent    ~ "Treatment Intent"
    )
  ) |>
  bold_labels() |> 
  italicize_levels() |> 
  bold_p() |> 
  add_global_p(keep = TRUE) |> 
  add_vif(statistic = "aGVIF")

tbl_no_site <- model_cycles_no_site |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      age ~ "Age (years)",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Type",
      tx_intent    ~ "Treatment Intent"
    )
  ) |>
  bold_labels() |> 
  italicize_levels() |> 
  bold_p() |> 
  add_global_p(keep = TRUE) |> 
  add_vif(statistic = "aGVIF")

lrt_p <- anova(model_cycles_no_site, model_cycles_full, test = "LRT")$`Pr(>Chi)`[2]


tbl_merge(
  list(tbl_no_site, tbl_full),
  tab_spanner = c("Without cancer site", "With cancer site")
) |>
  as_gt() |>
  tab_source_note(
    source_note = paste0(
      "Model without cancer site: AIC = ", round(AIC(model_cycles_no_site), 1),
      ", N = ", nobs(model_cycles_no_site),
      " | Model with cancer site: AIC = ", round(AIC(model_cycles_full), 1),
      ", N = ", nobs(model_cycles_full)
    )
  ) |>
  tab_source_note(
    source_note = paste0(
      "LRT (with vs without cancer site): p = ",
      round(anova(model_cycles_no_site, model_cycles_full, 
                  test = "LRT")$`Pr(>Chi)`[2], 3)
    )
  )
  
```


# Toxicity-Related Hospitalizations
```{r}
#| label: prep-hosp-days

hosp_cols <- names(gic_models)[str_detect(names(gic_models), "^hosp_days_([1-9]|1[0-2])$")]

hosp_long <- gic_models |>
  select(record_id, hiv_status, cycles_given, age, sex, cancer_stage, chemo_type, all_of(hosp_cols)) |>
  pivot_longer(
    cols = all_of(hosp_cols),
    names_to = "cycle_var",
    values_to = "hosp_days"
  ) |> 
  mutate(
    cycle = as.integer(str_extract(cycle_var, "\\d+"))
  )

# Patient-level summaries
hosp_pt <- hosp_long |> 
  group_by(record_id, hiv_status, cycles_given) |> 
  summarize(
    hosp_days_total = sum(hosp_days, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  mutate(
    hosp_days_per_cycle_given = if_else(is.na(cycles_given) | cycles_given <= 0,
                                        NA_real_,
                                        hosp_days_total / cycles_given)
  )

```

## Summary by HIV Status
```{r}
#| label: tbl-hosp-by-pt
#| tbl-cap: "Hospital days attributable to toxicity by HIV status"

hosp_pt |>
  group_by(hiv_status) |>
  summarize(
    n_patients = n(),

    hosp_days_total_mean_sd = summ_median_iqr(hosp_days_total, 1),
    hosp_days_per_cycle_given_mean_sd = summ_median_iqr(hosp_days_per_cycle_given, 1),
    .groups = "drop"
  ) |>
  mutate(
    `HIV Status` = hiv_status,
    `Patients (N)` = n_patients,
    `Total hosp days, mean (SD)` = hosp_days_total_mean_sd,
    `Hosp days / cycles_given, mean (SD)` = hosp_days_per_cycle_given_mean_sd
  ) |>
  select(
    `HIV Status`, `Patients (N)`,
    `Total hosp days, mean (SD)`,
    `Hosp days / cycles_given, mean (SD)`
  ) |>
  kbl(booktabs = TRUE, align = "lcccc") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
#| label: prep-hosp-cycle-data

# Identify cycle-level column groups needed for hospitalization analysis
severe_cols <- names(gic_models)[str_detect(names(gic_models), "^ae_n_severe_([1-9]|1[0-2])$")]
max_grade_cols <- names(gic_models)[str_detect(names(gic_models), "^ae_max_grade_([1-9]|1[0-2])$")]
measured_cols <- names(gic_models)[str_detect(names(gic_models), "^ae_n_markers_measured_([1-9]|1[0-2])$")]

# Pivot each set of cycle cols to long
pivot_cycle_cols <- function(df, cols, value_name) {
  df |>
    select(record_id, all_of(cols)) |>
    pivot_longer(cols = all_of(cols), names_to = "cycle_var", values_to = value_name) |>
    mutate(cycle = as.integer(str_extract(cycle_var, "\\d+$"))) |>
    select(-cycle_var)
}

severe_long <- pivot_cycle_cols(gic_models, severe_cols,    "ae_n_severe")
max_grade_long <- pivot_cycle_cols(gic_models, max_grade_cols, "ae_max_grade")
measured_long <- pivot_cycle_cols(gic_models, measured_cols,  "ae_n_markers_measured")

# Join AE variables to hospitalization data
hosp_severe_cycle <- hosp_long |>
  left_join(severe_long,   by = c("record_id", "cycle")) |>
  left_join(measured_long, by = c("record_id", "cycle")) |>
  mutate(ae_measured = ae_n_markers_measured > 0)

hosp_maxgrade_cycle <- hosp_long |>
  left_join(max_grade_long, by = c("record_id", "cycle")) |>
  left_join(measured_long,  by = c("record_id", "cycle")) |>
  mutate(ae_measured = ae_n_markers_measured > 0)
```

## Hospitalizations by AE Severity
```{r}
#| label: tbl-hosp-vs-severe

hosp_severe_cycle |> 
  filter(ae_measured) |> 
  group_by(hiv_status, ae_n_severe) |> 
  summarize(
    n_cycles = sum(!is.na(hosp_days)),
    hosp_days_mean_sd = summ_median_iqr(hosp_days),
    .groups = "drop"
  ) |> 
  mutate(
    `HIV Status` = hiv_status,
    `# severe AEs in cycle` = ae_n_severe,
    `# Cycles` = n_cycles,
    `Hosp days, mean (SD); range` = hosp_days_mean_sd
  ) |> 
  select(`HIV Status`, `# severe AEs in cycle`, `# Cycles`, `Hosp days, mean (SD); range`) |> 
  kbl(booktabs = TRUE, align = "lccc") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
#| label: plt-hosp-vs-severe
#| fig-cap: "Mean hospital days vs number of severe AEs in the cycle, by HIV status"

hosp_severe_summ <- hosp_severe_cycle |>
  filter(ae_measured) |>
  group_by(hiv_status, ae_n_severe) |>
  summarize(
    mean_hosp_days = mean(hosp_days, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(hosp_severe_summ, aes(x = ae_n_severe, y = mean_hosp_days, color = hiv_status, group = hiv_status)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    x = "Number of severe AEs in cycle",
    y = "Mean hospital days (toxicity) in cycle",
    color = "HIV status"
  ) +
  theme_minimal()
```

```{r}
#| label: tbl-hosp-vs-max-grade
#| tbl-cap: "Hospital days vs max AE grade in cycle, by HIV status (among cycles with any AE grading recorded)"

hosp_maxgrade_cycle |>
  filter(ae_measured) |>
  group_by(hiv_status, ae_max_grade) |>
  summarize(
    n_cycles = sum(!is.na(hosp_days)),
    hosp_days_median_iqr = summ_median_iqr(hosp_days, 1),
    .groups = "drop"
  ) |>
  mutate(
    `HIV Status` = hiv_status,
    `Max AE grade in cycle` = ae_max_grade,
    `# Cycles` = n_cycles,
    `Hosp days, median (IQR)` = hosp_days_median_iqr
  ) |>
  select(`HIV Status`, `Max AE grade in cycle`, `# Cycles`, `Hosp days, median (IQR)`) |>
  kbl(booktabs = TRUE, align = "lccc") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
#| label: plt-hosp-vs-max-grade
#| fig-cap: "Mean hospital days vs max AE grade in cycle, by HIV status (among measured cycles)"

hosp_maxgrade_summ <- hosp_maxgrade_cycle |>
  filter(ae_measured) |>
  group_by(hiv_status, ae_max_grade) |>
  summarize(
    mean_hosp_days = mean(hosp_days, na.rm = TRUE),
    .groups = "drop"
  )

plt_hosp_vs_max_grade <- ggplot(hosp_maxgrade_summ, aes(x = ae_max_grade, y = mean_hosp_days, color = hiv_status, group = hiv_status)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 0:5) +
  labs(
    x = "Max AE grade in cycle",
    y = "Mean hospital days (toxicity) in cycle",
    color = "HIV status"
  ) +
  theme_minimal()

plt_hosp_vs_max_grade

pub_objects_analysis$plt_hosp_vs_max_grade <- save_figure(plt_hosp_vs_max_grade, "plt_hosp_vs_max_grade")
```

## Regression Models


```{r}
#| label: prep-hosp-regression


hosp_ae_cycle <- hosp_long |>
  left_join(
    ae_cycle_long |> select(record_id, cycle, ae_n_severe, ae_max_grade, ae_n_markers_measured),
    by = c("record_id", "cycle")
  ) |>
  mutate(ae_measured = ae_n_markers_measured > 0)


hosp_model_data <- hosp_ae_cycle |>
  filter(ae_measured)

hosp_model_data <- hosp_model_data |>
  mutate(any_hosp = as.integer(hosp_days > 0))
```

```{r}
#| label: model-hosp-any

# Any hospitalization this cycle (logistic)
model_any_hosp <- glmer(
  any_hosp ~ ae_max_grade + hiv_status + age + sex + (1 | record_id),
  data = hosp_model_data,
  family = binomial
)
```

OR table for whether a patient was hospitalized at all in a given cycle.
```{r}
#| label: tbl-any-hosp
#| tbl-cap: "Odds of any toxicity-related hospitalization in a cycle"

tbl_any_hosp <- model_any_hosp |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      ae_max_grade ~ "Max AE Grade in Cycle",
      hiv_status   ~ "HIV Status",
      age          ~ "Age (years)",
      sex          ~ "Sex"
    ),
    pvalue_fun = label_style_pvalue(digits = 3)
  ) |>
  tbl_drop_ref_rows() |>
  bold_p() |>
  bold_labels() |>
  italicize_levels()

tbl_any_hosp

obj <- save_table(slug = "tbl_any_hosp", gtsum = tbl_any_hosp)
pub_objects_analysis$tbl_any_hosp <- obj
```

#### Validation: model_any_hosp
::: {.callout-note collapse="true" title="Validation: model_any_hosp"}
```{r}
#| label: chk-any-hosp-vif

check_vif(model_any_hosp)
```
:::

```{r}
#| label: model-hosp-days

# Length of stay given hospitalization: unadjusted vs adjusted
model_hosp_days_simple <- lmer(
  hosp_days ~ hiv_status + (1 | record_id),
  data = hosp_model_data |> filter(hosp_days > 0)
)

model_hosp_days_adjusted <- lmer(
  hosp_days ~ hiv_status + age + sex + chemo_type + cancer_stage + (1 | record_id),
  data = hosp_model_data |> filter(hosp_days > 0)
)
```
AE severity has minimal impact on length of stay for either group once someone is already hospitalized, although HIV positive patients are expected to spend slightly longer in the hospital than HIV negative patients for each unit increase in max adverse event experienced.

```{r}
#| label: tbl-hosp-days-comparison
#| tbl-cap: "Length of stay among hospitalized cycles: unadjusted vs adjusted"

tbl_hosp_simple <- model_hosp_days_simple |>
  tbl_regression(
    exponentiate = FALSE,
    label = list(hiv_status ~ "HIV Status")
  ) |>
  tbl_drop_ref_rows()

tbl_hosp_adjusted <- model_hosp_days_adjusted |>
  tbl_regression(
    exponentiate = FALSE,
    label = list(
      hiv_status   ~ "HIV Status",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Regimen",
      age          ~ "Age (years)",
      sex          ~ "Sex"
    )
  ) |>
  tbl_drop_ref_rows()

tbl_hosp_days_comparison <- tbl_merge(
  list(tbl_hosp_simple, tbl_hosp_adjusted),
  tab_spanner = c("**Unadjusted Analysis**", "**Adjusted Analysis**"),
  quiet = TRUE
) |>
  modify_header(
    estimate_1 ~ "**Beta (Days)**",
    ci_1       ~ "**95% CI**",
    estimate_2 ~ "**Beta (Days)**",
    ci_2       ~ "**95% CI**"
  ) |>
  bold_labels() |>
  italicize_levels()

tbl_hosp_days_comparison

obj <- save_table(slug = "tbl_hosp_days_comparison", gtsum = tbl_hosp_days_comparison)
pub_objects_analysis$tbl_hosp_days_comparison <- obj
```

#### Validation: model_hosp_days_adjusted
::: {.callout-note collapse="true" title="Validation: model_hosp_days_adjusted"}
```{r}
#| label: chk-hosp-days-adjusted-vif

check_vif(model_hosp_days_adjusted)
```
:::


# Early Treatment Dropout

```{r}
#| label: prep-early-dropout

# Add early dropout flag to gic_models
gic_models <- gic_models |>
  mutate(
    early_dropout = as.integer(
      cycles_given <= early_dropout_cycles & cycles_prescribed > early_dropout_cycles
    )
  )
```

## Descriptive Overview

This section looks at whether the types of adverse events patients experience in their first few cycles are associated with stopping treatment early, and whether that relationship looks different for HIV-positive vs HIV-negative patients.

```{r}
#| label: tbl-early-dropout-by-hiv
#| tbl-cap: "Early treatment discontinuation by HIV status at two cycle cutoffs"

map_df(c(2, 4), function(cutoff) {
  gic_models |>
    mutate(dropout = as.integer(cycles_given <= cutoff & cycles_prescribed > cutoff)) |>
    group_by(hiv_status) |>
    summarize(
      n_total     = n(),
      n_dropout   = sum(dropout == 1, na.rm = TRUE),
      pct_dropout = round(100 * n_dropout / n_total, 1),
      .groups     = "drop"
    ) |>
    mutate(
      cutoff            = cutoff,
      `Early dropout, n (%)` = paste0(n_dropout, " (", pct_dropout, "%)")
    )
}) |>
  select(hiv_status, cutoff, `Total patients` = n_total, `Early dropout, n (%)`) |>
  pivot_wider(
    names_from  = cutoff,
    values_from = `Early dropout, n (%)`,
    names_glue  = "Cutoff ≤{cutoff} cycles"
  ) |>
  rename(`HIV Status` = hiv_status, `Total patients` = `Total patients`) |>
  gt() |>
  tab_spanner(
    label   = "Early dropout, n (%)",
    columns = starts_with("Cutoff")
  ) |>
  tab_footnote(
    footnote = "Early dropout defined as cycles given ≤ cutoff among patients prescribed more than the cutoff number of cycles.",
    location = cells_column_spanners()
  ) |>
  opt_footnote_marks(marks = "extended") |>
  cols_align(align = "center", columns = -`HIV Status`) |>
  cols_align(align = "left",   columns = `HIV Status`)
```

For each AE type, the predictor used is the max grade recorded across cycles 1 to 4. Where no grade was recorded for a given AE type in that window, it's been set to 0, on the assumption that if nothing was flagged there was probably nothing to flag. That's a more reasonable assumption for the liver panel markers, which Dr. Mazhindu said are usually only ordered when there's a clinical reason to, than it is for something like hemoglobin or platelets where a missing value might just mean the data wasn't entered. It was done this way mainly because doing a complete case analysis instead would have left only 97 patients and 26 early dropout events, which isn't really enough to work with. This is worth coming back to before the analysis is finalized.


## Global test for early dropout by HIV status

```{r}
#| label: tst-early-dropout-hiv
#| tbl-cap: "Early treatment discontinuation by HIV status"

# 2x2 table: early dropout x HIV status
dropout_hiv_table <- table(
  dropout = factor(gic_models$early_dropout, levels = c(0, 1), labels = c("Completed", "Early dropout")),
  HIV = gic_models$hiv_status
)

dropout_hiv_table |>
  as.data.frame.matrix() |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE) |>
  add_header_above(c(" " = 1, "HIV Status" = 2))

# Check expected counts to decide test
dropout_expected <- suppressWarnings(chisq.test(dropout_hiv_table)$expected)

if (any(dropout_expected < 5)) {
  dropout_test <- fisher.test(dropout_hiv_table)
  cat("Fisher's exact test (expected count < 5 in at least one cell)\n")
} else {
  dropout_test <- chisq.test(dropout_hiv_table, correct = FALSE)
  cat("Chi-squared test (all expected counts >= 5)\n")
}

dropout_test
```

```{r}
#| label: prep-ae-early-dropout

# Max grade per AE type across cycles 1-early_dropout_cycles
# Patients with no recorded grade for a given AE type in this window
# are excluded from that AE type's predictor (not treated as grade 0)
ae_early <- ae_long_grouped |>
  filter(cycle <= early_dropout_cycles, !is.na(grade)) |>
  group_by(record_id, ae_group) |>
  summarize(max_grade_early = max(grade, na.rm = TRUE), .groups = "drop") |>
  rename(ae_type = ae_group) |>
  left_join(
    gic_models |> select(record_id, hiv_status, early_dropout),
    by = "record_id"
  ) |>
  filter(!is.na(early_dropout))

# Wide format for modeling — one row per patient, one col per AE type
ae_types <- unique(ae_early$ae_type)

ae_early_wide <- ae_early |>
  pivot_wider(
    names_from  = ae_type,
    values_from = max_grade_early
  ) |>
  mutate(
    # All missing AE grades imputed to 0 — absence of a recorded grade treated
    # as no adverse event. Less defensible for some markers (e.g. plt, hb) than
    # for liver panels. Revisit when time permits
    across(all_of(ae_types), ~ replace_na(.x, 0))
  ) |>
  left_join(
    gic_models |> select(record_id, hiv_status, early_dropout),
    by = c("record_id", "hiv_status", "early_dropout")
  )
```


<!-- ```{r} -->
<!-- #| label: tbl-ae-early-dropout -->
<!-- #| tbl-cap: "Adverse event grades in cycles 1–4 by early dropout status and HIV status" -->

<!-- ae_early |> -->
<!--   mutate(ae_label = ae_type) |> -->
<!--   group_by(hiv_status, early_dropout, ae_label) |> -->
<!--   summarize( -->
<!--     n_any    = sum(max_grade_early > 0, na.rm = TRUE), -->
<!--     med      = median(max_grade_early, na.rm = TRUE), -->
<!--     n_severe = sum(max_grade_early >= 3, na.rm = TRUE), -->
<!--     .groups  = "drop" -->
<!--   ) |> -->
<!--   mutate( -->
<!--     dropout_label = if_else(early_dropout == 1, "Early dropout", "Completed >4 cycles"), -->
<!--     cell          = paste0(n_any, " / ", round(med, 1), " / ", n_severe) -->
<!--   ) |> -->
<!--   select(hiv_status, ae_label, dropout_label, cell) |> -->
<!--   pivot_wider( -->
<!--     names_from  = c(hiv_status, dropout_label), -->
<!--     values_from = cell, -->
<!--     names_glue  = "{hiv_status}: {dropout_label}" -->
<!--   ) |> -->
<!--   select( -->
<!--     ae_label, -->
<!--     `Negative: Completed >4 cycles`, -->
<!--     `Negative: Early dropout`, -->
<!--     `Positive: Completed >4 cycles`, -->
<!--     `Positive: Early dropout` -->
<!--   ) |> -->
<!--   rename(`Adverse event` = ae_label) |> -->
<!--   gt() |> -->
<!--   tab_spanner( -->
<!--     label   = "HIV Negative", -->
<!--     columns = starts_with("Negative") -->
<!--   ) |> -->
<!--   tab_spanner( -->
<!--     label   = "HIV Positive", -->
<!--     columns = starts_with("Positive") -->
<!--   ) |> -->
<!--   cols_label( -->
<!--     `Negative: Completed >4 cycles` = "Completed >4", -->
<!--     `Negative: Early dropout`        = "Early dropout", -->
<!--     `Positive: Completed >4 cycles`  = "Completed >4", -->
<!--     `Positive: Early dropout`         = "Early dropout" -->
<!--   ) |> -->
<!--   cols_align(align = "center", columns = -`Adverse event`) |> -->
<!--   cols_align(align = "left",   columns = `Adverse event`) |> -->
<!--   tab_footnote( -->
<!--     footnote = "Cells show: n with any AE / median max grade / n with severe AE (grade ≥ 3)", -->
<!--     location = cells_column_spanners() -->
<!--   ) |> -->
<!--   opt_footnote_marks(marks = "extended") -->
<!-- ``` -->



```{r}
#| label: tst-dropout-by-ae-type
#| tbl-cap: "Association between experiencing each AE type (any grade) in cycles 1–4 and early dropout"

dropout_ae_tests <- map_dfr(ae_types, function(at) {
  dat <- ae_early_wide |>
    filter(!is.na(early_dropout)) |>
    mutate(has_ae = as.integer(.data[[at]] > 0))
  
  tab <- table(dropout = dat$early_dropout, has_ae = dat$has_ae)
  
  if (nrow(tab) < 2 | ncol(tab) < 2) {
    return(tibble(ae_type = at, p = NA_real_, test = "—",
                  n_ae = NA, n_ae_dropout = NA, n_total = NA))
  }
  
  expected <- suppressWarnings(chisq.test(tab)$expected)
  if (any(expected < 5)) {
    result <- fisher.test(tab)
    test_used <- "Fisher's"
  } else {
    result <- chisq.test(tab, correct = FALSE)
    test_used <- "Chi-squared"
  }
  
  tibble(
    ae_type = at,
    n_total = nrow(dat),
    n_ae = sum(dat$has_ae == 1),
    n_ae_dropout = sum(dat$has_ae == 1 & dat$early_dropout == 1),
    p = result$p.value,
    test = test_used
  )
}) |>
  mutate(
    ae_label = ae_type,
    p_adj = p.adjust(p, method = "BH"),
    `p (raw)` = map_chr(p, format_p),
    `p (adj)` = map_chr(p_adj, format_p),
    `With AE, n (dropout)` = paste0(n_ae, " (", n_ae_dropout, ")")
  )

dropout_ae_tests |>
  select(`Adverse event` = ae_label, `With AE, n (dropout)`, Test = test, `p (raw)`, `p (adj)`) |>
  gt() |>
  cols_align(align = "left", columns = `Adverse event`) |>
  cols_align(align = "center", columns = -`Adverse event`) |>
  tab_footnote(
    footnote = paste0(
      "Tests whether having any occurrence of each AE type (grade > 0) in cycles 1–",
      early_dropout_cycles,
      " is associated with early dropout, pooling HIV groups. ",
      "Chi-squared or Fisher's exact as appropriate. p-values BH-adjusted across AE types."
    ),
    locations = cells_column_spanners()
  ) |>
  opt_footnote_marks(marks = "extended")
```

Tests whether, among all patients in the early dropout analysis, is having any occurrence of a specific AE type (grade > 0) in cycles 1-4 associated with whether they dropped out? Multiple comparisons correction applied because testing the same population multiple times across adverse event types.


The heatmap below gives a quick visual overview of where the grade differences between early dropout and completed patients are largest, and whether those patterns look similar across HIV groups.

```{r}
#| label: plt-ae-early-dropout-heatmap
#| fig-cap: "Median max ADE grade in cycles 1–4 by ADE category, early dropout status, and HIV status"

ae_early |>
  group_by(hiv_status, early_dropout, ae_type) |>
  summarize(
    median_grade = median(max_grade_early, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    dropout_label = if_else(early_dropout == 1, "Early dropout", "Completed ≥5 cycles"),
    facet_label   = paste0("HIV ", hiv_status)
  ) |>
  ggplot(aes(x = dropout_label, y = ae_type, fill = median_grade)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(median_grade, 1)), size = 3) +
  scale_fill_viridis_c(option = "magma", direction = -1, 
                       limits = c(0, 4), name = "Median\nmax grade") +
  facet_wrap(~ facet_label) +
  labs(
    x = NULL,
    y = "Adverse event category"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1),
    strip.text  = element_text(face = "bold")
  )
```

Before moving to the models, it's worth checking how many patients in each HIV group actually experienced each type of adverse event. AE types with very few patients in either group are excluded from the univariate models since there isn't enough data to estimate anything reliable for them.

```{r}
#| label: tbl-ae-coverage
#| tbl-cap: "AE grade coverage in cycles 1–4: patients with any recorded adverse event by HIV status"

ae_early_wide |>
  select(hiv_status, early_dropout, all_of(ae_types)) |>
  pivot_longer(
    cols      = all_of(ae_types),
    names_to  = "ae_type",
    values_to = "grade"
  ) |>
  group_by(ae_type, hiv_status) |>
  summarize(
    n_total    = n(),
    n_any_ae   = sum(grade > 0, na.rm = TRUE),
    pct_any_ae = round(100 * n_any_ae / n_total, 1),
    .groups    = "drop"
  ) |>
  mutate(
    ae_label = ae_type,
    `n (%)` = paste0(n_any_ae, " (", pct_any_ae, "%)")
  ) |>
  select(ae_label, hiv_status, `n (%)`) |>
  pivot_wider(
    names_from  = hiv_status,
    values_from = `n (%)`,
    names_prefix = "HIV "
  ) |>
  rename(`Adverse event` = ae_label) |>
  gt() |>
  tab_spanner(
    label   = "Patients with any AE (grade > 0)",
    columns = starts_with("HIV")
  ) |>
  cols_align(align = "left",   columns = `Adverse event`) |>
  cols_align(align = "center", columns = -`Adverse event`) |>
  tab_footnote(
    footnote = paste0(
      "Among patients with cycles given ≤ ", early_dropout_cycles,
      " or prescribed > ", early_dropout_cycles, " cycles. ",
      "AE types with 0 patients in either group cannot be evaluated in univariate models."
    ),
    location = cells_column_spanners()
  ) |>
  opt_footnote_marks(marks = "extended")
```

```{r}
#| label: prep-sparse-ae

# AE types with fewer than 5 patients with any non-zero grade in either
# HIV group — too sparse for reliable estimation
ae_cols <- intersect(ae_types, names(ae_early_wide))

sparse_ae <- ae_early_wide |>
  select(hiv_status, all_of(ae_cols)) |>
  pivot_longer(cols = all_of(ae_cols), names_to = "ae_type", values_to = "grade") |>
  group_by(ae_type, hiv_status) |>
  summarize(n_any = sum(grade > 0, na.rm = TRUE), .groups = "drop") |>
  group_by(ae_type) |>
  filter(any(n_any < 5)) |>
  pull(ae_type) |>
  unique()
```


## Univariate Logistic Models

```{r}
#| label: fit-univariate-ae-models

# One logistic model per AE type — early dropout ~ max grade in cycles 1-4 + HIV status
# Patients with no recorded grade for a given AE type are excluded from that model
univariate_results <- map_df(ae_types, function(at) {
  dat <- ae_early_wide |>
    filter(!is.na(.data[[at]]), !is.na(early_dropout), !is.na(hiv_status))

  if (sum(dat$early_dropout) < 10) {
    return(tibble(ae_type = at, note = "Insufficient early dropout cases", .rows = 1))
  }

  if (at %in% sparse_ae) {
    return(tibble(ae_type = at, note = "Sparse — excluded per coverage check", .rows = 1))
  }

  mod <- glm(
    as.formula(paste0("early_dropout ~ `", at, "` + hiv_status")),
    data   = dat,
    family = binomial
  )

  max_gvif <- check_vif(mod) |> pull(gvif) |> max(na.rm = TRUE)

  tidy(mod, conf.int = TRUE, exponentiate = TRUE) |>
    filter(str_detect(term, fixed(at))) |>
    mutate(ae_type = at, max_gvif = max_gvif)
}) |>
  filter(!is.na(estimate)) |>
  mutate(p_adj = p.adjust(p.value, method = "BH"))
```

#### Validation: univariate early-dropout AE models
::: {.callout-note collapse="true" title="Validation: univariate early-dropout AE models"}
Multicollinearity check across all univariate AE models (one GVIF per model, between the AE type's grade and HIV status, rather than a separate diagnostic block per AE type).
```{r}
#| label: chk-univariate-ae-vif
#| tbl-cap: "Multicollinearity check: max GVIF per univariate early-dropout model"

univariate_results |>
  select(`Adverse event` = ae_type, `Max GVIF` = max_gvif) |>
  mutate(Flag = if_else(`Max GVIF` > 5, "Check", "OK")) |>
  kbl(booktabs = TRUE, digits = 2) |>
  kable_styling(full_width = FALSE, position = "left")
```
:::

```{r}
#| label: tbl-univariate-ae-results
#| tbl-cap: "Univariate logistic regression: association between AE type max grade and early dropout (adjusted for HIV status)"

# Sample sizes per AE type for context
# Sample sizes per AE category for context
ae_sample_sizes <- ae_types |>
  map(function(at) {
    ae_early_wide |>
      filter(!is.na(early_dropout), !is.na(hiv_status)) |>
      group_by(hiv_status) |>
      summarize(
        n_nonzero        = sum(.data[[at]] > 0),
        n_events_nonzero = sum(.data[[at]] > 0 & early_dropout == 1),
        .groups          = "drop"
      ) |>
      mutate(ae_type = at, cell = paste0(n_nonzero, " (", n_events_nonzero, ")")) |>
      select(ae_type, hiv_status, cell) |>
      pivot_wider(names_from = hiv_status, values_from = cell, names_glue = "n_{hiv_status}")
  }) |>
  list_rbind()

univariate_results |>
  left_join(ae_sample_sizes, by = "ae_type") |>
  mutate(
    `OR (95% CI)` = paste0(round(estimate, 2), " (", round(conf.low, 2), "–", round(conf.high, 2), ")"),
    `p (raw)`     = map_chr(p.value, format_p),
    `p (adj)`     = map_chr(p_adj, format_p)
  ) |>
  arrange(ae_type) |>
  select(
    `Adverse event`  = ae_type,
    `HIV-: n (drop)` = n_Negative,
    `HIV+: n (drop)` = n_Positive,
    `OR (95% CI)`,
    `p (raw)`,
    `p (adj)`
  ) |>
  gt() |>
  cols_align(align = "left",   columns = `Adverse event`) |>
  cols_align(align = "center", columns = -`Adverse event`) |>
  tab_footnote(
    footnote = "OR: change in odds of early dropout per one-unit increase in max AE grade (cycles 1–4), holding HIV status constant. n (drop) = patients with any grade > 0 for this AE category / early dropouts among those. p-values BH-adjusted across AE categories.",
    locations = cells_column_labels(columns = `p (adj)`)
  ) |>
  opt_footnote_marks(marks = "extended")
```

Publication version: descriptive breakdown by dropout status merged with the HIV-adjusted univariate associations, for the AE types that passed the sparsity check above.
```{r}
#| label: tbl-early-dropout-final
#| tbl-cap: "Early dropout: descriptive breakdown and HIV-adjusted univariate associations"

final_ae_columns <- setdiff(ae_types, sparse_ae)

tbl_dropout_descriptive <- ae_early_wide |>
  select(early_dropout, hiv_status, all_of(final_ae_columns)) |>
  mutate(early_dropout = factor(early_dropout, labels = c("Completed Treatment", "Early Dropout"))) |>
  tbl_summary(
    by = early_dropout,
    missing = "no",
    label = list(hiv_status ~ "HIV Status")
  )

tbl_dropout_regression <- ae_early_wide |>
  select(early_dropout, hiv_status, all_of(final_ae_columns)) |>
  tbl_uvregression(
    method = glm,
    y = early_dropout,
    method.args = list(family = binomial),
    exponentiate = TRUE,
    include = all_of(final_ae_columns),
    formula = "{y} ~ {x} + hiv_status",
    label = map(final_ae_columns, ~ as.formula(paste0(".x ~ '", .x, " Max Grade'")))
  ) |>
  add_q(method = "BH")

table_early_dropout_final <- tbl_merge(
  list(tbl_dropout_descriptive, tbl_dropout_regression),
  tab_spanner = c("**Descriptive Breakdown**", "**Univariate Association**")
) |>
  bold_labels()

table_early_dropout_final

obj <- save_table(slug = "table_early_dropout_final", gtsum = table_early_dropout_final)
pub_objects_analysis$table_early_dropout_final <- obj
```

Forest plot of the same univariate associations (max grade in cycles 1–`r early_dropout_cycles`, adjusted for HIV status).
```{r}
#| label: plt-early-dropout-forest
#| fig-cap: "Impact of early-cycle toxicities on early treatment dropout"

plt_early_dropout_forest <- univariate_results |>
  filter(ae_type %in% final_ae_columns) |>
  ggplot(aes(x = estimate, y = reorder(ae_type, estimate))) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), height = 0.2, color = "midnightblue", linewidth = 0.7) +
  geom_point(color = "firebrick", size = 3) +
  scale_x_log10(breaks = c(0.5, 1, 2, 5, 10)) +
  labs(
    x = "Odds Ratio (95% CI) on Log Scale\n[Adjusted for HIV Status]",
    y = "Adverse Event Type (Max Grade Cycles 1-4)",
    title = "Impact of Early Cycle Toxicities on Early Treatment Dropout",
    subtitle = "Odds of dropping out per 1-unit increase in toxicity grade"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

plt_early_dropout_forest

pub_objects_analysis$plt_early_dropout_forest <- save_figure(plt_early_dropout_forest, "plt_early_dropout_forest")
```

## Early dropout associations split by HIV status

Requested 5-Jun-2026 (Tinashe, agreed by Audrey). The forest plot above adjusts
for HIV status but reports a single pooled estimate per adverse event. This
version fits each association **separately within each HIV group**, so readers
can see whether the two groups track together or whether an apparent difference
is small-n noise on the HIV-positive side.

Kept as a **separate figure** rather than adding a second bar per row to the
plot above, so both versions can be compared before one is chosen for the paper.

::: {.callout-warning}
The HIV-positive group is small. Many of these stratum-specific models will
either fail to converge or return very wide intervals, and that is itself the
finding: a difference between groups is only interpretable where both intervals
are reasonably tight. Models that could not be fit are listed beneath the
figure rather than silently dropped.
:::

```{r}
#| label: fit-univariate-ae-models-byhiv

# One logistic model per (AE type x HIV group). HIV status is not a covariate
# here because it defines the stratum.
univariate_by_hiv <- expand_grid(
  ae_type    = ae_types,
  hiv_group  = levels(factor(ae_early_wide$hiv_status))
) |>
  pmap_dfr(function(ae_type, hiv_group) {
    base <- tibble(ae_type = ae_type, hiv_group = hiv_group)

    if (ae_type %in% sparse_ae) {
      return(base |> mutate(note = "Sparse — excluded per coverage check"))
    }

    dat <- ae_early_wide |>
      filter(hiv_status == hiv_group,
             !is.na(.data[[ae_type]]), !is.na(early_dropout))

    # Need both outcome variation and predictor variation to estimate anything.
    if (nrow(dat) < 10 ||
        sum(dat$early_dropout) < 5 ||
        sum(dat$early_dropout == 0) < 5 ||
        dplyr::n_distinct(dat[[ae_type]]) < 2) {
      return(base |> mutate(
        n     = nrow(dat),
        note  = "Too few events or no grade variation in this stratum"
      ))
    }

    fit <- tryCatch(
      glm(as.formula(paste0("early_dropout ~ `", ae_type, "`")),
          data = dat, family = binomial),
      error = function(e) e
    )
    if (inherits(fit, "error")) {
      return(base |> mutate(n = nrow(dat),
                            note = paste("Model failed:", conditionMessage(fit))))
    }

    # conf.int = TRUE profiles the likelihood and can fail on separated fits;
    # fall back to Wald intervals rather than losing the row entirely.
    td <- tryCatch(
      tidy(fit, conf.int = TRUE, exponentiate = TRUE),
      error = function(e) tidy(fit, conf.int = FALSE, exponentiate = TRUE) |>
        mutate(conf.low = NA_real_, conf.high = NA_real_)
    )

    td |>
      filter(str_detect(term, fixed(ae_type))) |>
      mutate(ae_type = ae_type, hiv_group = hiv_group, n = nrow(dat),
             note = NA_character_)
  })

# BH correction applied within each HIV group, since each group is its own
# family of tests.
stopifnot("p.value" %in% names(univariate_by_hiv))

univariate_by_hiv <- univariate_by_hiv |>
  group_by(hiv_group) |>
  # n is set explicitly so the strata that could not be fit do not inflate the
  # denominator; p.adjust()'s default n = length(p) would count them as tests.
  mutate(p_adj = p.adjust(p.value, method = "BH",
                          n = sum(!is.na(p.value)))) |>
  ungroup()
```

```{r}
#| label: plt-early-dropout-forest-byhiv
#| fig-cap: "Association between early-cycle toxicity grade and early treatment dropout, estimated separately within each HIV group. Points are odds ratios per one-grade increase; bars are 95% confidence intervals. Compare with the pooled, HIV-adjusted figure above — overlapping intervals across the two groups indicate the pooled estimate is not masking a difference."
#| fig-height: 6

forest_byhiv_data <- univariate_by_hiv |>
  filter(is.na(note), !is.na(estimate), ae_type %in% final_ae_columns)

plt_early_dropout_forest_byhiv <- forest_byhiv_data |>
  ggplot(aes(x = estimate, y = reorder(ae_type, estimate), color = hiv_group)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high),
                height = 0.2, linewidth = 0.7,
                position = position_dodge(width = 0.6)) +
  geom_point(size = 2.6, position = position_dodge(width = 0.6)) +
  scale_x_log10() +
  labs(
    x = "Odds Ratio (95% CI) on Log Scale",
    y = "Adverse Event Type (Max Grade Cycles 1-4)",
    title = "Early Cycle Toxicities and Early Dropout, by HIV Status",
    subtitle = "Odds of dropping out per 1-unit increase in toxicity grade, estimated within each group",
    color = "HIV status"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

plt_early_dropout_forest_byhiv

pub_objects_analysis$plt_early_dropout_forest_byhiv <-
  save_figure(plt_early_dropout_forest_byhiv,
              "plt_early_dropout_forest_byhiv", height = 6)
```

```{r}
#| label: tbl-early-dropout-byhiv-notfit
#| tbl-cap: "Adverse event / HIV group combinations that could not be estimated, with the reason. Read alongside the figure above: an adverse event missing from one group is not evidence of no association there."

univariate_by_hiv |>
  filter(!is.na(note)) |>
  select(`Adverse event` = ae_type, `HIV group` = hiv_group,
         n, Reason = note) |>
  arrange(`Adverse event`, `HIV group`) |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left")
```



# Management Interventions: Transfusions and G-CSF

```{r}
#| label: prep-mgmt-analysis

# Load lmerTest to get p-values for linear mixed models
library(lmerTest)

# G-CSF can be indicated for either neutropenia (ANC) or leukopenia (WBC).
# Pull WBC grade per cycle and compute the per-cycle max of the two.
# Assigned to a new object rather than overwriting mgmt_cycle, so that
# re-running this chunk in an interactive session does not join wbc_by_cycle on
# twice and produce wbc_grade.x / wbc_grade.y.
wbc_by_cycle <- ae_long |>
  filter(ae_type == "wbc") |>
  select(record_id, cycle, wbc_grade = grade)

mgmt_cycle_mgmt <- mgmt_cycle |>
  left_join(wbc_by_cycle, by = c("record_id", "cycle")) |>
  mutate(gcsf_grade = pmax(anc_grade, wbc_grade, na.rm = TRUE))

mgmt_cycle <- mgmt_cycle_mgmt

# Anemia grade as a collapsed factor. Grades 3 and 4 are pooled throughout
# because grade 4 is rare and near-universally transfused.
add_anemia_grade_f <- function(d) {
  d |>
    mutate(
      anemia_grade_f = factor(anemia_grade),
      anemia_grade_f = fct_collapse(anemia_grade_f, "3-4" = c("3", "4"))
    )
}

mgmt_cycle_transfusion <- mgmt_cycle |>
  filter(!is.na(anemia_grade), !is.na(transfusion_given), anemia_grade >= 1) |>
  add_anemia_grade_f()

# --- Transfusion analysis strata (13-Jul-2026) -----------------------------
# Stratum 1: every cycle where the patient had a transfusion indication.
mgmt_grade2plus <- mgmt_cycle |>
  filter(anemia_grade >= 2, !is.na(transfusion_given)) |>
  add_anemia_grade_f()

# Stratum 2: every cycle belonging to a patient who was transfused at least
# once. Restricting to *cycles* where transfusion_given == 1 would leave the
# outcome constant and unmodellable, so the restriction is applied at the
# patient level: among patients for whom transfusion was demonstrably
# available, does HIV status predict receiving one in a given cycle?
tx_ever_ids <- mgmt_cycle |>
  filter(transfusion_given == 1) |>
  distinct(record_id) |>
  pull(record_id)

mgmt_tx_patients <- mgmt_cycle |>
  filter(record_id %in% tx_ever_ids, !is.na(transfusion_given),
         !is.na(anemia_grade)) |>
  add_anemia_grade_f()

# Retained sensitivity stratum: the 19-Jun-2026 grade-2-exactly specification.
mgmt_grade2 <- mgmt_cycle |>
  filter(anemia_grade == 2, !is.na(transfusion_given)) |>
  add_anemia_grade_f()

c(grade_2plus  = nrow(mgmt_grade2plus),
  ever_transfused = nrow(mgmt_tx_patients),
  grade_2_only = nrow(mgmt_grade2))
```

```{r}
mgmt_cycle_transfusion |>
  count(anemia_grade, hiv_status, transfusion_given) |>
  pivot_wider(names_from = transfusion_given, values_from = n, values_fill = 0) |>
  rename(`No transfusion` = `0`, `Transfusion given` = `1`) |>
  mutate(
    total = `No transfusion` + `Transfusion given`,
    pct_transfused = round(100 * `Transfusion given` / total, 1)
  ) |>
  arrange(anemia_grade, hiv_status)
```

```{r}
mgmt_cycle |>
  filter(!is.na(gcsf_grade), !is.na(csfg_indicated)) |>
  count(gcsf_grade, hiv_status, csfg_indicated) |>
  pivot_wider(names_from = csfg_indicated, values_from = n, values_fill = 0) |>
  rename(`No G-CSF` = `0`, `G-CSF given` = `1`) |>
  mutate(
    total = `No G-CSF` + `G-CSF given`,
    pct_gcsf = round(100 * `G-CSF given` / total, 1)
  ) |>
  arrange(gcsf_grade, hiv_status)
```





## Transfusion Received

Per the 13-Jul-2026 meeting, the transfusion analysis is fit on **two separate
strata** and the effect estimates compared for consistency, rather than pooled
into a single model:

1. **Cycles with anemia grade 2 or higher** — patients who had a clinical
   indication for transfusion.
2. **Cycles belonging to patients transfused at least once** — patients for whom
   transfusion was demonstrably available, so that variation in
   `transfusion_given` reflects a treatment decision rather than the absence of
   any indication at all.

The point of running both is that they fail in different directions. Stratum 1
conditions on indication but includes patients who were never going to be
transfused because blood was not available; stratum 2 conditions on
availability but selects on a consequence of the exposure. If the HIV estimate
is stable across the two, that consistency is the result worth reporting.

::: {.callout-warning collapse="true" title="Interpretation caveat: indication is not the same as eligibility"}
Grade 2+ patients are *supposed* to receive a transfusion, but blood supply
constraints in this setting mean a substantial share do not. The grade 2+
stratum should therefore be read as **"had an indication"**, not "was eligible
and available for transfusion". A null HIV effect in this stratum is consistent
with either equal treatment or with supply constraints dominating clinical
decision-making. This is stated in the table footnote as well.
:::

The previous grade-2-*exactly* model (19-Jun-2026) is retained below as a
sensitivity check, since it is the version the earlier results were based on.

```{r}
#| label: prep-tx-strata

# Levels that separate are identified from the data for each stratum rather
# than hard-coded, because the exclusions derived for the grade-2-only subset
# (LCV/5-FU, Gallbladder/Biliary/Pancreas) are not necessarily the levels that
# separate in a different stratum.
tx_adj_vars <- c("cancer_stage", "chemo_type", "cancer_site")

mgmt_grade2plus_adj <- mgmt_grade2plus |>
  drop_separated_levels("transfusion_given", tx_adj_vars)

mgmt_tx_patients_adj <- mgmt_tx_patients |>
  drop_separated_levels("transfusion_given", tx_adj_vars)

mgmt_grade2_adj <- mgmt_grade2 |>
  drop_separated_levels("transfusion_given", tx_adj_vars)
```

```{r}
#| label: fit-tx-strata

tx_formula <- transfusion_given ~ hiv_status + anemia_grade_f + cancer_stage +
  chemo_type + cancer_site + (1 | record_id)

model_tx_grade2plus <- glmer(tx_formula, data = mgmt_grade2plus_adj,
                             family = binomial)

model_tx_patients  <- glmer(tx_formula, data = mgmt_tx_patients_adj,
                            family = binomial)

# A *named list of strings*, not a list of formulas. gtsummary accepts either,
# but only this form can be subset by name -- which is needed below, where the
# anemia-grade entry is dropped for the sensitivity model (anemia grade is
# constant there, so the variable is not in that model and passing a label for
# it would error).
#
# Do not rewrite this as `list(hiv_status = hiv_status ~ "HIV Status", ...)`.
# A named list whose *values* are formulas makes broom.helpers read the value
# as the label itself, and it fails with "Can't combine <list> and <character>".
tx_labels <- list(
  hiv_status     = "HIV Status",
  anemia_grade_f = "Anemia Grade",
  cancer_stage   = "Cancer Stage",
  chemo_type     = "Chemotherapy Type",
  cancer_site    = "Cancer Site"
)

tbl_tx_grade2plus <- model_tx_grade2plus |>
  tbl_regression(exponentiate = TRUE, label = tx_labels)

tbl_tx_patients <- model_tx_patients |>
  tbl_regression(exponentiate = TRUE, label = tx_labels)
```

```{r}
#| label: tbl-tx-strata-comparison
#| tbl-cap: "Odds of Receiving a Blood Transfusion, by Analysis Stratum. Stratum 1 restricts to cycles with a transfusion indication (anemia grade 2+); stratum 2 restricts to patients transfused at least once. Consistency of the HIV estimate across strata, rather than either estimate alone, is the quantity of interest."

tbl_merge(
  list(tbl_tx_grade2plus, tbl_tx_patients),
  tab_spanner = c("**Anemia grade 2+**", "**Ever transfused**")
) |>
  bold_labels()
```

Levels excluded for separation, by stratum:

```{r}
#| label: tbl-tx-strata-exclusions

tibble(
  stratum = c("Anemia grade 2+", "Ever transfused", "Anemia grade 2 only"),
  n_cycles = c(nrow(mgmt_grade2plus_adj), nrow(mgmt_tx_patients_adj),
               nrow(mgmt_grade2_adj)),
  excluded = c(describe_dropped_levels(mgmt_grade2plus_adj),
               describe_dropped_levels(mgmt_tx_patients_adj),
               describe_dropped_levels(mgmt_grade2_adj))
)
```

#### Validation: stratified transfusion models

::: {.callout-note collapse="true" title="Validation: model_tx_grade2plus / model_tx_patients"}
```{r}
#| label: chk-tx-strata-vif

bind_rows(
  check_vif(model_tx_grade2plus) |> mutate(stratum = "Anemia grade 2+"),
  check_vif(model_tx_patients)   |> mutate(stratum = "Ever transfused")
) |>
  relocate(stratum)
```

A "VIF not computable" result most likely reflects residual sparse-cell
rank-deficiency in these small subsets rather than a coding error.

```{r}
#| label: chk-tx-strata-separation

bind_rows(
  check_separation(model_tx_grade2plus)$coefs |>
    mutate(stratum = "Anemia grade 2+"),
  check_separation(model_tx_patients)$coefs |>
    mutate(stratum = "Ever transfused")
) |>
  relocate(stratum) |>
  arrange(desc(extreme))
```
:::

### Sensitivity: anemia grade 2 only

The 19-Jun-2026 specification, kept so the grade 2+ result can be checked
against the estimates the team has already seen. Anemia grade is constant here
and so is dropped from the model.

```{r}
#| label: model-transfusion-given
#| tbl-cap: "Sensitivity Analysis: Odds of Receiving a Blood Transfusion, Anemia Grade 2 Only (Unadjusted vs Adjusted)"

model_tx_unadj <- glmer(
  transfusion_given ~ hiv_status + (1 | record_id),
  data = mgmt_grade2, family = binomial
)

model_tx_grade2 <- glmer(
  transfusion_given ~ hiv_status + cancer_stage + chemo_type + cancer_site +
    (1 | record_id),
  data = mgmt_grade2_adj, family = binomial
)

tbl_merge(
  list(
    model_tx_unadj |>
      tbl_regression(exponentiate = TRUE,
                     label = list(hiv_status ~ "HIV Status")),
    model_tx_grade2 |>
      tbl_regression(exponentiate = TRUE,
                     label = tx_labels[setdiff(names(tx_labels), "anemia_grade_f")])
  ),
  tab_spanner = c("**Unadjusted**", "**Adjusted**")
) |>
  bold_labels()
```

#### Validation: model_tx_grade2

::: {.callout-note collapse="true" title="Validation: model_tx_grade2"}
```{r}
#| label: chk-tx-grade2-vif

check_vif(model_tx_grade2)
```
:::


## Blood Units Given

For patients who were given transfusions, is HIV status associated with the number of units of blood given, adjusting for level of anemia

```{r}
#| label: model-transfusion-units
#| tbl-cap: "Linear Mixed Model: Transfusion Units Received (Among Transfused Cycles)"

model_tx_units <- lmer(
  transfusion_units ~ hiv_status + anemia_grade + (1 | record_id),
  data = mgmt_cycle |> filter(transfusion_given == 1, !is.na(anemia_grade))
)

model_tx_units |>
  tbl_regression(
    label = list(
      hiv_status ~ "HIV Status",
      anemia_grade ~ "Anemia Grade"
    )
  ) |>
  bold_labels() |>
  bold_p()
```

#### Validation: model_tx_units
::: {.callout-note collapse="true" title="Validation: model_tx_units"}
```{r}
#| label: chk-tx-units-vif

check_vif(model_tx_units)
```
:::

## G-CSF Administration

Modeled among cycles with a recorded neutropenia or leukopenia grade. The
myelosuppression severity predictor is `gcsf_grade`, the per-cycle maximum of
neutropenia (ANC) and leukopenia (WBC) grades, since G-CSF can be indicated by
either condition. `pmax(..., na.rm = TRUE)` is used so that a cycle with only
one grade recorded still contributes.

Age is included per the 13-Jul-2026 meeting: elderly patients are more likely to
have a count drop and therefore to be boosted. The remaining covariates (stage,
chemotherapy type, cancer site) mirror the transfusion model and follow
Tinashe's 5-Jun-2026 comment that cancer type, stage, and chemotherapy regimen
all plausibly drive neutropenia and hence G-CSF indication, whereas HIV's
influence is expected to be behavioural — clinicians may over-indicate G-CSF for
HIV-positive patients out of perceived immune risk.

```{r}
#| label: prep-gcsf-data

# Build the analytic dataset once so the separation screen, the model, and the
# validation checks all operate on exactly the same rows. Dropping incomplete
# cases here rather than letting glmer() do it silently means the n reported
# below is the n actually modelled.
gcsf_vars <- c("csfg_indicated", "hiv_status", "gcsf_grade", "age",
               "cancer_stage", "chemo_type", "cancer_site", "record_id")

mgmt_gcsf <- mgmt_cycle |>
  select(all_of(gcsf_vars)) |>
  drop_na()

# Grade 2+ subset, for parallelism with the transfusion analysis. 13-Jul noted
# that G-CSF indication rises sharply at grade 2 and above for both leukopenia
# and neutropenia.
mgmt_gcsf_grade2plus <- mgmt_gcsf |>
  filter(gcsf_grade >= 2)

c(all_cycles = nrow(mgmt_gcsf), grade_2plus = nrow(mgmt_gcsf_grade2plus))
```

#### Separation screen (do this before trusting the adjusted model)

::: {.callout-important collapse="true" title="Separation screen: G-CSF"}
Carried over from `docs/journal.md` (open since 19-Jun-2026): the transfusion
model turned out to be completely separated across `chemo_type` and
`cancer_site`, which is why it is now fit on a restricted stratum. The same
check has to happen here **before** the extra covariates are trusted, because
adding covariates to a separated model makes the fit worse, not better.

Any level flagged `separated = TRUE` below contributes no information and will
produce an odds ratio of 0 or Inf. If levels are flagged, the options are the
same as for transfusion: drop those levels at the model-fitting stage, collapse
them into a coarser category, or move to Firth's penalised likelihood
(`logistf`) / a weakly-informative Bayesian fit.

```{r}
#| label: chk-gcsf-separation

# Cell-level screen on the analytic data, before fitting anything.
mgmt_gcsf |>
  mutate(gcsf_grade_f = factor(gcsf_grade)) |>
  select(csfg_indicated, hiv_status, gcsf_grade_f,
         cancer_stage, chemo_type, cancer_site) |>
  pivot_longer(-csfg_indicated, names_to = "term", values_to = "level",
               values_transform = as.character) |>
  group_by(term, level) |>
  summarise(
    n         = n(),
    n_gcsf    = sum(csfg_indicated == 1, na.rm = TRUE),
    .groups   = "drop"
  ) |>
  mutate(
    pct_gcsf  = round(100 * n_gcsf / n, 1),
    separated = n_gcsf == 0 | n_gcsf == n
  ) |>
  arrange(desc(separated), term, level)
```
:::

```{r}
#| label: model-gcsf-given
#| tbl-cap: "Logistic Mixed Model: Odds of Receiving G-CSF (All Graded Cycles)"

model_gcsf_given <- glmer(
  csfg_indicated ~ hiv_status + gcsf_grade + age +
    cancer_stage + chemo_type + cancer_site + (1 | record_id),
  data   = mgmt_gcsf,
  family = binomial
)

tbl_gcsf_all <- model_gcsf_given |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      gcsf_grade   ~ "Max Neutropenia/Leukopenia Grade",
      age          ~ "Age (years)",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Type",
      cancer_site  ~ "Cancer Site"
    )
  )

tbl_gcsf_all |>
  bold_labels() |>
  bold_p()
```

#### Validation: model_gcsf_given

::: {.callout-note collapse="true" title="Validation: model_gcsf_given"}
```{r}
#| label: chk-gcsf-given-vif

check_vif(model_gcsf_given)
```

Post-fit separation signal — coefficients with an implausibly large magnitude or
standard error. These should agree with the pre-fit cell screen above.

```{r}
#| label: chk-gcsf-given-separation

check_separation(model_gcsf_given)$coefs |>
  arrange(desc(extreme))
```
:::

### G-CSF at grade 2 and above

Parallel to the transfusion analysis: restricted to cycles where the patient had
a grade 2 or higher neutropenia/leukopenia, i.e. cycles where G-CSF had a
plausible indication. Compare the HIV estimate here against the all-cycles model
— if they disagree substantially, the all-cycles estimate is being driven by the
large number of low-grade cycles where G-CSF was never on the table.

```{r}
#| label: model-gcsf-grade2plus
#| tbl-cap: "Odds of Receiving G-CSF: All Graded Cycles vs Grade 2+ Cycles"

model_gcsf_grade2plus <- glmer(
  csfg_indicated ~ hiv_status + gcsf_grade + age +
    cancer_stage + chemo_type + cancer_site + (1 | record_id),
  data   = mgmt_gcsf_grade2plus,
  family = binomial
)

tbl_gcsf_grade2plus <- model_gcsf_grade2plus |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      gcsf_grade   ~ "Max Neutropenia/Leukopenia Grade",
      age          ~ "Age (years)",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Type",
      cancer_site  ~ "Cancer Site"
    )
  )

tbl_merge(
  list(tbl_gcsf_all, tbl_gcsf_grade2plus),
  tab_spanner = c("**All graded cycles**", "**Grade 2+ only**")
) |>
  bold_labels()
```

#### Validation: model_gcsf_grade2plus

::: {.callout-note collapse="true" title="Validation: model_gcsf_grade2plus"}
```{r}
#| label: chk-gcsf-grade2plus-vif

check_vif(model_gcsf_grade2plus)
```

```{r}
#| label: chk-gcsf-grade2plus-separation

check_separation(model_gcsf_grade2plus)$cells |>
  filter(separated)
```
:::

```{r}
#| label: save-pub-objects-analysis

save_rds_safe(pub_objects_analysis, onedrive_data_path("processed", "pub_objects_analysis.rds"))
```

