---
title: "GI Cancer: Analysis"
bibliography: references.bib
execute:
  # Collaborators read this document for results, not code. Warnings and
  # messages are suppressed here and the information that matters -- model
  # formulas, convergence status, separation screens -- is printed deliberately
  # instead. When working on the analysis, run chunks interactively or flip
  # these to true; do not rely on the rendered document to surface problems.
  echo: false
  warning: false
  message: false
format:
  html:
    toc-depth: 4
    code-links:
      - text: GitHub Page
        icon: github
        href: "https://github.com/EDukeChase/gi-cancer_project/blob/main/4%20-%20Statistical%20Analysis.qmd"
---
```{r}
#| label: setup
#| warning: false

source("R/setup.R")

# Load Data
gic_descriptive <- readRDS(onedrive_data_path("processed", "gic_descriptive.rds"))
gic_models <- gic_descriptive$gic_models
ae_long <- gic_descriptive$ae_long
ae_long_grouped <- gic_descriptive$ae_long_grouped
ae_cycle_long <- gic_descriptive$ae_cycle_long
ae_cycle_measured <- gic_descriptive$ae_cycle_measured
cycle_summary <- gic_descriptive$cycle_summary
cycle_summary_grouped <- gic_descriptive$cycle_summary_grouped
grade_cols <- gic_descriptive$grade_cols
mgmt_long <- gic_descriptive$mgmt_long
mgmt_cycle <- gic_descriptive$mgmt_cycle
rm(gic_descriptive)

# Finalized tables/figures destined for the publication doc (05), keyed by slug.
# Each save_table()/save_figure() call below also appends its result here.
pub_objects_analysis <- list()
```

# Model Building
## Severe Adverse Events as Outcome
### Unadjusted model
```{r}
#| label: model-any-simple

model_any_simple <- glm(ae_any_pt ~ hiv_status,
                        data = gic_models, family = binomial)

# summary(model_any_simple)
```
Invalid model due to having a 100% adverse event rate in the HIV-positive group.

### Adjusted model
```{r}
#| label: model-severe-simple


model_severe_simple <- glm(ae_severe_any_pt ~ hiv_status,
                           data = gic_models, family = binomial)
```

```{r}
#| label: tbl-severe-simple-or-clean
#| tbl-cap: "Association between HIV Status and Severe Adverse Events (Grades 3-4)"

tidy(model_severe_simple, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  # Clean up the name and prepare for the reference row
  mutate(
    Term = str_replace(term, "hiv_statusPositive", "HIV Positive"),
    `OR (95% CI)` = paste0(round(estimate, 2), " (",
                           round(conf.low, 2), " - ",
                           round(conf.high, 2), ")"),
    p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))),
    std.error = round(std.error, 3)
  ) |>
  select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |>
  # Add the reference row at the top
  add_row(Term = "HIV Negative (Reference)",
          `OR (95% CI)` = "1.00",
          `Std. Error` = NA,
          `P-value` = NA,
          .before = 1) |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left")
```



### Multivariate Logistic: Full treatment period
#### Basic logistic model of any adverse events by HIV status
::: {.callout-important title="Deprecated model — not a result"}
`model_any_basic` **is not a valid model and must not be read as a finding.**
Every patient in the HIV-positive group experienced at least one adverse event,
so the outcome is completely separated on HIV status and the coefficient is
uninterpretable.

It is retained only because the 100% event rate is itself worth stating in the
paper. Note that this is why age was *not* added to it under the 15-May-2026
"add age to remaining models" item: the problem is separation, not a missing
covariate, and adding covariates to a separated model makes the fit worse.

**Open decision for the team:** delete this chunk outright, or keep it as the
source of the "100% of HIV-positive patients experienced an adverse event"
sentence. If kept, it should not appear in the publication figures document.
:::

```{r}
#| label: tbl-any-ade-rate
#| tbl-cap: "Patients experiencing any adverse event at any point during treatment, by HIV status. This is the substantive content of the deprecated model above: the HIV-positive rate is 100%, which is precisely why that model cannot be estimated."

model_any_basic <- glm(ae_any_pt ~ hiv_status + cancer_stage + chemo_type + tx_intent,
                       data = gic_models, family = binomial)

# The model fit is retained, but a raw glm summary is not what a reader needs
# here -- it renders as a wall of console text describing a model that is not
# interpretable anyway. The rate itself is the reportable fact.
gic_models |>
  group_by(hiv_status) |>
  summarize(
    `Patients (n)`        = n(),
    `With any ADE, n (%)` = paste0(sum(ae_any_pt == 1, na.rm = TRUE), " (",
                                   round(100 * mean(ae_any_pt == 1, na.rm = TRUE), 1), "%)"),
    .groups = "drop"
  ) |>
  rename(`HIV status` = hiv_status) |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left")
```

#### Basic logicstic model of any severe adverse events by HIV status


```{r}
#| label: model-severe-basic

model_severe_basic <- glm(ae_severe_any_pt ~ hiv_status + cancer_stage + chemo_type + tx_intent + age + sex,
                       data = gic_models, family = binomial)
```

```{r}
#| label: show-formula-severe-basic
#| results: asis

show_formula(model_severe_basic)
```

```{r}
#| label: tbl-severe-multivariate-or
#| tbl-cap: "Adjusted Odds Ratios for Severe Adverse Events (Grades 3-4)"

model_severe_basic |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      age ~ "Age (years)",
      sex ~ "Sex",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Regimen",
      tx_intent    ~ "Treatment Intent"
    ),
    pvalue_fun = label_style_pvalue(
      digits = 3
    )
  ) |>
  tbl_drop_ref_rows() |>
  bold_labels() |>
  italicize_levels() |>
  bold_p() |>
  add_global_p(keep = TRUE) |>
  add_glance_source_note(include = c(nobs, AIC))
```

#### Validation: model_severe_basic
::: {.callout-note collapse="true" title="Validation: model_severe_basic"}
```{r}
#| label: chk-severe-basic-vif

check_vif(model_severe_basic)
```

GVIF is the generalised version of VIF. Ordinary VIF gives one number per
model *coefficient*, so a categorical variable like cancer site gets a separate
value for each of its levels and the answer changes depending on which level you
picked as the reference — GVIF gives one number per *variable* instead, which is
what you actually want. Anything above roughly 5 is worth a look, and here it
usually means two clinical variables are carrying the same information (cancer
site and chemotherapy regimen, for instance, since regimen is largely chosen by
site).
:::

Publication version: adjusted odds ratios for severe ADE. The unadjusted estimate is quoted in the caption.
```{r}
#| label: show-formula-severe-comparison
#| results: asis

show_formula(model_severe_simple, "Unadjusted")
show_formula(model_severe_basic, "Adjusted")

# Built here, not in the table chunk below: knitr evaluates a chunk's options
# (including !expr captions) before running that chunk's body, so a caption
# variable assigned in the same chunk would not exist yet.
severe_caption <- paste0(
  "Adjusted odds of a severe (grade 3-4) adverse event by HIV status. ",
  "The unadjusted model, containing HIV status only, was closely consistent: ",
  hiv_effect_text(model_severe_simple), "."
)

pub_objects_analysis$severe_caption <- severe_caption
```

```{r}
#| label: tbl-severe-or-comparison
#| tbl-cap: !expr severe_caption

tbl_severe_simple <- model_severe_simple |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(hiv_status ~ "HIV Status"),
    pvalue_fun = label_style_pvalue(digits = 3)
  ) |>
  tbl_drop_ref_rows() |>
  bold_p()

tbl_severe_adjusted <- model_severe_basic |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Regimen",
      tx_intent    ~ "Treatment Intent",
      age          ~ "Age (years)",
      sex          ~ "Sex"
    ),
    pvalue_fun = label_style_pvalue(digits = 3)
  ) |>
  tbl_drop_ref_rows() |>
  bold_p()

# 5-Jun-2026: the unadjusted column is not shown, because it was consistent
# with the adjusted one and took a full column to say so. The unadjusted model
# is still fit, and its HIV estimate is quoted in the caption below so the
# "consistent with" claim carries an actual number rather than an assertion.
tbl_severe_or_comparison <- tbl_severe_adjusted |>
  bold_labels() |>
  italicize_levels()


tbl_severe_or_comparison

obj <- save_table(slug = "tbl_severe_or_comparison", gtsum = tbl_severe_or_comparison)
pub_objects_analysis$tbl_severe_or_comparison <- obj
```


<!-- Commented out 2026-07-28: Audrey prefers a single test type across the
     analysis rather than a mix, and these chi-squared tests duplicate what the
     Fisher-based tests elsewhere already cover. Contingency tables and the
     tests that go with them are retained in "3 - Descriptive Statistics.qmd".
     Uncomment only if a specific question needs them. -->

<!-- ### $\chi^2$ Tests (HIV vs Early Events) -->

<!-- ```{r} -->
<!-- #| label: tbl-any-early-contingency -->
<!-- #| tbl-cap: "Contingency table of early adverse events (cycles 1-2) vs HIV status" -->

<!-- table_any_early <- table(early_ae = factor(gic_models$ae_any_early, levels = c(0, 1), labels = c("No", "Yes")), HIV = gic_models$hiv_status) -->

<!-- table_any_early |> -->
<!--   as.data.frame.matrix() |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") |> -->
<!--   add_header_above(c("Any Early ADE" = 1, "HIV Status" = 2)) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- #| label: tbl-severe-early-contingency -->
<!-- #| tbl-cap: "Contingency table of severe (grade 3-4) adverse events in cycles 1-2 versus HIV status." -->

<!-- table_severe_early <- table( -->
<!--   severe_early_ae = factor(gic_models$ae_severe_any_early, -->
<!--                            levels = c(0, 1), labels = c("No", "Yes")), -->
<!--   HIV = gic_models$hiv_status -->
<!-- ) -->

<!-- table_severe_early |> -->
<!--   as.data.frame.matrix() |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") |> -->
<!--   add_header_above(c("Severe Early ADE" = 1, "HIV Status" = 2)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- #| label: tbl-chi-sq-tests -->
<!-- #| tbl-cap: "Tests of association between early adverse events and HIV status. Test choice follows the expected-cell-count rule used throughout: Fisher's exact where any expected count is below 5, chi-squared otherwise." -->

<!-- # These tests were previously commented out, so the section rendered the -->
<!-- # contingency tables with no result attached to them. -->
<!-- bind_rows( -->
<!--   test_contingency(table_any_early, -->
<!--                    "Any ADE in cycles 1-2 vs HIV status"), -->
<!--   test_contingency(table_severe_early, -->
<!--                    "Severe ADE in cycles 1-2 vs HIV status") -->
<!-- ) |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") -->
<!-- ``` -->

## HIV Status as Outcome
### Unadjusted model
```{r}
#| label: tbl-hiv-simple
#| tbl-cap: "Unadjusted logistic regression of HIV status on maximum adverse event grade across all treatment cycles."

model_hiv_simple <- glm(hiv_status ~ ae_max_grade_pt, data = gic_models, family = binomial)

model_hiv_simple |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      ae_max_grade_pt ~ "Max ADE grade"
    ),
    pvalue_fun = label_style_pvalue(
      digits = 3
    )
  ) |>
  tbl_drop_ref_rows() |>
  bold_labels()
```


<!-- ```{r} -->
<!-- #| label: tbl-hiv-simple-or -->
<!-- #| tbl-cap: "Logistic regression: HIV status vs mean max ADE grade (unadjusted)" -->

<!-- tidy(model_hiv_simple, conf.int = TRUE, exponentiate = TRUE) |> -->
<!--   filter(term != "(Intercept)") |> -->
<!--   mutate( -->
<!--     Term = case_when( -->
<!--       term == "ae_mean_max_grade" ~ "Mean max ADE grade per measured cycle", -->
<!--       TRUE ~ term -->
<!--     ), -->
<!--     `OR (95% CI)` = paste0( -->
<!--       round(estimate, 2), " (", -->
<!--       round(conf.low, 2), " - ", -->
<!--       round(conf.high, 2), ")" -->
<!--     ), -->
<!--     p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))), -->
<!--     std.error = round(std.error, 3) -->
<!--   ) |> -->
<!--   select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") -->
<!-- ``` -->




### Adjusted model

```{r}
#| label: model-hiv-basic

model_hiv_basic <- glm(
  hiv_status ~ ae_max_grade_pt + cancer_stage + chemo_type + tx_intent + sex + age,
  data = gic_models,
  family = binomial
)
```

<!-- ```{r} -->
<!-- #| label: tbl-hiv-basic-or -->
<!-- #| tbl-cap: "Logistic regression: HIV status vs mean max ADE grade (adjusted for stage, chemo type, and intent)" -->

<!-- tidy(model_hiv_basic, conf.int = TRUE, exponentiate = TRUE) |> -->
<!--   filter(term != "(Intercept)") |> -->
<!--   mutate( -->
<!--     Term = case_when( -->
<!--       term == "ae_mean_max_grade" ~ "Mean max ADE grade per measured cycle", -->
<!--       TRUE ~ term -->
<!--     ), -->
<!--     `OR (95% CI)` = paste0( -->
<!--       round(estimate, 2), " (", -->
<!--       round(conf.low, 2), " - ", -->
<!--       round(conf.high, 2), ")" -->
<!--     ), -->
<!--     p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))), -->
<!--     std.error = round(std.error, 3) -->
<!--   ) |> -->
<!--   select(Term, `OR (95% CI)`, `Std. Error` = std.error, `P-value` = p.value) |> -->
<!--   kbl(booktabs = TRUE) |> -->
<!--   kable_styling(full_width = FALSE, position = "left") -->
<!-- ``` -->

### Model comparison
```{r}
#| label: show-formula-hiv
#| results: asis

show_formula(model_hiv_simple, "Unadjusted")
show_formula(model_hiv_basic, "Adjusted")
```

```{r}
#| label: tbl-hiv-or-comparison
#| tbl-cap: "Logistic regression: HIV status — unadjusted vs adjusted odds ratios"

tbl_simple <- model_hiv_simple |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(ae_max_grade_pt ~ "Max ADE grade (patient)")
  ) |>
  tbl_drop_ref_rows()

tbl_basic <- model_hiv_basic |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      ae_max_grade_pt ~ "Max ADE grade (patient)",
      cancer_stage ~ "Cancer stage",
      chemo_type ~ "Chemo type",
      tx_intent ~ "Treatment intent",
      age ~ "Age",
      sex ~ "Sex"
    )
  ) |>
  tbl_drop_ref_rows()

tbl_merge(
  list(tbl_simple, tbl_basic),
  tab_spanner = c("Unadjusted", "Adjusted")
)
```

**Summary:** Both the unadjusted and adjusted models with HIV status as outcome indicate that each one unit increase in max ADE grade across all treatment cycles is associated with between 28-34% higher odds of the patient being HIV positive. In the adjusted model, higher cancer stages were not reliably associated with being HIV positive. For chemo type there does seem to be a pretty clear difference in treatment options associated with HIV positive status. There does not appear to be any association between treatment intent and HIV status.

### Validation: model_hiv_basic

::: {.callout-note collapse="true" title="Validation: model_hiv_basic"}
```{r}
#| label: chk-hiv-assumptions

# Linearity of the log-odds
gic_models |>
  filter(!is.na(ae_max_grade_pt)) |>
  group_by(ae_max_grade_pt) |>
  summarize(prop_hiv = mean(hiv_status == "Positive", na.rm = TRUE)) |>
  mutate(log_odds = log(prop_hiv / (1 - prop_hiv))) |>
  ggplot(aes(x = ae_max_grade_pt, y = log_odds)) +
  geom_point() +
  geom_line() +
  labs(x = "Max ADE grade (patient)", y = "Empirical log-odds of HIV+")

# Multicollinearity
check_vif(model_hiv_basic)

# Influential observations
plot(model_hiv_basic, which = 4)  # Cook's distance
```

Linearity of the log-odds is not perfect, but a straight-line fit is reasonable
over the observed range of grades.

```{r}
#| label: prep-influential-obs

# Cook's D > 4/n, the usual rule-of-thumb cutoff.
n_hiv_basic     <- nrow(model_hiv_basic$model)
cooks_threshold <- 4 / n_hiv_basic
influential_idx <- which(cooks.distance(model_hiv_basic) > cooks_threshold)
influential_ids <- gic_models$record_id[influential_idx]

influential_caption <- paste0(
  "Observations flagged as influential in the adjusted HIV-status model, by ",
  "Cook's distance above the 4/n cutoff (", round(cooks_threshold, 4), "). ",
  length(influential_ids), " of ", n_hiv_basic, " observations flagged. ",
  "Flagged is not the same as erroneous; this is a prompt to look, not a ",
  "reason to exclude."
)
```

```{r}
#| label: tbl-influential-obs
#| tbl-cap: !expr influential_caption

# Cook's distance above the conventional 4/n cutoff. Previously reported with
# cat(), which rendered as unlabelled console text once code stopped being
# shown; the threshold and counts now live in the caption instead.
gic_models |>
  filter(record_id %in% influential_ids) |>
  select(record_id, hiv_status, cancer_stage, chemo_type, tx_intent,
         ae_max_grade_pt) |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left")
```

All pts with a large Cook's D are HIV positive, but I haven't immediately spotted any other commonality amongst them to explain the group collectively being so influential, will have to do more investigation.
:::



## Cycles Prescribed as Outcome
```{r}
#| label: model-cycles-full

# Full cycles prescribed model
model_cycles_full <- glm(
  cycles_prescribed ~ hiv_status + cancer_site + cancer_stage + chemo_type + tx_intent + age,
  data   = gic_models,
  family = poisson
)
```

```{r}
#| label: model-cycles-no-site

# Cycles prescribed model without cancer site
model_cycles_no_site <- glm(
  cycles_prescribed ~ hiv_status + cancer_stage + chemo_type + tx_intent + age,
  data   = gic_models,
  family = poisson
)
```

```{r}
#| label: show-formula-cycles
#| results: asis

show_formula(model_cycles_no_site, "Without cancer site")
show_formula(model_cycles_full, "With cancer site")
```

```{r}
#| label: tbl-cycles-models
#| tbl-cap: "Poisson regression: cycles prescribed (exponentiated coefficients = rate ratios)"

tbl_full <- model_cycles_full |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      age ~ "Age (years)",
      cancer_site  ~ "Cancer Site",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Type",
      tx_intent    ~ "Treatment Intent"
    )
  ) |>
  tbl_drop_ref_rows() |>
  bold_labels() |> 
  italicize_levels() |> 
  bold_p() |> 
  add_global_p(keep = TRUE) |> 
  add_vif(statistic = "aGVIF")

tbl_no_site <- model_cycles_no_site |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      hiv_status   ~ "HIV Status",
      age ~ "Age (years)",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Type",
      tx_intent    ~ "Treatment Intent"
    )
  ) |>
  tbl_drop_ref_rows() |>
  bold_labels() |> 
  italicize_levels() |> 
  bold_p() |> 
  add_global_p(keep = TRUE) |> 
  add_vif(statistic = "aGVIF")

lrt_p <- anova(model_cycles_no_site, model_cycles_full, test = "LRT")$`Pr(>Chi)`[2]


tbl_merge(
  list(tbl_no_site, tbl_full),
  tab_spanner = c("Without cancer site", "With cancer site")
) |>
  as_gt() |>
  tab_source_note(
    source_note = paste0(
      "Model without cancer site: AIC = ", round(AIC(model_cycles_no_site), 1),
      ", N = ", nobs(model_cycles_no_site),
      " | Model with cancer site: AIC = ", round(AIC(model_cycles_full), 1),
      ", N = ", nobs(model_cycles_full)
    )
  ) |>
  tab_source_note(
    source_note = paste0(
      "LRT (with vs without cancer site): p = ",
      round(anova(model_cycles_no_site, model_cycles_full, 
                  test = "LRT")$`Pr(>Chi)`[2], 3)
    )
  )
  
```


# Toxicity-Related Hospitalizations
```{r}
#| label: prep-hosp-days

hosp_cols <- names(gic_models)[str_detect(names(gic_models), "^hosp_days_([1-9]|1[0-2])$")]

hosp_long <- gic_models |>
  select(record_id, hiv_status, cycles_given, age, sex, cancer_stage, chemo_type, all_of(hosp_cols)) |>
  pivot_longer(
    cols = all_of(hosp_cols),
    names_to = "cycle_var",
    values_to = "hosp_days"
  ) |> 
  mutate(
    cycle = as.integer(str_extract(cycle_var, "\\d+"))
  )

# Patient-level summaries
hosp_pt <- hosp_long |> 
  group_by(record_id, hiv_status, cycles_given) |> 
  summarize(
    hosp_days_total = sum(hosp_days, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  mutate(
    hosp_days_per_cycle_given = if_else(is.na(cycles_given) | cycles_given <= 0,
                                        NA_real_,
                                        hosp_days_total / cycles_given)
  )

```

## Summary by HIV Status
```{r}
#| label: tbl-hosp-by-pt
#| tbl-cap: "Hospital days attributable to toxicity, per patient, by HIV status."
#| results: asis

# Was calling summ_median_iqr() under headers that said "mean (SD)". Hospital
# days are heavily zero-inflated, so the median was 0 (0-0) in every cell --
# true, but uninformative. Switched to mean (SD) with the range, which matches
# the headers and does separate the groups. The distribution is right-skewed;
# read the mean alongside the range rather than on its own.
hosp_by_pt <- hosp_pt |>
  group_by(hiv_status) |>
  summarize(
    `Patients (N)`                        = as.character(n()),
    `Total hosp days, mean (SD); range`   = summ_mean_sd_range(hosp_days_total, 1),
    `Hosp days per cycle given, mean (SD); range` =
      summ_mean_sd_range(hosp_days_per_cycle_given, 1),
    `Patients ever hospitalized, n (%)`   =
      paste0(sum(hosp_days_total > 0), " (",
             round(100 * mean(hosp_days_total > 0), 1), "%)"),
    .groups = "drop"
  ) |>
  pivot_longer(-hiv_status, names_to = "Measure", values_to = "value") |>
  pivot_wider(names_from = hiv_status, values_from = value)

ft <- hosp_by_pt |>
  flextable() |>
  add_header_row(values = c("", "HIV status"), colwidths = c(1, ncol(hosp_by_pt) - 1)) |>
  theme_booktabs() |>
  align(align = "center", part = "all") |>
  align(j = 1, align = "left", part = "all") |>
  autofit()

obj <- save_table(slug = "hosp_days_by_pt_hiv", df = hosp_by_pt, ft = ft)
pub_objects_analysis$hosp_days_by_pt_hiv <- obj

obj$ft
emit_buttons(obj)
```

```{r}
#| label: prep-hosp-cycle-data

# Identify cycle-level column groups needed for hospitalization analysis
severe_cols <- names(gic_models)[str_detect(names(gic_models), "^ae_n_severe_([1-9]|1[0-2])$")]
max_grade_cols <- names(gic_models)[str_detect(names(gic_models), "^ae_max_grade_([1-9]|1[0-2])$")]
measured_cols <- names(gic_models)[str_detect(names(gic_models), "^ae_n_markers_measured_([1-9]|1[0-2])$")]

# Pivot each set of cycle cols to long
pivot_cycle_cols <- function(df, cols, value_name) {
  df |>
    select(record_id, all_of(cols)) |>
    pivot_longer(cols = all_of(cols), names_to = "cycle_var", values_to = value_name) |>
    mutate(cycle = as.integer(str_extract(cycle_var, "\\d+$"))) |>
    select(-cycle_var)
}

severe_long <- pivot_cycle_cols(gic_models, severe_cols,    "ae_n_severe")
max_grade_long <- pivot_cycle_cols(gic_models, max_grade_cols, "ae_max_grade")
measured_long <- pivot_cycle_cols(gic_models, measured_cols,  "ae_n_markers_measured")

# Join ADE variables to hospitalization data
hosp_severe_cycle <- hosp_long |>
  left_join(severe_long,   by = c("record_id", "cycle")) |>
  left_join(measured_long, by = c("record_id", "cycle")) |>
  mutate(ae_measured = ae_n_markers_measured > 0)

hosp_maxgrade_cycle <- hosp_long |>
  left_join(max_grade_long, by = c("record_id", "cycle")) |>
  left_join(measured_long,  by = c("record_id", "cycle")) |>
  mutate(ae_measured = ae_n_markers_measured > 0)
```

## Hospitalizations by ADE Severity
```{r}
#| label: tbl-hosp-vs-severe
#| tbl-cap: "Hospital days by number of severe adverse events in the cycle and HIV status, among cycles with adverse event grading recorded."
#| results: asis

# HIV status belongs in the columns. Previously grouped into the rows, which
# doubled the row count and made the two groups impossible to compare at a
# glance.
hosp_vs_severe <- hosp_severe_cycle |>
  filter(ae_measured) |>
  group_by(hiv_status, ae_n_severe) |>
  summarize(
    cell = paste0(summ_mean_sd_range(hosp_days, 1),
                  " [n=", sum(!is.na(hosp_days)), "]"),
    .groups = "drop"
  ) |>
  select(ae_n_severe, hiv_status, cell) |>
  pivot_wider(names_from = hiv_status, values_from = cell, values_fill = "—") |>
  arrange(ae_n_severe) |>
  rename(`Severe ADEs in cycle` = ae_n_severe)

ft <- hosp_vs_severe |>
  flextable() |>
  add_header_row(values = c("", "Hospital days, mean (SD); range [n cycles]"),
                 colwidths = c(1, ncol(hosp_vs_severe) - 1)) |>
  theme_booktabs() |>
  align(align = "center", part = "all") |>
  align(j = 1, align = "left", part = "all") |>
  autofit()

obj <- save_table(slug = "hosp_days_by_severe_hiv", df = hosp_vs_severe, ft = ft)
pub_objects_analysis$hosp_days_by_severe_hiv <- obj

obj$ft
emit_buttons(obj)
```

```{r}
#| label: plt-hosp-vs-severe
#| fig-cap: "Mean hospital days vs number of severe ADEs in the cycle, by HIV status"

hosp_severe_summ <- hosp_severe_cycle |>
  filter(ae_measured) |>
  group_by(hiv_status, ae_n_severe) |>
  summarize(
    mean_hosp_days = mean(hosp_days, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(hosp_severe_summ, aes(x = ae_n_severe, y = mean_hosp_days, color = hiv_status, group = hiv_status)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    x = "Number of severe ADEs in cycle",
    y = "Mean hospital days (toxicity) in cycle",
    color = "HIV status"
  ) +
  theme_minimal()
```

```{r}
#| label: tbl-hosp-vs-max-grade
#| tbl-cap: "Hospital days by maximum adverse event grade in the cycle and HIV status, among cycles with adverse event grading recorded."
#| results: asis

hosp_vs_max_grade <- hosp_maxgrade_cycle |>
  filter(ae_measured) |>
  group_by(hiv_status, ae_max_grade) |>
  summarize(
    cell = paste0(summ_mean_sd_range(hosp_days, 1),
                  " [n=", sum(!is.na(hosp_days)), "]"),
    .groups = "drop"
  ) |>
  select(ae_max_grade, hiv_status, cell) |>
  pivot_wider(names_from = hiv_status, values_from = cell, values_fill = "—") |>
  arrange(ae_max_grade) |>
  rename(`Max ADE grade in cycle` = ae_max_grade)

ft <- hosp_vs_max_grade |>
  flextable() |>
  add_header_row(values = c("", "Hospital days, mean (SD); range [n cycles]"),
                 colwidths = c(1, ncol(hosp_vs_max_grade) - 1)) |>
  theme_booktabs() |>
  align(align = "center", part = "all") |>
  align(j = 1, align = "left", part = "all") |>
  autofit()

obj <- save_table(slug = "hosp_days_by_max_grade_hiv", df = hosp_vs_max_grade, ft = ft)
pub_objects_analysis$hosp_days_by_max_grade_hiv <- obj

obj$ft
emit_buttons(obj)
```

```{r}
#| label: plt-hosp-vs-max-grade
#| fig-cap: "Mean hospital days vs max ADE grade in cycle, by HIV status (among measured cycles)"

hosp_maxgrade_summ <- hosp_maxgrade_cycle |>
  filter(ae_measured) |>
  group_by(hiv_status, ae_max_grade) |>
  summarize(
    mean_hosp_days = mean(hosp_days, na.rm = TRUE),
    .groups = "drop"
  )

plt_hosp_vs_max_grade <- ggplot(hosp_maxgrade_summ, aes(x = ae_max_grade, y = mean_hosp_days, color = hiv_status, group = hiv_status)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 0:5) +
  labs(
    x = "Max ADE grade in cycle",
    y = "Mean hospital days (toxicity) in cycle",
    color = "HIV status"
  ) +
  theme_minimal()

plt_hosp_vs_max_grade

pub_objects_analysis$plt_hosp_vs_max_grade <- save_figure(plt_hosp_vs_max_grade, "plt_hosp_vs_max_grade")
```

## Regression Models


```{r}
#| label: prep-hosp-regression


hosp_ae_cycle <- hosp_long |>
  left_join(
    ae_cycle_long |> select(record_id, cycle, ae_n_severe, ae_max_grade, ae_n_markers_measured),
    by = c("record_id", "cycle")
  ) |>
  mutate(ae_measured = ae_n_markers_measured > 0)


hosp_model_data <- hosp_ae_cycle |>
  filter(ae_measured)

hosp_model_data <- hosp_model_data |>
  mutate(any_hosp = as.integer(hosp_days > 0))
```

```{r}
#| label: model-hosp-any

# Any hospitalization this cycle (logistic)
model_any_hosp <- glmer(
  any_hosp ~ ae_max_grade + hiv_status + age + sex + (1 | record_id),
  data = hosp_model_data,
  family = binomial
)
```

OR table for whether a patient was hospitalized at all in a given cycle.
```{r}
#| label: show-formula-any-hosp
#| results: asis

show_formula(model_any_hosp)
```

```{r}
#| label: tbl-any-hosp
#| tbl-cap: "Odds of any toxicity-related hospitalization in a cycle"

tbl_any_hosp <- model_any_hosp |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      ae_max_grade ~ "Max ADE Grade in Cycle",
      hiv_status   ~ "HIV Status",
      age          ~ "Age (years)",
      sex          ~ "Sex"
    ),
    pvalue_fun = label_style_pvalue(digits = 3)
  ) |>
  tbl_drop_ref_rows() |>
  bold_p() |>
  bold_labels() |>
  italicize_levels()

tbl_any_hosp

obj <- save_table(slug = "tbl_any_hosp", gtsum = tbl_any_hosp)
pub_objects_analysis$tbl_any_hosp <- obj
```

#### Validation: model_any_hosp
::: {.callout-note collapse="true" title="Validation: model_any_hosp"}
```{r}
#| label: chk-any-hosp-vif

check_vif(model_any_hosp)
```
:::

```{r}
#| label: model-hosp-days

# Length of stay given hospitalization: unadjusted vs adjusted
model_hosp_days_simple <- lmer(
  hosp_days ~ hiv_status + (1 | record_id),
  data = hosp_model_data |> filter(hosp_days > 0)
)

model_hosp_days_adjusted <- lmer(
  hosp_days ~ hiv_status + age + sex + chemo_type + cancer_stage + (1 | record_id),
  data = hosp_model_data |> filter(hosp_days > 0)
)
```
ADE severity has minimal impact on length of stay for either group once someone is already hospitalized, although HIV positive patients are expected to spend slightly longer in the hospital than HIV negative patients for each unit increase in max adverse event experienced.

```{r}
#| label: show-formula-hosp-days
#| results: asis

show_formula(model_hosp_days_simple, "Unadjusted")
show_formula(model_hosp_days_adjusted, "Adjusted")
```

```{r}
#| label: tbl-hosp-days-comparison
#| tbl-cap: "Length of stay among hospitalized cycles: unadjusted vs adjusted"

tbl_hosp_simple <- model_hosp_days_simple |>
  tbl_regression(
    exponentiate = FALSE,
    label = list(hiv_status ~ "HIV Status")
  ) |>
  tbl_drop_ref_rows()

tbl_hosp_adjusted <- model_hosp_days_adjusted |>
  tbl_regression(
    exponentiate = FALSE,
    label = list(
      hiv_status   ~ "HIV Status",
      cancer_stage ~ "Cancer Stage",
      chemo_type   ~ "Chemotherapy Regimen",
      age          ~ "Age (years)",
      sex          ~ "Sex"
    )
  ) |>
  tbl_drop_ref_rows()

tbl_hosp_days_comparison <- tbl_merge(
  list(tbl_hosp_simple, tbl_hosp_adjusted),
  tab_spanner = c("**Unadjusted Analysis**", "**Adjusted Analysis**"),
  quiet = TRUE
) |>
  modify_header(
    estimate_1 ~ "**Beta (Days)**",
    ci_1       ~ "**95% CI**",
    estimate_2 ~ "**Beta (Days)**",
    ci_2       ~ "**95% CI**"
  ) |>
  bold_labels() |>
  italicize_levels()

tbl_hosp_days_comparison

obj <- save_table(slug = "tbl_hosp_days_comparison", gtsum = tbl_hosp_days_comparison)
pub_objects_analysis$tbl_hosp_days_comparison <- obj
```

#### Validation: model_hosp_days_adjusted
::: {.callout-note collapse="true" title="Validation: model_hosp_days_adjusted"}
```{r}
#| label: chk-hosp-days-adjusted-vif

check_vif(model_hosp_days_adjusted)
```
:::


# Early Treatment Dropout

```{r}
#| label: prep-early-dropout

# Add early dropout flag to gic_models
gic_models <- gic_models |>
  mutate(
    early_dropout = as.integer(
      cycles_given <= early_dropout_cycles & cycles_prescribed > early_dropout_cycles
    )
  )
```

## Descriptive Overview

This section looks at whether the types of adverse events patients experience in their first few cycles are associated with stopping treatment early, and whether that relationship looks different for HIV-positive vs HIV-negative patients.

```{r}
#| label: tbl-early-dropout-by-hiv
#| tbl-cap: "Early treatment discontinuation by HIV status at two cycle cutoffs"

map_df(c(2, 4), function(cutoff) {
  gic_models |>
    mutate(dropout = as.integer(cycles_given <= cutoff & cycles_prescribed > cutoff)) |>
    group_by(hiv_status) |>
    summarize(
      n_total     = n(),
      n_dropout   = sum(dropout == 1, na.rm = TRUE),
      pct_dropout = round(100 * n_dropout / n_total, 1),
      .groups     = "drop"
    ) |>
    mutate(
      cutoff            = cutoff,
      `Early dropout, n (%)` = paste0(n_dropout, " (", pct_dropout, "%)")
    )
}) |>
  select(hiv_status, cutoff, `Total patients` = n_total, `Early dropout, n (%)`) |>
  pivot_wider(
    names_from  = cutoff,
    values_from = `Early dropout, n (%)`,
    names_glue  = "Cutoff ≤{cutoff} cycles"
  ) |>
  rename(`HIV Status` = hiv_status, `Total patients` = `Total patients`) |>
  gt() |>
  tab_spanner(
    label   = "Early dropout, n (%)",
    columns = starts_with("Cutoff")
  ) |>
  tab_footnote(
    footnote = "Early dropout defined as cycles given ≤ cutoff among patients prescribed more than the cutoff number of cycles.",
    location = cells_column_spanners()
  ) |>
  opt_footnote_marks(marks = "extended") |>
  cols_align(align = "center", columns = -`HIV Status`) |>
  cols_align(align = "left",   columns = `HIV Status`)
```

For each ADE type, the predictor used is the max grade recorded across cycles 1 to 4. Where no grade was recorded for a given ADE type in that window, it's been set to 0, on the assumption that if nothing was flagged there was probably nothing to flag. That's a more reasonable assumption for the liver panel markers, which Dr. Mazhindu said are usually only ordered when there's a clinical reason to, than it is for something like hemoglobin or platelets where a missing value might just mean the data wasn't entered. It was done this way mainly because doing a complete case analysis instead would have left only 97 patients and 26 early dropout events, which isn't really enough to work with. This is worth coming back to before the analysis is finalized.


## Global test for early dropout by HIV status

```{r}
#| label: tbl-early-dropout-hiv
#| tbl-cap: "Early treatment discontinuation by HIV status"

# 2x2 table: early dropout x HIV status
dropout_hiv_table <- table(
  dropout = factor(gic_models$early_dropout, levels = c(0, 1), labels = c("Completed", "Early dropout")),
  HIV = gic_models$hiv_status
)

dropout_hiv_table |>
  as.data.frame.matrix() |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE) |>
  add_header_above(c(" " = 1, "HIV Status" = 2))
```

```{r}
#| label: tbl-early-dropout-hiv-test
#| tbl-cap: "Test of association between early treatment discontinuation and HIV status."

# Was a cat() line plus a raw htest print-out, which rendered as unlabelled
# console text once the code stopped being shown.
test_contingency(dropout_hiv_table,
                 "Early discontinuation vs HIV status") |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
#| label: prep-ae-early-dropout

# Max grade per ADE type across cycles 1-early_dropout_cycles
# Patients with no recorded grade for a given ADE type in this window
# are excluded from that ADE type's predictor (not treated as grade 0)
ae_early <- ae_long_grouped |>
  filter(cycle <= early_dropout_cycles, !is.na(grade)) |>
  group_by(record_id, ae_group) |>
  summarize(max_grade_early = max(grade, na.rm = TRUE), .groups = "drop") |>
  rename(ae_type = ae_group) |>
  left_join(
    gic_models |> select(record_id, hiv_status, early_dropout),
    by = "record_id"
  ) |>
  filter(!is.na(early_dropout))

# Wide format for modeling — one row per patient, one col per ADE type
ae_types <- unique(ae_early$ae_type)

ae_early_wide <- ae_early |>
  pivot_wider(
    names_from  = ae_type,
    values_from = max_grade_early
  ) |>
  mutate(
    # All missing ADE grades imputed to 0 — absence of a recorded grade treated
    # as no adverse event. Less defensible for some markers (e.g. plt, hb) than
    # for liver panels. Revisit when time permits
    across(all_of(ae_types), ~ replace_na(.x, 0))
  ) |>
  left_join(
    gic_models |> select(record_id, hiv_status, early_dropout),
    by = c("record_id", "hiv_status", "early_dropout")
  )
```


<!-- ```{r} -->
<!-- #| label: tbl-ae-early-dropout -->
<!-- #| tbl-cap: "Adverse event grades in cycles 1–4 by early dropout status and HIV status" -->

<!-- ae_early |> -->
<!--   mutate(ae_label = ae_type) |> -->
<!--   group_by(hiv_status, early_dropout, ae_label) |> -->
<!--   summarize( -->
<!--     n_any    = sum(max_grade_early > 0, na.rm = TRUE), -->
<!--     med      = median(max_grade_early, na.rm = TRUE), -->
<!--     n_severe = sum(max_grade_early >= 3, na.rm = TRUE), -->
<!--     .groups  = "drop" -->
<!--   ) |> -->
<!--   mutate( -->
<!--     dropout_label = if_else(early_dropout == 1, "Early dropout", "Completed >4 cycles"), -->
<!--     cell          = paste0(n_any, " / ", round(med, 1), " / ", n_severe) -->
<!--   ) |> -->
<!--   select(hiv_status, ae_label, dropout_label, cell) |> -->
<!--   pivot_wider( -->
<!--     names_from  = c(hiv_status, dropout_label), -->
<!--     values_from = cell, -->
<!--     names_glue  = "{hiv_status}: {dropout_label}" -->
<!--   ) |> -->
<!--   select( -->
<!--     ae_label, -->
<!--     `Negative: Completed >4 cycles`, -->
<!--     `Negative: Early dropout`, -->
<!--     `Positive: Completed >4 cycles`, -->
<!--     `Positive: Early dropout` -->
<!--   ) |> -->
<!--   rename(`Adverse event` = ae_label) |> -->
<!--   gt() |> -->
<!--   tab_spanner( -->
<!--     label   = "HIV Negative", -->
<!--     columns = starts_with("Negative") -->
<!--   ) |> -->
<!--   tab_spanner( -->
<!--     label   = "HIV Positive", -->
<!--     columns = starts_with("Positive") -->
<!--   ) |> -->
<!--   cols_label( -->
<!--     `Negative: Completed >4 cycles` = "Completed >4", -->
<!--     `Negative: Early dropout`        = "Early dropout", -->
<!--     `Positive: Completed >4 cycles`  = "Completed >4", -->
<!--     `Positive: Early dropout`         = "Early dropout" -->
<!--   ) |> -->
<!--   cols_align(align = "center", columns = -`Adverse event`) |> -->
<!--   cols_align(align = "left",   columns = `Adverse event`) |> -->
<!--   tab_footnote( -->
<!--     footnote = "Cells show: n with any ADE / median max grade / n with severe ADE (grade ≥ 3)", -->
<!--     location = cells_column_spanners() -->
<!--   ) |> -->
<!--   opt_footnote_marks(marks = "extended") -->
<!-- ``` -->



```{r}
#| label: tbl-dropout-by-ae-type
#| tbl-cap: "Association between experiencing each ADE type (any grade) in cycles 1–4 and early dropout"

dropout_ae_tests <- map_dfr(ae_types, function(at) {
  dat <- ae_early_wide |>
    filter(!is.na(early_dropout)) |>
    mutate(has_ae = as.integer(.data[[at]] > 0))
  
  tab <- table(dropout = dat$early_dropout, has_ae = dat$has_ae)
  
  if (nrow(tab) < 2 | ncol(tab) < 2) {
    return(tibble(ae_type = at, p = NA_real_, test = "—",
                  n_ae = NA, n_ae_dropout = NA, n_total = NA))
  }
  
  expected <- suppressWarnings(chisq.test(tab)$expected)
  if (any(expected < 5)) {
    result <- fisher.test(tab)
    test_used <- "Fisher's"
  } else {
    result <- chisq.test(tab, correct = FALSE)
    test_used <- "Chi-squared"
  }
  
  tibble(
    ae_type = at,
    n_total = nrow(dat),
    n_ae = sum(dat$has_ae == 1),
    n_ae_dropout = sum(dat$has_ae == 1 & dat$early_dropout == 1),
    p = result$p.value,
    test = test_used
  )
}) |>
  mutate(
    ae_label = ae_type,
    p_adj = p.adjust(p, method = "BH"),
    `p (raw)` = map_chr(p, format_p),
    `p (adj)` = map_chr(p_adj, format_p),
    `With ADE, n (dropout)` = paste0(n_ae, " (", n_ae_dropout, ")")
  )

dropout_ae_tests |>
  select(`Adverse event` = ae_label, `With ADE, n (dropout)`, Test = test, `p (raw)`, `p (adj)`) |>
  gt() |>
  cols_align(align = "left", columns = `Adverse event`) |>
  cols_align(align = "center", columns = -`Adverse event`) |>
  tab_footnote(
    footnote = paste0(
      "Tests whether having any occurrence of each ADE type (grade > 0) in cycles 1–",
      early_dropout_cycles,
      " is associated with early dropout, pooling HIV groups. ",
      "Chi-squared or Fisher's exact as appropriate. p-values BH-adjusted across ADE types."
    ),
    locations = cells_column_spanners()
  ) |>
  opt_footnote_marks(marks = "extended")
```

Tests whether, among all patients in the early dropout analysis, is having any occurrence of a specific ADE type (grade > 0) in cycles 1-`r early_dropout_cycles` associated with whether they dropped out? Multiple comparisons correction applied because testing the same population multiple times across adverse event types.


The heatmap below gives a quick visual overview of where the grade differences between early dropout and completed patients are largest, and whether those patterns look similar across HIV groups.

```{r}
#| label: plt-ae-early-dropout-heatmap
#| fig-cap: "Median max ADE grade in cycles 1–4 by ADE category, early dropout status, and HIV status"

ae_early |>
  group_by(hiv_status, early_dropout, ae_type) |>
  summarize(
    median_grade = median(max_grade_early, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    dropout_label = if_else(early_dropout == 1, "Early dropout", "Completed ≥5 cycles"),
    facet_label   = paste0("HIV ", hiv_status)
  ) |>
  ggplot(aes(x = dropout_label, y = ae_type, fill = median_grade)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(median_grade, 1)), size = 3) +
  scale_fill_viridis_c(option = "magma", direction = -1, 
                       limits = c(0, 4), name = "Median\nmax grade") +
  facet_wrap(~ facet_label) +
  labs(
    x = NULL,
    y = "Adverse event category"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1),
    strip.text  = element_text(face = "bold")
  )
```

Before moving to the models, it's worth checking how many patients in each HIV group actually experienced each type of adverse event. ADE types with very few patients in either group are excluded from the univariate models since there isn't enough data to estimate anything reliable for them.

```{r}
#| label: tbl-ae-coverage
#| tbl-cap: "ADE grade coverage in cycles 1–4: patients with any recorded adverse event by HIV status"

ae_early_wide |>
  select(hiv_status, early_dropout, all_of(ae_types)) |>
  pivot_longer(
    cols      = all_of(ae_types),
    names_to  = "ae_type",
    values_to = "grade"
  ) |>
  group_by(ae_type, hiv_status) |>
  summarize(
    n_total    = n(),
    n_any_ae   = sum(grade > 0, na.rm = TRUE),
    pct_any_ae = round(100 * n_any_ae / n_total, 1),
    .groups    = "drop"
  ) |>
  mutate(
    ae_label = ae_type,
    `n (%)` = paste0(n_any_ae, " (", pct_any_ae, "%)")
  ) |>
  select(ae_label, hiv_status, `n (%)`) |>
  pivot_wider(
    names_from  = hiv_status,
    values_from = `n (%)`,
    names_prefix = "HIV "
  ) |>
  rename(`Adverse event` = ae_label) |>
  gt() |>
  tab_spanner(
    label   = "Patients with any ADE (grade > 0)",
    columns = starts_with("HIV")
  ) |>
  cols_align(align = "left",   columns = `Adverse event`) |>
  cols_align(align = "center", columns = -`Adverse event`) |>
  tab_footnote(
    footnote = paste0(
      "Among patients with cycles given ≤ ", early_dropout_cycles,
      " or prescribed > ", early_dropout_cycles, " cycles. ",
      "ADE types with 0 patients in either group cannot be evaluated in univariate models."
    ),
    location = cells_column_spanners()
  ) |>
  opt_footnote_marks(marks = "extended")
```

```{r}
#| label: prep-sparse-ae

# ADE types with fewer than 5 patients with any non-zero grade in either
# HIV group — too sparse for reliable estimation
ae_cols <- intersect(ae_types, names(ae_early_wide))

sparse_ae <- ae_early_wide |>
  select(hiv_status, all_of(ae_cols)) |>
  pivot_longer(cols = all_of(ae_cols), names_to = "ae_type", values_to = "grade") |>
  group_by(ae_type, hiv_status) |>
  summarize(n_any = sum(grade > 0, na.rm = TRUE), .groups = "drop") |>
  group_by(ae_type) |>
  filter(any(n_any < 5)) |>
  pull(ae_type) |>
  unique()
```


## Univariate Logistic Models

```{r}
#| label: fit-univariate-ae-models

# One logistic model per ADE type — early dropout ~ max grade in cycles 1-4 + HIV status
# Patients with no recorded grade for a given ADE type are excluded from that model
univariate_results <- map_df(ae_types, function(at) {
  dat <- ae_early_wide |>
    filter(!is.na(.data[[at]]), !is.na(early_dropout), !is.na(hiv_status))

  if (sum(dat$early_dropout) < 10) {
    return(tibble(ae_type = at, note = "Insufficient early dropout cases", .rows = 1))
  }

  if (at %in% sparse_ae) {
    return(tibble(ae_type = at, note = "Sparse — excluded per coverage check", .rows = 1))
  }

  mod <- glm(
    as.formula(paste0("early_dropout ~ `", at, "` + hiv_status")),
    data   = dat,
    family = binomial
  )

  max_gvif <- check_vif(mod) |> pull(gvif) |> max(na.rm = TRUE)

  tidy(mod, conf.int = TRUE, exponentiate = TRUE) |>
    filter(str_detect(term, fixed(at))) |>
    mutate(ae_type = at, max_gvif = max_gvif)
}) |>
  filter(!is.na(estimate)) |>
  mutate(p_adj = p.adjust(p.value, method = "BH"))
```

#### Validation: univariate early-dropout ADE models
::: {.callout-note collapse="true" title="Validation: univariate early-dropout ADE models"}
Multicollinearity check across all univariate ADE models (one GVIF per model, between the ADE type's grade and HIV status, rather than a separate diagnostic block per ADE type).
```{r}
#| label: tbl-chk-univariate-ae-vif
#| tbl-cap: "Multicollinearity check: max GVIF per univariate early-dropout model"

univariate_results |>
  select(`Adverse event` = ae_type, `Max GVIF` = max_gvif) |>
  mutate(Flag = if_else(`Max GVIF` > 5, "Check", "OK")) |>
  kbl(booktabs = TRUE, digits = 2) |>
  kable_styling(full_width = FALSE, position = "left")
```
:::

```{r}
#| label: tbl-univariate-ae-results
#| tbl-cap: "Univariate logistic regression: association between ADE type max grade and early dropout (adjusted for HIV status)"

# Sample sizes per ADE type for context
# Sample sizes per ADE category for context
ae_sample_sizes <- ae_types |>
  map(function(at) {
    ae_early_wide |>
      filter(!is.na(early_dropout), !is.na(hiv_status)) |>
      group_by(hiv_status) |>
      summarize(
        n_nonzero        = sum(.data[[at]] > 0),
        n_events_nonzero = sum(.data[[at]] > 0 & early_dropout == 1),
        .groups          = "drop"
      ) |>
      mutate(ae_type = at, cell = paste0(n_nonzero, " (", n_events_nonzero, ")")) |>
      select(ae_type, hiv_status, cell) |>
      pivot_wider(names_from = hiv_status, values_from = cell, names_glue = "n_{hiv_status}")
  }) |>
  list_rbind()

univariate_results |>
  left_join(ae_sample_sizes, by = "ae_type") |>
  mutate(
    `OR (95% CI)` = paste0(round(estimate, 2), " (", round(conf.low, 2), "–", round(conf.high, 2), ")"),
    `p (raw)`     = map_chr(p.value, format_p),
    `p (adj)`     = map_chr(p_adj, format_p)
  ) |>
  arrange(ae_type) |>
  select(
    `Adverse event`  = ae_type,
    `HIV-: n (drop)` = n_Negative,
    `HIV+: n (drop)` = n_Positive,
    `OR (95% CI)`,
    `p (raw)`,
    `p (adj)`
  ) |>
  gt() |>
  cols_align(align = "left",   columns = `Adverse event`) |>
  cols_align(align = "center", columns = -`Adverse event`) |>
  tab_footnote(
    footnote = paste0("OR: change in odds of early dropout per one-unit increase in max ADE grade (cycles 1–", early_dropout_cycles, "), holding HIV status constant. n (drop) = patients with any grade > 0 for this ADE category / early dropouts among those. p-values BH-adjusted across ADE categories."),
    locations = cells_column_labels(columns = `p (adj)`)
  ) |>
  opt_footnote_marks(marks = "extended")
```

Publication version: descriptive breakdown by dropout status merged with the HIV-adjusted univariate associations, for the ADE types that passed the sparsity check above.
```{r}
#| label: prep-early-dropout-final

# These were previously merged into a single table with two spanners. They
# answer different questions and share only their row names: the descriptive
# half summarises grade distributions by dropout status, the model half reports
# associations. HIV status also appeared as a row in one half and not the
# other, which is what made the merged version look like two tables stacked.
final_ae_columns <- setdiff(ae_types, sparse_ae)

dropout_dat <- ae_early_wide |>
  select(early_dropout, hiv_status, all_of(final_ae_columns))
```

```{r}
#| label: tbl-early-dropout-descriptive
#| tbl-cap: "Adverse event grades in the early treatment window, and HIV status, by whether the patient discontinued early."
#| results: asis

table_early_dropout_desc <- dropout_dat |>
  mutate(early_dropout = factor(early_dropout,
                                labels = c("Completed treatment", "Early dropout"))) |>
  tbl_summary(
    by        = early_dropout,
    missing   = "no",
    statistic = list(all_continuous()  ~ "{median} ({p25}-{p75})",
                     all_categorical() ~ "{n} ({p}%)"),
    label     = c(list(hiv_status = "HIV Status"),
                  setNames(as.list(final_ae_columns), final_ae_columns))
  ) |>
  add_overall(last = FALSE) |>
  # Spell out what each cell is. Percentages are *within* the column, i.e. of
  # the patients in that dropout group -- not of the whole cohort and not of
  # the row.
  add_stat_label(label = list(all_categorical() ~ "n (% of column)",
                              all_continuous()  ~ "Median (IQR)")) |>
  modify_spanning_header(everything() ~ NA_character_) |>
  bold_labels() |>
  add_source_note("Percentages are column percentages: the share of patients within that dropout group, not of the whole cohort. Adverse event rows give the maximum grade recorded in the early treatment window.")

obj <- save_table(slug = "table_early_dropout_descriptive", gtsum = table_early_dropout_desc)
pub_objects_analysis$table_early_dropout_descriptive <- obj

obj$ft
emit_buttons(obj)
```

```{r}
#| label: tbl-early-dropout-final
#| tbl-cap: "Univariate association between maximum adverse event grade in the early treatment window and early treatment discontinuation, each model adjusted for HIV status."
#| results: asis

table_early_dropout_final <- dropout_dat |>
  tbl_uvregression(
    method       = glm,
    y            = early_dropout,
    method.args  = list(family = binomial),
    exponentiate = TRUE,
    include      = all_of(final_ae_columns),
    formula      = "{y} ~ {x} + hiv_status",
    label = map(final_ae_columns, ~ as.formula(paste0(".x ~ '", .x, " max grade'")))
  ) |>
  add_q(method = "BH") |>
  bold_labels() |>
  add_source_note(paste0(
    "One logistic model per adverse event category: early dropout ~ max grade in cycles 1-",
    early_dropout_cycles,
    " + HIV status. Odds ratios are per one-grade increase in that category."
  )) |>
  add_source_note(paste0(
    "q-value: the p-value after Benjamini-Hochberg correction for having run ",
    "one test per adverse event category. Testing ", length(final_ae_columns),
    " categories means some will look significant by chance alone; the q-value ",
    "is the threshold to read instead of the raw p. Interpret it as a false ",
    "discovery rate -- calling everything with q < 0.05 a finding means roughly ",
    "5% of those calls are expected to be false."
  ))

obj <- save_table(slug = "table_early_dropout_final", gtsum = table_early_dropout_final)
pub_objects_analysis$table_early_dropout_final <- obj

obj$ft
emit_buttons(obj)
```

Forest plot of the same univariate associations (max grade in cycles 1–`r early_dropout_cycles`, adjusted for HIV status).
```{r}
#| label: plt-early-dropout-forest
#| fig-cap: "Impact of early-cycle toxicities on early treatment dropout"

plt_early_dropout_forest <- univariate_results |>
  filter(ae_type %in% final_ae_columns) |>
  ggplot(aes(x = estimate, y = reorder(ae_type, estimate))) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.2, color = "midnightblue", linewidth = 0.7) +
  geom_point(color = "firebrick", size = 3) +
  scale_x_log10(breaks = c(0.5, 1, 2, 5, 10)) +
  labs(
    x = "Odds Ratio (95% CI) on Log Scale\n[Adjusted for HIV Status]",
    y = paste0("Adverse Event Type (Max Grade Cycles 1-", early_dropout_cycles, ")"),
    title = "Impact of Early Cycle Toxicities on Early Treatment Dropout",
    subtitle = "Odds of dropping out per 1-unit increase in toxicity grade"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

plt_early_dropout_forest

pub_objects_analysis$plt_early_dropout_forest <- save_figure(plt_early_dropout_forest, "plt_early_dropout_forest")
```

## Early dropout associations split by HIV status

Requested 5-Jun-2026 (Tinashe, agreed by Audrey). The forest plot above adjusts
for HIV status but reports a single pooled estimate per adverse event. This
version fits each association **separately within each HIV group**, so readers
can see whether the two groups track together or whether an apparent difference
is small-n noise on the HIV-positive side.

Kept as a **separate figure** rather than adding a second bar per row to the
plot above, so both versions can be compared before one is chosen for the paper.

Adverse event categories that failed the sparsity coverage check are excluded
here exactly as they are from the pooled analysis — that exclusion is global and
has nothing to do with the HIV split. The table beneath the figure reports only
categories that could not be estimated *within a particular HIV group*.

::: {.callout-warning}
The HIV-positive group is small. Many of these stratum-specific models will
either fail to converge or return very wide intervals, and that is itself the
finding: a difference between groups is only interpretable where both intervals
are reasonably tight. Models that could not be fit are listed beneath the
figure rather than silently dropped.
:::

```{r}
#| label: fit-univariate-ae-models-byhiv

# One logistic model per (ADE type x HIV group). HIV status is not a covariate
# here because it defines the stratum.
#
# Iterates over final_ae_columns, i.e. the categories that already passed the
# sparsity coverage check, rather than over every ADE type. Previously it looped
# over all of them and recorded "Sparse -- excluded per coverage check" once per
# HIV group, which filled the exclusions table below with rows that had nothing
# to do with stratification: that exclusion is global, applies to both groups
# equally, and already removes those categories from the pooled analysis too.
# Listing it here implied those categories failed *because* of the split.
univariate_by_hiv <- expand_grid(
  ae_type    = final_ae_columns,
  hiv_group  = levels(factor(ae_early_wide$hiv_status))
) |>
  pmap_dfr(function(ae_type, hiv_group) {
    base <- tibble(ae_type = ae_type, hiv_group = hiv_group)

    dat <- ae_early_wide |>
      filter(hiv_status == hiv_group,
             !is.na(.data[[ae_type]]), !is.na(early_dropout))

    # Need both outcome variation and predictor variation to estimate anything.
    if (nrow(dat) < 10 ||
        sum(dat$early_dropout) < 5 ||
        sum(dat$early_dropout == 0) < 5 ||
        dplyr::n_distinct(dat[[ae_type]]) < 2) {
      return(base |> mutate(
        n     = nrow(dat),
        note  = "Too few events or no grade variation in this stratum"
      ))
    }

    fit <- tryCatch(
      glm(as.formula(paste0("early_dropout ~ `", ae_type, "`")),
          data = dat, family = binomial),
      error = function(e) e
    )
    if (inherits(fit, "error")) {
      return(base |> mutate(n = nrow(dat),
                            note = paste("Model failed:", conditionMessage(fit))))
    }

    # conf.int = TRUE profiles the likelihood and can fail on separated fits;
    # fall back to Wald intervals rather than losing the row entirely.
    td <- tryCatch(
      tidy(fit, conf.int = TRUE, exponentiate = TRUE),
      error = function(e) tidy(fit, conf.int = FALSE, exponentiate = TRUE) |>
        mutate(conf.low = NA_real_, conf.high = NA_real_)
    )

    td |>
      filter(str_detect(term, fixed(ae_type))) |>
      mutate(ae_type = ae_type, hiv_group = hiv_group, n = nrow(dat),
             note = NA_character_)
  })

# BH correction applied within each HIV group, since each group is its own
# family of tests.
stopifnot("p.value" %in% names(univariate_by_hiv))

univariate_by_hiv <- univariate_by_hiv |>
  group_by(hiv_group) |>
  # n is set explicitly so the strata that could not be fit do not inflate the
  # denominator; p.adjust()'s default n = length(p) would count them as tests.
  mutate(p_adj = p.adjust(p.value, method = "BH",
                          n = sum(!is.na(p.value)))) |>
  ungroup()
```

```{r}
#| label: plt-early-dropout-forest-byhiv
#| fig-cap: "Association between early-cycle toxicity grade and early treatment dropout, estimated separately within each HIV group. Points are odds ratios per one-grade increase; bars are 95% confidence intervals. Compare with the pooled, HIV-adjusted figure above — overlapping intervals across the two groups indicate the pooled estimate is not masking a difference."
#| fig-height: 6

forest_byhiv_data <- univariate_by_hiv |>
  filter(is.na(note), !is.na(estimate), ae_type %in% final_ae_columns)

plt_early_dropout_forest_byhiv <- forest_byhiv_data |>
  ggplot(aes(x = estimate, y = reorder(ae_type, estimate), color = hiv_group)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  # `width`, not `height`: ggplot 4.x renamed the argument for vertical
  # errorbars and emits "`height` was translated to `width`" otherwise.
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high),
                width = 0.2, linewidth = 0.7,
                position = position_dodge(width = 0.6)) +
  geom_point(size = 2.6, position = position_dodge(width = 0.6)) +
  scale_x_log10() +
  labs(
    x = "Odds Ratio (95% CI) on Log Scale",
    y = paste0("Adverse Event Type (Max Grade Cycles 1-", early_dropout_cycles, ")"),
    title = "Early Cycle Toxicities and Early Dropout, by HIV Status",
    subtitle = "Odds of dropping out per 1-unit increase in toxicity grade, estimated within each group",
    color = "HIV status"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

plt_early_dropout_forest_byhiv

pub_objects_analysis$plt_early_dropout_forest_byhiv <-
  save_figure(plt_early_dropout_forest_byhiv,
              "plt_early_dropout_forest_byhiv", height = 6)
```

```{r}
#| label: tbl-early-dropout-byhiv-notfit
#| tbl-cap: "Adverse event categories that could not be estimated within a specific HIV group, and why. An adverse event missing from one panel of the figure above is not evidence of no association in that group."

# Only stratum-specific failures belong here. Categories dropped by the global
# sparsity check are excluded from the pooled analysis as well, so they are
# reported once above rather than repeated per HIV group.
byhiv_notfit <- univariate_by_hiv |>
  filter(!is.na(note)) |>
  select(`Adverse event` = ae_type, `HIV group` = hiv_group,
         n, Reason = note) |>
  arrange(`Adverse event`, `HIV group`)

if (nrow(byhiv_notfit) == 0) {
  byhiv_notfit <- tibble(
    `Adverse event` = "—",
    `HIV group`     = "—",
    n               = NA_integer_,
    Reason          = "None. Every category that passed the coverage check was estimable in both HIV groups."
  )
}

byhiv_notfit |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left")
```



# Management Interventions: Transfusions and G-CSF

```{r}
#| label: prep-mgmt-analysis

# Load lmerTest to get p-values for linear mixed models
library(lmerTest)

# G-CSF can be indicated for either neutropenia (ANC) or leukopenia (WBC).
# Pull WBC grade per cycle and compute the per-cycle max of the two.
# Assigned to a new object rather than overwriting mgmt_cycle, so that
# re-running this chunk in an interactive session does not join wbc_by_cycle on
# twice and produce wbc_grade.x / wbc_grade.y.
wbc_by_cycle <- ae_long |>
  filter(ae_type == "wbc") |>
  select(record_id, cycle, wbc_grade = grade)

mgmt_cycle_mgmt <- mgmt_cycle |>
  left_join(wbc_by_cycle, by = c("record_id", "cycle")) |>
  mutate(gcsf_grade = pmax(anc_grade, wbc_grade, na.rm = TRUE))

mgmt_cycle <- mgmt_cycle_mgmt

# Anemia grade as a collapsed factor. Grades 3 and 4 are pooled throughout
# because grade 4 is rare and near-universally transfused.
add_anemia_grade_f <- function(d) {
  d |>
    mutate(
      anemia_grade_f = factor(anemia_grade),
      anemia_grade_f = fct_collapse(anemia_grade_f, "3-4" = c("3", "4"))
    )
}

mgmt_cycle_transfusion <- mgmt_cycle |>
  filter(!is.na(anemia_grade), !is.na(transfusion_given), anemia_grade >= 1) |>
  add_anemia_grade_f()

# --- Transfusion analysis strata (13-Jul-2026) -----------------------------
# Stratum 1: every cycle where the patient had a transfusion indication.
mgmt_grade2plus <- mgmt_cycle |>
  filter(anemia_grade >= 2, !is.na(transfusion_given)) |>
  add_anemia_grade_f()

# Stratum 2: every cycle belonging to a patient who was transfused at least
# once. Restricting to *cycles* where transfusion_given == 1 would leave the
# outcome constant and unmodellable, so the restriction is applied at the
# patient level: among patients for whom transfusion was demonstrably
# available, does HIV status predict receiving one in a given cycle?
tx_ever_ids <- mgmt_cycle |>
  filter(transfusion_given == 1) |>
  distinct(record_id) |>
  pull(record_id)

mgmt_tx_patients <- mgmt_cycle |>
  filter(record_id %in% tx_ever_ids, !is.na(transfusion_given),
         !is.na(anemia_grade)) |>
  add_anemia_grade_f()

# Retained sensitivity stratum: the 19-Jun-2026 grade-2-exactly specification.
mgmt_grade2 <- mgmt_cycle |>
  filter(anemia_grade == 2, !is.na(transfusion_given)) |>
  add_anemia_grade_f()

c(grade_2plus  = nrow(mgmt_grade2plus),
  ever_transfused = nrow(mgmt_tx_patients),
  grade_2_only = nrow(mgmt_grade2))
```

These two tables show how transfusion and G-CSF receipt vary with the severity
grade that is supposed to indicate them. They set up the modelling decisions
that follow: where the outcome barely varies within a grade, that grade carries
no information and the model cannot use it.

```{r}
#| label: tbl-transfusion-rate-by-grade
#| tbl-cap: "Blood transfusion receipt per treatment cycle, by anemia grade and HIV status."
#| results: asis

transfusion_by_grade <- mgmt_rate_table(
  mgmt_cycle_transfusion, "anemia_grade", "transfusion_given", "Anemia grade"
)

ft <- mgmt_rate_flextable(
  transfusion_by_grade,
  spanner  = "Transfused / cycles (%)",
  footnote = "Denominator is cycles with a recorded transfusion decision."
)

obj <- save_table(slug = "transfusion_rate_by_grade_hiv", df = transfusion_by_grade, ft = ft)
pub_objects_analysis$transfusion_rate_by_grade_hiv <- obj

obj$ft
emit_buttons(obj)
```

```{r}
#| label: tbl-gcsf-rate-by-grade
#| tbl-cap: "G-CSF receipt per treatment cycle, by maximum neutropenia/leukopenia grade and HIV status."
#| results: asis

gcsf_by_grade <- mgmt_rate_table(
  mgmt_cycle, "gcsf_grade", "csfg_indicated", "Max neutropenia/leukopenia grade"
)

ft <- mgmt_rate_flextable(
  gcsf_by_grade,
  spanner  = "G-CSF given / cycles (%)",
  footnote = "Denominator is cycles with a recorded G-CSF decision."
)

obj <- save_table(slug = "gcsf_rate_by_grade_hiv", df = gcsf_by_grade, ft = ft)
pub_objects_analysis$gcsf_rate_by_grade_hiv <- obj

obj$ft
emit_buttons(obj)
```





## Transfusion Received

Per the 13-Jul-2026 meeting, the transfusion analysis is fit on **two separate
strata** and the effect estimates compared for consistency, rather than pooled:

1. **Cycles with anemia grade 2 or higher** — patients who had a clinical
   indication for transfusion.
2. **Cycles belonging to patients transfused at least once** — patients for whom
   transfusion was demonstrably available.

Consistency of the HIV estimate across the two, rather than either estimate
alone, is the result worth reporting.

::: {.callout-warning collapse="true" title="Interpretation caveat: indication is not the same as eligibility"}
Grade 2+ patients are *supposed* to receive a transfusion, but blood supply
constraints in this setting mean a substantial share do not. The grade 2+
stratum should therefore be read as **"had an indication"**, not "was eligible
and available for transfusion".
:::

```{r}
#| label: prep-tx-strata

# anemia_grade_f is screened alongside the other covariates. It was previously
# left out, and that was the cause of the odds ratios of 1e21 and Inf in this
# section: at grade 3-4 essentially every cycle is transfused, so the level is
# perfectly separated and its coefficient diverges.
tx_adj_vars <- c("anemia_grade_f", "cancer_stage", "chemo_type", "cancer_site")

mgmt_grade2plus_adj  <- mgmt_grade2plus  |> drop_separated_levels("transfusion_given", tx_adj_vars)
mgmt_tx_patients_adj <- mgmt_tx_patients |> drop_separated_levels("transfusion_given", tx_adj_vars)
mgmt_grade2_adj      <- mgmt_grade2      |> drop_separated_levels("transfusion_given", tx_adj_vars)

# Dropping a separated level removes rows, which can leave a covariate with a
# single level. glmer errors on those, so the formula is built from whatever
# still varies within each stratum rather than being fixed in advance. This is
# also why the two strata do not necessarily carry the same covariates.
tx_formula_for <- function(d) {
  candidates <- c("hiv_status", "anemia_grade_f", "cancer_stage",
                  "chemo_type", "cancer_site")
  keep <- candidates[vapply(candidates,
                            function(v) dplyr::n_distinct(d[[v]]) > 1,
                            logical(1))]
  reformulate(c(keep, "(1 | record_id)"), response = "transfusion_given")
}
```

```{r}
#| label: fit-tx-strata

model_tx_grade2plus <- glmer(tx_formula_for(mgmt_grade2plus_adj),
                             data = mgmt_grade2plus_adj, family = binomial)

model_tx_patients   <- glmer(tx_formula_for(mgmt_tx_patients_adj),
                             data = mgmt_tx_patients_adj, family = binomial)

# A named list of strings, not formulas -- see the note in the sensitivity
# section below. Subset per model, since the two strata may carry different
# covariates after separation screening.
tx_labels <- list(
  hiv_status     = "HIV Status",
  anemia_grade_f = "Anemia Grade",
  cancer_stage   = "Cancer Stage",
  chemo_type     = "Chemotherapy Type",
  cancer_site    = "Cancer Site"
)

tx_labels_for <- function(model) {
  tx_labels[intersect(names(tx_labels), all.vars(formula(model)))]
}
```

```{r}
#| label: tbl-tx-hiv-comparison
#| tbl-cap: "HIV status and the odds of receiving a blood transfusion, compared across analysis strata. This is the comparison the 13-Jul-2026 meeting asked for: whether the HIV estimate is stable across two differently-constructed populations."
#| results: asis

# Reported as a purpose-built comparison rather than a tbl_merge() of the two
# full models. The strata legitimately carry different covariate levels -- one
# is restricted to anemia grade 2+, the other is not -- so merging them row-wise
# produced misaligned rows and the "number of rows do not match" warning. Only
# the HIV row is actually comparable, so only that is compared.
tx_hiv_comparison <- bind_rows(
  hiv_estimate(model_tx_grade2plus, "Anemia grade 2+ (had an indication)"),
  hiv_estimate(model_tx_patients,   "Ever transfused (transfusion available)")
)

ft <- tx_hiv_comparison |>
  flextable() |>
  theme_booktabs() |>
  align(align = "center", part = "all") |>
  align(j = 1, align = "left", part = "all") |>
  flextable::add_footer_lines(
    "Odds of transfusion for HIV-positive vs HIV-negative patients, from a mixed-effects logistic model fit separately within each stratum. Covariates differ between strata where separation forced a level to be dropped; see the full model tables below."
  ) |>
  autofit()

obj <- save_table(slug = "tx_hiv_comparison_by_stratum", df = tx_hiv_comparison, ft = ft)
pub_objects_analysis$tx_hiv_comparison_by_stratum <- obj

obj$ft
emit_buttons(obj)
```

### Full model: anemia grade 2+

```{r}
#| label: show-formula-tx-grade2plus
#| results: asis

show_formula(model_tx_grade2plus)
```

```{r}
#| label: tbl-tx-grade2plus
#| tbl-cap: "Odds of receiving a blood transfusion among cycles with anemia grade 2 or higher."

model_tx_grade2plus |>
  tbl_regression(exponentiate = TRUE, label = tx_labels_for(model_tx_grade2plus)) |>
  tbl_drop_ref_rows() |>
  add_exclusion_note(mgmt_grade2plus_adj) |>
  bold_labels() |>
  bold_p()
```

### Full model: patients transfused at least once

```{r}
#| label: show-formula-tx-patients
#| results: asis

show_formula(model_tx_patients)
```

```{r}
#| label: tbl-tx-patients
#| tbl-cap: "Odds of receiving a blood transfusion among cycles belonging to patients transfused at least once."

model_tx_patients |>
  tbl_regression(exponentiate = TRUE, label = tx_labels_for(model_tx_patients)) |>
  tbl_drop_ref_rows() |>
  add_exclusion_note(mgmt_tx_patients_adj) |>
  bold_labels() |>
  bold_p()
```

```{r}
#| label: tbl-tx-strata-exclusions
#| tbl-cap: "Covariate levels dropped from each transfusion model, and the resulting analytic sample size. A level is dropped when the outcome does not vary within it — every cycle transfused, or none — because no finite odds ratio exists for such a level. Cycles belonging to a dropped level leave the model with it, which is why the sample sizes differ."

tibble(
  Stratum      = c("Anemia grade 2+", "Ever transfused", "Anemia grade 2 only"),
  `Cycles (n)` = c(nrow(mgmt_grade2plus_adj), nrow(mgmt_tx_patients_adj),
                   nrow(mgmt_grade2_adj)),
  `Levels dropped` = c(describe_dropped_levels(mgmt_grade2plus_adj),
                       describe_dropped_levels(mgmt_tx_patients_adj),
                       describe_dropped_levels(mgmt_grade2_adj))
) |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left")
```

#### Validation: stratified transfusion models

::: {.callout-note collapse="true" title="Validation: model_tx_grade2plus / model_tx_patients"}
```{r}
#| label: chk-tx-strata-convergence

bind_rows(
  check_convergence(model_tx_grade2plus) |> mutate(stratum = "Anemia grade 2+"),
  check_convergence(model_tx_patients)   |> mutate(stratum = "Ever transfused")
) |>
  relocate(stratum)
```

```{r}
#| label: chk-tx-strata-vif

bind_rows(
  check_vif(model_tx_grade2plus) |> mutate(stratum = "Anemia grade 2+"),
  check_vif(model_tx_patients)   |> mutate(stratum = "Ever transfused")
) |>
  relocate(stratum)
```

A "VIF not computable" result most likely reflects residual sparse-cell
rank-deficiency in these small subsets rather than a coding error.

```{r}
#| label: chk-tx-strata-separation

bind_rows(
  check_separation(model_tx_grade2plus)$coefs |> mutate(stratum = "Anemia grade 2+"),
  check_separation(model_tx_patients)$coefs   |> mutate(stratum = "Ever transfused")
) |>
  relocate(stratum) |>
  arrange(desc(extreme))
```

Any row still flagged `extreme` here indicates separation the pre-fit screen did
not catch, and means that coefficient should not be interpreted.
:::

### Sensitivity: anemia grade 2 only

The 19-Jun-2026 specification, kept so the grade 2+ result can be checked
against the estimates the team has already seen. Anemia grade is constant here
and so does not appear in the model.

```{r}
#| label: show-formula-tx-grade2
#| results: asis

model_tx_unadj <- glmer(
  transfusion_given ~ hiv_status + (1 | record_id),
  data = mgmt_grade2, family = binomial
)

model_tx_grade2 <- glmer(
  transfusion_given ~ hiv_status + cancer_stage + chemo_type + cancer_site +
    (1 | record_id),
  data = mgmt_grade2_adj, family = binomial
)

show_formula(model_tx_unadj, "Unadjusted")
show_formula(model_tx_grade2, "Adjusted")
```

```{r}
#| label: tbl-transfusion-sensitivity
#| tbl-cap: "Sensitivity analysis: odds of receiving a blood transfusion, anemia grade 2 only, unadjusted versus adjusted."

tbl_merge(
  list(
    model_tx_unadj |>
      tbl_regression(exponentiate = TRUE, label = list(hiv_status = "HIV Status")) |>
      tbl_drop_ref_rows(),
    model_tx_grade2 |>
      tbl_regression(exponentiate = TRUE,
                     label = tx_labels[setdiff(names(tx_labels), "anemia_grade_f")]) |>
      tbl_drop_ref_rows()
  ),
  tab_spanner = c("**Unadjusted**", "**Adjusted**")
) |>
  bold_labels()
```

#### Validation: model_tx_grade2

::: {.callout-note collapse="true" title="Validation: model_tx_grade2"}
```{r}
#| label: chk-tx-grade2-vif

check_vif(model_tx_grade2)
```

```{r}
#| label: chk-tx-grade2-convergence

check_convergence(model_tx_grade2)
```
:::


## Blood Units Given

For patients who were given transfusions, is HIV status associated with the number of units of blood given, adjusting for level of anemia

```{r}
#| label: show-formula-tx-units
#| results: asis

model_tx_units <- lmer(
  transfusion_units ~ hiv_status + anemia_grade + (1 | record_id),
  data = mgmt_cycle |> filter(transfusion_given == 1, !is.na(anemia_grade))
)

show_formula(model_tx_units)
```

```{r}
#| label: tbl-transfusion-units
#| tbl-cap: "Linear mixed model: units of blood received among transfused cycles."

model_tx_units |>
  tbl_regression(
    label = list(
      hiv_status   = "HIV Status",
      anemia_grade = "Anemia Grade"
    )
  ) |>
  tbl_drop_ref_rows() |>
  bold_labels() |>
  bold_p()
```

#### Validation: model_tx_units
::: {.callout-note collapse="true" title="Validation: model_tx_units"}
```{r}
#| label: chk-tx-units-vif

check_vif(model_tx_units)
```
:::

## G-CSF Administration

Modeled among cycles with a recorded neutropenia or leukopenia grade. The
severity predictor is `gcsf_grade`, the per-cycle maximum of neutropenia (ANC)
and leukopenia (WBC) grade, since either can indicate G-CSF.

Age is included per the 13-Jul-2026 meeting: elderly patients are more likely to
have a count drop and therefore to be supported. Stage, chemotherapy type and
cancer site mirror the transfusion model.

```{r}
#| label: prep-gcsf-data

gcsf_vars <- c("csfg_indicated", "hiv_status", "gcsf_grade", "age",
               "cancer_stage", "chemo_type", "cancer_site", "record_id")

mgmt_gcsf_raw <- mgmt_cycle |>
  select(all_of(gcsf_vars)) |>
  drop_na()

mgmt_gcsf_grade2plus_raw <- mgmt_gcsf_raw |>
  filter(gcsf_grade >= 2)

# Separated levels are dropped, not merely reported. The screen below was
# previously diagnostic only, which left Gall bladder/Biliary/Pancreas (no
# cycle in that group received G-CSF) in the model. That is the source of the
# "Hessian is numerically singular" and "unable to evaluate scaled gradient"
# warnings: the coefficient for a perfectly separated level has no finite
# maximum-likelihood estimate, so the optimiser cannot converge on it.
gcsf_adj_vars <- c("cancer_stage", "chemo_type", "cancer_site")

mgmt_gcsf            <- drop_separated_levels(mgmt_gcsf_raw, "csfg_indicated", gcsf_adj_vars)
mgmt_gcsf_grade2plus <- drop_separated_levels(mgmt_gcsf_grade2plus_raw, "csfg_indicated", gcsf_adj_vars)

# As with the transfusion strata, dropping rows can leave a covariate constant.
gcsf_formula_for <- function(d) {
  candidates <- c("hiv_status", "gcsf_grade", "age", "cancer_stage",
                  "chemo_type", "cancer_site")
  keep <- candidates[vapply(candidates,
                            function(v) dplyr::n_distinct(d[[v]]) > 1,
                            logical(1))]
  reformulate(c(keep, "(1 | record_id)"), response = "csfg_indicated")
}

gcsf_labels <- list(
  hiv_status   = "HIV Status",
  gcsf_grade   = "Max Neutropenia/Leukopenia Grade",
  age          = "Age (years)",
  cancer_stage = "Cancer Stage",
  chemo_type   = "Chemotherapy Type",
  cancer_site  = "Cancer Site"
)

gcsf_labels_for <- function(model) {
  gcsf_labels[intersect(names(gcsf_labels), all.vars(formula(model)))]
}

```

```{r}
#| label: tbl-gcsf-exclusions
#| tbl-cap: "Covariate levels dropped from each G-CSF model. A level is dropped when the outcome does not vary within it. The two strata drop different levels: a cancer site can have variation across all graded cycles and none once restricted to grade 2+."

tibble(
  Stratum      = c("All graded cycles", "Grade 2+ cycles"),
  `Cycles (n)` = c(nrow(mgmt_gcsf), nrow(mgmt_gcsf_grade2plus)),
  `Levels dropped` = c(describe_dropped_levels(mgmt_gcsf),
                       describe_dropped_levels(mgmt_gcsf_grade2plus))
) |>
  kbl(booktabs = TRUE) |>
  kable_styling(full_width = FALSE, position = "left")
```

::: {.callout-important collapse="true" title="Separation screen: G-CSF"}
Any level flagged `separated` contributes no information and would produce an
odds ratio of 0 or infinity. Those levels are removed from the model before
fitting; descriptive tables retain them.

**Screened separately for each stratum**, because a level can have perfectly
good variation across all graded cycles and none at all once restricted to
grade 2+. Screening only the full data would make the grade-2+ exclusions look
unjustified.

```{r}
#| label: chk-gcsf-separation

gcsf_screen <- function(d, stratum) {
  d |>
    mutate(gcsf_grade_f = factor(gcsf_grade)) |>
    select(csfg_indicated, hiv_status, gcsf_grade_f,
           cancer_stage, chemo_type, cancer_site) |>
    pivot_longer(-csfg_indicated, names_to = "term", values_to = "level",
                 values_transform = as.character) |>
    group_by(term, level) |>
    summarise(n = n(), n_gcsf = sum(csfg_indicated == 1, na.rm = TRUE),
              .groups = "drop") |>
    mutate(Stratum   = stratum,
           Variable  = pretty_var(term),
           pct_gcsf  = round(100 * n_gcsf / n, 1),
           separated = n_gcsf == 0 | n_gcsf == n) |>
    select(Stratum, Variable, level, n, n_gcsf, pct_gcsf, separated)
}

bind_rows(
  gcsf_screen(mgmt_gcsf_raw,            "All graded cycles"),
  gcsf_screen(mgmt_gcsf_grade2plus_raw, "Grade 2+ cycles")
) |>
  arrange(desc(separated), Stratum, Variable, level)
```
:::

### All graded cycles

```{r}
#| label: show-formula-gcsf-all
#| results: asis

model_gcsf_given <- glmer(gcsf_formula_for(mgmt_gcsf),
                          data = mgmt_gcsf, family = binomial)

show_formula(model_gcsf_given)
```

```{r}
#| label: tbl-gcsf-given
#| tbl-cap: "Odds of receiving G-CSF across all cycles with a recorded neutropenia or leukopenia grade."

model_gcsf_given |>
  tbl_regression(exponentiate = TRUE, label = gcsf_labels_for(model_gcsf_given)) |>
  tbl_drop_ref_rows() |>
  add_exclusion_note(mgmt_gcsf) |>
  bold_labels() |>
  bold_p()
```

#### Validation: model_gcsf_given

::: {.callout-note collapse="true" title="Validation: model_gcsf_given"}
```{r}
#| label: chk-gcsf-given-convergence

check_convergence(model_gcsf_given)
```

```{r}
#| label: chk-gcsf-given-vif

check_vif(model_gcsf_given)
```

```{r}
#| label: chk-gcsf-given-separation

check_separation(model_gcsf_given)$coefs |>
  arrange(desc(extreme))
```
:::

### G-CSF at grade 2 and above

Restricted to cycles where the patient had grade 2 or higher
neutropenia/leukopenia — cycles where G-CSF had a plausible indication.

```{r}
#| label: show-formula-gcsf-grade2plus
#| results: asis

model_gcsf_grade2plus <- glmer(gcsf_formula_for(mgmt_gcsf_grade2plus),
                               data = mgmt_gcsf_grade2plus, family = binomial)

show_formula(model_gcsf_grade2plus)
```

```{r}
#| label: tbl-gcsf-grade2plus
#| tbl-cap: "Odds of receiving G-CSF among cycles with grade 2 or higher neutropenia/leukopenia."

model_gcsf_grade2plus |>
  tbl_regression(exponentiate = TRUE,
                 label = gcsf_labels_for(model_gcsf_grade2plus)) |>
  tbl_drop_ref_rows() |>
  add_exclusion_note(mgmt_gcsf_grade2plus) |>
  bold_labels() |>
  bold_p()
```

```{r}
#| label: tbl-gcsf-hiv-comparison
#| tbl-cap: "HIV status and the odds of receiving G-CSF, compared across the two analysis populations."
#| results: asis

# Presented as a targeted comparison rather than tbl_merge(). The two models do
# not necessarily carry the same covariate levels once separated levels are
# removed, and merging tables with different row sets produced the "number of
# rows in the tables to be merged do not match" warning and silently
# misaligned rows.
gcsf_hiv_comparison <- bind_rows(
  hiv_estimate(model_gcsf_given,      "All graded cycles"),
  hiv_estimate(model_gcsf_grade2plus, "Grade 2+ cycles only")
)

ft <- gcsf_hiv_comparison |>
  flextable() |>
  theme_booktabs() |>
  align(align = "center", part = "all") |>
  align(j = 1, align = "left", part = "all") |>
  flextable::add_footer_lines(
    "If these estimates disagree substantially, the all-cycles result is being driven by the large number of low-grade cycles where G-CSF was never under consideration."
  ) |>
  autofit()

obj <- save_table(slug = "gcsf_hiv_comparison_by_stratum", df = gcsf_hiv_comparison, ft = ft)
pub_objects_analysis$gcsf_hiv_comparison_by_stratum <- obj

obj$ft
emit_buttons(obj)
```

#### Validation: model_gcsf_grade2plus

::: {.callout-note collapse="true" title="Validation: model_gcsf_grade2plus"}
```{r}
#| label: chk-gcsf-grade2plus-convergence

check_convergence(model_gcsf_grade2plus)
```

A non-converged or singular fit here is a real limitation of the restricted
sample, not a formatting problem. Do not report estimates from a model flagged
`converged = FALSE` without saying so.

```{r}
#| label: chk-gcsf-grade2plus-vif

check_vif(model_gcsf_grade2plus)
```

```{r}
#| label: chk-gcsf-grade2plus-separation

check_separation(model_gcsf_grade2plus)$coefs |>
  arrange(desc(extreme))
```
:::


```{r}
#| label: save-pub-objects-analysis

save_rds_safe(pub_objects_analysis, onedrive_data_path("processed", "pub_objects_analysis.rds"))
```

