---
title: "GI Cancer: Publication Figures & Tables"
bibliography: references.bib
execute:
  echo: false
  warning: true
  message: true
format:
  html:
    code-links:
      - text: GitHub Page
        icon: github
        href: "https://github.com/EDukeChase/gi-cancer_project/blob/main/5%20-%20Publication%20Figures.qmd"
---

```{r}
#| label: setup
#| include: false

source("R/setup.R")

# Finalized tables/figures saved incrementally by 03 and 04. This document
# only loads, applies light final formatting, and displays/exports — it
# performs no statistical computation of its own.
pub_objects_desc <- readRDS(onedrive_data_path("processed", "pub_objects_desc.rds"))
pub_objects_analysis <- readRDS(onedrive_data_path("processed", "pub_objects_analysis.rds"))
```


# Comprehensive Descriptive Table

```{r}
#| label: tbl-desc
#| tbl-cap: "Patient, disease, and treatment characteristics by HIV status"
#| results: asis
#| echo: false

obj <- pub_objects_desc$table1_descriptive
obj$ft
emit_buttons(obj)
```

```{r}
#| label: tbl-desc-ae
#| tbl-cap: "Adverse drug events (any grade) by category and HIV status"
#| results: asis
#| echo: false

obj <- pub_objects_desc$table_ae_by_category
obj$ft
emit_buttons(obj)
```


# Completion Rate by HIV Status

Proportion of patients who completed treatment cycle $k$, out of the patients
who were prescribed at least $k$ cycles.

::: {.callout-note}
This figure resembles a Kaplan-Meier curve but is **not** one — there is no
censoring and no time axis. Each point is a simple proportion among the
patients prescribed at least that many cycles.
:::

Two versions are shown. The **stratified** version is the one to look at first:
patients prescribed six cycles necessarily stop contributing after cycle 6, so
the pooled curve mixes that structural drop-off with genuine early
discontinuation. Splitting on the number of cycles prescribed separates the two.
Please tell us which version you would rather see in the paper — we do not need
both.

```{r}
#| label: plt-cycle-completion-split
#| fig-cap: "Chemotherapy cycle completion by HIV status, stratified by number of cycles prescribed. Shaded bands are pointwise 95% confidence intervals."
#| fig-width: 10
#| fig-height: 4.5

pub_objects_desc$plt_cycle_completion_split$plot
```

```{r}
#| label: plt-cycle-completion
#| fig-cap: "Chemotherapy cycle completion by HIV status, all patients pooled. Shaded bands are pointwise 95% confidence intervals; the counts beneath the panel give the number of patients prescribed at least k cycles."

pub_objects_desc$plt_cycle_completion$plot
```


# Severe Adverse Events

Odds of experiencing a grade 3-4 adverse event at any point during treatment,
adjusted for cancer stage, chemotherapy regimen, treatment intent, age and sex.
Reference levels are listed in a note beneath the table. The unadjusted model
was closely consistent and is quoted in the caption rather than tabulated.

```{r}
#| label: tbl-severe-or-comparison
#| tbl-cap: !expr pub_objects_analysis$severe_caption

pub_objects_analysis$tbl_severe_or_comparison$ft
```


# Hospitalization

Whether a patient was hospitalized at all in a given cycle, and — among the
cycles where a hospitalization did occur — how long it lasted.

```{r}
#| label: tbl-any-hosp
#| tbl-cap: "Odds of any toxicity-related hospitalization within a treatment cycle, adjusted for maximum adverse event grade in that cycle, age and sex"

pub_objects_analysis$tbl_any_hosp$ft
```

```{r}
#| label: plt-hosp-vs-max-grade
#| fig-cap: "Mean hospital days vs maximum adverse event grade in the cycle, by HIV status, among cycles where hospital days were recorded"

pub_objects_analysis$plt_hosp_vs_max_grade$plot
```

```{r}
#| label: tbl-hosp-days
#| tbl-cap: "Length of stay among hospitalized cycles: unadjusted vs adjusted linear mixed models. Estimates are differences in hospital days, not odds ratios."

pub_objects_analysis$tbl_hosp_days_comparison$ft
```

# Early Dropout

"Early dropout" means receiving four or fewer cycles among patients who were
prescribed more than four. The question is whether toxicity in the first four
cycles predicts stopping early.

```{r}
#| label: tbl-early-dropout-descriptive
#| tbl-cap: "Adverse event grades in the early treatment window, and HIV status, by whether the patient discontinued early."

pub_objects_analysis$table_early_dropout_descriptive$ft
```

```{r}
#| label: tbl-early-dropout-final
#| tbl-cap: "Association between early-cycle adverse events and early treatment discontinuation. P-values are Benjamini-Hochberg adjusted across adverse event types."

pub_objects_analysis$table_early_dropout_final$ft
```

The forest plot below shows the same associations. Odds ratios are per one-grade
increase in toxicity; a value above 1 means more toxicity was associated with a
higher chance of stopping early.

```{r}
#| label: plt-early-dropout-forest
#| fig-cap: "Impact of early-cycle toxicities on early treatment dropout, pooled across HIV groups and adjusted for HIV status"

pub_objects_analysis$plt_early_dropout_forest$plot
```

The version below fits each association separately within each HIV group, so the
two groups can be compared directly rather than pooled.

::: {.callout-warning}
The HIV-positive group is small, so several of these intervals are very wide. A
gap between the two groups is only meaningful where both intervals are
reasonably tight — please read this figure alongside the pooled one above rather
than instead of it.
:::

```{r}
#| label: plt-early-dropout-forest-byhiv
#| fig-cap: "Association between early-cycle toxicity grade and early treatment dropout, estimated separately within each HIV group"
#| fig-height: 6

pub_objects_analysis$plt_early_dropout_forest_byhiv$plot
```
