---
title: "GI cancer"
output: html_notebook
---

Comparing  real world gastrointestinal cancer chemotherapy treatment related adverse events incidence and severity between HIV positive and HIV negative patients in Zimbabwe- Hematological and hepatic functions

# libraries
```{r}
# Load libraries
library(ggplot2)
library(tidyr)
library(dplyr)
library(DT)
library(readxl)
library(here)
here()
```

# data
```{r}
IPROTECT <- read_excel("/Users/audreyhendricks/Desktop/CU Denver/Research/Active/Tinashe/project 1/IPROTECTARetrospecti_DATA_2025-08-19_2157 FINAL SHEET.xlsx")

dim(IPROTECT)
```

# checking duplicates - ok to remove?
```{r}
plot(IPROTECT$ANC.3...108, IPROTECT$ANC.3...109)
abline(0,1)

plot(IPROTECT$`HyperK+.3...114`, IPROTECT$`HyperK+.3...115`)
abline(0,1)
```

# summary of completion rates and cycles completed
```{r}
summary(IPROTECT$`Chemotherapy.completion.rates.(%)`)
summary(IPROTECT$Total.number.chemo.cycles.prescribed)
summary(IPROTECT$Total.number.of.chemotherapy.administered)

plot(IPROTECT$Total.number.chemo.cycles.prescribed, IPROTECT$Total.number.of.chemotherapy.administered)

# Scatterplot heat map
ggplot(IPROTECT, aes(Total.number.chemo.cycles.prescribed, Total.number.of.chemotherapy.administered)) +
  geom_bin2d(bins = 8) +         # adjust bins for granularity
  scale_fill_gradient(low = "lightblue", high = "red") +
  theme_minimal() +
  labs(
    title = "Scatterplot Heat Map",
    x = "Prescribed",
    y = "Administered",
    fill = "Count"
  )


ggplot(IPROTECT, aes(Total.number.chemo.cycles.prescribed, `Chemotherapy.completion.rates.(%)`)) +
  geom_bin2d(bins = 10) +         # adjust bins for granularity
  scale_fill_gradient(low = "lightblue", high = "red") +
  theme_minimal() +
  labs(
    title = "Scatterplot Heat Map",
    x = "Prescribed",
    y = "Completion Rate",
    fill = "Count"
  )


```
```{r}
ggplot(IPROTECT, aes(Total.number.chemo.cycles.prescribed, Total.number.of.chemotherapy.administered)) +
  stat_density_2d(aes(fill = ..density..), geom = "raster", contour = FALSE) +
  scale_fill_viridis_c(option = "plasma") +
  theme_minimal() +
  labs(
    title = "Density-based Scatterplot Heat Map",
    x = "X-axis",
    y = "Y-axis",
    fill = "Density"
  )
```


# recalculate completion rates and cycles completed
```{r}
IPROTECT$completion.rate<-IPROTECT$Total.number.of.chemotherapy.administered/IPROTECT$Total.number.chemo.cycles.prescribed

plot(IPROTECT$`Chemotherapy.completion.rates.(%)`, IPROTECT$completion.rate)
abline(0,1)

IPROTECT$administered<-apply(IPROTECT[,c("Was.the.cycle.delayed.due..to.toxicity.1","Was.the.cycle.delayed.due..to.toxicity.2","Was.the.cycle.delayed.due..to.toxicity.3","Was.the.cycle.delayed.due..to.toxicity.4","Was.the.cycle.delayed.due..to.toxicity.5","Was.the.cycle.delayed.due..to.toxicity.6","Was.the.cycle.delayed.due..to.toxicity.7","Was.the.cycle.delayed.due..to.toxicity.8","Was.the.cycle.delayed.due..to.toxicity.9","Was.the.cycle.delayed.due..to.toxicity.10","Was.the.cycle.delayed.due..to.toxicity.11","Was.the.cycle.delayed.due..to.toxicity.12")], 1, function(x){sum(I(1-x),na.rm=T)})

IPROTECT$administered2<-apply(IPROTECT[,c("Was.the.cycle.delayed.due..to.toxicity.1","Was.the.cycle.delayed.due..to.toxicity.2","Was.the.cycle.delayed.due..to.toxicity.3","Was.the.cycle.delayed.due..to.toxicity.4","Was.the.cycle.delayed.due..to.toxicity.5","Was.the.cycle.delayed.due..to.toxicity.6","Was.the.cycle.delayed.due..to.toxicity.7","Was.the.cycle.delayed.due..to.toxicity.8","Was.the.cycle.delayed.due..to.toxicity.9","Was.the.cycle.delayed.due..to.toxicity.10","Was.the.cycle.delayed.due..to.toxicity.11","Was.the.cycle.delayed.due..to.toxicity.12")], 1, function(x){sum(!is.na(x))})


##summary(apply(IPROTECT[,c("Was.the.cycle.delayed.due..to.toxicity.1","Was.the.cycle.delayed.due..to.toxicity.2","Was.the.cycle.delayed.due..to.toxicity.3")], 1, function(x){sum(x,na.rm=T)}))

plot(IPROTECT$Total.number.of.chemotherapy.administered, IPROTECT$administered2)
abline(0,1)


# Scatterplot heat map
ggplot(IPROTECT, aes(Total.number.of.chemotherapy.administered, administered2)) +
  geom_bin2d(bins = 8) +         # adjust bins for granularity
  scale_fill_gradient(low = "lightblue", high = "red") +
  theme_minimal() +
  labs(
    title = "Scatterplot Heat Map",
    x = "previous administered",
    y = "Administered",
    fill = "Count"
  )

temp<-IPROTECT[IPROTECT$administered2!=IPROTECT$Total.number.of.chemotherapy.administered,c("pgh_number", "Total.number.of.chemotherapy.administered", "administered", "administered2", "Was.the.cycle.delayed.due..to.toxicity.1","Was.the.cycle.delayed.due..to.toxicity.2","Was.the.cycle.delayed.due..to.toxicity.3","Was.the.cycle.delayed.due..to.toxicity.4","Was.the.cycle.delayed.due..to.toxicity.5","Was.the.cycle.delayed.due..to.toxicity.6","Was.the.cycle.delayed.due..to.toxicity.7","Was.the.cycle.delayed.due..to.toxicity.8","Was.the.cycle.delayed.due..to.toxicity.9","Was.the.cycle.delayed.due..to.toxicity.10","Was.the.cycle.delayed.due..to.toxicity.11","Was.the.cycle.delayed.due..to.toxicity.12")]
dim(IPROTECT)
dim(temp)

write.csv(temp, "~/Desktop/CU Denver/Research/Active/Tinashe/project 1/different_administered.csv", row.names=F)
```
# baseline descriptive statistics
## HIV status
```{r}

table(IPROTECT$`HIV.Status:.1,.Negative.|.2,.Positive`)

temp<-table(IPROTECT$`HIV.Status:.1,.Negative.|.2,.Positive`, IPROTECT$`Sex.Male.1/Female.2`)
temp

```
```{r}
# --- 1) Clean & label variables ----------------------------------------------
IPROTECT_clean <- IPROTECT |>
  dplyr::rename(
    HIV_Status_raw = `HIV.Status:.1,.Negative.|.2,.Positive`,
    Sex_raw        = `Sex.Male.1/Female.2`
  ) |>
  mutate(
    HIV_Status = dplyr::recode(as.character(HIV_Status_raw),
                               `1` = "Negative", `2` = "Positive", .default = NA_character_),
    Sex        = dplyr::recode(as.character(Sex_raw),
                               `1` = "Male", `2` = "Female", `3` = "3", .default = NA_character_)
  )

# Utility for percent formatting
pct_fmt <- function(x) scales::percent(x, accuracy = 0.1, trim = TRUE)


# --- 3) Two-way summary (Sex × HIV Status; counts & row %) --------------------
twoway_counts <- IPROTECT_clean |>
  count(Sex, HIV_Status, name = "Count") |>
  group_by(Sex) |>
  mutate(RowProp = Count / sum(Count)) |>
  ungroup() |>
  mutate(Cell = paste0(Count, " (", pct_fmt(RowProp), ")")) |>
  select(Sex, HIV_Status, Cell) |>
  tidyr::pivot_wider(names_from = HIV_Status, values_from = Cell) |>
  identity()

datatable(
  twoway_counts,
  rownames = FALSE,
  class = "cell-border stripe hover",
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-weight: bold;',
    'Table 2. Sex × HIV Status (counts with row %)'
  ),
  options = list(pageLength = 5, autoWidth = TRUE, dom = 't')
)
```

```{r}
# install.packages(c("dplyr","tidyr","DT","scales"))
library(dplyr)
library(tidyr)
library(DT)
library(scales)

# Helper for percent formatting
pct_fmt <- function(x) percent(x, accuracy = 0.1, trim = TRUE)

# --- Clean variable names & recode -------------------------------------------
IPROTECT_clean <- IPROTECT |>
  rename(
    HIV_Status_raw = `HIV.Status:.1,.Negative.|.2,.Positive`,
    Sex_raw        = `Sex.Male.1/Female.2`
  ) |>
  mutate(
    HIV_Status = recode(as.character(HIV_Status_raw),
                        `1` = "Negative", `2` = "Positive", .default = NA_character_),
    Sex        = recode(as.character(Sex_raw),
                        `1` = "Male", `2` = "Female", .default = NA_character_)
  )

# --- Chi-squared test ---------------------------------------------------------
ct <- IPROTECT_clean |>
  filter(!is.na(Sex), !is.na(HIV_Status)) |>
  count(Sex, HIV_Status, name = "Count") |>
  pivot_wider(names_from = HIV_Status, values_from = Count, values_fill = 0) |>
  tibble::column_to_rownames("Sex") |>
  as.matrix()

chisq <- suppressWarnings(chisq.test(ct, correct = FALSE))
p_str <- format.pval(chisq$p.value, digits = 3, eps = 1e-4)
x2_str <- sprintf("%.2f", unname(chisq$statistic))
df_str <- sprintf("%d", unname(chisq$parameter))

# --- Create summary table with p-value column ---------------------------------
twoway_counts <- IPROTECT_clean |>
  filter(!is.na(Sex), !is.na(HIV_Status)) |>
  count(Sex, HIV_Status, name = "Count") |>
  group_by(Sex) |>
  mutate(RowProp = Count / sum(Count)) |>
  ungroup() |>
  mutate(Cell = paste0(Count, " (", pct_fmt(RowProp), ")")) |>
  select(Sex, HIV_Status, Cell) |>
  pivot_wider(names_from = HIV_Status, values_from = Cell) |>
  mutate(`p-value` = ifelse(row_number() == 1,
                            paste0("X² = ", x2_str, ", df = ", df_str, ", p = ", p_str),
                            ""))

# --- Render the DataTable -----------------------------------------------------
datatable(
  twoway_counts,
  rownames = FALSE,
  class = "cell-border stripe hover",
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-weight: bold;',
    'Table 2. Sex × HIV Status (counts with row % and chi-squared p-value)'
  ),
  options = list(
    pageLength = 5,
    autoWidth = TRUE,
    dom = 't',
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  )
)
```


# summary statistics for side effects - at least one?
## make long dataset
```{r}
IPROTECT_long <- IPROTECT %>%
  pivot_longer(
    cols = starts_with(c("HB", "MCV")),    # which columns to make long
    names_to = "appointment",        # name of new 'variable name' column
    values_to = c("HB", "MCV")            # name of new 'value' column
  )
```

# grades and summary statistics for side effects - at least one?
```{r}
##HB
# make wide dataset long
ggplot(IPROTECT, aes(x = replicates, y = Nfoods, fill = replicates)) +
  geom_violin(alpha = 0.5) +   # Density distribution (Pirate plot style)
  geom_boxplot(width = 0.2, outlier.shape = NA, 
               coef = 1.5, color = "black", lwd = 1.2) +  # Adjust whiskers
  geom_jitter(width = 0.1, size = 1.5, alpha = 0.7) +  # Raw data points
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel1") +  
  labs(title = "N foods in each compound")
```




